### 2021-08-02 supplement.tex
all figures

\section{Supplementary Material}
\beginsupplement

\begin{figure}[H]
  \centering
  \fbox{
   \includegraphics[width=0.36\paperwidth]{fig/hcp_iq_nonnorm_hist.pdf}
   \includegraphics[width=0.36\paperwidth]{fig/hcp_iq_quanttrf_hist.pdf}
   }
  \caption{Histogram of fluid intelligence score in the HPC dataset, before (left) and after (right) quantile transformation.}
  \label{fig:hcp-hist}
\end{figure}

\begin{figure}[H]
  \centering
  \fbox{
   \includegraphics[width=0.36\paperwidth]{fig/abide_motion_hist.pdf}
   \includegraphics[width=0.36\paperwidth]{fig/abide_motion_quanttrf_hist.pdf}
   }
  \caption{Histogram of mean framewise displacement in the ABIDE dataset, before (left) and after (right) quantile transformation.}
  \label{fig:abide-hist}
\end{figure}

### 2021-08-03 supplement.tex
results ready

@@
   \fbox{
-   \includegraphics[width=0.36\paperwidth]{fig/hcp_iq_nonnorm_hist.pdf}
-   \includegraphics[width=0.36\paperwidth]{fig/hcp_iq_quanttrf_hist.pdf}
+   \includegraphics[width=0.36\paperwidth]{fig/supplement/hcp_iq_nonnorm_hist.pdf}
+   \includegraphics[width=0.36\paperwidth]{fig/supplement/hcp_iq_quanttrf_hist.pdf}
    }
@@
   \fbox{
-   \includegraphics[width=0.36\paperwidth]{fig/abide_motion_hist.pdf}
-   \includegraphics[width=0.36\paperwidth]{fig/abide_motion_quanttrf_hist.pdf}
+   \includegraphics[width=0.36\paperwidth]{fig/supplement/abide_motion_hist.pdf}
+   \includegraphics[width=0.36\paperwidth]{fig/supplement/abide_motion_quanttrf_hist.pdf}
    }
@@
 \end{figure}
+
+\begin{figure}[H]
+  \centering
+  \fbox{
+   \includegraphics[width=0.75\paperwidth]{fig/supplement/sim_ccc_full_normal_all_heatmap.pdf}}
+  \caption{Heatmaps showing the positive rates of the 'full' confounder test, with normally distributed numerical simulated variables.}
+  \label{fig:sim-ccc-full}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \fbox{
+   \includegraphics[width=0.75\paperwidth]{fig/supplement/sim_ccb_partial_normal_all_heatmap.pdf}}
+  \caption{Heatmaps showing the positive rates of the 'partial' confounder test, with normally distributed numeric $Y$ and $\hat{Y}$ and categorical $C$.}
+  \label{fig:sim-ccb-partial}
+\end{figure}

### 2021-08-08 manuscript.tex
patched

\documentclass{article}
\usepackage{arxiv}

\captionsetup[figure]{labelfont=it,textfont={it}}

\title{A conditional permutation-based approach to test confounder effect and center-bias in machine learning models}

\author{
  Tamas~Spisak \\
  Institute for Diagnostic and Interventional Radiology and Neuroradiology \\
  University Hospital Essen\\
  Hufelandstrasse 55, 45147 Essen \\
  \texttt{tamas.spisak@uk-essen.de} \\
  %% examples of more authors
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

\begin{document}
\maketitle

\begin{abstract} % 150 words
The lack of rigorous non-parametric statistical tests of confoudner-effects significantly hampers the development of robust, valid ans generalizable predictive models in many fields of research.
Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recent theoretical framework of conditional independence testing and test the null hypothesis of \emph{no bias} and \emph{fully biased model}, respectively.
The proposed tests set no assumptions about the distribution of the predictive model output that is often non-normal.
As shown by simulations, the test are statistically valid, robust and display a high statistical power.
Applying the tests on models based on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange databases shows that the proposed tests can significantly improve the assessment of the robustness, generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.

The tests are implemented in the python package \emph{mlconfound}\footnote{https://github.com/pni-lab/mlconfound}.
\end{abstract}

% keywords can be removed
\keywords{machine learning, predictive modelling, confounder test, conditional independence, conditional permutation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Predictive modelling and supervised learning methods have recently became increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized}. The usefulness and applicability of the such markers is to be evaluated not solely on the biomedical relevance of the predictive performance (i.e. prognostic/diagnostic value); generalizability across contexts and populations and biomedical validity (i.e. the model is driven by biomedical relevant signal) are also highly important requirements towards potential biomarkers \citep{woo2017building}.

Variables that can induce spurious associations between the predictor variables (features) and the prediction target - often referred to as confounders - can be detrimental to the model's biomedical validity and generalizability. Many types of confounders can be distinguished.
First, measurement artifacts can obviously be considered as confounders. For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with movement kinematics or motor function, like previously demonstrated, among others, in case of e.g Alzheimer's disease \citep{rao2017predictive} or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}. Similarly, eye-blink artifacts can result in confound-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.
Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling. E.g. resting-state functional connectivity is known to be strongly predictive to  demographic age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}. As fluid intelligence is known to decline with aging, \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by simply picking up on age-related variance \cite{lohmann2021predicting, dubois2018distributed}. 
Third, center-effects (or in general, batch effect) in multi-center studies can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance}. In this case, predictive models can utilize center-effects to explain (random or true) center-differences in the predictive target. Such models will display worse-than-expected performance when applied on data from new centers.

Recently, various data harmonization approaches have been suggested to mitigate confounder-bias, ranging from simple matching \citep{rao2017predictive} or regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} or sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}). However, it is often unclear which variables should be considered as confounders during harmonization and considering many variables as confounder holds risks of eliminating signal-of-interest \citep{wachinger2021detect}.

Powerful statistical tests to quantify the confounding effect of certain variables in predictive models could largely foster both the the identification of confounders to correct for and the assessment of the effectiveness of various harmonization or other confound-mitigation approaches. Such methods, however, have to tackle the fact that the outputs of predictive models often have non-normal distributions (due to e.g. regularization) \citep{garcia2009study, kristensen2017whole}.

Recently three dedicated techniques \citep{chaibub2019permutation, ferrari2020measuring, wachinger2021detect} have been proposed for quantifying confounder bias in predictive modelling.

The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via restricted permutations. It's p-values, however, might be invalid in certain circumstances (e.g. biased classification problems) and, as shown by \cite{ferrari2020measuring}, it does not guarantee a proper type I error control.
The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, by design, does not provide any quantification of statistical significance.
Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high computational complexity (especially when hyperparameters are optimized in a nested cross-validation).
The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the the predictor(s) are a direct cause of the target variable or, alternatively, there exists an unobserved random variable that is the cause of both. It's computation assumes 'faithfulness' (if two variables are independent, there is no direct influence between the two in the underlying graph) a strong condition that might not be fulfilled in many real applications (e.g. functional MRI is an \emph{indirect} measure of neural activation, rendering the alternative hypothesis true in absence of any real confounder).

% todo: RO and CI: classification, KC: regression

In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}. This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ("no free lunch" theorem).

Next, I consider the conditional permutation test, a statistical test of conditional independence, recently introduced by \cite{berrett2020conditional}, as a test for evaluating confounder-bias of predictive models.

Within this framework, various confounder-related null hypotheses can be tested, so that no assumption is to be made about the distribution of one of the conditioned variables. As normality is typically a reasonable assumption for the target and confounder variables but not for the predictions of the model, the focus is put on the cases allowing inference without assumptions on the model predictions' distribution.
Two cases satisfy this requirement, one testing whether the model's explained variance can be exclusively attributed to the confounder and the other testing whether the model utilizes any confounder-variance, above what's already present in the target variable.
Both tests require only on the target variable, the model predictions and the putative confounder, that is, they can be performed without needing to re-fit the model.
Additionally, a regression-based estimation of the conditional distribution of $c|y$ is proposed, which allows testing confounder bias without a-priory knowledge about the conditional distribution and without any assumptions on the distribution of the predicted values (which are often non-normal in practice).
The validity of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normal model output).
The proposed tests are then applied on functional brain connectivity data to provide evidence of motion-bias in a model trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} and to evaluate the age dependence of fluid intelligence predictions on data from the Human Connectome Project.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}

\subsection{Notation and Background}

In a predictive modelling setting, let $Y$ denote the target variable, $\hat{Y}$ denote model output, i.e. the predictions for $Y$ and let $C$ denote a variable, which is considered as a confounder. Depending on the research question, the influence of $C$ on the model predictions $\hat{Y}$ is subject to be kept at a negligible level, or it is to be proven that the model is not completely driven by $C$. Below, I summarize how testing conditional independence \citep{dawid1979conditional} among these variables allows addressing the above questions and build on the conditional permutation test of \cite{berrett2020conditional} to perform inference without assumptions on the distribution of the model output, where normality is often violated \citep{garcia2009study, kristensen2017whole}.

\subsubsection*{Testing conditional independence}

Let us shall assume that the data consist of independent and identically distributed (IID) data points $(x_i, y_i, z_i) \in X \times Y \times Z$ for $i=1, \dots , n$ and write $X=(x_1, \dots ,x_n)$, $Y=(y_1, \dots, y_n)$ and $Z=(Z_1, \dots, Zn)$.

$X$ is \emph{conditionally independent} of $Y$ given $Z$ $X \independent Y | Z$, if for all $x, y, z$ with $p(z) > 0$, we have $p(x, y|z) = p(x|z)p(y|z)$ \citep{dawid1979conditional}.

While conditional independence lies at the heart of several fundamental concepts is statistics and  plays an increasingly important role in various fields of applied statistics (and particularly in biomedical applications, \citep{spirtes2000causation, peters2016causal, fiedler2011mediation, candes2016panning}, recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
In their "no free luch" theorem, Shah and Peters has shown that, without placing some assumptions on the joint distribution of $(X, Y, Z)$, conditional testing is effectively impossible. In other words, no test exists that - for all distributions - provides a valid type I error control and, at the same time, has a non-trivial statistical power.

This result is in strong contrast to unconditional independence testing, for which permutation tests  \citep{pitman1937significance, fisher1942189} allow for the correct calibration of any hypothesis test. Even if the null hypothesis $H0 :X \independent Y|Z$ is true, $X$ and $Y$ might be highly marginally dependent due to their dependence on $Z$,  thus the permutation null-distribution will be invalid.

The "no-free-lunch" theorem has important implications for confound-testing in predictive modelling where the distribution of the model outputs - depending on the applied machine learning model - is \emph{unknown}.

Nevertheless, as shown by \cite{berrett2020conditional}, if information about the conditional distribution $Q(X|Z)$ is available, sampling from a non-uniform distribution over the set of possible permutations $\pi$ can incorporate available distributional information into permutation-based inference and achieve a valid type I error control and a non-trivial power.
Below we give a short description of the procedure of \cite{berrett2020conditional}.

\subsubsection*{The conditional permutation test for independence}

The conditional permutation test (CPT) of \cite{berrett2020conditional} build on the conditional randomization test of \cite{candes2016panning} and ensures that \emph{no assumptions} are made about the distribution of the response variable Y but, instead, it is assumed that the conditional distribution of X given Z is known.

CPT generates a null-distribution for an arbitrary predefined test statistic $T(X,Y,Z)$ by sampling permutation based "copies" of $X$,

$$x_i^{(j)} \sim Q(\cdot|z_i) \ $$

where, $Q(.|z)$ denotes the conditional distribution of $X$ given $Z=z$ and $j=1,\dots, m$ indexes the copies of $X$ and so that so that $X^{(i)} = (x_1^{(j)}, \dots, x_n^{(j)})$ is a permutation of the original vector $X = (x_1, \dots, x_n)$. This mechanism creates copies $X^{(1)}, \dots ,X^{(m)}$ that are exchangeable with the original vector X under the null hypothesis that $X \independent Y | Z$.

Under the null hypothesis, the triples $(X,Y,Z), (X^{(1)},Y,Z),\dots,(X^{(m)},Y,Z)$ are all identically distributed and exchangeable, and so are the 
$$T(X,Y,Z), T(X^{(1)},Y,Z),\dots,T(X^{(m)},Y,Z)$$
test statistics, as well.

The p-value under the null hypothesis is then obtained as
$$ p= 1+\frac{\sum_{j=1}^m \mathbb{1} \{T(X^{(j)},Y,Z) \geq T(X,Y,Z) \}  }{1+m}$$

Let $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$ and $X_\pi = (x_{\pi_1}, \dots, x_{\pi_n})$ the vector $X$ with its elements reordered according to the permutation $\pi \in S_i$.
In the CPT procedure, the permutation-based copies of $X$ are consist of $x_i^{(j)} = x_{\pi^{(j)}}$ which are drawn so that 
so that:

\begin{equation}
    \mathbb{P}(\pi^{(j)} = \pi | X,Y,Z) = \frac{q^n(X_\pi | Z)}{\sum_{\pi' \in S_n} q^n(X_{\pi'} | Z)}
    \label{eq-pperm}
\end{equation}

that is, according to the $q^n(\cdot|Z) := q(\cdot | z_1, \dots q(\cdot|z_n)$ product density corresponding of the conditional distribution $Q(\cdot|z)$. Note that Eq. \ref{eq-pperm} does not necessarily assume a continuous distribution.
For a verification of the valid type I error control of this approach refer to Theorem 1 in \citep{berrett2020conditional}.

While the required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$, at random, the acceptance ratio will be extremely low, even for moderate $n$ (except there is very low dependence of X on Z), resulting in slow mixing times. \cite{berrett2020conditional} (Algorithm 1) propose a parallelized pairwise Markov-Chain Monte Carlo sampler for CPT, that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data. In their Theorem 2, \cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.

\subsection{Conditional Permutation-based confounder-bias test}

% fig:schematic
\begin{figure}
  \centering
  \resizebox{0.75\columnwidth}{!}{%
\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

\tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt

\begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-1,xscale=1]
%uncomment if require: \path (0,300); %set diagram left start at 0, and has height of 300

%Shape: Circle [id:dp12907748114570317] 
\draw   (299.5,107) .. controls (299.5,93.19) and (310.69,82) .. (324.5,82) .. controls (338.31,82) and (349.5,93.19) .. (349.5,107) .. controls (349.5,120.81) and (338.31,132) .. (324.5,132) .. controls (310.69,132) and (299.5,120.81) .. (299.5,107) -- cycle ;
%Shape: Circle [id:dp12318976880062338] 
\draw   (397,39) .. controls (397,25.19) and (408.19,14) .. (422,14) .. controls (435.81,14) and (447,25.19) .. (447,39) .. controls (447,52.81) and (435.81,64) .. (422,64) .. controls (408.19,64) and (397,52.81) .. (397,39) -- cycle ;
%Shape: Circle [id:dp2655574667622329] 
\draw   (202,39) .. controls (202,25.19) and (213.19,14) .. (227,14) .. controls (240.81,14) and (252,25.19) .. (252,39) .. controls (252,52.81) and (240.81,64) .. (227,64) .. controls (213.19,64) and (202,52.81) .. (202,39) -- cycle ;
%Curve Lines [id:da9340239406716428] 
\draw    (250,28) .. controls (292,20) and (364,22) .. (400,28) ;
%Curve Lines [id:da6958426380393512] 
\draw    (239,61) .. controls (249,76) and (276,99) .. (299.5,107) ;
%Curve Lines [id:da8565840155201192] 
\draw    (349.5,107) .. controls (365,101) and (403,71) .. (410,60) ;

%Shape: Circle [id:dp9753164043561875] 
\draw   (475.5,257) .. controls (475.5,243.19) and (486.69,232) .. (500.5,232) .. controls (514.31,232) and (525.5,243.19) .. (525.5,257) .. controls (525.5,270.81) and (514.31,282) .. (500.5,282) .. controls (486.69,282) and (475.5,270.81) .. (475.5,257) -- cycle ;
%Shape: Circle [id:dp6395662810469516] 
\draw   (573,189) .. controls (573,175.19) and (584.19,164) .. (598,164) .. controls (611.81,164) and (623,175.19) .. (623,189) .. controls (623,202.81) and (611.81,214) .. (598,214) .. controls (584.19,214) and (573,202.81) .. (573,189) -- cycle ;
%Shape: Circle [id:dp786541411329625] 
\draw   (378,189) .. controls (378,175.19) and (389.19,164) .. (403,164) .. controls (416.81,164) and (428,175.19) .. (428,189) .. controls (428,202.81) and (416.81,214) .. (403,214) .. controls (389.19,214) and (378,202.81) .. (378,189) -- cycle ;
%Curve Lines [id:da22693900064180195] 
\draw    (426,178) .. controls (468,170) and (540,172) .. (576,178) ;
%Curve Lines [id:da6720989194231959] 
\draw    (415,211) .. controls (425,226) and (452,249) .. (475.5,257) ;
%Curve Lines [id:da8140666812849395] 
\draw    (525.5,257) .. controls (541,251) and (579,221) .. (586,210) ;
%Shape: Circle [id:dp471386481785639] 
\draw   (50,256) .. controls (50,242.19) and (61.19,231) .. (75,231) .. controls (88.81,231) and (100,242.19) .. (100,256) .. controls (100,269.81) and (88.81,281) .. (75,281) .. controls (61.19,281) and (50,269.81) .. (50,256) -- cycle ;
%Shape: Rectangle [id:dp8617145222488287] 
\draw   (156,237) -- (240,237) -- (240,277) -- (156,277) -- cycle ;
%Shape: Circle [id:dp8776576784951928] 
\draw   (290,256) .. controls (290,242.19) and (301.19,231) .. (315,231) .. controls (328.81,231) and (340,242.19) .. (340,256) .. controls (340,269.81) and (328.81,281) .. (315,281) .. controls (301.19,281) and (290,269.81) .. (290,256) -- cycle ;
%Shape: Circle [id:dp2868603849402189] 
\draw   (174,190) .. controls (174,176.19) and (185.19,165) .. (199,165) .. controls (212.81,165) and (224,176.19) .. (224,190) .. controls (224,203.81) and (212.81,215) .. (199,215) .. controls (185.19,215) and (174,203.81) .. (174,190) -- cycle ;
%Callout Right Arrow [id:dp26184503306970397] 
\draw   (46.55,225.27) -- (106,225.27) -- (106,250) -- (133,250) -- (133,241) -- (155.55,255.27) -- (133,269.55) -- (133,260.55) -- (106,260.55) -- (106,285.27) -- (46.55,285.27) -- cycle ;
%Right Arrow [id:dp7986276189609152] 
\draw   (240,250) -- (268,250) -- (268,241) -- (290,255.5) -- (268,270) -- (268,261) -- (240,261) -- cycle ;
%Shape: Circle [id:dp9535423837564185] 
\draw   (50,190) .. controls (50,176.19) and (61.19,165) .. (75,165) .. controls (88.81,165) and (100,176.19) .. (100,190) .. controls (100,203.81) and (88.81,215) .. (75,215) .. controls (61.19,215) and (50,203.81) .. (50,190) -- cycle ;
%Straight Lines [id:da6172567079022628] 
\draw    (75,215) -- (75,231) ;
%Straight Lines [id:da8731068989736062] 
\draw    (198,238) -- (198,215) ;

% Text Node
\draw (165,251) node [anchor=north west][inner sep=0.75pt]   [align=left] {ML model};
% Text Node
\draw (9,6) node [anchor=north west][inner sep=0.75pt]   [align=left] {\textbf{A}};
% Text Node
\draw (9,146) node [anchor=north west][inner sep=0.75pt]   [align=left] {\textbf{B}};
% Text Node
\draw (494,252.4) node [anchor=north west][inner sep=0.75pt]    {$C$};
% Text Node
\draw (591.5,180.4) node [anchor=north west][inner sep=0.75pt]    {$\hat{Y}$};
% Text Node
\draw (396,184.4) node [anchor=north west][inner sep=0.75pt]    {$Y$};
% Text Node
\draw (484,198) node [anchor=north west][inner sep=0.75pt]   [align=left] {CPT};
% Text Node
\draw (67,251.4) node [anchor=north west][inner sep=0.75pt]    {$X$};
% Text Node
\draw (308.5,247.4) node [anchor=north west][inner sep=0.75pt]    {$\hat{Y}$};
% Text Node
\draw (191,185.4) node [anchor=north west][inner sep=0.75pt]    {$Y$};
% Text Node
\draw (318,102.4) node [anchor=north west][inner sep=0.75pt]    {$Z$};
% Text Node
\draw (415.5,34.4) node [anchor=north west][inner sep=0.75pt]    {$Y$};
% Text Node
\draw (219,34.4) node [anchor=north west][inner sep=0.75pt]    {$X$};
% Text Node
\draw (308,48) node [anchor=north west][inner sep=0.75pt]   [align=left] {CPT};
% Text Node
\draw (68,185.4) node [anchor=north west][inner sep=0.75pt]    {$C$};

\end{tikzpicture}

    }
  \caption{Schematic diagram of using CPT in a predictive modelling context. \\ \textbf{(A)} CPT was originally proposed to be used on the feature variable $X$, target variable $Y$ and confounders $Z$. \textbf{(B)} The proposed use of CPT in predictive modelling requires the model to be fitted first, to obtain the model's prediction $\hat{Y}$ on $Y$. CPT is then utilized on the triplet $(Y, \hat{Y}, Z)$, to test hypotheses $\hat{y} \independent c | y$ or $y \independent \hat{y} | c$.}
  \label{fig:schematic}
\end{figure}

% tab:conditional-independence-cases
\renewcommand{\arraystretch}{2}
\begin{table}[]
\centering
\begin{tabular}{l|rp{60mm}|c|>{\centering\arraybackslash}m{30mm}}
 &  & H0  & assumption needed for: & no assumptions about the distribution of: \\
\hline
\textbf{1.} & $Y \independent \hat{Y} | C$ \quad  & model exclusively driven by confounder & $Q(\cdot|C)$ & $\hat{Y}$ or $Y$ \\
%\hline
\textbf{2.} & $Y \independent C | \hat{Y}$ \quad & model captures all variance in confounder & $Q(\cdot|\hat{Y})$ & $Y$ or $C$ \\
%\hline
\textbf{3.} & $\hat{Y} \independent C | Y$  \quad &  model does not capture more confound than what is present in the target & $Q(\cdot|Y)$ & $\hat{Y}$ or $C$ \\
\end{tabular}
\caption{\label{tab:conditional-independence-cases} Possibilities when testing conditional independence in potentially biased predictive models. \\The table lists the three possible null hypotheses (H0), and the variables where assumption about the joint/conditional distributions is required/not required.   ($Y$: prediction target, $\hat{Y}$: predictions, $C$: confounder variable) }
\end{table}

As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is the predictor(s) (features), $Y$ is the target variable and $Z$ is a potential confounder. In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether $X$ affects $Y$, when controlling for $Z$.
For instance, in genome-wide association studies, CPT can be used o determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.

Here I consider a different setting, where supervised learning model is already fitted and we are focusing on model diagnostics, by testing the triplet $(Y,\hat{Y}, C)$, where $Y$ is the target variable, $\hat{Y}$ is the model's prediction for Y, an $C$ is the (possibly multivariate) confounder. In such a setting, we are interested in how much variance the fitted model captures from the confounder $C$ or, in other words, to what extent the model is biased by the confounder variable $C$ when predicting $Y$ with $\hat{Y}$. The conceptual difference between the original and the proposed application of CPT is depicted on Fig. \ref{fig:schematic}.

Conditional independence testing and, specifically, the framework of conditional permutation testing allows investigating three different null hypotheses corresponding to the $(Y, \hat{y}, C)$ triplet, as listed in Table \ref{tab:conditional-independence-cases}. Testing the null hypothesis $Y \independent \hat{Y} | C$ (option 1) investigates whether the model predictions are likely explainable solely with the confounder, i.e. whether the model is exclusively confounder-driven. Testing $Y \independent C | \hat{Y}$ (option 2) aims the question whether the model captures all the variance in $C$ when predicting $Y$. Testing the null hypothesis $\hat{Y} \independent C | Y$ (option 3) examines, whether the dependence of the model output on the confounder can likely be explained by the confounder's dependence on the target variable, i.e. whether there is any confounder bias in the model.

While option 3. is typically of interest when testing confounder bias of predictive models, option 1 might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction. Option 2 seems less appealing for model diagnostics and as detailed below, the proposed variety of the CPT framework does not allow forming a non-parametric method in this case.

Obtaining a relatively accurate, independent estimate of $D(\cdot|Z)$ (of any shape) for CPT inference is important. Berrett and colleagues recommend to use a large independent sample to obtain the log-likelihood matrix of the conditional distribution $D(\cdot|Z)$, which might be often viable in semi-supervised learning settings where unlabelled data $(X, Z)$ are easier to obtain than labelled data $(X, Y, Z)$.
Since in the proposed predictive modelling setting normality might be often a reasonable assumption for $Y$ and $C$ (but not for $\hat{Y}$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|Z)$.

Here I propose to fit a linear regression model (or in case of categorical X, a logistic regression model):

\begin{equation}
    \label{eq:linreg}
    X = \alpha + \beta Z + \epsilon
\end{equation}

with the assumption that the residual error follows a Gaussian distribution and let $\mu = \alpha + \beta Z$ and $\sigma^2 = E(\epsilon^2)$. Then, for the conditional Normal distribution $ (X|Z=z) \sim \mathcal{N}\{\mu(z), \sigma^2(z)\}$ and the log-likelihood, that is to be used in Eq. \ref{eq-pperm}, can be computed as follows:

$$ \mathcal{L}(X|Z) = -{X^2}^T * \frac{1}{2} \sigma^2 + X^T \frac{\mu}{\sigma^2}$$

where $T$ means transposing row vectors to column vectors.

Applying this Gaussian likelihood estimation in the case predictive model diagnostics requires that the two variables in Eq. \ref{eq:linreg} are substituted with $Y$ and $C$, but not $\hat{Y}$, for which the Gaussian assumption of the proposed likelihood estimation might not hold.

Therefore, from the three options for conditional independence-based null hypotheses, enumerated in Table \ref{tab:conditional-independence-cases}, this is possible for options 1 and 3, but not option 2, as the variable on which the independence is conditional must be always the predictor variable in Eq. \ref{eq:linreg}.

For options 1 and 3, the proposed simple Gaussian conditional likelihood estimation allows for deploying CPT without a large, independent sample to estimate the conditional distribution.

---------------------------------

Todo: test statistic: $R^2$

The validity of the type I error control of CPT has been justified when using a true $D(\cdot|Z)$. While \cite{berrett2020conditional} gave bounds on the validity of the test given deviations from the true conditional distribution (which proved to be relatively tolerant against deviations from the true distribution), these bounds assume that the $D(\cdot|Z)$ has been estimated from an independent sample.

To provide foundations for the use of the proposed method we perfromed simulations...

-----------------------------

% tab:substitution
\renewcommand{\arraystretch}{2}
\begin{table}[]
\centering
\begin{tabular}{rl|l|l|l}
original & CPT & $X$ & $Y$ & $Z$ \\
\hline
%\hline
option 1: & full confounder test & $Y$ & $\hat{Y}$ & $C$ \\
%\hline
option 3: & partial confounder test & $C$ & $\hat{Y}$ & $Y$ \\
\end{tabular}
\caption{\label{tab:substitution} Substitutions of the symbols describing the variables in the original formulation of the CPT method and its two possible applications in predictive modelling diagnostics. }
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Validation on simulated data}

Using CPT to test confounder-bias in predictive modelling allows relaxing assumptions on $\hat{Y}$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions - on the joint distribution of the other two variables ($Y$ and $C$). 
\cite{berrett2020conditional} give a detailed analysis of the robustness of their CPT approach when estimating the conditional distribution with re-using the tested data and, also, against misspecifying the conditional distribution of interest.

Here I extend their results by performing validation that aids estimating the power of the test in a form that is accessible for power calculations in predictive modelling (considering various weights of the confounder and the target signals in $\hat{y}$ and in the case of varying levels of covariance between $Y$ and $C$).
Moreover, I perform a thorough investigation of the robustness of the test against the violation of normality (for both $(Y, C)$ and $\hat{Y}$).

Simulations are performed separately for the 'partial' and 'full' conditional confound tests.

\subsubsection*{Gaussian Simulations}
The target variable $Y$ and the confounder variable $C$ were drawn randomly from a multivariate normal distribution:

$$ Y, C \sim \mathcal{N}(\mu_0, \Sigma_\sigma) $$,

where $\mu=(0, 0)$ and the covariance matrix is given as:
$\Sigma_\sigma = \big(\begin{smallmatrix}
  1 & \sigma\\
  \sigma & 1
\end{smallmatrix}\big)$.

Next, we construct the simulated predicted values:

$$ \hat{Y} = w_Y Y + w_C C + w_e e$$

so that $w_Y + w_C + w_e = 1$, where $e \sim \mathcal{N}(0,1)$ is the noise component.

To test the implementation for categorical variables, simulated $Y$, $\hat{Y}$ and $C$ variables are binarized by thresholding at 0.

For more straightforward interpretation and visualization of the simulation results, the 'confound-to-target' and the 'signal-in-prediction' parameters were defined as

$$ w_{X/Y} = w_C / w_Y $$

and

$$ w_{C+Y} = w_C + w_Y $$,

respectively.

\subsubsection*{Introducing non-normality}

Non-normality is introduced by applying the \emph{sinh-arcsinh} transformation of \cite{jones2009sinh} on the simulated variables:

$$X' = sinh(\delta sinh^{-1}(X) - \epsilon)$$

where the parameters $\delta$ and $\epsilon$ control the kurtosis and skewness of the resulting \emph{sinh-arcsinh} distribution, with $\delta=1$ and $\epsilon=0$ producing the identity function (i.e. no non-normality introduced).

\subsubsection*{Simulation cases}

In the simulation procedure for the 'partial' confound test, 100 repetitions were performed of all combination of the following parameter values: 
\begin{itemize}
    \item covariance of $C$ and $Y$: $\sigma \in \{0, 0.2, 0.4, 0.6, 0.8\}$
    \item confound-to-target: $w_{C/Y} \in \{0, 0.1, 0.3, 0.5, 1\}$
    \item signal-in-prediction: $w_{C+Y} \in \{0, 0.3, 0.6, 0.9\}$
    \item sample size: $N \in \{50, 100, 500, 1000\}$
\end{itemize}

For testing 'full' confounder bias, identical parameters were used, except that instead of the 'confound-to-target' parameter $ w_{C/Y}$ we used the 'target-to-confound' parameter $ w_{Y/C} = w_Y / w_C $, but with the same parameter values $w_{Y/C} \in \{0, 0.1, 0.3, 0.5, 1\}$.

Next to the normally distributed case, 4 \emph{sinh-arcsinh} transformed cases were also investigated, with $\delta \in \{1, 1.05, 1.5, 5\}$ $\epsilon \in \{1, 3, 5, 10\}$

The 'partial' and 'full' confound tests, as implemented in version 0.7.1 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.

Analysis code used for simulations is available\footnote{https://github.com/pni-lab/mlconfound/blob/master/validation/simulation.py} in the github repository of the 'mlconfound' package.

\subsection{Validation on functional brain connectivity data}

The usefulness of the proposed confounder tests is demonstrated by applying them for predictive classification and regression models based on functional brain connectivity data, with different confound-mitigation approaches.

'Partial' and 'full' confound testing was used with default parameters (1000 permutations, 50 Markov-chain Monte Carlo steps) as implemented in version 0.7 of the package \emph{'mlconfound'}.

\subsubsection*{HCP: testing age- and acquisition batch-bias in fluid intelligence prediction}

Human Connectome Project contains imaging and behavioral data of more than 1000 healthy subjects \citep{van2013wu}. Preprocessed resting state fMRI connectivity data (partial correlation matrices) \citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores (IQ), as measured with Penn Progressive Matrices \citep{duncan2000neural}.

To ensure normality, fluid intelligence scores were non-linearly transformed to normal distribution with the quantile transformation \citep{beasley2009rank} as implemented in \emph{scikit-learn} \citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:hcp-hist} for details).

Features (functional concavities across 100 group independent component analysis based regions) were either (i) considered in their raw form or were subject to confound mitigation approaches by (ii) feature regression \citep{rao2017predictive} or (iii) COMBAT \citep{johnson2007adjusting, fortin2018harmonization}.
The feature mitigation strategies were separately applied for acquisition batch and age group as confounder variable.

The total of 5 types of features (raw, regressing out acquisition batch, regressing out age group, COMBAT with acquisition batch, COMBAT with age group) were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the individual fluid intelligence scores with a ridge-regression \citep{hoerl1970ridge}. The $\alpha$ parameter of the ridge was considered as a hyperparameter ($\alpha \in \{0.000001, 0.01, 0.1, 1, 10, 100, 1000000\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric. Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.

Analysis code is available as jupyter notebook as part\footnote{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb} of the github repository of the 'mlconfound' package.

\subsubsection*{ABIDE: testing motion- and center-bias in predictive models of autism spectrum disorder diagnosis}

The proposed method is applied to provide evidence of motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464. Preprocessed regional timeseries data was obtained as shared with the paper \citep{dadi2019benchmarking} which was based on preprocessed image data provided by the Preprocessed Connectome Project \citep{craddock2013neuro}.

Tangent correlation across the timeseries of the n=122 regions of the BASC \citep{bellec2010multi} atlas) was computed with nilearn\footnote{http://nilearn.github.io/} \citep{huntenburg2017loading, esteve2015big}.

The resulting functional connectivity estimates were considered as features either (i) in their raw form or after applying (ii) feature regression \citep{rao2017predictive} or (iii) COMBAT \citep{johnson2007adjusting, fortin2018harmonization}.
The investigated confounder variables were 'imaging center' and in-scanner motion, as measured by the mean framewise displacement of \cite{power2014methods}).
Mean FD was non-linearly transformed to normal distribution with the quantile transformation \citep{beasley2009rank} as implemented in \emph{scikit-learn} \citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:abide-hist} for details).

As COMBAT is not able to handle continuous variables (since it was primarily designed to remove "batch-effects"), motion was binned into 10 groups, based on equdistant data quantiles ranging from 0 to 1.

The total of 5 types (raw, feature regression of site, feature regression of motion, COMBAT with site, COMBAT with motion) of features were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the diagnosis (DX: ASD vs. HC) with a L2-regularized logistic regression (as e.g. in \citep{dadi2019benchmarking}). The $C$ parameter of the model was considered as a hyperparameter ($C \in \{0.1, 1, 10\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric. Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.

Analysis code is available as jupyter notebook\footnote{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_abide.ipynb} in the github repository of the 'mlconfound' package.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
  \caption{\textbf{Type-I error control and power of the conditional permutation-based confound testing for predictive modelling in normally distributed simulated data}. \\
  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette at the bottom) in various simulations settings (100 simulations per tile) with different covariance between the target and the confounder (given as $R^2_{y,c}$ on the vertical axes of the heatmaps), for different sample sizes (N, horizontal axes).
  Rows and columns correspond to simulations with different signal-in-prediction (y+c) and confound-to-signal values, respectively. The first row and the first column contain simulations under H0 (no confounder bias) and separated form the simulations under the alternative hypothesis with dashed line. Positive rates for the simulations under the null and alternative hypotheses can be interpreted as type I error rate and statistical power, respectively. The higher confidence limit for the type I error rates is 0.11 for each tile, with 5 tiles expected to pass it form all H0 tiles.
  The actual weights $w_{Y}$, $w_{C}$ and $w_{e}$ are depicted at the bars on top of each heatmap and color coded as shown in the legend at the bottom. Above each heatmap, the strengths of the associations of $\hat{y}$ to $Y$ and $C$, respectively, are given with $R^2$ values averaged over all simulation cases of the heatmap.
  }
  \label{fig:sim-normal}
\end{figure}

The conditional permutation based 'partial' and 'full' confound tests for predictive modelling have been implemented in the python package \emph{'mlconfound'}. Simulations showed valid type I control and a high, practically relevant statistical power in a wide variety of data characteristics. Analysis of real functional brain connectivity data demonstrates the usefulness of the tests in guiding confound-mitigation strategies.

\subsection{Simulations}

\subsubsection*{Type I error}

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.40\paperwidth]{fig/sim_non-norm.eps}
  \caption{\textbf{Robustness of conditional permutation based confound testing to the violation of normality assumptions for $Y$ and $C$.} \\
  Simulations included variables with five different degrees of non-normality (A), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation, from normal (left) to severely skewed and heavy tailed (right). Fisher's kurtosis and skewness is given for each at the bottom of panel A. Panel (B) depicts false positive rates in the simulations under H0, separately for different signal-in-confounder ratios (here identical to $w_{Y}$).
  Vertical axis shows type I error rate. The nominal type I error rate of $alpha=0.05$ is shown with a vertical dashed line. Panel (C) shows how violation of the normality assumption affects power. Boxplots show the mean power over simulation cases, grouped by confound-to-target ratio and sample size.}
  \label{fig:sim-non-normal}
\end{figure}

Simulations based on normally distributed data justify that the proposed confounder tests provide valid type I error control (For detailed results, see first row and first column on Figure \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-full}, \ref{fig:sim-ccb-partial}, \ref{fig:sim-ccb-full}, \ref{fig:sim-bbc-partial}, \ref{fig:sim-bbc-full}, \ref{fig:sim-bbb-partial} and \ref{fig:sim-bbc-full} for the partial test categorical data and the full tests on both numeric and categorical data). The higher confidence limit for the type I error rates for one simulation case with 1
00 repetitions is 0.11 (approximately 5 simulation cases are expected to pass this limit across all 200 H0 simulations on one figure).
The results suggest that the proposed tests are somewhat overly conservative if the model has a high predictive performance.

Investigations of the tolerance to violation of normality assumptions showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $Y$ and the confounder $C$ (and completely non-parametric regarding the distribution of the predictions $\hat{y}$). The six simulation cases of normality violation , shown on Figure \ref{fig:sim-non-normal} involved various \emph{sinh-arcsinh} distributions, form totally normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to a kurtosis and skewness values as high as 5 and 2 respectively. Furthermore, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).

\subsubsection*{Power}

When the normality assumptions hold, the test generally possess a statistical power of high practical relevance. Power estimates for the 'partial' test (with normality) is shown in Figure \ref{fig:sim-normal}. Notably, with sample sizes as large as 1000, a confounder contributing to only 0.4-0.6\% to the variance of the predictions can already be successfully detected ($w_c = 0.05 - 0.07$) with a power of 60-90\%. With a sample size of 500, a confounder contributing to only 1.4-3\% to the variance of the predictions can be detected ($w_c = 0.1 - 0.14$) with a power greater than 80\% in most of the simulation cases. A sample size of 100 yields an approximate power of 60-100\% for a confounder-variance of 6-7\% ($w_c = 0.2 - 0.21$) in most of the cases. Finally, even with a low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%. If the confoudner explains more than 15\% of variance, it is almost certainly detected with N>=50.

Violation of the normality assumptions for $Y$ and $C$ has a negative effect on the power of the tests, however the loss of power is low or even negligible, if normality is not extremely violated (first 4 distributions on Fig. \ref{fig:sim-non-normal}A) and the confounder-to-signal ratio is greater or equal then 0.3 (Fig. \ref{fig:sim-non-normal}C).

\subsection{Neuroimaging data}
\subsubsection*{HCP dataset}

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.eps}
  \caption{\textbf{Acquisition- and age-bias of fluid IQ prediction in the HCP dataset.} \\
  Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies. Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data. This bias is robustly detected by the proposed 'partial' confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
  Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the 'partial confound test'. P-values of the 'full' confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:hcp}
\end{figure}

Acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$a and $p=0.0014$ and $0.001$, respectively). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
However, as revealed by the partial confounder test the model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.

After applying the confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p < 0.05$ for all). Both feature regression and COMBAT slightly improved the predictive performance in the case of both confounders, with COMBAT providing better performances ($R^2=122$ and $121$ when applied to remove the effect of acquisition and age, respectively).

The full confounder test was highly significant ($p<0.0001$) for all models, indicating that neither of the models were exclusively driven by the confounds.

\subsubsection*{ABIDE dataset}

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.eps}
  \caption{\textbf{Center- and motion age-bias of ASD diagnosis prediction in the ABIDE dataset.} \\
  Boxplots and points show the predicted class probabilities (0: HC, 1: ASD), separately for the HC and ASD groups. In the top panel, predictions are plotted for each center separately and color indicates the true diagnosis (DX). AT the bottom plot, color indicates the normalized index of in-scanner motion (based on mean framewise displacement, mean FD). The proposed confounder test reveals significant center- and motion-bias in the model trained on the raw data (p<0.0001). While center bias seems to be effectively mitigated by feature regression, the same technique fails to remove motion-bias (actually introduces a paradoxical negative bias). COMBAT seems to effectively remove the effect of both confounders on the predictions and improves the predictive performance of the models in both cases.
  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the 'partial confound test'. P-values of the 'full' confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:abide}
\end{figure}

Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.027$ and $0.028$, respectively, $p<0.001$ for both). The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.095$, $ROC AUC = 0.71$, $p<0.001$).
However, as revealed by the partial confounder test the raw model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with several sites (e.g. Carnegine Mellon University or Caltech) and higher motion being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to a higher probability of the model to assign the participant to the ASD group.

After applying the confound mitigation approaches to remove the effect of center (feature regression or COMBAT) the partial confound test did not provide evidence of center-bias anymore ($p < 0.05$ for both techniques). When trying to mitigate the effect of in-scanner motion, feature regression failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates. The partial confounder test successfully detected this bias ($p<0.0001$). Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, effectively mitigated motion bias, as suggested by the partial confound test ($p=0.54$).

Both feature regression and COMBAT slightly improved the predictive performance when mitigating center-effects ($R^2=0.111$ and $0.132$ and $AUC=0.76$ and $0.75$, respectively). With both feature regression and COMBAT, the effort of mitigating motion effects happened at the cost of predictive performance ($R^2=0.009$ and $0.087$ and $AUC=0.59$ and $0.70$, respectively)

The full confounder test was highly significant ($p<0.0001$) for all models, indicating that neither of the models were exclusively driven by the confounds.

% tab:unconditional-pvals
\renewcommand{\arraystretch}{1.2}
\begin{table}
\centering
\begin{tabular}{lll|ll|ll|ll|ll} 
dataset & conf. & method & $R^2_{y, c}$ & $p_{y, c}$ & $R^2_{\hat{y}, c}$ & $p_{\hat{y}, c}$ & $R^2_{\hat{y}, y}$ & $p_{\hat{y}, y}$ & partial test & full test  \\
\hline
HCP & acq.  & raw      & 0.032 & <0.001 & 0.084 & <0.001  & 0.082 & <0.001 & <0.001 & <0.001 \\
    &              & f.reg.    & & &0.003 & 0.99 & 0.066  & <0.001  & 0.99 & <0.001 \\
    &              & COMBAT    & & &0.008 & 0.75 & 0.075  & <0.001  & 0.88 & <0.001\\
\hline
    & age   & raw       & 0.011 & 0.001  & 0.041 & <0.001  & 0.082  & <0.001 & <0.001 & <0.001 \\
    &       & f.reg.    & & &0.001 & 0.3 & 0.07  & <0.001  & 0.5 & <    0.001 \\
    &       & COMBAT    & & &0.004 & 0.055 & 0.076 & <0.001 & 0.12 & <0.001 \\
\hline
ABIDE   & center   & raw       & 0.005  & 0.203 &  0.169 & <0.001& 0.126     & <0.001 & <0.001 & <0.001 \\
        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & 0.04 & <0.001 \\
        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & 0.01 & <0.001 \\
\hline
        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & <0.001& <0.001 \\
        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & <0.001 & 0.09 \\
        &          & COMBAT    & & &  0.002 & 0.202 & 0.111     & <0.001 & 0.64 & <0.001 \\

\end{tabular}
\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination and the corresponding p-values and the p-values of the partial and full confounder tests for all investigated models.}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models. Testing confoudner-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confoudner variable. These two scenarios result in two different test, testing whether the model is partially or fully driven by the given confounder.
Importantly, in predictive modelling, the model predictions are often follow a non-normal distribution, thus the any test for confounder bias must be, at least partly, non-parametric.
However, as shown by in \citet{shah2020hardness} their the 'no free lunch' theorem, it is effectively impossible to establishing a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
Here I propose the use of the Markov-chain Monte-Carlo based conditional permutation test (CPT) of \cite{berrett2020conditional} to test confoudner bias in predictive models. Adapting this approach for the confoudner testing problem makes it possible to place assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confoudner tests so that no assumption is made about the distribution of the model output. 
Here I propose a computationally effective parallel implementation of CPT that utilizes the parallel pairwise sampler of \cite{berrett2020conditional}, coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confoudner or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confoudner is normal. The implementation provides a test that can be uniformly applied for any predictive model, given that realistic (e.g. cross-validated) model predictions can be obtained.

The validity and the power of the tests has been investigated in simulations with normally distributed data and with various degrees of non-normality.

The simulations have confirmed that the proposed implementation provides a valid type I error control for both the full and the partial confound test and revealed that both the tests have a high power for various configurations of the association between the target and the confounder, predictive performance, sample size and the confound-to-signal for the partial or signal-to-confound ratios for the full test. 
Both tests become somewhat conservative (i.e. lower than expected type I error rates ) in the case of very powerful predictive models (when the $R^2$ between the predicted and observed values is greater than 0.7). However, the power of the test remains high with sufficiently large ($\ge 500$) sample sizes.

Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confoudner bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confoudner explaining 1-2\% of the variance in the prediction). A currently somewhat more common sample size of 100 will very likely identify medium sized confoudner-biases (around 9\%). Strong biases  (>25\%) can be very robustly detected with sample sizes as low as 50.

As expected, the power of the full confoudner test is very similar to the partial test, when investigating the contribution of the target to the prediction, instead of the confounder.

While different biomedical applications might consider different amounts of bias to be relevant, an a-priory power calculation can help identifying the necessary sample size in order to exclude the possibility of a relevant confoudner bias with a high probability.

Importantly, the simulations strongly suggest that both tests are very robust to the violation of the normality assumption for the conditional distribution between the target and the confoudner. Type I error is well controlled even in case of highly skewed and heavy tailed distributions (kurtosis > 5, skewness > 2) and gets significantly inflated only with extremely degenerate distributions (see the purple distribution on Fig. \ref{fig:sim-non-normal}). Due to the conservatives of the tests in case of very accurate models, even this extremely skewed and heavy tailed distribution resulted in lower than the nominal type I error rates if the predictions were strongly associated to the target variable.

Non-normality has a negative effect on the power of the both tests, however, the decrease in power is in many simulation cases negligible and gets only relevant if the sample size and/or the confounder-effect is low.

Given the power and the robustness of the proposed tests, they might become a useful tool in diagnosing predictive models, not exclusively in the biomedical filed.

In such studies, confounder-bias has often been investigated (if investigated at all) by testing the association between the confoudner variable and the predictive values, see e.g. \citep{spisak2020pain}.
However, the significance of this association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable. In this case, namely, the model still might not be directly driven by the confounder, i.e. the dependence of the predictions on the confounder can be explained solely by the confoudner-target association. 
Arguably, in certain applications, the researcher might aim to fully eliminate confoudner-related variance of the predictions and in this case testing the association (unconditional independence) between the confoudner and the predictions might be sufficient. However, a much more realistic aim is to construct models that allow a reasonable degree of confoudner-association of the predictions, but are likely not directly driven by confounder-information, i.e. the prediction-confoudner association is proportional to, and can likely be fully explained by, the target-confoudner association.
Put simply, this is what is tested by the proposed partial confounder test.

Obviously, a true confoudner bias might still be present in the model, even if the the target-confoudner association likely explains the prediction-confoudner, i.e. the model might still use the confoudner effect in the features (e.g artifacts) to present that pert of the variance in the prediction. Nevertheless, in most of the cases, it does not seem likely that exclusively those confoudner components are captured the model that are needed to mimic the unbiased model and it might be often a much more plausible explanation that the confounder-variance is simply not captured by the model (or it was effectively mitigated by feature cleaning/harmonization).

Among the many potential areas of applications, the proposed tests could be certainly useful in translational neuroscience. Applying predictive modelling and machine learning, for instance, on functional neuroimaging data holds a great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools \citep{woo2017building, wager2013fmri, spisak2020pain}. However, the presence of confounders typical for biomedical research (e.g. sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g. imaging artifacts) present a great challenge to these efforts and are, currently, often neglected.

To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two research settings where specific types of confounding effects have previously been described. 
First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effects of age \cite{lohmann2021predicting, dubois2018distributed} and also a somewhat ignored aspect, the acquisition data within the course of the data acquisition process.
Second, functional connectivity data from the ABIDE \citep{di2014autism} was used to investigate the potential motion- and center-bias (as previously reported by e.g. \cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.

When trained on the raw (unadjusted) connectivity features, in all four cases, the confounders were statistically significantly associated to both the target variable and the model predictions (with weak-to-moderate effect sizes) and the proposed partial confounder test provided significant evidence against the null-hypothesis of no confounder bias (i.e. conditional independence between the confoudner and the predictions, given the target).

The detected confounder bias is most probably of practical relevance for all four confounders.

In case of the HCP dataset, the weak, but statistically significant association between age and IQ predictions could likely exaggerate to a serious bias when testing the model on data of participants outside of the age range of the training sample (that is, the bias could likely significantly harm out-of-sample generalizability). The bias of the model for the acquisition batch (phase of the study the acquisition belongs to) can also be problematic and - to my best knowledge - it was not yet reported in the HCP dataset. There can be manifold reasons for the observed associations. Fluid intelligence of the participants can be affected by a change of selection bias during requirement (e.g. as the human connectome project got more public interest during its course the the sociodemographic and educational characteristics of the volunteer sample might have changed). The model can utilise changing characteristics of the data (e.g. gradual changes in the calibration settings of the measuring equipment or the changing level of experience/motivation of the research staff) to predict this part of the effect. This is obviously undesirable for any brain-based predictive model of fluid intelligence.

In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously undesired for a diagnostic model of autism. For instance the model trained on the raw features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g. the CMU (Carnegie Mellon University) center as neurotypical control participants. It's easy to see that this bias can possibly lead to a very low predictive performance on out-of-center data. Similarly, a model biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with higher motion or altered motor characteristics (as known e.g. for ADHD \cite{eloyan2012automated}).

The investigated confound regression techniques, feature regression  \citep{spisak2014voxel, dukart2011age, abdulkadir2014reduction, rao2017predictive} and COMBAT \citep{johnson2007adjusting}, did not fully eliminate the association (unconditional dependence) between the confoudner and the predictions (neither for the HCP nor for the ABIDE dataset, all the corresponding p-values were smaller or equal than 0.003). However, as discussed above, fully eliminating this association is not necessarily a requirement for a model without confounder-bias. Indeed, the partial confoudner test revealed that except one, all models become likely free of confounder bias or - more precisely - the bias was suppressed to such an extent that the test was not able to detect it anymore. Given the high power of the test at these sample sizes (N=999 and 866 for the HCP and the ABIDE datasets, respectively), any confoudner remaining, but not detectable confounder bias is most probably negligible or is not of practical relevance.
The exception, where confound mitigation did not successfully eliminate confoudner-bias, is confoudner regression of motion in the ABIDE dataset. Interestingly, in this case, the association between the confoudner and the predictions become stronger but with an opposite sign, while the prediction performance significantly dropped. COMBAT, on the other hand still effectively mitigated confound-bias in this case, although with a slight decrease of predictive performance, as well. A possible explanation to these phenomena is that in scanner motion is not simply a confounder but, as previously described \citep{fournier2010motor, anzulewicz2016toward}, it might have a complex relationship to ASD. Therefore, regressing out motion estimates from the connectivity features might eliminate significant amount of signal-of-interest, especially in functional connections that are associated with motor deficits in ASD. 
In such cases, applying feature regression might be a too "aggressive" solution. As COMBAT was originally developed for harmonizing effects of categorical variables in the dataset (e.g. center or batch), our method of inputting the binned "motion groups" into COMBAT - together with the ability of the empirical Bayes method to incorporate information about sample variances - might have resulted in "less aggressive" data harmonization and preserved more signal-of-interest. 
The behavior of existing confound elimination techniques in such situations and the applicability of COMBAT with numeric confiders is an interesting target of further research, where the proposed conditional permutation based confoudner tests can be deployed.

The full confoudner test - as expected - confirmed that none of the model is not fully driven by the confoudner variable or, in other words, the target and the prediction were not independent, given the confounders. While the full confoudner test might seem less useful in the presented examples, it might become useful in exploratory efforts, where the questions is simply if it is possible at all to train an unbiased predictive model.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}

The lack of rigorous statistical tests of confoudner-bias significantly hampers the development of predictive models in many fields of research. Even tough several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confoudner and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias in the resulting model.

To fill this critical gap in predictive model development, I propose two novel tests, the partial and the full confounder tests, which test the null hypotheses of 'no confounder-bias' of 'full confounder-bias', respectively. Both approaches are based on a recently proposed conditional permutation framework, with a mathematical justification of valid type I error control. The tests were constructed so that they do not place any assumption about the distribution of the model predictions, thereby being applicable with arbitrary machine learning algorithms.

Given the high statistical power and the high robustness to the violation of the assumptions on the data (normality of the conditional distribution of the target and the confoudner variables), the tests can be useful in a wide variety of applications.

As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of confound mitigation and expected to improve the assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research, thereby paving the way for the development of clinically useful machine learning biomarkers.

\bibliographystyle{apalike}  
\bibliography{references}

\newpage
\input{supplement}

\end{document}

### 2021-08-09 manuscript.tex
figures refreshed

@@
 \begin{abstract} % 150 words
-The lack of rigorous non-parametric statistical tests of confoudner-effects significantly hampers the development of robust, valid ans generalizable predictive models in many fields of research.
-Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recent theoretical framework of conditional independence testing and test the null hypothesis of \emph{no bias} and \emph{fully biased model}, respectively.
+The lack of rigorous non-parametric statistical tests of confoudner-effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
+Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recent theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unbiased} and \emph{fully biased model}, respectively.
 The proposed tests set no assumptions about the distribution of the predictive model output that is often non-normal.
-As shown by simulations, the test are statistically valid, robust and display a high statistical power.
-Applying the tests on models based on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange databases shows that the proposed tests can significantly improve the assessment of the robustness, generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
-
-The tests are implemented in the python package \emph{mlconfound}\footnote{https://github.com/pni-lab/mlconfound}.
+Simulations suggests that the tests are remarkably robust for the violation of the assumptions and display a high statistical power.
+Applying the proposed tests on models based on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange databases reveals significant confoudner-biases.
+Some of these biases were previously unreported and hard to mitigate with state-of-the-art confound mitigation approaches.
+The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
 \end{abstract}
@@
   \centering
-  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.eps}
+  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.pdf}
   \caption{\textbf{Acquisition- and age-bias of fluid IQ prediction in the HCP dataset.} \\
@@
   \centering
-  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.eps}
+  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.pdf}
   \caption{\textbf{Center- and motion age-bias of ASD diagnosis prediction in the ABIDE dataset.} \\
@@
 \end{tabular}
-\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination and the corresponding p-values and the p-values of the partial and full confounder tests for all investigated models.}
+\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated models.}
 \end{table}

### 2021-08-09 manuscript.tex
polished

@@
 The lack of rigorous non-parametric statistical tests of confoudner-effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
-Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recent theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unbiased} and \emph{fully biased model}, respectively.
+Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unbiased} and \emph{fully biased model}, respectively.
 The proposed tests set no assumptions about the distribution of the predictive model output that is often non-normal.
-Simulations suggests that the tests are remarkably robust for the violation of the assumptions and display a high statistical power.
-Applying the proposed tests on models based on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange databases reveals significant confoudner-biases.
-Some of these biases were previously unreported and hard to mitigate with state-of-the-art confound mitigation approaches.
+Simulations show that the tests are applicable in a wide variety of cases with a remarkable statistical power.
+Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset identified significantly biased predictions.
+Some of these biases were previously unreported and found to be hard to mitigate with state-of-the-art confound mitigation approaches.
 The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
@@
 Predictive modelling and supervised learning methods have recently became increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized}.
-The usefulness and applicability of the such markers is to be evaluated not solely on the biomedical relevance of the predictive performance (i.e.
-prognostic/diagnostic value); generalizability across contexts and populations and biomedical validity (i.e.
-the model is driven by biomedical relevant signal) are also highly important requirements towards potential biomarkers \citep{woo2017building}.
+When evaluate the usefulness and applicability of such markers, predictive performance (i.e.
+prognostic/diagnostic value) i snot the only consideration.
+Generalizability across contexts and populations and biomedical validity (i.e.
+whether the model is driven by biomedical relevant signal) are also highly important requirements for candidate biomarkers \citep{woo2017building}.
 
@@
 First, measurement artifacts can obviously be considered as confounders.
-For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with movement kinematics or motor function, like previously demonstrated, among others, in case of e.g Alzheimer's disease \citep{rao2017predictive} or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}.
-Similarly, eye-blink artifacts can result in confound-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.
+For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with movement kinematics or motor function, like previously demonstrated, among others, in case of e.g Alzheimer's disease \citep{rao2017predictive} attention deficit hyperactivity disorder (ADHD) \citep{couvy2016head, eloyan2012automated}er or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}.
+Similarly, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.
 Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling.
-E.g.
-resting-state functional connectivity is known to be strongly predictive to  demographic age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}.
-As fluid intelligence is known to decline with aging, \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by simply picking up on age-related variance \cite{lohmann2021predicting, dubois2018distributed}.
-Third, center-effects (or in general, batch effect) in multi-center studies can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance}.
-In this case, predictive models can utilize center-effects to explain (random or true) center-differences in the predictive target.
-Such models will display worse-than-expected performance when applied on data from new centers.
+For instance, resting-state functional connectivity is known to be strongly predictive to  demographic age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}.
+As fluid intelligence is known to decline with aging, \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by simply picking up on age-related variance \citep{lohmann2021predicting, dubois2018distributed}.
+Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance}.
+In this case, predictive models can utilize center/batch-effects to explain (random or true) center/batch-differences in the predictive target.
+Such models will display worse-than-expected performance when applied on data from new centers/batches.
 
-Recently, various data harmonization approaches have been suggested to mitigate confounder-bias, ranging from simple matching \citep{rao2017predictive} or regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} or sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}).
+Recently, various data harmonization approaches have been suggested to mitigate confounder-bias, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}.
 However, it is often unclear which variables should be considered as confounders during harmonization and considering many variables as confounder holds risks of eliminating signal-of-interest \citep{wachinger2021detect}.
 
-Powerful statistical tests to quantify the confounding effect of certain variables in predictive models could largely foster both the the identification of confounders to correct for and the assessment of the effectiveness of various harmonization or other confound-mitigation approaches.
-Such methods, however, have to tackle the fact that the outputs of predictive models often have non-normal distributions (due to e.g.
+Powerful statistical tests to quantify the confounding effect of certain variables in predictive models could largely foster both the the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation and data harmonization approaches.
+Such methods, however, have to tackle the fact that the outputs of predictive models often follow non-normal distributions (due to e.g.
 regularization) \citep{garcia2009study, kristensen2017whole}.
@@
 Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high computational complexity (especially when hyperparameters are optimized in a nested cross-validation).
-The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the the predictor(s) are a direct cause of the target variable or, alternatively, there exists an unobserved random variable that is the cause of both.
+The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the the predictors are a direct cause of the target variable or, alternatively, there exists an unobserved random variable that is the cause of both.
 It's computation assumes 'faithfulness' (if two variables are independent, there is no direct influence between the two in the underlying graph) a strong condition that might not be fulfilled in many real applications (e.g.
-functional MRI is an \emph{indirect} measure of neural activation, rendering the alternative hypothesis true in absence of any real confounder).
+functional MRI is an \emph{indirect} measure of neural activation, rendering the alternative hypothesis of this test true in absence of any real confounder).
 
@@
 
-Next, I consider the conditional permutation test, a statistical test of conditional independence, recently introduced by \cite{berrett2020conditional}, as a test for evaluating confounder-bias of predictive models.
-
-Within this framework, various confounder-related null hypotheses can be tested, so that no assumption is to be made about the distribution of one of the conditioned variables.
-As normality is typically a reasonable assumption for the target and confounder variables but not for the predictions of the model, the focus is put on the cases allowing inference without assumptions on the model predictions' distribution.
-Two cases satisfy this requirement, one testing whether the model's explained variance can be exclusively attributed to the confounder and the other testing whether the model utilizes any confounder-variance, above what's already present in the target variable.
-Both tests require only on the target variable, the model predictions and the putative confounder, that is, they can be performed without needing to re-fit the model.
-Additionally, a regression-based estimation of the conditional distribution of $c|y$ is proposed, which allows testing confounder bias without a-priory knowledge about the conditional distribution and without any assumptions on the distribution of the predicted values (which are often non-normal in practice).
+Next, I utilize the permutation-based conditional independence framework of \cite{berrett2020conditional}, to construct two novel tests for evaluating the confounder-bias in predictive models.
+The \emph{full confounder test} probes whether the model's explained variance can be attributed exclusively to the confounder
+The \emph{partial confounder test} investigates whether the model utilizes any confounder-variance, above what is already present in the target variable.
+Both tests require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed without needing to re-fit the model.
+The tests set no assumption about the distribution of the model predictions.
 The validity of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normal model output).
@@
 
-In a predictive modelling setting, let $Y$ denote the target variable, $\hat{Y}$ denote model output, i.e.
-the predictions for $Y$ and let $C$ denote a variable, which is considered as a confounder.
-Depending on the research question, the influence of $C$ on the model predictions $\hat{Y}$ is subject to be kept at a negligible level, or it is to be proven that the model is not completely driven by $C$.
-Below, I summarize how testing conditional independence \citep{dawid1979conditional} among these variables allows addressing the above questions and build on the conditional permutation test of \cite{berrett2020conditional} to perform inference without assumptions on the distribution of the model output, where normality is often violated \citep{garcia2009study, kristensen2017whole}.
+In a predictive modelling setting, let $\y$ denote the target variable, $\yhat$ denote model output, i.e.
+the predictions for $\y$ and let $\c$ denote a variable which is considered as a confounder.
+Depending on the research question, the influence of $\c$ on the model predictions $\yhat$ is subject to be kept at a negligible level or it is to be proven that, at least, the model is not completely driven by $\c$.
+Below, I summarize how testing conditional independence \citep{dawid1979conditional} across these variables allows addressing the above questions and build on the conditional permutation test of \cite{berrett2020conditional} to perform inference without assumptions on the distribution of the model output, where normality is often violated \citep{garcia2009study, kristensen2017whole}.
 
-\subsubsection*{Testing conditional independence}
+\subsubsection*{Conditional independence}
 
-Let us shall assume that the data consist of independent and identically distributed (IID) data points $(x_i, y_i, z_i) \in X \times Y \times Z$ for $i=1, \dots , n$ and write $X=(x_1, \dots ,x_n)$, $Y=(y_1, \dots, y_n)$ and $Z=(Z_1, \dots, Zn)$.
+Let us assume that $\y$, $\yhat$ and $\c$ consist of independent and identically distributed (IID) data points $(y_i, \hat{y}_i, c_i)$ for $i=1, \dots , n$ and write $\y=(y_1, \dots ,y_n)$, $\yhat=(\hat{y}_1, \dots, \hat{y}_n)$ and $\c=(c_1, \dots, c_n)$.
 
-$X$ is \emph{conditionally independent} of $Y$ given $Z$ $X \independent Y | Z$, if for all $x, y, z$ with $p(z) > 0$, we have $p(x, y|z) = p(x|z)p(y|z)$ \citep{dawid1979conditional}.
+Obviously, a strong association between $\yhat$ and $\c$ can imply that the model is biased; it's predictions are driven by the confounder rather than information about the target variable.
 
-While conditional independence lies at the heart of several fundamental concepts is statistics and  plays an increasingly important role in various fields of applied statistics (and particularly in biomedical applications, \citep{spirtes2000causation, peters2016causal, fiedler2011mediation, candes2016panning}, recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
-In their "no free luch" theorem, Shah and Peters has shown that, without placing some assumptions on the joint distribution of $(X, Y, Z)$, conditional testing is effectively impossible.
-In other words, no test exists that - for all distributions - provides a valid type I error control and, at the same time, has a non-trivial statistical power.
+Testing unconditional independence between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confoudner-bias in predictive modelling.
+For instance, even if the null hypothesis $H0 : \yhat \independent \c$ is false, $\yhat$ and $\c$ might be only marginally dependent on $\c$, due to their dependence on $\y$.
+In other words, if the target variable $\y$ is associated to the confoudner variable $\c$, a model that is completely blind to $\c$ (and, therefore, unbiased) might still provide outputs $\yhat$ that are significantly associated with $\c$.
 
-This result is in strong contrast to unconditional independence testing, for which permutation tests  \citep{pitman1937significance, fisher1942189} allow for the correct calibration of any hypothesis test.
-Even if the null hypothesis $H0 :X \independent Y|Z$ is true, $X$ and $Y$ might be highly marginally dependent due to their dependence on $Z$,  thus the permutation null-distribution will be invalid.
+Our primary interest is, therefore, the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ ($\yhat \independent \c | \y$) which, according to \cite{dawid1979conditional}, implies that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(X|Z)\mathbb{P}(Y|Z)$.
+The test with this null hypothesis will be referred to as the \emph{partial confounder test}.
 
-The "no-free-lunch" theorem has important implications for confound-testing in predictive modelling where the distribution of the model outputs - depending on the applied machine learning model - is \emph{unknown}.
+Alternatively, one might be interested in testing $\yhat \independent \c | \y$. We will refer to the corresponding test as the \emph{full confoudner test}.
 
-Nevertheless, as shown by \cite{berrett2020conditional}, if information about the conditional distribution $Q(X|Z)$ is available, sampling from a non-uniform distribution over the set of possible permutations $\pi$ can incorporate available distributional information into permutation-based inference and achieve a valid type I error control and a non-trivial power.
-Below we give a short description of the procedure of \cite{berrett2020conditional}.
+ Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications \citep{spirtes2000causation, peters2016causal, fiedler2011mediation, candes2016panning}.
+Recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
+ Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional Independence testing is effectively impossible.
+In other words, neither the full nor the partial confoudner tests can be constructed so that - for all distributions - they provides a valid type I error control and, at the same time, a non-trivial statistical power.
 
-\subsubsection*{The conditional permutation test for independence}
+This result is in strong contrast to unconditional independence testing - where permutation tests  \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs - depending on the applied machine learning model - is unknown and often heavily non-normal.
 
-The conditional permutation test (CPT) of \cite{berrett2020conditional} build on the conditional randomization test of \cite{candes2016panning} and ensures that \emph{no assumptions} are made about the distribution of the response variable Y but, instead, it is assumed that the conditional distribution of X given Z is known.
+Nevertheless, as shown by \cite{berrett2020conditional}, conditional independence testing is possible with information only about the conditional distribution of two out of the three variables.
+In this case, sampling from a non-uniform distribution over the set of possible permutations $\pi$ of the one of the variables can incorporate available distributional information into permutation-based inference and achieve a valid type I error control and a non-trivial power.
+Below we adapt the formalization of the procedure of \cite{berrett2020conditional} for the confoudner problem.
 
-CPT generates a null-distribution for an arbitrary predefined test statistic $T(X,Y,Z)$ by sampling permutation based "copies" of $X$,
-
-$$x_i^{(j)} \sim Q(\cdot|z_i) \ $$
-
-where, $Q(.|z)$ denotes the conditional distribution of $X$ given $Z=z$ and $j=1,\dots, m$ indexes the copies of $X$ and so that so that $X^{(i)} = (x_1^{(j)}, \dots, x_n^{(j)})$ is a permutation of the original vector $X = (x_1, \dots, x_n)$.
-This mechanism creates copies $X^{(1)}, \dots ,X^{(m)}$ that are exchangeable with the original vector X under the null hypothesis that $X \independent Y | Z$.
-
-Under the null hypothesis, the triples $(X,Y,Z), (X^{(1)},Y,Z),\dots,(X^{(m)},Y,Z)$ are all identically distributed and exchangeable, and so are the 
-$$T(X,Y,Z), T(X^{(1)},Y,Z),\dots,T(X^{(m)},Y,Z)$$
-test statistics, as well.
-
-The p-value under the null hypothesis is then obtained as
-$$ p= 1+\frac{\sum_{j=1}^m \mathbb{1} \{T(X^{(j)},Y,Z) \geq T(X,Y,Z) \}  }{1+m}$$
-
-Let $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$ and $X_\pi = (x_{\pi_1}, \dots, x_{\pi_n})$ the vector $X$ with its elements reordered according to the permutation $\pi \in S_i$.
-In the CPT procedure, the permutation-based copies of $X$ are consist of $x_i^{(j)} = x_{\pi^{(j)}}$ which are drawn so that 
-so that:
-
-\begin{equation}
-    \mathbb{P}(\pi^{(j)} = \pi | X,Y,Z) = \frac{q^n(X_\pi | Z)}{\sum_{\pi' \in S_n} q^n(X_{\pi'} | Z)}
-    \label{eq-pperm}
-\end{equation}
-
-that is, according to the $q^n(\cdot|Z) := q(\cdot | z_1, \dots q(\cdot|z_n)$ product density corresponding of the conditional distribution $Q(\cdot|z)$.
-Note that Eq.
-\ref{eq-pperm} does not necessarily assume a continuous distribution.
-For a verification of the valid type I error control of this approach refer to Theorem 1 in \citep{berrett2020conditional}.
-
-While the required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$, at random, the acceptance ratio will be extremely low, even for moderate $n$ (except there is very low dependence of X on Z), resulting in slow mixing times.
-\cite{berrett2020conditional} (Algorithm 1) propose a parallelized pairwise Markov-Chain Monte Carlo sampler for CPT, that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data.
-In their Theorem 2, \cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.
-
-\subsection{Conditional Permutation-based confounder-bias test}
+\subsubsection*{The conditional permutation test}
 
@@
 % Text Node
-\draw (494,252.4) node [anchor=north west][inner sep=0.75pt]    {$C$};
+\draw (494,252.4) node [anchor=north west][inner sep=0.75pt]    {$\c$};
 % Text Node
-\draw (591.5,180.4) node [anchor=north west][inner sep=0.75pt]    {$\hat{Y}$};
+\draw (591.5,180.4) node [anchor=north west][inner sep=0.75pt]    {$\yhat$};
 % Text Node
-\draw (396,184.4) node [anchor=north west][inner sep=0.75pt]    {$Y$};
+\draw (396,184.4) node [anchor=north west][inner sep=0.75pt]    {$\y$};
 % Text Node
@@
 % Text Node
-\draw (308.5,247.4) node [anchor=north west][inner sep=0.75pt]    {$\hat{Y}$};
+\draw (308.5,247.4) node [anchor=north west][inner sep=0.75pt]    {$\yhat$};
 % Text Node
-\draw (191,185.4) node [anchor=north west][inner sep=0.75pt]    {$Y$};
+\draw (191,185.4) node [anchor=north west][inner sep=0.75pt]    {$\y$};
 % Text Node
@@
 % Text Node
-\draw (68,185.4) node [anchor=north west][inner sep=0.75pt]    {$C$};
+\draw (68,185.4) node [anchor=north west][inner sep=0.75pt]    {$\c$};
 
@@
 \\ \textbf{(A)} CPT was originally proposed to be used on the feature variable $X$, target variable $Y$ and confounders $Z$.
-\textbf{(B)} The proposed use of CPT in predictive modelling requires the model to be fitted first, to obtain the model's prediction $\hat{Y}$ on $Y$.
-CPT is then utilized on the triplet $(Y, \hat{Y}, Z)$, to test hypotheses $\hat{y} \independent c | y$ or $y \independent \hat{y} | c$.}
+\textbf{(B)} The proposed use of CPT in predictive modelling requires the model to be fitted first, to obtain the model's prediction $\yhat$ on $\y$.
+CPT is then utilized on the triplet $(\y, \yhat, \c)$, to test hypotheses $\yhat \independent \c | \y$ or $\y \independent \yhat | \c$.}
   \label{fig:schematic}
@@
 
+The conditional permutation test (CPT) of \cite{berrett2020conditional} builds on the conditional randomization test of \cite{candes2016panning} and allows that \emph{no assumptions} are made about the distribution of one of the variables but, instead, it is assumed that the conditional distribution of the other two variables is known.
+
+As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is the predictor(s) (features), $Y$ is the target variable and $Z$ is a potential confounder.
+In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether $X$ affects $Y$, when controlling for $Z$.
+For instance, in genome-wide association studies, CPT can be used o determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.
+
+Here, different setting is considered, where supervised learning model is already fitted and we are focusing on model diagnostics, by testing the triplet $(\y,\yhat, \c)$, as introduced above.
+In such a setting, we are interested in how much variance the fitted model captures from the confounder $C$ or, in other words, to what extent the model is biased by the confounder variable $\c$ when predicting $\y$ with $\yhat$.
+The conceptual difference between the original and the proposed application of CPT is depicted on Fig.
+\ref{fig:schematic}.
+
+Within this setting, conditional independence testing and, specifically, the framework of conditional permutation testing allows investigating three different null hypotheses corresponding to the $(\y, \yhat, \c)$ triplet, as listed in Table \ref{tab:conditional-independence-cases}.
+Testing the null hypothesis $\y \independent \yhat | \c$ (option 1) investigates whether the model predictions are likely explainable solely with the confounder, i.e.
+whether the model is exclusively confounder-driven.
+Testing $\y \independent \c | \yhat$ (option 2) addresses the question whether the model captures all the variance in $c$ when predicting $y$.
+Testing the null hypothesis $\yhat \independent \c | \y$ (option 3) examines, whether the dependence of the model output on the confounder can likely be explained by the confounder's dependence on the target variable, i.e.
+whether there is any confounder bias in the model.
+
 % tab:conditional-independence-cases
-\renewcommand{\arraystretch}{2}
+\renewcommand{\arraystretch}{1.2}
 \begin{table}[]
@@
 \hline
-\textbf{1.} & $Y \independent \hat{Y} | C$ \quad  & model exclusively driven by confounder & $Q(\cdot|C)$ & $\hat{Y}$ or $Y$ \\
+\textbf{1.} & $\y \independent \yhat | \c$ \quad  & model exclusively driven by confounder (full confounder test) & $Q(\cdot|\c)$ & $\yhat$ or $\y$ \\
 %\hline
-\textbf{2.} & $Y \independent C | \hat{Y}$ \quad & model captures all variance in confounder & $Q(\cdot|\hat{Y})$ & $Y$ or $C$ \\
+\textbf{2.} & $\y \independent \c | \yhat$ \quad & model captures all variance in confounder (not of interest) & $Q(\cdot|\yhat)$ & $\y$ or $\c$ \\
 %\hline
-\textbf{3.} & $\hat{Y} \independent C | Y$  \quad &  model does not capture more confound than what is present in the target & $Q(\cdot|Y)$ & $\hat{Y}$ or $C$ \\
+\textbf{3.} & $\yhat \independent \c | \y$  \quad &  model does not capture more confound than what is present in the target (partial test) & $Q(\cdot|\y)$ & $\yhat$ or $\c$ \\
 \end{tabular}
@@
 \\The table lists the three possible null hypotheses (H0), and the variables where assumption about the joint/conditional distributions is required/not required.
-($Y$: prediction target, $\hat{Y}$: predictions, $C$: confounder variable) }
+($y$: prediction target, $\yhat$: predictions, $\c$: confounder variable) }
 \end{table}
-
-As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is the predictor(s) (features), $Y$ is the target variable and $Z$ is a potential confounder.
-In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether $X$ affects $Y$, when controlling for $Z$.
-For instance, in genome-wide association studies, CPT can be used o determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.
-
-Here I consider a different setting, where supervised learning model is already fitted and we are focusing on model diagnostics, by testing the triplet $(Y,\hat{Y}, C)$, where $Y$ is the target variable, $\hat{Y}$ is the model's prediction for Y, an $C$ is the (possibly multivariate) confounder.
-In such a setting, we are interested in how much variance the fitted model captures from the confounder $C$ or, in other words, to what extent the model is biased by the confounder variable $C$ when predicting $Y$ with $\hat{Y}$.
-The conceptual difference between the original and the proposed application of CPT is depicted on Fig.
-\ref{fig:schematic}.
-
-Conditional independence testing and, specifically, the framework of conditional permutation testing allows investigating three different null hypotheses corresponding to the $(Y, \hat{y}, C)$ triplet, as listed in Table \ref{tab:conditional-independence-cases}.
-Testing the null hypothesis $Y \independent \hat{Y} | C$ (option 1) investigates whether the model predictions are likely explainable solely with the confounder, i.e.
-whether the model is exclusively confounder-driven.
-Testing $Y \independent C | \hat{Y}$ (option 2) aims the question whether the model captures all the variance in $C$ when predicting $Y$.
-Testing the null hypothesis $\hat{Y} \independent C | Y$ (option 3) examines, whether the dependence of the model output on the confounder can likely be explained by the confounder's dependence on the target variable, i.e.
-whether there is any confounder bias in the model.
 
@@
 
-Obtaining a relatively accurate, independent estimate of $D(\cdot|Z)$ (of any shape) for CPT inference is important.
+Below, CPT is re-formalized for option 3, i.e. partial confoudner testing. The formulation of the full confoudner test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
+
+\subsubsection*{The partial confoudner test}
+
+The partial confoudner test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based "copies" of $\c$,
+
+$$\c_i^{(j)} \sim Q(\cdot|y_i) \ $$
+
+where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y$ and $j=1,\dots, m$ indexes the "copy" of $\c$ so that $\c^{(i)} = (c_1^{(j)}, \dots, c_n^{(j)})$ is a permutation of the original vector $\c = (c_1, \dots, c_n)$.
+This mechanism creates copies $\c^{(1)}, \dots ,\c^{(m)}$ that are exchangeable with the original vector $\c$ under the null hypothesis that $\yhat \independent \c | \y$.
+
+Under the null hypothesis, the triples $$(\y,\yhat,\c), (\y, \yhat, \c^{(1)}),\dots, (\y, \yhat, \c^{(m)})$$ are all identically distributed and exchangeable, and so are the 
+$$T(\y,\yhat,\c), T(\y, \yhat, \c^{(1)}),\dots,T(\y, \yhat, \c^{(m)})$$
+test statistics, as well.
+
+The p-value under the null hypothesis is then obtained as
+$$ p= \frac{\sum_{j=1}^m \mathbb{1} \{T(\y, \yhat, \c^{(j)}) \geq T(\y, \yhat, \c) \}  }{m}$$
+
+Let $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$ and $\c_\pi = (c_{\pi_1}, \dots, c_{\pi_n})$ the vector $\c$ with its elements reordered according to the permutation $\pi \in S_i$.
+The permutation-based copies of $\c$ are then of $\c^{(j)} = \c_{\pi^{(j)}}$ which are drawn so that:
+
+\begin{equation}
+    \mathbb{P}(\pi^{(j)} = \pi | \y,\yhat,\c) = \frac{q^n(\c_\pi | \y)}{\sum_{\pi' \in S_n} q^n(\c_{\pi'} | \y)}
+    \label{eq-pperm}
+\end{equation}
+
+that is, according to the $q^n(\cdot|\y) := q(\cdot | y_1) \dots q(\cdot|y_n)$ product density corresponding to the conditional distribution $Q(\cdot|\y)$.
+Note that Eq.
+\ref{eq-pperm} does not necessarily assume a continuous distribution.
+For a verification of the valid type I error control of this approach refer to Theorem 1 in \citep{berrett2020conditional}.
+
+While the required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$ at random, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times.
+\cite{berrett2020conditional} (Algorithm 1) propose a parallelized pairwise Markov-Chain Monte Carlo sampler for CPT, that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data.
+In their Theorem 2, \cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.
+
+\subsubsection*{Conditional log-likelihood matrix}
+
+Obtaining a relatively accurate, independent estimate of $D(\cdot|\y)$ (of any shape) for CPT inference is important.
 Berrett and colleagues recommend to use a large independent sample to obtain the log-likelihood matrix of the conditional distribution $D(\cdot|Z)$, which might be often viable in semi-supervised learning settings where unlabelled data $(X, Z)$ are easier to obtain than labelled data $(X, Y, Z)$.
-Since in the proposed predictive modelling setting normality might be often a reasonable assumption for $Y$ and $C$ (but not for $\hat{Y}$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|Z)$.
+Since in the proposed predictive modelling , normality might be often a reasonable assumption for $\y$ and $\c$ (but not for $\yhat$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|\y)$.
 
-Here I propose to fit a linear regression model (or in case of categorical X, a logistic regression model):
+In case of continuous target, a linear regression model  can be fitted:
 
@@
     \label{eq:linreg}
-    X = \alpha + \beta Z + \epsilon
+    \c = \alpha + \beta \y + \epsilon
 \end{equation}
 
-with the assumption that the residual error follows a Gaussian distribution and let $\mu = \alpha + \beta Z$ and $\sigma^2 = E(\epsilon^2)$.
-Then, for the conditional Normal distribution $ (X|Z=z) \sim \mathcal{N}\{\mu(z), \sigma^2(z)\}$ and the log-likelihood, that is to be used in Eq.
+with the assumption that the residual error follows a Gaussian distribution and let $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma^2 = E(\epsilon^2)$.
+Then, for the conditional Normal distribution $ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$ and the log-likelihood, that is to be used in Eq.
 \ref{eq-pperm}, can be computed as follows:
 
-$$ \mathcal{L}(X|Z) = -{X^2}^T * \frac{1}{2} \sigma^2 + X^T \frac{\mu}{\sigma^2}$$
+$$ log\mathcal{L}(\c|\y) = - \frac{n}{2} ln(2 \pi) -  \frac{n}{2} ln( \sigma^2) - \frac{1}{2 \sigma^2} \sum_{i=1}^n (c_i - \mu_i)^2 $$
 
-where $T$ means transposing row vectors to column vectors.
+In the case of categorical target, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$.
 
-Applying this Gaussian likelihood estimation in the case predictive model diagnostics requires that the two variables in Eq.
-\ref{eq:linreg} are substituted with $Y$ and $C$, but not $\hat{Y}$, for which the Gaussian assumption of the proposed likelihood estimation might not hold.
+A straightforward choice of test statistic with the proposed Gaussian likelihood estimation is coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confoudner or classification):
 
-Therefore, from the three options for conditional independence-based null hypotheses, enumerated in Table \ref{tab:conditional-independence-cases}, this is possible for options 1 and 3, but not option 2, as the variable on which the independence is conditional must be always the predictor variable in Eq.
+$$T(\y, \yhat, \c) = R^2(\yhat, \c)$$
+
+and
+
+$$T(\y, \yhat, \c_\pi) = R^2(\yhat, \c_\pi)$$
+
+for any $\pi \in S_n$
+
+Notably, from the three options for conditional independence-based null hypotheses, enumerated in Table \ref{tab:conditional-independence-cases}, the proposed Gaussian likelihood estimation is possible for options 1 and 3, but not option 2, as the variable on which the independence is conditional must be always the predictor variable in Eq.
 \ref{eq:linreg}.
@@
 For options 1 and 3, the proposed simple Gaussian conditional likelihood estimation allows for deploying CPT without a large, independent sample to estimate the conditional distribution.
-
----------------------------------
-
-Todo: test statistic: $R^2$
-
-The validity of the type I error control of CPT has been justified when using a true $D(\cdot|Z)$.
-While \cite{berrett2020conditional} gave bounds on the validity of the test given deviations from the true conditional distribution (which proved to be relatively tolerant against deviations from the true distribution), these bounds assume that the $D(\cdot|Z)$ has been estimated from an independent sample.
-
-To provide foundations for the use of the proposed method we perfromed simulations...
-
------------------------------
-
-% tab:substitution
-\renewcommand{\arraystretch}{2}
-\begin{table}[]
-\centering
-\begin{tabular}{rl|l|l|l}
-original & CPT & $X$ & $Y$ & $Z$ \\
-\hline
-%\hline
-option 1: & full confounder test & $Y$ & $\hat{Y}$ & $C$ \\
-%\hline
-option 3: & partial confounder test & $C$ & $\hat{Y}$ & $Y$ \\
-\end{tabular}
-\caption{\label{tab:substitution} Substitutions of the symbols describing the variables in the original formulation of the CPT method and its two possible applications in predictive modelling diagnostics.
-}
-\end{table}
 
@@
 
+-------------------
+Both tests require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed without needing to re-fit the model.
+-----------------
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-08-09 supplement.tex
polished

@@
 \beginsupplement
+
+\section{Full confounder test}
+\label{sup:full-test}
+
+The full confoudner test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based "copies" of $\y$,
+
+$$\y_i^{(j)} \sim Q(\cdot|c_i) \ $$
+
+where, $Q(.|c)$ denotes the conditional distribution of $\y$ given $\c=c$ and $j=1,\dots, m$ indexes the "copy" of $\y$ so that $\y^{(i)} = (y_1^{(j)}, \dots, y_n^{(j)})$ is a permutation of the original vector $\y = (y_1, \dots, y_n)$.
+This mechanism creates copies $\y^{(1)}, \dots ,\y^{(m)}$ that are exchangeable with the original vector $\y$ under the null hypothesis that $\yhat \independent \y | \c$.
+
+Under the null hypothesis, the triples $$(\y,\yhat,\c), (\y^{(1)}, \yhat, \c),\dots, (\y^{(1)}, \yhat, \c)$$ are all identically distributed and exchangeable, and so are the 
+$$T(\y,\yhat,\c), T(\y^{(1)}, \yhat, \c),\dots,T(\y^{(m)}, \yhat, \c)$$
+test statistics, as well.
+
+The p-value under the null hypothesis is then obtained as
+$$ p= \frac{\sum_{j=1}^m \mathbb{1} \{T(\y^{(j)}, \yhat, \c) \geq T(\y, \yhat, \c) \}  }{m}$$
+
+Let $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$ and $\y_\pi = (y_{\pi_1}, \dots, y_{\pi_n})$ the vector $\y$ with its elements reordered according to the permutation $\pi \in S_i$.
+The permutation-based copies of $\y$ are then of $\y^{(j)} = \y_{\pi^{(j)}}$ which are drawn so that:
+
+\begin{equation}
+    \mathbb{P}(\pi^{(j)} = \pi | \y,\yhat,\c) = \frac{q^n(\y_\pi | \c)}{\sum_{\pi' \in S_n} q^n(\y_{\pi'} | \c)}
+    \label{eq-pperm}
+\end{equation}
+
+that is, according to the $q^n(\cdot|\c) := q(\cdot | c_1) \dots q(\cdot|c_n)$ product density corresponding to the conditional distribution $Q(\cdot|\c)$.
+Note that Eq.
+\ref{eq-pperm} does not necessarily assume a continuous distribution.
+For a verification of the valid type I error control of this approach refer to Theorem 1 in \citep{berrett2020conditional}.
 

### 2021-08-10 manuscript.tex
polished

@@
 \begin{abstract} % 150 words
-The lack of rigorous non-parametric statistical tests of confoudner-effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
+The lack of rigorous non-parametric statistical tests of confounder-effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
 Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unbiased} and \emph{fully biased model}, respectively.
@@
 In a predictive modelling setting, let $\y$ denote the target variable, $\yhat$ denote model output, i.e.
-the predictions for $\y$ and let $\c$ denote a variable which is considered as a confounder.
+the predictions for $\y$ and let $\c$ denote a variable which is considered as a confounder (see Fig.
+\ref{fig:schematic}B).
+Let us assume, furthermore, that $\y$, $\yhat$ and $\c$ consist of independent and identically distributed (IID) data points $(y_i, \hat{y}_i, c_i)$ for $i=1, \dots , n$ so that $\y=(y_1, \dots ,y_n)$, $\yhat=(\hat{y}_1, \dots, \hat{y}_n)$ and $\c=(c_1, \dots, c_n)$.
+
+Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; it's predictions are driven by the confounder rather than information about the target variable.
+Testing unconditional independence between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
+For instance, even if the null hypothesis $H0 : \yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to their dependence on $\y$.
+In other words, if the target variable $\y$ is associated to the confounder variable $\c$, a model that is completely blind to $\c$ (and, therefore, unbiased) might still provide outputs $\yhat$ that are significantly associated with $\c$.
+
+ Below, I summarize how testing conditional independence \citep{dawid1979conditional} with conditional permutation test \citep{berrett2020conditional} allows evaluating confounder-bias in predictive modelling, without placing assumptions on the distribution of the model output, where normality is often violated \citep{garcia2009study, kristensen2017whole}.
+
+\subsubsection*{Conditional independence for testing confounder bias}
+
 Depending on the research question, the influence of $\c$ on the model predictions $\yhat$ is subject to be kept at a negligible level or it is to be proven that, at least, the model is not completely driven by $\c$.
-Below, I summarize how testing conditional independence \citep{dawid1979conditional} across these variables allows addressing the above questions and build on the conditional permutation test of \cite{berrett2020conditional} to perform inference without assumptions on the distribution of the model output, where normality is often violated \citep{garcia2009study, kristensen2017whole}.
 
-\subsubsection*{Conditional independence}
-
-Let us assume that $\y$, $\yhat$ and $\c$ consist of independent and identically distributed (IID) data points $(y_i, \hat{y}_i, c_i)$ for $i=1, \dots , n$ and write $\y=(y_1, \dots ,y_n)$, $\yhat=(\hat{y}_1, \dots, \hat{y}_n)$ and $\c=(c_1, \dots, c_n)$.
-
-Obviously, a strong association between $\yhat$ and $\c$ can imply that the model is biased; it's predictions are driven by the confounder rather than information about the target variable.
-
-Testing unconditional independence between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confoudner-bias in predictive modelling.
-For instance, even if the null hypothesis $H0 : \yhat \independent \c$ is false, $\yhat$ and $\c$ might be only marginally dependent on $\c$, due to their dependence on $\y$.
-In other words, if the target variable $\y$ is associated to the confoudner variable $\c$, a model that is completely blind to $\c$ (and, therefore, unbiased) might still provide outputs $\yhat$ that are significantly associated with $\c$.
-
-Our primary interest is, therefore, the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ ($\yhat \independent \c | \y$) which, according to \cite{dawid1979conditional}, implies that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(X|Z)\mathbb{P}(Y|Z)$.
+Our primary interest is, therefore, the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, according to \cite{dawid1979conditional}, implies that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$.
 The test with this null hypothesis will be referred to as the \emph{partial confounder test}.
 
-Alternatively, one might be interested in testing $\yhat \independent \c | \y$. We will refer to the corresponding test as the \emph{full confoudner test}.
+Alternatively, one might be interested in testing $\yhat \independent \c | \y$. We will refer to the corresponding test as the \emph{full confounder test}.
 
@@
  Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional Independence testing is effectively impossible.
-In other words, neither the full nor the partial confoudner tests can be constructed so that - for all distributions - they provides a valid type I error control and, at the same time, a non-trivial statistical power.
+In other words, neither the full nor the partial confounder tests can be constructed so that - for all distributions - they provides a valid type I error control and, at the same time, a non-trivial statistical power.
 
-This result is in strong contrast to unconditional independence testing - where permutation tests  \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs - depending on the applied machine learning model - is unknown and often heavily non-normal.
+This result is in strong contrast to unconditional independence testing - where permutation tests  \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs, depending on the applied machine learning model, is unknown and often heavily non-normal.
 
 Nevertheless, as shown by \cite{berrett2020conditional}, conditional independence testing is possible with information only about the conditional distribution of two out of the three variables.
-In this case, sampling from a non-uniform distribution over the set of possible permutations $\pi$ of the one of the variables can incorporate available distributional information into permutation-based inference and achieve a valid type I error control and a non-trivial power.
-Below we adapt the formalization of the procedure of \cite{berrett2020conditional} for the confoudner problem.
-
-\subsubsection*{The conditional permutation test}
+The conditional permutation test (CPT) of Berret and colleagues builds on the conditional randomization test of \cite{candes2016panning} and samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, so that it incorporates available distributional information into permutation-based inference.
 
@@
 
-The conditional permutation test (CPT) of \cite{berrett2020conditional} builds on the conditional randomization test of \cite{candes2016panning} and allows that \emph{no assumptions} are made about the distribution of one of the variables but, instead, it is assumed that the conditional distribution of the other two variables is known.
-
 As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is the predictor(s) (features), $Y$ is the target variable and $Z$ is a potential confounder.
 In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether $X$ affects $Y$, when controlling for $Z$.
-For instance, in genome-wide association studies, CPT can be used o determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.
+For instance, in genome-wide association studies, CPT can be used to determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.
 
-Here, different setting is considered, where supervised learning model is already fitted and we are focusing on model diagnostics, by testing the triplet $(\y,\yhat, \c)$, as introduced above.
-In such a setting, we are interested in how much variance the fitted model captures from the confounder $C$ or, in other words, to what extent the model is biased by the confounder variable $\c$ when predicting $\y$ with $\yhat$.
+In this paper, a different setting is considered, where the supervised learning model is already fitted and we are focusing on model diagnostics, by testing the triplet $(\y,\yhat, \c)$.
 The conceptual difference between the original and the proposed application of CPT is depicted on Fig.
@@
 
-While option 3.
-is typically of interest when testing confounder bias of predictive models, option 1 might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction.
-Option 2 seems less appealing for model diagnostics and as detailed below, the proposed variety of the CPT framework does not allow forming a non-parametric method in this case.
+Option 3, i.e.
+partial confounder testing is typically of interest when testing confounder bias of predictive models.
+Option 1, i.e.
+full confounder testing, might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction.
+Option 2 seems less appealing for model diagnostics and importantly, the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.
 
-Below, CPT is re-formalized for option 3, i.e. partial confoudner testing. The formulation of the full confoudner test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
+Below, CPT is adapted for partial confounder testing (option 3). The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
 
-\subsubsection*{The partial confoudner test}
+\subsubsection*{The partial confounder test}
 
-The partial confoudner test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based "copies" of $\c$,
+The partial confounder test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based "copies" of $\c$,
 
@@
 
-where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y$ and $j=1,\dots, m$ indexes the "copy" of $\c$ so that $\c^{(i)} = (c_1^{(j)}, \dots, c_n^{(j)})$ is a permutation of the original vector $\c = (c_1, \dots, c_n)$.
-This mechanism creates copies $\c^{(1)}, \dots ,\c^{(m)}$ that are exchangeable with the original vector $\c$ under the null hypothesis that $\yhat \independent \c | \y$.
+where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y$ and $j=1,\dots, m$ indexes the "copy" of $\c$ so that $\c^{(j)} = (c_1^{(j)}, \dots, c_n^{(j)})$ is a permutation of the original vector $\c = (c_1, \dots, c_n)$.
 
-Under the null hypothesis, the triples $$(\y,\yhat,\c), (\y, \yhat, \c^{(1)}),\dots, (\y, \yhat, \c^{(m)})$$ are all identically distributed and exchangeable, and so are the 
+This mechanism creates copies $\c^{(1)}, \dots ,\c^{(m)}$ so that under the null hypothesis ($\yhat \independent \c | \y$), the triples $$(\y,\yhat,\c), (\y, \yhat, \c^{(1)}),\dots, (\y, \yhat, \c^{(m)})$$ are all identically distributed and exchangeable, and so are the
 $$T(\y,\yhat,\c), T(\y, \yhat, \c^{(1)}),\dots,T(\y, \yhat, \c^{(m)})$$
@@
 
-The p-value under the null hypothesis is then obtained as
-$$ p= \frac{\sum_{j=1}^m \mathbb{1} \{T(\y, \yhat, \c^{(j)}) \geq T(\y, \yhat, \c) \}  }{m}$$
+Let us denote the permutation-based copies of
 
-Let $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$ and $\c_\pi = (c_{\pi_1}, \dots, c_{\pi_n})$ the vector $\c$ with its elements reordered according to the permutation $\pi \in S_i$.
-The permutation-based copies of $\c$ are then of $\c^{(j)} = \c_{\pi^{(j)}}$ which are drawn so that:
+$$\c^{(j)} = \c_{\pi^{(j)}}, \quad \pi \in S_n$$
+
+where $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$ and $\c_\pi = (c_{\pi_1}, \dots, c_{\pi_n})$ the vector $\c$ with its elements reordered according to the permutation .
+
+In the partial confoudner test, the $\c_{\pi^{(j)}}$ copies are drawn so that:
 
@@
 
+The p-value under the null hypothesis is then obtained as
+$$ p= \frac{\sum_{j=1}^m \mathbb{1} \{T(\y, \yhat, \c^{(j)}) \geq T(\y, \yhat, \c) \}  }{m}$$
+
 While the required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$ at random, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times.
-\cite{berrett2020conditional} (Algorithm 1) propose a parallelized pairwise Markov-Chain Monte Carlo sampler for CPT, that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data.
+The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of \cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data.
 In their Theorem 2, \cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.
@@
 Berrett and colleagues recommend to use a large independent sample to obtain the log-likelihood matrix of the conditional distribution $D(\cdot|Z)$, which might be often viable in semi-supervised learning settings where unlabelled data $(X, Z)$ are easier to obtain than labelled data $(X, Y, Z)$.
-Since in the proposed predictive modelling , normality might be often a reasonable assumption for $\y$ and $\c$ (but not for $\yhat$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|\y)$.
+Since in the proposed predictive modelling, normality might be often a reasonable assumption for $\y$ and $\c$ (but not for $\yhat$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|\y)$.
 
@@
     \label{eq:linreg}
-    \c = \alpha + \beta \y + \epsilon
+    \c = \alpha + \beta \y + \boldsymbol{e}
 \end{equation}
 
-with the assumption that the residual error follows a Gaussian distribution and let $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma^2 = E(\epsilon^2)$.
-Then, for the conditional Normal distribution $ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$ and the log-likelihood, that is to be used in Eq.
+with the assumption that the residual error follows a Gaussian distribution.
+If we write $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma^2 = E(\boldsymbol{e}^2)$ then, for the conditional Normal distribution $ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$ and the log-likelihood, that is to be used in Eq.
 \ref{eq-pperm}, can be computed as follows:
@@
 
-A straightforward choice of test statistic with the proposed Gaussian likelihood estimation is coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confoudner or classification):
+A straightforward choice of test statistic with the proposed Gaussian likelihood estimation is the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification):
 
@@
 
-$$T(\y, \yhat, \c_\pi) = R^2(\yhat, \c_\pi)$$
+$$T(\y, \yhat, \c^{(j)}) = R^2(\yhat, \c^{(j)})$$
 
@@
 
-Using CPT to test confounder-bias in predictive modelling allows relaxing assumptions on $\hat{Y}$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions - on the joint distribution of the other two variables ($Y$ and $C$).
+Using CPT to test confounder-bias in predictive modelling allows relaxing assumptions on $\yhat$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions on - the joint distribution of the other two variables ($\yhat$ and $\c$).
 \cite{berrett2020conditional} give a detailed analysis of the robustness of their CPT approach when estimating the conditional distribution with re-using the tested data and, also, against misspecifying the conditional distribution of interest.
 
-Here I extend their results by performing validation that aids estimating the power of the test in a form that is accessible for power calculations in predictive modelling (considering various weights of the confounder and the target signals in $\hat{y}$ and in the case of varying levels of covariance between $Y$ and $C$).
-Moreover, I perform a thorough investigation of the robustness of the test against the violation of normality (for both $(Y, C)$ and $\hat{Y}$).
+Here I extend their results by performing simulations for estimating the power of the test in a form that is accessible for power calculations in predictive modelling (considering various weights of the confounder and the target signals in $\yhat$ and in the case of varying levels of covariance between $\y$ and $\c$).
+Moreover, I perform a thorough investigation of the robustness of the test against the violation of normality (for both $(\y, \c)$ and $\yhat$).
 
@@
 \subsubsection*{Gaussian Simulations}
-The target variable $Y$ and the confounder variable $C$ were drawn randomly from a multivariate normal distribution:
+The target variable $\y$ and the confounder variable $\c$ were drawn randomly from a multivariate normal distribution:
 
-$$ Y, C \sim \mathcal{N}(\mu_0, \Sigma_\sigma) $$,
+$$ \y, \c \sim \mathcal{N}(\mu_0, \Sigma_\sigma) $$
 
@@
 
-Next, we construct the simulated predicted values:
+Next, the simulated predicted values are constructed:
 
-$$ \hat{Y} = w_Y Y + w_C C + w_e e$$
+$$ \yhat = w_y \y + w_c \c + w_e \boldsymbol{e}$$
 
-so that $w_Y + w_C + w_e = 1$, where $e \sim \mathcal{N}(0,1)$ is the noise component.
+so that $w_y + w_c + w_e = 1$, where $\boldsymbol{e} \sim \mathcal{N}(0,1)$ is the noise component.
 
-To test the implementation for categorical variables, simulated $Y$, $\hat{Y}$ and $C$ variables are binarized by thresholding at 0.
+To test the implementation for categorical variables, simulated $\y$, $\yhat$ and $\c$ variables are binarized by thresholding at 0.
 
@@
 
-$$ w_{X/Y} = w_C / w_Y $$
+$$ w_{c/y} = w_c / w_y $$
 
@@
 
-$$ w_{C+Y} = w_C + w_Y $$,
+$$ w_{c+y} = w_c + w_y $$
 
@@
 
-$$X' = sinh(\delta sinh^{-1}(X) - \epsilon)$$
+$$\boldsymbol{x}' = sinh(\delta sinh^{-1}(\boldsymbol{x}) - \epsilon)$$
 
@@
 \begin{itemize}
-    \item covariance of $C$ and $Y$: $\sigma \in \{0, 0.2, 0.4, 0.6, 0.8\}$
-    \item confound-to-target: $w_{C/Y} \in \{0, 0.1, 0.3, 0.5, 1\}$
-    \item signal-in-prediction: $w_{C+Y} \in \{0, 0.3, 0.6, 0.9\}$
-    \item sample size: $N \in \{50, 100, 500, 1000\}$
+    \item covariance of $\c$ and $\y$: $\sigma \in \{0, 0.2, 0.4, 0.6, 0.8\}$
+    \item confound-to-target: $w_{c/y} \in \{0, 0.1, 0.3, 0.5, 1\}$
+    \item signal-in-prediction: $w_{c+y} \in \{0, 0.3, 0.6, 0.9\}$
+    \item sample size: $n \in \{50, 100, 500, 1000\}$
 \end{itemize}
 
-For testing 'full' confounder bias, identical parameters were used, except that instead of the 'confound-to-target' parameter $ w_{C/Y}$ we used the 'target-to-confound' parameter $ w_{Y/C} = w_Y / w_C $, but with the same parameter values $w_{Y/C} \in \{0, 0.1, 0.3, 0.5, 1\}$.
+For testing 'full' confounder bias, identical parameters were used, except that instead of the 'confound-to-target' parameter $ w_{c/y}$ we used the 'target-to-confound' parameter $ w_{y/c} = w_y/ w_c $, but with the same parameter values $w_{y/c} \in \{0, 0.1, 0.3, 0.5, 1\}$.
 
-Next to the normally distributed case, 4 \emph{sinh-arcsinh} transformed cases were also investigated, with $\delta \in \{1, 1.05, 1.5, 5\}$ $\epsilon \in \{1, 3, 5, 10\}$
+Next to the normally distributed case, four \emph{sinh-arcsinh} transformed cases were also investigated, with $\delta \in \{1, 1.05, 1.5, 5\}$ $\epsilon \in \{1, 3, 5, 10\}$
 
-The 'partial' and 'full' confound tests, as implemented in version 0.7.1 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.
+The 'partial' and 'full' confound tests, as implemented in version 0.9.0 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.
 
@@
 
-'Partial' and 'full' confound testing was used with default parameters (1000 permutations, 50 Markov-chain Monte Carlo steps) as implemented in version 0.7 of the package \emph{'mlconfound'}.
+'Partial' and 'full' confound testing was performed with default parameters (1000 permutations, 50 Markov-chain Monte Carlo steps) as implemented in version 0.9.0 of the package \emph{'mlconfound'}.
 
@@
 
-Human Connectome Project contains imaging and behavioral data of more than 1000 healthy subjects \citep{van2013wu}.
-Preprocessed resting state fMRI connectivity data (partial correlation matrices) \citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores (IQ), as measured with Penn Progressive Matrices \citep{duncan2000neural}.
+Human Connectome Project contains imaging and behavioral data of approximately 1200 healthy subjects \citep{van2013wu}.
+Preprocessed resting state fMRI connectivity data (partial correlation matrices) \citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores (IQ), measured with Penn Progressive Matrices \citep{duncan2000neural}.
 
@@
 
-Features (functional concavities across 100 group independent component analysis based regions) were either (i) considered in their raw form or were subject to confound mitigation approaches by (ii) feature regression \citep{rao2017predictive} or (iii) COMBAT \citep{johnson2007adjusting, fortin2018harmonization}.
+Features (functional connectivities across 100 group-independent component analysis based regions) were either (i) considered in their raw form or were subject to confound mitigation approaches by (ii) feature regression \citep{rao2017predictive} or (iii) COMBAT \citep{johnson2007adjusting, fortin2018harmonization}.
 The feature mitigation strategies were separately applied for acquisition batch and age group as confounder variable.
@@
 The total of 5 types of features (raw, regressing out acquisition batch, regressing out age group, COMBAT with acquisition batch, COMBAT with age group) were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the individual fluid intelligence scores with a ridge-regression \citep{hoerl1970ridge}.
-The $\alpha$ parameter of the ridge was considered as a hyperparameter ($\alpha \in \{0.000001, 0.01, 0.1, 1, 10, 100, 1000000\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric.
+The $\alpha$ parameter of the ridge model was considered as a hyperparameter ($\alpha \in \{0.00001, 0.0001, 0.001, 0.01, 0.1, 1, 10, 100, 1000, 10000, 100000\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric.
 Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.
 
-Analysis code is available as jupyter notebook as part\footnote{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb} of the github repository of the 'mlconfound' package.
+Analysis code is available as jupyter notebook\footnote{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb} in the github repository of the 'mlconfound' package.
 
@@
 
-The proposed method is applied to provide evidence of motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464.
+The proposed tests were applied to provide evidence of motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464.
 Preprocessed regional timeseries data was obtained as shared with the paper \citep{dadi2019benchmarking} which was based on preprocessed image data provided by the Preprocessed Connectome Project \citep{craddock2013neuro}.
@@
 
-The total of 5 types (raw, feature regression of site, feature regression of motion, COMBAT with site, COMBAT with motion) of features were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the diagnosis (DX: ASD vs.
+The total of five types (raw, feature regression of site, feature regression of motion, COMBAT with site, COMBAT with motion) of features were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the diagnosis (DX: ASD vs.
 HC) with a L2-regularized logistic regression (as e.g.
 in \citep{dadi2019benchmarking}).
-The $C$ parameter of the model was considered as a hyperparameter ($C \in \{0.1, 1, 10\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric.
+The $C$ parameter of the model was considered as a hyperparameter ($C \in \{0.1, 1, 10\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with area under the receiver operator curve (AUC under ROC) as optimization metric.
 Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.
+
+Confounder testing was performed on the class probabilities (instead of the predicted labels).
 
@@
 
-The conditional permutation based 'partial' and 'full' confound tests for predictive modelling have been implemented in the python package \emph{'mlconfound'}.
-Simulations showed valid type I control and a high, practically relevant statistical power in a wide variety of data characteristics.
-Analysis of real functional brain connectivity data demonstrates the usefulness of the tests in guiding confound-mitigation strategies.
+The conditional permutation based 'partial' and 'full' confound tests for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.
 
@@
 
-Simulations based on normally distributed data justify that the proposed confounder tests provide valid type I error control (For detailed results, see first row and first column on Figure \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-full}, \ref{fig:sim-ccb-partial}, \ref{fig:sim-ccb-full}, \ref{fig:sim-bbc-partial}, \ref{fig:sim-bbc-full}, \ref{fig:sim-bbb-partial} and \ref{fig:sim-bbc-full} for the partial test categorical data and the full tests on both numeric and categorical data).
-The higher confidence limit for the type I error rates for one simulation case with 1
-00 repetitions is 0.11 (approximately 5 simulation cases are expected to pass this limit across all 200 H0 simulations on one figure).
-The results suggest that the proposed tests are somewhat overly conservative if the model has a high predictive performance.
+Simulations based on normally distributed data justified that the proposed confounder tests - in line with theory - provide a valid type I error control (For detailed results, see first row and first column on Figure \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-full}, \ref{fig:sim-ccb-partial}, \ref{fig:sim-ccb-full}, \ref{fig:sim-bbc-partial}, \ref{fig:sim-bbc-full}, \ref{fig:sim-bbb-partial} and \ref{fig:sim-bbc-full} for the partial test categorical data and the full tests on both numeric and categorical data).
+The higher 95\% confidence limit for the type I error rates for one simulation case (with 100 repetitions) is 0.11.
+If the test is valid, five simulation instances are expected to pass this limit across 200 H0 simulations (all null hypotheses for all simulations depicted on the same figure).
+The number of simulation cases with type I error rate greater than 0.11 was always smaller then this expected value.
+The results suggest, moreover, that the proposed tests are somewhat overly conservative if the model has a high predictive performance (type I error rates closer to 0 than expected).
 
-Investigations of the tolerance to violation of normality assumptions showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $Y$ and the confounder $C$ (and completely non-parametric regarding the distribution of the predictions $\hat{y}$).
-The six simulation cases of normality violation , shown on Figure \ref{fig:sim-non-normal} involved various \emph{sinh-arcsinh} distributions, form totally normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
+Simulations with non-normal data showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $\y$ and the confounder $\c$ and, as expected - completely non-parametric regarding the distribution of the predictions $\yhat$).
+The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal} involved various \emph{sinh-arcsinh} distributions, form totally normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
 The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to a kurtosis and skewness values as high as 5 and 2 respectively.
@@
 
-When the normality assumptions hold, the test generally possess a statistical power of high practical relevance.
-Power estimates for the 'partial' test (with normality) is shown in Figure \ref{fig:sim-normal}.
+When normality assumptions hold, the test generally possess a statistical power of high practical relevance.
+Power estimates for the 'partial' test (with normality) are shown on Figure \ref{fig:sim-normal}.
 Notably, with sample sizes as large as 1000, a confounder contributing to only 0.4-0.6\% to the variance of the predictions can already be successfully detected ($w_c = 0.05 - 0.07$) with a power of 60-90\%.
@@
 Finally, even with a low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%.
-If the confoudner explains more than 15\% of variance, it is almost certainly detected with N>=50.
+If the confounder explains more than 15\% of variance, it is almost certainly detected with $n \geq 50$.
 
-Violation of the normality assumptions for $Y$ and $C$ has a negative effect on the power of the tests, however the loss of power is low or even negligible, if normality is not extremely violated (first 4 distributions on Fig.
+Violation of the normality assumptions for $\y$ and $\c$ has a negative effect on the power of the tests, however the loss of power is low or even negligible, if normality is not extremely violated (first 4 distributions on Fig.
 \ref{fig:sim-non-normal}A) and the confounder-to-signal ratio is greater or equal then 0.3 (Fig.
@@
 
-Acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$a and $p=0.0014$ and $0.001$, respectively).
+Acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p=0.0014$ and $0.001$, respectively).
 The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
-However, as revealed by the partial confounder test the model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.
+However, as revealed by the partial confounder test, the model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.
 
@@
 Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models.
-Testing confoudner-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confoudner variable.
+Testing confounder-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confounder variable.
 These two scenarios result in two different test, testing whether the model is partially or fully driven by the given confounder.
@@
 However, as shown by in \citet{shah2020hardness} their the 'no free lunch' theorem, it is effectively impossible to establishing a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
-Here I propose the use of the Markov-chain Monte-Carlo based conditional permutation test (CPT) of \cite{berrett2020conditional} to test confoudner bias in predictive models.
-Adapting this approach for the confoudner testing problem makes it possible to place assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confoudner tests so that no assumption is made about the distribution of the model output.
-Here I propose a computationally effective parallel implementation of CPT that utilizes the parallel pairwise sampler of \cite{berrett2020conditional}, coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confoudner or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confoudner is normal.
+Here I propose the use of the Markov-chain Monte-Carlo based conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models.
+Adapting this approach for the confounder testing problem makes it possible to place assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output.
+Here I propose a computationally effective parallel implementation of CPT that utilizes the parallel pairwise sampler of \cite{berrett2020conditional}, coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal.
 The implementation provides a test that can be uniformly applied for any predictive model, given that realistic (e.g.
@@
 
-Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confoudner bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confoudner explaining 1-2\% of the variance in the prediction).
-A currently somewhat more common sample size of 100 will very likely identify medium sized confoudner-biases (around 9\%).
+Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confounder bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confounder explaining 1-2\% of the variance in the prediction).
+A currently somewhat more common sample size of 100 will very likely identify medium sized confounder-biases (around 9\%).
 Strong biases  (>25\%) can be very robustly detected with sample sizes as low as 50.
 
-As expected, the power of the full confoudner test is very similar to the partial test, when investigating the contribution of the target to the prediction, instead of the confounder.
+As expected, the power of the full confounder test is very similar to the partial test, when investigating the contribution of the target to the prediction, instead of the confounder.
 
-While different biomedical applications might consider different amounts of bias to be relevant, an a-priory power calculation can help identifying the necessary sample size in order to exclude the possibility of a relevant confoudner bias with a high probability.
+While different biomedical applications might consider different amounts of bias to be relevant, an a-priory power calculation can help identifying the necessary sample size in order to exclude the possibility of a relevant confounder bias with a high probability.
 
-Importantly, the simulations strongly suggest that both tests are very robust to the violation of the normality assumption for the conditional distribution between the target and the confoudner.
+Importantly, the simulations strongly suggest that both tests are very robust to the violation of the normality assumption for the conditional distribution between the target and the confounder.
 Type I error is well controlled even in case of highly skewed and heavy tailed distributions (kurtosis > 5, skewness > 2) and gets significantly inflated only with extremely degenerate distributions (see the purple distribution on Fig.
@@
 
-In such studies, confounder-bias has often been investigated (if investigated at all) by testing the association between the confoudner variable and the predictive values, see e.g.
+In such studies, confounder-bias has often been investigated (if investigated at all) by testing the association between the confounder variable and the predictive values, see e.g.
 \citep{spisak2020pain}.
@@
 In this case, namely, the model still might not be directly driven by the confounder, i.e.
-the dependence of the predictions on the confounder can be explained solely by the confoudner-target association.
-Arguably, in certain applications, the researcher might aim to fully eliminate confoudner-related variance of the predictions and in this case testing the association (unconditional independence) between the confoudner and the predictions might be sufficient.
-However, a much more realistic aim is to construct models that allow a reasonable degree of confoudner-association of the predictions, but are likely not directly driven by confounder-information, i.e.
-the prediction-confoudner association is proportional to, and can likely be fully explained by, the target-confoudner association.
+the dependence of the predictions on the confounder can be explained solely by the confounder-target association.
+Arguably, in certain applications, the researcher might aim to fully eliminate confounder-related variance of the predictions and in this case testing the association (unconditional independence) between the confounder and the predictions might be sufficient.
+However, a much more realistic aim is to construct models that allow a reasonable degree of confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e.
+the prediction-confounder association is proportional to, and can likely be fully explained by, the target-confounder association.
 Put simply, this is what is tested by the proposed partial confounder test.
 
-Obviously, a true confoudner bias might still be present in the model, even if the the target-confoudner association likely explains the prediction-confoudner, i.e.
-the model might still use the confoudner effect in the features (e.g artifacts) to present that pert of the variance in the prediction.
-Nevertheless, in most of the cases, it does not seem likely that exclusively those confoudner components are captured the model that are needed to mimic the unbiased model and it might be often a much more plausible explanation that the confounder-variance is simply not captured by the model (or it was effectively mitigated by feature cleaning/harmonization).
+Obviously, a true confounder bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder, i.e.
+the model might still use the confounder effect in the features (e.g artifacts) to present that pert of the variance in the prediction.
+Nevertheless, in most of the cases, it does not seem likely that exclusively those confounder components are captured the model that are needed to mimic the unbiased model and it might be often a much more plausible explanation that the confounder-variance is simply not captured by the model (or it was effectively mitigated by feature cleaning/harmonization).
 
@@
 When trained on the raw (unadjusted) connectivity features, in all four cases, the confounders were statistically significantly associated to both the target variable and the model predictions (with weak-to-moderate effect sizes) and the proposed partial confounder test provided significant evidence against the null-hypothesis of no confounder bias (i.e.
-conditional independence between the confoudner and the predictions, given the target).
+conditional independence between the confounder and the predictions, given the target).
 
@@
 
-The investigated confound regression techniques, feature regression  \citep{spisak2014voxel, dukart2011age, abdulkadir2014reduction, rao2017predictive} and COMBAT \citep{johnson2007adjusting}, did not fully eliminate the association (unconditional dependence) between the confoudner and the predictions (neither for the HCP nor for the ABIDE dataset, all the corresponding p-values were smaller or equal than 0.003).
+The investigated confound regression techniques, feature regression  \citep{spisak2014voxel, dukart2011age, abdulkadir2014reduction, rao2017predictive} and COMBAT \citep{johnson2007adjusting}, did not fully eliminate the association (unconditional dependence) between the confounder and the predictions (neither for the HCP nor for the ABIDE dataset, all the corresponding p-values were smaller or equal than 0.003).
 However, as discussed above, fully eliminating this association is not necessarily a requirement for a model without confounder-bias.
-Indeed, the partial confoudner test revealed that except one, all models become likely free of confounder bias or - more precisely - the bias was suppressed to such an extent that the test was not able to detect it anymore.
-Given the high power of the test at these sample sizes (N=999 and 866 for the HCP and the ABIDE datasets, respectively), any confoudner remaining, but not detectable confounder bias is most probably negligible or is not of practical relevance.
-The exception, where confound mitigation did not successfully eliminate confoudner-bias, is confoudner regression of motion in the ABIDE dataset.
-Interestingly, in this case, the association between the confoudner and the predictions become stronger but with an opposite sign, while the prediction performance significantly dropped.
+Indeed, the partial confounder test revealed that except one, all models become likely free of confounder bias or - more precisely - the bias was suppressed to such an extent that the test was not able to detect it anymore.
+Given the high power of the test at these sample sizes (N=999 and 866 for the HCP and the ABIDE datasets, respectively), any confounder remaining, but not detectable confounder bias is most probably negligible or is not of practical relevance.
+The exception, where confound mitigation did not successfully eliminate confounder-bias, is confounder regression of motion in the ABIDE dataset.
+Interestingly, in this case, the association between the confounder and the predictions become stronger but with an opposite sign, while the prediction performance significantly dropped.
 COMBAT, on the other hand still effectively mitigated confound-bias in this case, although with a slight decrease of predictive performance, as well.
@@
 center or batch), our method of inputting the binned "motion groups" into COMBAT - together with the ability of the empirical Bayes method to incorporate information about sample variances - might have resulted in "less aggressive" data harmonization and preserved more signal-of-interest.
-The behavior of existing confound elimination techniques in such situations and the applicability of COMBAT with numeric confiders is an interesting target of further research, where the proposed conditional permutation based confoudner tests can be deployed.
+The behavior of existing confound elimination techniques in such situations and the applicability of COMBAT with numeric confiders is an interesting target of further research, where the proposed conditional permutation based confounder tests can be deployed.
 
-The full confoudner test - as expected - confirmed that none of the model is not fully driven by the confoudner variable or, in other words, the target and the prediction were not independent, given the confounders.
-While the full confoudner test might seem less useful in the presented examples, it might become useful in exploratory efforts, where the questions is simply if it is possible at all to train an unbiased predictive model.
+The full confounder test - as expected - confirmed that none of the model is not fully driven by the confounder variable or, in other words, the target and the prediction were not independent, given the confounders.
+While the full confounder test might seem less useful in the presented examples, it might become useful in exploratory efforts, where the questions is simply if it is possible at all to train an unbiased predictive model.
 
@@
 
-The lack of rigorous statistical tests of confoudner-bias significantly hampers the development of predictive models in many fields of research.
-Even tough several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confoudner and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias in the resulting model.
+The lack of rigorous statistical tests of confounder-bias significantly hampers the development of predictive models in many fields of research.
+Even tough several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias in the resulting model.
 
@@
 
-Given the high statistical power and the high robustness to the violation of the assumptions on the data (normality of the conditional distribution of the target and the confoudner variables), the tests can be useful in a wide variety of applications.
+Given the high statistical power and the high robustness to the violation of the assumptions on the data (normality of the conditional distribution of the target and the confounder variables), the tests can be useful in a wide variety of applications.
 

### 2021-08-11 manuscript.tex
polished

@@
 
+Depending on the research question, the direct influence of $\c$ on the model predictions $\yhat$ is subject to be kept at a negligible level or it is to be proven that, at least, the model is not completely driven by $\c$.
+
 Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; it's predictions are driven by the confounder rather than information about the target variable.
-Testing unconditional independence between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
-For instance, even if the null hypothesis $H0 : \yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to their dependence on $\y$.
-In other words, if the target variable $\y$ is associated to the confounder variable $\c$, a model that is completely blind to $\c$ (and, therefore, unbiased) might still provide outputs $\yhat$ that are significantly associated with $\c$.
-
- Below, I summarize how testing conditional independence \citep{dawid1979conditional} with conditional permutation test \citep{berrett2020conditional} allows evaluating confounder-bias in predictive modelling, without placing assumptions on the distribution of the model output, where normality is often violated \citep{garcia2009study, kristensen2017whole}.
+Testing unconditional independence ($\yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
+For instance, even if $\yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to the dependence of both on $\y$.
+In other words, if the target variable $\y$ is significantly associated to the confounder variable $\c$, a model that is completely blind to $\c$ (and, therefore, unbiased) might still provide outputs $\yhat$ that are significantly associated with $\c$.
 
@@
 
-Depending on the research question, the influence of $\c$ on the model predictions $\yhat$ is subject to be kept at a negligible level or it is to be proven that, at least, the model is not completely driven by $\c$.
-
-Our primary interest is, therefore, the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, according to \cite{dawid1979conditional}, implies that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$.
+Instead of focusing on the unconditional independence between the confoudner and the predictions, we shall consider the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, according to \cite{dawid1979conditional}, implies that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$.
 The test with this null hypothesis will be referred to as the \emph{partial confounder test}.
 
-Alternatively, one might be interested in testing $\yhat \independent \c | \y$. We will refer to the corresponding test as the \emph{full confounder test}.
+Alternatively, one might be also interested in testing $\yhat \independent \c | \y$. We will refer to the corresponding test as the \emph{full confounder test}.
 
@@
 Recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
- Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional Independence testing is effectively impossible.
+ Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional independence testing is effectively impossible.
 In other words, neither the full nor the partial confounder tests can be constructed so that - for all distributions - they provides a valid type I error control and, at the same time, a non-trivial statistical power.
@@
 
-Nevertheless, as shown by \cite{berrett2020conditional}, conditional independence testing is possible with information only about the conditional distribution of two out of the three variables.
-The conditional permutation test (CPT) of Berret and colleagues builds on the conditional randomization test of \cite{candes2016panning} and samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, so that it incorporates available distributional information into permutation-based inference.
+Even if constructing totally non-parametric confoudner tests was proven to be impossible, as shown by \cite{berrett2020conditional} and \citep{candes2016panning}, valid and powerful conditional independence tests can be constructed with inputting information about the distribution of only two (out of the three) variables.
+Specifically, the conditional permutation test (CPT) of Berret and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable, thereby it incorporating available distributional information into permutation-based inference.
 
@@
 
-As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is the predictor(s) (features), $Y$ is the target variable and $Z$ is a potential confounder.
-In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether $X$ affects $Y$, when controlling for $Z$.
+As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is a set of predictors (features), $Y$ is the target variable and $Z$ is a potential confounder.
+In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether the features $X$ still affect $Y$, when controlling for $Z$.
 For instance, in genome-wide association studies, CPT can be used to determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.
@@
 Within this setting, conditional independence testing and, specifically, the framework of conditional permutation testing allows investigating three different null hypotheses corresponding to the $(\y, \yhat, \c)$ triplet, as listed in Table \ref{tab:conditional-independence-cases}.
-Testing the null hypothesis $\y \independent \yhat | \c$ (option 1) investigates whether the model predictions are likely explainable solely with the confounder, i.e.
+Testing the null hypothesis $\y \independent \yhat | \c$ (option 1, full confoudner testing) investigates whether the model predictions are likely explainable solely with the confounder, i.e.
 whether the model is exclusively confounder-driven.
 Testing $\y \independent \c | \yhat$ (option 2) addresses the question whether the model captures all the variance in $c$ when predicting $y$.
-Testing the null hypothesis $\yhat \independent \c | \y$ (option 3) examines, whether the dependence of the model output on the confounder can likely be explained by the confounder's dependence on the target variable, i.e.
+Testing the null hypothesis $\yhat \independent \c | \y$ (option 3, partial confounder testing) examines, whether the dependence of the model output on the confounder can likely be explained by the confounder's dependence on the target variable, i.e.
 whether there is any confounder bias in the model.
@@
 full confounder testing, might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction.
-Option 2 seems less appealing for model diagnostics and importantly, the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.
+Option 2 seems less appealing for model diagnostics and importantly, in this case the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.
 
 Below, CPT is adapted for partial confounder testing (option 3). The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
-
 \subsubsection*{The partial confounder test}
@@
 
-$$\c_i^{(j)} \sim Q(\cdot|y_i) \ $$
+\begin{equation}
+    c_i^{(j)} \sim Q(\cdot|y_i)
+     \label{eq:cond-copy}
+\end{equation}
 
-where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y$ and $j=1,\dots, m$ indexes the "copy" of $\c$ so that $\c^{(j)} = (c_1^{(j)}, \dots, c_n^{(j)})$ is a permutation of the original vector $\c = (c_1, \dots, c_n)$.
+where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y$ and $j=1,\dots, m$ indexes the "copy" of $\c$ so that
 
-This mechanism creates copies $\c^{(1)}, \dots ,\c^{(m)}$ so that under the null hypothesis ($\yhat \independent \c | \y$), the triples $$(\y,\yhat,\c), (\y, \yhat, \c^{(1)}),\dots, (\y, \yhat, \c^{(m)})$$ are all identically distributed and exchangeable, and so are the
-$$T(\y,\yhat,\c), T(\y, \yhat, \c^{(1)}),\dots,T(\y, \yhat, \c^{(m)})$$
-test statistics, as well.
+$$ \c^{(j)} = (c_1^{(j)}, \dots, c_n^{(j)}) = (c_{\pi_1^{(j)}}, \dots, c_{\pi_n^{(j)}}) =  \c_{\boldsymbol{\pi}^{(j)}} $$
 
-Let us denote the permutation-based copies of
+is a permutation of the original vector $\c = (c_1, \dots, c_n)$, with its elements reordered according to the permutation $\quad \boldsymbol{\pi} \in S_n$ where $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$.
 
-$$\c^{(j)} = \c_{\pi^{(j)}}, \quad \pi \in S_n$$
-
-where $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$ and $\c_\pi = (c_{\pi_1}, \dots, c_{\pi_n})$ the vector $\c$ with its elements reordered according to the permutation .
-
-In the partial confoudner test, the $\c_{\pi^{(j)}}$ copies are drawn so that:
+To ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:
 
 \begin{equation}
-    \mathbb{P}(\pi^{(j)} = \pi | \y,\yhat,\c) = \frac{q^n(\c_\pi | \y)}{\sum_{\pi' \in S_n} q^n(\c_{\pi'} | \y)}
     \label{eq-pperm}
+    \mathbb{P}(\boldsymbol{ \pi }^{(j)} = \boldsymbol{ \pi } | \y,\yhat,\c) = \frac{q^n(\c_{\boldsymbol{ \pi }} | \y)}{\sum_{\boldsymbol{ \pi } ' \in S_n} q^n(\c_{\boldsymbol{ \pi } ' } | \y)}
 \end{equation}
@@
 \ref{eq-pperm} does not necessarily assume a continuous distribution.
-For a verification of the valid type I error control of this approach refer to Theorem 1 in \citep{berrett2020conditional}.
+
+This mechanism creates copies $\c^{(1)}, \dots ,\c^{(m)}$ so that under the null hypothesis ($\yhat \independent \c | \y$), the triples
+
+$$(\y,\yhat,\c), (\y, \yhat, \c^{(1)}),\dots, (\y, \yhat, \c^{(m)})$$
+
+are all identically distributed and exchangeable, and so are the
+
+$$T(\y,\yhat,\c), T(\y, \yhat, \c^{(1)}),\dots,T(\y, \yhat, \c^{(m)})$$
+
+test statistics, as well.
 
@@
 
+For a verification of the valid type I error control of this approach refer to Theorem 1 in \citep{berrett2020conditional}.
+
 While the required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$ at random, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times.
 The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of \cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data.
+The acceptance odds ratio of swapping indices $i$ and $j$ is:
+
+\begin{equation}
+\frac{ q(c_j | y_i) q(c_i | y_j)}{q(c_i | y_i) q(c_j | y_j) }
+=
+log \mathcal{L}(c_j | y_i) + log \mathcal{L}(c_i | y_j) - log \mathcal{L}(c_i | y_i) - log \mathcal{L}(c_j | y_j) 
+\label{eq:accept-odds}
+\end{equation}
+
 In their Theorem 2, \cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.
 
-\subsubsection*{Conditional log-likelihood matrix}
+\subsubsection*{Gaussian conditional log-likelihood}
 
@@
 with the assumption that the residual error follows a Gaussian distribution.
-If we write $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma^2 = E(\boldsymbol{e}^2)$ then, for the conditional Normal distribution $ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$ and the log-likelihood, that is to be used in Eq.
-\ref{eq-pperm}, can be computed as follows:
+If we write $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma$ is the standard deviation of the residual $\boldsymbol{e}$, then the conditional distribution of interest is given as $ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$ and the log-likelihood, that is to be used in Eq.
+\ref{eq:accept-odds}, can be computed simply as the log of the corresponding probability density function:
 
-$$ log\mathcal{L}(\c|\y) = - \frac{n}{2} ln(2 \pi) -  \frac{n}{2} ln( \sigma^2) - \frac{1}{2 \sigma^2} \sum_{i=1}^n (c_i - \mu_i)^2 $$
+$$ log\mathcal{L}(c_i|y_j) = - \frac{1}{2} \Big(\frac{c_i-\mu_j}{\sigma}\Big)^2 - ln(2 \pi \sigma)   $$
 
@@
 
-A straightforward choice of test statistic with the proposed Gaussian likelihood estimation is the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification):
+Notably, from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, the proposed Gaussian likelihood estimation is possible for options 1 and 3, but not option 2, as the variable on which the independence is conditional must be always the predictor variable in Eq.
+\ref{eq:linreg}.
+For options 1 and 3, the proposed simple Gaussian conditional likelihood estimation allows for deploying CPT without a large, independent sample to estimate the conditional distribution.
+
+While, in theory, any predefined test statistic $T$ can be used, for the proposed Gaussian likelihood estimation, the python package \emph{mlconfound} implements only the two-sided variant of the tests (as on-sided testing is out-of-interest in the confoudner problem) and uses the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic:
 
@@
 
-for any $\pi \in S_n$
-
-Notably, from the three options for conditional independence-based null hypotheses, enumerated in Table \ref{tab:conditional-independence-cases}, the proposed Gaussian likelihood estimation is possible for options 1 and 3, but not option 2, as the variable on which the independence is conditional must be always the predictor variable in Eq.
-\ref{eq:linreg}.
-
-For options 1 and 3, the proposed simple Gaussian conditional likelihood estimation allows for deploying CPT without a large, independent sample to estimate the conditional distribution.
+for any $\pi \in S_n$.
 
@@
 
-'Partial' and 'full' confound testing was performed with default parameters (1000 permutations, 50 Markov-chain Monte Carlo steps) as implemented in version 0.9.0 of the package \emph{'mlconfound'}.
+'Partial' and 'full' confound testing was performed with default parameters (10000 permutations, 50 Markov-chain Monte Carlo steps) as implemented in version 0.9.0 of the package \emph{'mlconfound'}.
+Unconditional dependence across the involved variables was investigated with conventional permutation test on the $R^2$ values, with 1000 permutations.
 
@@
 
-Simulations based on normally distributed data justified that the proposed confounder tests - in line with theory - provide a valid type I error control (For detailed results, see first row and first column on Figure \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-full}, \ref{fig:sim-ccb-partial}, \ref{fig:sim-ccb-full}, \ref{fig:sim-bbc-partial}, \ref{fig:sim-bbc-full}, \ref{fig:sim-bbb-partial} and \ref{fig:sim-bbc-full} for the partial test categorical data and the full tests on both numeric and categorical data).
-The higher 95\% confidence limit for the type I error rates for one simulation case (with 100 repetitions) is 0.11.
-If the test is valid, five simulation instances are expected to pass this limit across 200 H0 simulations (all null hypotheses for all simulations depicted on the same figure).
-The number of simulation cases with type I error rate greater than 0.11 was always smaller then this expected value.
-The results suggest, moreover, that the proposed tests are somewhat overly conservative if the model has a high predictive performance (type I error rates closer to 0 than expected).
+Simulations based on normally distributed data justified that the proposed confounder tests - in line with theory - provide a valid type I error control (For detailed results, see \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-full}, \ref{fig:sim-ccb-partial}, \ref{fig:sim-ccb-full}, \ref{fig:sim-bbc-partial}, \ref{fig:sim-bbc-full}, \ref{fig:sim-bbb-partial} and \ref{fig:sim-bbc-full} for the partial test on categorical data and the full tests on both numeric and categorical data).
+In case of a valid test, the expected type I error rate in the null hypothesis simulation cases (first row and first column of the figures) is less then or equal to the alpha level (0.05).
+The upper 95\% confidence limit for the type I error rates for one simulation case (i.e.
+a single tile on the figures) is 0.11 (binomial confidence intervals with 100 repetitions).
+The results suggest, that the proposed tests are somewhat overly conservative if the model has a high predictive performance (type I error rates closer to 0 than expected).
 
-Simulations with non-normal data showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $\y$ and the confounder $\c$ and, as expected - completely non-parametric regarding the distribution of the predictions $\yhat$).
-The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal} involved various \emph{sinh-arcsinh} distributions, form totally normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
-The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to a kurtosis and skewness values as high as 5 and 2 respectively.
+Simulations with non-normal data showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $\y$ and the confounder $\c$ and, as expected, completely non-parametric regarding the distribution of the predictions $\yhat$).
+The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal}, involved various \emph{sinh-arcsinh} distributions, form normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
+The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to kurtosis and skewness values as high as 5 and 2 respectively.
 Furthermore, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).
@@
 
-When normality assumptions hold, the test generally possess a statistical power of high practical relevance.
 Power estimates for the 'partial' test (with normality) are shown on Figure \ref{fig:sim-normal}.
@@
 A sample size of 100 yields an approximate power of 60-100\% for a confounder-variance of 6-7\% ($w_c = 0.2 - 0.21$) in most of the cases.
-Finally, even with a low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%.
+Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%.
 If the confounder explains more than 15\% of variance, it is almost certainly detected with $n \geq 50$.
@@
 
-Acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p=0.0014$ and $0.001$, respectively).
+Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}).
 The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
-However, as revealed by the partial confounder test, the model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.
+The partial confounder test revealed that, the model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.084$ and $0.041$, respectively) associated to larger predicted values.
 
-After applying the confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p < 0.05$ for all).
-Both feature regression and COMBAT slightly improved the predictive performance in the case of both confounders, with COMBAT providing better performances ($R^2=122$ and $121$ when applied to remove the effect of acquisition and age, respectively).
+After applying confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p > 0.05$ for all).
+Both feature regression and COMBAT decreased the predictive performance in the case of both confounders, with COMBAT providing better performances ($R^2=0.75$ and $0.76$ when applied to remove the effect of acquisition and age, respectively).
 
@@
 
-Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.027$ and $0.028$, respectively, $p<0.001$ for both).
-The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.095$, $ROC AUC = 0.71$, $p<0.001$).
-However, as revealed by the partial confounder test the raw model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with several sites (e.g.
-Carnegine Mellon University or Caltech) and higher motion being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to a higher probability of the model to assign the participant to the ASD group.
+Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.019$ and $0.028$, respectively, $p<0.001$ for both, see also Table \ref{tab:unconditional-pvals}).
+The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).
+The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$).
+Several sites (e.g.
+Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to a lower probability for the participant to be assigned to the ASD group by the model.
 
-After applying the confound mitigation approaches to remove the effect of center (feature regression or COMBAT) the partial confound test did not provide evidence of center-bias anymore ($p < 0.05$ for both techniques).
-When trying to mitigate the effect of in-scanner motion, feature regression failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates.
-The partial confounder test successfully detected this bias ($p<0.0001$).
-Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, effectively mitigated motion bias, as suggested by the partial confound test ($p=0.54$).
+Both feature regression and COMBAT seemed to significantly attenuate the effect of center, however the partial confound test still provided evidence for a significant center-bias ($p < 0.04$ and $0.01$ for feature regression and COMBAT, respectively).
+
+When trying to mitigate the effect of in-scanner motion, feature regression obviously failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates.
+The partial confounder successfully detected this strong bias ($p<0.0001$).
+Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$).
 
 Both feature regression and COMBAT slightly improved the predictive performance when mitigating center-effects ($R^2=0.111$ and $0.132$ and $AUC=0.76$ and $0.75$, respectively).
-With both feature regression and COMBAT, the effort of mitigating motion effects happened at the cost of predictive performance ($R^2=0.009$ and $0.087$ and $AUC=0.59$ and $0.70$, respectively)
+With both feature regression and COMBAT, the effort of mitigating motion effects happened at the cost of a drop in predictive performance ($R^2=0.02$ and $0.111$ and $AUC=0.59$ and $0.70$, respectively)
 
-The full confounder test was highly significant ($p<0.0001$) for all models, indicating that neither of the models were exclusively driven by the confounds.
+The full confounder test was highly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$) Indicating that this model might have been (almost) exclusively driven by motion artifacts.
 
@@
 \hline
-HCP & acq.  & raw      & 0.032 & <0.001 & 0.084 & <0.001  & 0.082 & <0.001 & <0.001 & <0.001 \\
+HCP & acq.  & raw      & 0.032 & <0.001 & 0.084 & <0.001  & 0.095 & <0.001 & <0.001 & <0.001 \\
     &              & f.reg.    & & &0.003 & 0.99 & 0.066  & <0.001  & 0.99 & <0.001 \\
@@
 \hline
-    & age   & raw       & 0.011 & 0.001  & 0.041 & <0.001  & 0.082  & <0.001 & <0.001 & <0.001 \\
-    &       & f.reg.    & & &0.001 & 0.3 & 0.07  & <0.001  & 0.5 & <    0.001 \\
+    & age   & raw       & 0.011 & 0.001  & 0.041 & <0.001  & 0.095  & <0.001 & <0.001 & <0.001 \\
+    &       & f.reg.    & & &0.001 & 0.3 & 0.07  & <0.001  & 0.5 & <0.001 \\
     &       & COMBAT    & & &0.004 & 0.055 & 0.076 & <0.001 & 0.12 & <0.001 \\
 \hline
-ABIDE   & center   & raw       & 0.005  & 0.203 &  0.169 & <0.001& 0.126     & <0.001 & <0.001 & <0.001 \\
+ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & <0.001 & <0.001 \\
         &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & 0.04 & <0.001 \\
@@
 These two scenarios result in two different test, testing whether the model is partially or fully driven by the given confounder.
-Importantly, in predictive modelling, the model predictions are often follow a non-normal distribution, thus the any test for confounder bias must be, at least partly, non-parametric.
-However, as shown by in \citet{shah2020hardness} their the 'no free lunch' theorem, it is effectively impossible to establishing a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
-Here I propose the use of the Markov-chain Monte-Carlo based conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models.
+Importantly, in predictive modelling, the model predictions are often follow a non-normal distribution, thus any test for confounder bias must be, at least partly, non-parametric.
+However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
+Here I propose the use of the conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models.
 Adapting this approach for the confounder testing problem makes it possible to place assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output.
-Here I propose a computationally effective parallel implementation of CPT that utilizes the parallel pairwise sampler of \cite{berrett2020conditional}, coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal.
-The implementation provides a test that can be uniformly applied for any predictive model, given that realistic (e.g.
+The tests are made available in the python package \emph{mlconfound} which builds upon a parallelized, computationally effective implementation of the Markov-chain Monte-Carlo based parallel pairwise sampler of \cite{berrett2020conditional}.
+It utilizes, coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal.
+The partial and full confoudner tests available in \emph{mlconfound} can be uniformly applied for any predictive model, given that realistic (e.g.
 cross-validated) model predictions can be obtained.
@@
 
-The simulations have confirmed that the proposed implementation provides a valid type I error control for both the full and the partial confound test and revealed that both the tests have a high power for various configurations of the association between the target and the confounder, predictive performance, sample size and the confound-to-signal for the partial or signal-to-confound ratios for the full test.
-Both tests become somewhat conservative (i.e.
+These simulations have confirmed that the proposed implementation provides a valid type I error control for both the full and the partial confound test and revealed that both tests have a high power for a wide variety of scenarios, with different strength of the association between the target and the confounder, varying predictive performance, various sample sizes and different true confound-to-signal ratios in the predictions.
+Both tests were found to become somewhat conservative (i.e.
 lower than expected type I error rates ) in the case of very powerful predictive models (when the $R^2$ between the predicted and observed values is greater than 0.7).
@@
 Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confounder bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confounder explaining 1-2\% of the variance in the prediction).
-A currently somewhat more common sample size of 100 will very likely identify medium sized confounder-biases (around 9\%).
+A lower sample size of 100 will still very likely identify medium sized confounder-biases (around 9\%).
 Strong biases  (>25\%) can be very robustly detected with sample sizes as low as 50.
 
-As expected, the power of the full confounder test is very similar to the partial test, when investigating the contribution of the target to the prediction, instead of the confounder.
+As expected, the power of the full confounder test is very similar to the partial test, hence the only difference is that the conditional contribution of the target to the prediction is investigated, instead that of the confounder.
 
@@
 Importantly, the simulations strongly suggest that both tests are very robust to the violation of the normality assumption for the conditional distribution between the target and the confounder.
-Type I error is well controlled even in case of highly skewed and heavy tailed distributions (kurtosis > 5, skewness > 2) and gets significantly inflated only with extremely degenerate distributions (see the purple distribution on Fig.
+Type I error is well controlled even in case of highly skewed and heavy tailed distributions (kurtosis > 5, skewness > 2) and gets significantly inflated only with extreme violations of normality (see the purple distribution on Fig.
 \ref{fig:sim-non-normal}).
-Due to the conservatives of the tests in case of very accurate models, even this extremely skewed and heavy tailed distribution resulted in lower than the nominal type I error rates if the predictions were strongly associated to the target variable.
+Due to the conservatives of the tests, if predictive model is very accurate, even this extremely skewed and heavy tailed distribution resulted in lower than the nominal type I error rates.
 
@@
 
-Given the power and the robustness of the proposed tests, they might become a useful tool in diagnosing predictive models, not exclusively in the biomedical filed.
+Given the power and the robustness of the proposed tests, they might become a useful tool in diagnosing predictive models in a wide spectrum of applications.
 
-In such studies, confounder-bias has often been investigated (if investigated at all) by testing the association between the confounder variable and the predictive values, see e.g.
+For instance, in biomedical research, confounder-bias has often been investigated (if investigated at all) by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.
 \citep{spisak2020pain}.
@@
 the dependence of the predictions on the confounder can be explained solely by the confounder-target association.
+Two examples for this situation can be found in the presented analysis of real data (when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data).
+
 Arguably, in certain applications, the researcher might aim to fully eliminate confounder-related variance of the predictions and in this case testing the association (unconditional independence) between the confounder and the predictions might be sufficient.
-However, a much more realistic aim is to construct models that allow a reasonable degree of confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e.
+However, a much more realistic aim is to construct models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e.
 the prediction-confounder association is proportional to, and can likely be fully explained by, the target-confounder association.
@@
 
-Obviously, a true confounder bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder, i.e.
-the model might still use the confounder effect in the features (e.g artifacts) to present that pert of the variance in the prediction.
-Nevertheless, in most of the cases, it does not seem likely that exclusively those confounder components are captured the model that are needed to mimic the unbiased model and it might be often a much more plausible explanation that the confounder-variance is simply not captured by the model (or it was effectively mitigated by feature cleaning/harmonization).
+Obviously, a true confounder bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder association, i.e.
+the model might still use the confounder's effect on the features (e.g artifacts) to construct the corresponding part of the variance in the prediction.
+Nevertheless, in most of the cases, it does not seem likely that the model at the same time (i) misses that specific "part" of the target variable's variance in the features that is shared with the confoudner (ii) captures exclusively those components of the confounder's variance in the features that is needed to mimic the unbiased model.
+Usually a much more plausible explanation is that the direct confoudner effect is simply not present in the features or not captured by the model.
 
 Among the many potential areas of applications, the proposed tests could be certainly useful in translational neuroscience.
-Applying predictive modelling and machine learning, for instance, on functional neuroimaging data holds a great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools \citep{woo2017building, wager2013fmri, spisak2020pain}.
+A characteristic example is the field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds a great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools \citep{woo2017building, wager2013fmri, spisak2020pain}.
 However, the presence of confounders typical for biomedical research (e.g.
@@
 To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two research settings where specific types of confounding effects have previously been described.
-First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effects of age \cite{lohmann2021predicting, dubois2018distributed} and also a somewhat ignored aspect, the acquisition data within the course of the data acquisition process.
-Second, functional connectivity data from the ABIDE \citep{di2014autism} was used to investigate the potential motion- and center-bias (as previously reported by e.g.
+First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \cite{lohmann2021predicting, dubois2018distributed} and also a somewhat ignored aspect, the acquisition date of the data within the course of the data acquisition process.
+Second, functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.
 \cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
 
-When trained on the raw (unadjusted) connectivity features, in all four cases, the confounders were statistically significantly associated to both the target variable and the model predictions (with weak-to-moderate effect sizes) and the proposed partial confounder test provided significant evidence against the null-hypothesis of no confounder bias (i.e.
-conditional independence between the confounder and the predictions, given the target).
+The hypothesised confound effects were clearly present in both datasets.
 
-The detected confounder bias is most probably of practical relevance for all four confounders.
-
-In case of the HCP dataset, the weak, but statistically significant association between age and IQ predictions could likely exaggerate to a serious bias when testing the model on data of participants outside of the age range of the training sample (that is, the bias could likely significantly harm out-of-sample generalizability).
-The bias of the model for the acquisition batch (phase of the study the acquisition belongs to) can also be problematic and - to my best knowledge - it was not yet reported in the HCP dataset.
+In case of the HCP dataset, the weak but statistically significant association between age and IQ predictions could likely exaggerate to a serious bias when testing the model on data of participants outside of the (relatively narrow)
+age range of the training sample.
+In this case, the bias could significantly harm out-of-sample generalizability of the model.
+The bias of the model for the acquisition batch (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset.
 There can be manifold reasons for the observed associations.
-Fluid intelligence of the participants can be affected by a change of selection bias during requirement (e.g.
-as the human connectome project got more public interest during its course the the sociodemographic and educational characteristics of the volunteer sample might have changed).
-The model can utilise changing characteristics of the data (e.g.
-gradual changes in the calibration settings of the measuring equipment or the changing level of experience/motivation of the research staff) to predict this part of the effect.
+Fluid intelligence of the participants can be affected by a changing selection bias during requirement (e.g.
+as the human connectome project gradually received more public interest during its course the sociodemographic and educational characteristics of the volunteer sample might have changed).
+The model can utilise such changing characteristics of the data (caused by e.g.
+gradual changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff) to predict this part of the effect.
 This is obviously undesirable for any brain-based predictive model of fluid intelligence.
 
-In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously undesired for a diagnostic model of autism.
+In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of autism.
 For instance the model trained on the raw features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g.
@@
 
-The investigated confound regression techniques, feature regression  \citep{spisak2014voxel, dukart2011age, abdulkadir2014reduction, rao2017predictive} and COMBAT \citep{johnson2007adjusting}, did not fully eliminate the association (unconditional dependence) between the confounder and the predictions (neither for the HCP nor for the ABIDE dataset, all the corresponding p-values were smaller or equal than 0.003).
-However, as discussed above, fully eliminating this association is not necessarily a requirement for a model without confounder-bias.
-Indeed, the partial confounder test revealed that except one, all models become likely free of confounder bias or - more precisely - the bias was suppressed to such an extent that the test was not able to detect it anymore.
-Given the high power of the test at these sample sizes (N=999 and 866 for the HCP and the ABIDE datasets, respectively), any confounder remaining, but not detectable confounder bias is most probably negligible or is not of practical relevance.
-The exception, where confound mitigation did not successfully eliminate confounder-bias, is confounder regression of motion in the ABIDE dataset.
-Interestingly, in this case, the association between the confounder and the predictions become stronger but with an opposite sign, while the prediction performance significantly dropped.
-COMBAT, on the other hand still effectively mitigated confound-bias in this case, although with a slight decrease of predictive performance, as well.
+Confounder mitigation techniques, with one exception, tended to decrease confounder bias.
+The partial confoudner test provided detailed insights about the differences of the techniques for each confounder.
+It revealed that, in the case of both the acquisition-bias and the age-bias was very effectively removed by both feature regression and COMBAT (p>0.05 for all).
+Given the high power of the test at these sample sizes (N=999 and 866 for the HCP and the ABIDE datasets, respectively), any remaining confounder-bias is most probably negligible and out of the range of practical relevance.
+
+The performance of the investigated confound mitigation techniques is, however, less convincing for the ABIDE dataset.
+The center-bias of the classification in this massively multi-center dataset was, namely, found to be significant, even after feature regression and COMBAT, although the p-values were larger with one order of magnitude, suggesting that a large part of the center effects was still successfully removed by both approaches.
+While determining the relevance of the remaining bias is to be performed based on the experimental context and the requirements towards the predictive marker, significant p-values of the partial confoudner test generally suggest that this model would either require more effective data harmonization or a validation on external data, in order to construct realistic estimates of the amount of bias.
+
+Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
+Interestingly, in this case, the association between the confounder and the predictions become even stronger than before but with an opposite sign.
+Meanwhile, the prediction performance significantly dropped.
  A possible explanation to these phenomena is that in scanner motion is not simply a confounder but, as previously described \citep{fournier2010motor, anzulewicz2016toward}, it might have a complex relationship to ASD.
-Therefore, regressing out motion estimates from the connectivity features might eliminate significant amount of signal-of-interest, especially in functional connections that are associated with motor deficits in ASD.
-In such cases, applying feature regression might be a too "aggressive" solution.
+Therefore, regressing out motion estimates from the connectivity features might eliminate significant amount of signal-of-interest (especially in functional connections that are associated with motor deficits in ASD).
+ On the other hand, motion artefacts were previously shown to display complex spatiotemporal patterns in fMRI-based functional connectivity estimates \cite{spisak2014voxel} and simply regressing them out of the features is far from being sufficient for removing their effect on the model.
+The previously described "residual artefacts" and their regional interactions might be responsible for the paradox behavior of the predictive model and raises caution for using feature regression to mitigate the effect of similarly complex confounders.
+
  As COMBAT was originally developed for harmonizing effects of categorical variables in the dataset (e.g.
-center or batch), our method of inputting the binned "motion groups" into COMBAT - together with the ability of the empirical Bayes method to incorporate information about sample variances - might have resulted in "less aggressive" data harmonization and preserved more signal-of-interest.
-The behavior of existing confound elimination techniques in such situations and the applicability of COMBAT with numeric confiders is an interesting target of further research, where the proposed conditional permutation based confounder tests can be deployed.
+center or batch), using it to remove the effect of continuous variables is not trivial.
+The presented method of inputting the binned "motion groups" into COMBAT resulted in an efficient removal of motion-bias and, at the same time, preserved a significant amount of signal-of-interest.
+ Nevertheless, inputting discretized versions of continuous variables into COMBAT might be sub-optimal and raises further questions e.g regarding the optimal number of bins used during the discretization.
+ The application of the partial confounder test has, therefore, pointed out that 
+ behavior of existing confound elimination techniques in such situations and the applicability of COMBAT with continuous numeric confiders is an important target of further research, where the proposed conditional permutation based confounder tests can be deployed.
 
-The full confounder test - as expected - confirmed that none of the model is not fully driven by the confounder variable or, in other words, the target and the prediction were not independent, given the confounders.
-While the full confounder test might seem less useful in the presented examples, it might become useful in exploratory efforts, where the questions is simply if it is possible at all to train an unbiased predictive model.
-
--------------------
-Both tests require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed without needing to re-fit the model.
------------------
+The full confounder test did not provide any evidence that the paradox motion-regression model in the ABIDE dataset captures any extra variance over that explained by the confoudner.
+In other words, this model might be exclusively driven by motion artifacts.
+For all other models, the full confound test confirmed that - even if there was a significant confounder-bias, the model was not fully driven by the confounder (in other words, the target and the prediction were not independent, given the confounders).
+This example highlight that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased and the questions is rather if there is any biomedical relevant signal driving the model.
 
@@
 
-To fill this critical gap in predictive model development, I propose two novel tests, the partial and the full confounder tests, which test the null hypotheses of 'no confounder-bias' of 'full confounder-bias', respectively.
+To fill this critical gap in predictive model development, I propose two novel tests, the partial and the full confounder tests, which probe the null hypotheses of 'no confounder-bias' and 'full confounder-bias', respectively.
 Both approaches are based on a recently proposed conditional permutation framework, with a mathematical justification of valid type I error control.
-The tests were constructed so that they do not place any assumption about the distribution of the model predictions, thereby being applicable with arbitrary machine learning algorithms.
+The tests were constructed so that they do not place any assumption about the distribution of the model predictions and require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed without needing to re-fit the model.
 
-Given the high statistical power and the high robustness to the violation of the assumptions on the data (normality of the conditional distribution of the target and the confounder variables), the tests can be useful in a wide variety of applications.
+Given the easy application, the high statistical power, the robustness to the violation of the assumptions on the data and the computationally effective implementation (python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}) the tests are deployable in a wide variety of applications.
 

### 2021-08-13 manuscript.tex
patched
ready for inner review

@@
 \begin{abstract} % 150 words
-The lack of rigorous non-parametric statistical tests of confounder-effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
+The lack of rigorous non-parametric statistical tests of confounder-bias significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
 Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unbiased} and \emph{fully biased model}, respectively.
-The proposed tests set no assumptions about the distribution of the predictive model output that is often non-normal.
-Simulations show that the tests are applicable in a wide variety of cases with a remarkable statistical power.
-Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset identified significantly biased predictions.
-Some of these biases were previously unreported and found to be hard to mitigate with state-of-the-art confound mitigation approaches.
-The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
+The proposed tests set no assumptions about the distribution of the predictive model output that is often non-normal and display a remarkable robustness and a high statistical power in simulations.
+Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset provided statistical evidence for several models being significantly confounded.
+Some of these confounder-biases were previously unreported and found to be hard to fully mitigate with state-of-the-art confound mitigation approaches.
+The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment and improvement of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
 \end{abstract}
@@
 Predictive modelling and supervised learning methods have recently became increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized}.
-When evaluate the usefulness and applicability of such markers, predictive performance (i.e.
-prognostic/diagnostic value) i snot the only consideration.
+When evaluating the usefulness and applicability of such markers, predictive performance (i.e.
+prognostic/diagnostic value) is far from being the only important consideration.
 Generalizability across contexts and populations and biomedical validity (i.e.
-whether the model is driven by biomedical relevant signal) are also highly important requirements for candidate biomarkers \citep{woo2017building}.
+whether the model is driven by biomedically relevant signal) are also crucial requirements for candidate biomarkers \citep{woo2017building}.
 
@@
 First, measurement artifacts can obviously be considered as confounders.
-For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with movement kinematics or motor function, like previously demonstrated, among others, in case of e.g Alzheimer's disease \citep{rao2017predictive} attention deficit hyperactivity disorder (ADHD) \citep{couvy2016head, eloyan2012automated}er or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}.
+For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with movement kinematics or motor function, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive} attention deficit hyperactivity disorder (ADHD) \citep{couvy2016head, eloyan2012automated} or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}.
 Similarly, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.
@@
 For instance, resting-state functional connectivity is known to be strongly predictive to  demographic age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}.
-As fluid intelligence is known to decline with aging, \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by simply picking up on age-related variance \citep{lohmann2021predicting, dubois2018distributed}.
-Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance}.
-In this case, predictive models can utilize center/batch-effects to explain (random or true) center/batch-differences in the predictive target.
+As fluid intelligence is known to decline with aging, \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by picking up on solely age-related variance \citep{lohmann2021predicting, dubois2018distributed}.
+Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance} as the models can utilize such effects to explain (random or true) center/batch-differences in the predictive target.
 Such models will display worse-than-expected performance when applied on data from new centers/batches.
 
-Recently, various data harmonization approaches have been suggested to mitigate confounder-bias, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}.
-However, it is often unclear which variables should be considered as confounders during harmonization and considering many variables as confounder holds risks of eliminating signal-of-interest \citep{wachinger2021detect}.
+Various data harmonization approaches have been proposed to mitigate confounder-bias, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}.
+However, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.
 
-Powerful statistical tests to quantify the confounding effect of certain variables in predictive models could largely foster both the the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation and data harmonization approaches.
+Powerful and robust statistical tests to quantify the confounding effect of certain variables in predictive models could largely foster both the the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation and data harmonization approaches.
 Such methods, however, have to tackle the fact that the outputs of predictive models often follow non-normal distributions (due to e.g.
@@
 
-Recently three dedicated techniques \citep{chaibub2019permutation, ferrari2020measuring, wachinger2021detect} have been proposed for quantifying confounder bias in predictive modelling.
+Recently, three dedicated techniques \citep{chaibub2019permutation, ferrari2020measuring, wachinger2021detect} have been proposed for quantifying confounder bias in predictive modelling.
 
-The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via restricted permutations.
-It's p-values, however, might be invalid in certain circumstances (e.g.
-biased classification problems) and, as shown by \cite{ferrari2020measuring}, it does not guarantee a proper type I error control.
+The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via "restricted permutations" that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature) However, as shown by \cite{southworth2009properties} and further discussed by \cite{hemerik2018exact}, balanced permutations do not provide a correct null behavior and can be invalid in certain circumstances.
+Indeed, \cite{ferrari2020measuring} demonstrated that this method does not guarantee a proper type I error control with biased classification problems.
 The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, by design, does not provide any quantification of statistical significance.
@@
 Next, I utilize the permutation-based conditional independence framework of \cite{berrett2020conditional}, to construct two novel tests for evaluating the confounder-bias in predictive models.
-The \emph{full confounder test} probes whether the model's explained variance can be attributed exclusively to the confounder
-The \emph{partial confounder test} investigates whether the model utilizes any confounder-variance, above what is already present in the target variable.
-Both tests require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed without needing to re-fit the model.
+The \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder.
+The \emph{partial confounder test} investigates whether the model utilizes any confounder-variance, when controlled for the target variable.
+Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
 The tests set no assumption about the distribution of the model predictions.
-The validity of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normal model output).
-The proposed tests are then applied on functional brain connectivity data to provide evidence of motion-bias in a model trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} and to evaluate the age dependence of fluid intelligence predictions on data from the Human Connectome Project.
+The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality).
+The proposed tests are then applied on functional brain connectivity data to provide evidence of motion- and center-bias in a model trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} and to evaluate the age and acquisition-batch dependence of fluid intelligence predictions on data from the Human Connectome Project.
 
@@
 \ref{fig:schematic}B).
-Let us assume, furthermore, that $\y$, $\yhat$ and $\c$ consist of independent and identically distributed (IID) data points $(y_i, \hat{y}_i, c_i)$ for $i=1, \dots , n$ so that $\y=(y_1, \dots ,y_n)$, $\yhat=(\hat{y}_1, \dots, \hat{y}_n)$ and $\c=(c_1, \dots, c_n)$.
+Let us assume, furthermore, that $\y$, $\yhat$ and $\c$ consist of independent and identically distributed (IID) data points $(y_i, \hat{y}_i, c_i) \in Y \times \\hat{Y} \times C$ for $i=1, \dots , n$ so that $\y=(y_1, \dots ,y_n)$, $\yhat=(\hat{y}_1, \dots, \hat{y}_n)$ and $\c=(c_1, \dots, c_n)$.
 
@@
 
-Instead of focusing on the unconditional independence between the confoudner and the predictions, we shall consider the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, according to \cite{dawid1979conditional}, implies that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$.
-The test with this null hypothesis will be referred to as the \emph{partial confounder test}.
+Instead of focusing on the unconditional independence between the confounder and the predictions, we shall consider the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, by definition  \citep{dawid1979conditional}, means that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$.
+The statistical test with this null hypothesis will be referred to as the \emph{partial confounder test}.
 
-Alternatively, one might be also interested in testing $\yhat \independent \c | \y$. We will refer to the corresponding test as the \emph{full confounder test}.
+Alternatively, one might be also interested in testing $\yhat \independent \y | \c$. We will refer to the corresponding test as the \emph{full confounder test}.
 
@@
 
-Even if constructing totally non-parametric confoudner tests was proven to be impossible, as shown by \cite{berrett2020conditional} and \citep{candes2016panning}, valid and powerful conditional independence tests can be constructed with inputting information about the distribution of only two (out of the three) variables.
-Specifically, the conditional permutation test (CPT) of Berret and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable, thereby it incorporating available distributional information into permutation-based inference.
+Even if constructing fully non-parametric confounder tests was proven to be impossible, as demonstrated by \citep{candes2016panning} and \cite{berrett2020conditional}, valid and powerful conditional independence tests can be constructed with inputting distributional information about only two (out of the three) variables.
+Specifically, the conditional permutation test (CPT) of Berret and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable.
+Thereby, it incorporates the information available on the conditional distribution of interest into the permutation-based inference in a statistically valid manner.
 
@@
 
-Within this setting, conditional independence testing and, specifically, the framework of conditional permutation testing allows investigating three different null hypotheses corresponding to the $(\y, \yhat, \c)$ triplet, as listed in Table \ref{tab:conditional-independence-cases}.
-Testing the null hypothesis $\y \independent \yhat | \c$ (option 1, full confoudner testing) investigates whether the model predictions are likely explainable solely with the confounder, i.e.
+Within this setting, conditional independence testing and, specifically, the framework of conditional permutation testing allows investigating three different null hypotheses corresponding to the $(\y, \yhat, \c)$ triplet.
+As listed in Table \ref{tab:conditional-independence-cases}, testing the null hypothesis $\y \independent \yhat | \c$ (option 1, full confounder testing) investigates whether the model predictions are likely explainable solely with the confounder, i.e.
 whether the model is exclusively confounder-driven.
@@
 full confounder testing, might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction.
-Option 2 seems less appealing for model diagnostics and importantly, in this case the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.
+Option 2 seems less appealing for model diagnostics and importantly, in this case the proposed variety of the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.
 
@@
 
-To ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:
+As shown in \citep{berrett2020conditional}, to ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:
 
@@
 
-are all identically distributed and exchangeable, and so are the
+are all identically distributed and so are the
 
@@
 
-The p-value under the null hypothesis is then obtained as
+As long as the numerator of Eq.
+\ref{eq-pperm} is non-zero for all $c_\pi \in C$ and $y \in Y$, the conditional permutations constitute an algebraic group, thus, as shown by \citep{hemerik2018exact}, an unbiased estimate of the p-value under the null can be obtained as:
 $$ p= \frac{\sum_{j=1}^m \mathbb{1} \{T(\y, \yhat, \c^{(j)}) \geq T(\y, \yhat, \c) \}  }{m}$$
 
-For a verification of the valid type I error control of this approach refer to Theorem 1 in \citep{berrett2020conditional}.
+While the group property of the conditioned permutations provides a straightforward proof for the validity of the approach, for an alternative verification see the proof of Theorem 1 in \citep{berrett2020conditional}.
 
@@
 =
-log \mathcal{L}(c_j | y_i) + log \mathcal{L}(c_i | y_j) - log \mathcal{L}(c_i | y_i) - log \mathcal{L}(c_j | y_j) 
+\ell(c_j | y_i) + \ell(c_i | y_j) - \ell(c_i | y_i) - \ell(c_j | y_j) 
 \label{eq:accept-odds}
 \end{equation}
+
+where $\ell$ denotes the log-likelihood.
 
@@
 Obtaining a relatively accurate, independent estimate of $D(\cdot|\y)$ (of any shape) for CPT inference is important.
-Berrett and colleagues recommend to use a large independent sample to obtain the log-likelihood matrix of the conditional distribution $D(\cdot|Z)$, which might be often viable in semi-supervised learning settings where unlabelled data $(X, Z)$ are easier to obtain than labelled data $(X, Y, Z)$.
-Since in the proposed predictive modelling, normality might be often a reasonable assumption for $\y$ and $\c$ (but not for $\yhat$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|\y)$.
+Berrett and colleagues recommend to use a large independent sample to obtain the log-likelihood matrix that represents the conditional distribution $D(\cdot|Z)$.
+This approach might be often viable in semi-supervised learning settings where unlabelled data $(X, Z)$ are easier to obtain than labelled data $(X, Y, Z)$.
+Since in the proposed predictive modelling setting, normality might be often a reasonable assumption for $\y$ and $\c$ (but not for $\yhat$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|\y)$.
 
@@
 with the assumption that the residual error follows a Gaussian distribution.
-If we write $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma$ is the standard deviation of the residual $\boldsymbol{e}$, then the conditional distribution of interest is given as $ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$ and the log-likelihood, that is to be used in Eq.
-\ref{eq:accept-odds}, can be computed simply as the log of the corresponding probability density function:
+If we write $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma$ denotes the standard deviation of the residual $\boldsymbol{e}$, then the conditional distribution of interest can also be assumed to be normal with the parameters:
 
-$$ log\mathcal{L}(c_i|y_j) = - \frac{1}{2} \Big(\frac{c_i-\mu_j}{\sigma}\Big)^2 - ln(2 \pi \sigma)   $$
+$$ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$$
 
-In the case of categorical target, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$.
+and the log-likelihood, that is to be used in Eq. \ref{eq:accept-odds}, can be computed simply as the log of the corresponding probability density function:
 
-Notably, from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, the proposed Gaussian likelihood estimation is possible for options 1 and 3, but not option 2, as the variable on which the independence is conditional must be always the predictor variable in Eq.
+$$ \ell(c_i|y_j) = - \frac{1}{2} \Big(\frac{c_i-\mu_j}{\sigma}\Big)^2 - ln(2 \pi \sigma)   $$
+
+In the case of categorical $c$, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$, with the extra assumption of non-complete separation if $y$ is also categorical (for an invertible Hessian, see e.g.
+\citep{bohning1992multinomial}).
+
+Importantly, both the linear and the multinomial logistic regression approaches guarantee that the numerator of Eq.
+\ref{eq-pperm} is always greater than zero and the group property for the permutations holds.
+
+Note that from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, for option 2, the proposed approach can not provide a test that is assumption-free about $yhat$, as the variable on which the independence is conditional must be always the predictor variable in Eq.
 \ref{eq:linreg}.
-For options 1 and 3, the proposed simple Gaussian conditional likelihood estimation allows for deploying CPT without a large, independent sample to estimate the conditional distribution.
+However, as discussed above, this option is of low practical relevance, anyway.
+Pleasingly,  the proposed Gaussian regression-based conditional likelihood estimation ensures that no assumptions on $\yhat$ have to be made for the practically relevant options 1 and 3, i.e.
+for the full and partial confounder tests.
 
-While, in theory, any predefined test statistic $T$ can be used, for the proposed Gaussian likelihood estimation, the python package \emph{mlconfound} implements only the two-sided variant of the tests (as on-sided testing is out-of-interest in the confoudner problem) and uses the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic:
+In theory, any predefined test statistic $T$ can be used with the proposed approach.
+The python package \emph{mlconfound}, implementing the proposed full and partial confoudner tests, utilizes the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification \citep{starkweather2011multinomial}) as a test statistic:
 
@@
 
-for any $\pi \in S_n$.
+which allows interpretable, two tailed inference about confoudner-bias in predictive modelling.
 
@@
 
-Here I extend their results by performing simulations for estimating the power of the test in a form that is accessible for power calculations in predictive modelling (considering various weights of the confounder and the target signals in $\yhat$ and in the case of varying levels of covariance between $\y$ and $\c$).
+While their findings are largely applicable to the proposed tests, here I extend those results by performing simulations in a form that is accessible for power calculations in predictive modelling (considering various weights of the confounder and the target signals in $\yhat$ and in the case of varying levels of covariance between $\y$ and $\c$).
 Moreover, I perform a thorough investigation of the robustness of the test against the violation of normality (for both $(\y, \c)$ and $\yhat$).
@@
 \subsubsection*{Gaussian Simulations}
-The target variable $\y$ and the confounder variable $\c$ were drawn randomly from a multivariate normal distribution:
+As  a first step, the target variable $\y$ and the confounder variable $\c$ are drawn randomly from a multivariate normal distribution:
 
@@
 
-where $\mu=(0, 0)$ and the covariance matrix is given as:
+where $\mu_0=(0, 0)$ and the covariance matrix is given as:
 $\Sigma_\sigma = \big(\begin{smallmatrix}
@@
   \sigma & 1
-\end{smallmatrix}\big)$.
+\end{smallmatrix}\big)$ with $\sigma$ (i.e. the covariance between $\y$ and $\c$) being a free parameter of the simulation.
 
 Next, the simulated predicted values are constructed:
-
+ with an additive model:
 $$ \yhat = w_y \y + w_c \c + w_e \boldsymbol{e}$$
@@
 
-For more straightforward interpretation and visualization of the simulation results, the 'confound-to-target' and the 'signal-in-prediction' parameters were defined as
+For more straightforward interpretation and visualization of the simulation results, the 'confound-to-target' and the 'signal-in-prediction' parameters are defined as
 
@@
 
-Next to the normally distributed case, four \emph{sinh-arcsinh} transformed cases were also investigated, with $\delta \in \{1, 1.05, 1.5, 5\}$ $\epsilon \in \{1, 3, 5, 10\}$
+Next to the above described normally distributed simulation scenario, four analogous, but \emph{sinh-arcsinh} transformed scenarios were also investigated, with $\delta \in \{1, 1.05, 1.5, 5\}$ $\epsilon \in \{1, 3, 5, 10\}$
 
@@
 
-\subsection{Validation on functional brain connectivity data}
+\subsection{Application on functional brain connectivity data}
 
-The usefulness of the proposed confounder tests is demonstrated by applying them for predictive classification and regression models based on functional brain connectivity data, with different confound-mitigation approaches.
+The usefulness of the proposed confounder tests is demonstrated by applying them for predictive classification and regression models based on functional brain connectivity data, processed with different confound-mitigation approaches.
 
@@
 
-The proposed tests were applied to provide evidence of motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464.
+The proposed tests were applied to provide evidence of center- and  motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464.
 Preprocessed regional timeseries data was obtained as shared with the paper \citep{dadi2019benchmarking} which was based on preprocessed image data provided by the Preprocessed Connectome Project \citep{craddock2013neuro}.
@@
   \includegraphics[width=0.40\paperwidth]{fig/sim_non-norm.eps}
-  \caption{\textbf{Robustness of conditional permutation based confound testing to the violation of normality assumptions for $Y$ and $C$.} \\
+  \caption{\textbf{Robustness of conditional permutation based confound testing to the violation of normality assumptions for $\y$ and $\c$.} \\
   Simulations included variables with five different degrees of non-normality (A), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation, from normal (left) to severely skewed and heavy tailed (right).
 Fisher's kurtosis and skewness is given for each at the bottom of panel A.
-Panel (B) depicts false positive rates in the simulations under H0, separately for different signal-in-confounder ratios (here identical to $w_{Y}$).
+Panel (B) depicts false positive rates in the simulations under H0, separately for different signal-in-confounder ratios (here identical to $w_{y}$).
   Vertical axis shows type I error rate.
@@
 The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to kurtosis and skewness values as high as 5 and 2 respectively.
-Furthermore, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).
+Furthermore, in terms of type I error rate, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).
 
@@
 The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
-The partial confounder test revealed that, the model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.084$ and $0.041$, respectively) associated to larger predicted values.
+The partial confounder test revealed that the model was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig.
+\ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.084$ and $0.041$, respectively) associated to larger predicted values.
 
-After applying confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p > 0.05$ for all).
+After applying confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p > 0.05$ for all, second and third columns of Fig.
+\ref{fig:hcp}).
 Both feature regression and COMBAT decreased the predictive performance in the case of both confounders, with COMBAT providing better performances ($R^2=0.75$ and $0.76$ when applied to remove the effect of acquisition and age, respectively).
@@
 The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).
-The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$).
+The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$, see first column on Fig.
+\ref{fig:abide}).
 Several sites (e.g.
@@
 
-Both feature regression and COMBAT seemed to significantly attenuate the effect of center, however the partial confound test still provided evidence for a significant center-bias ($p < 0.04$ and $0.01$ for feature regression and COMBAT, respectively).
+Both feature regression and COMBAT seemed to significantly attenuate the effect of center, however the partial confound test still provided evidence for a significant center-bias ($p < 0.04$ and $0.01$ for feature regression and COMBAT, respectively, see the second and third columns of Fig \ref{fig:abide}).
 
-When trying to mitigate the effect of in-scanner motion, feature regression obviously failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates.
+When trying to mitigate the effect of in-scanner motion (bottom row on Fig \ref{fig:abide}), feature regression obviously failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates.
 The partial confounder successfully detected this strong bias ($p<0.0001$).
-Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$).
+Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$, bottom right panel of Fig \ref{fig:abide}).
 
@@
 These two scenarios result in two different test, testing whether the model is partially or fully driven by the given confounder.
-Importantly, in predictive modelling, the model predictions are often follow a non-normal distribution, thus any test for confounder bias must be, at least partly, non-parametric.
+Importantly, in predictive modelling, the model predictions often follow a non-normal distribution, thus any test for confounder bias must be, at least partly, non-parametric.
 However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
@@
 It utilizes, coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal.
-The partial and full confoudner tests available in \emph{mlconfound} can be uniformly applied for any predictive model, given that realistic (e.g.
+The partial and full confounder tests available in \emph{mlconfound} can be uniformly applied for any predictive model, given that realistic (e.g.
 cross-validated) model predictions can be obtained.
+Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost; re-fitting the model is not required.
 
-The validity and the power of the tests has been investigated in simulations with normally distributed data and with various degrees of non-normality.
+As expected from theoretical justification, both tests displayed a valid type I error control in the simulation with the necessary assumptions, that is, even if $\yhat$ is not normally distributed.
+In fact, the tests were found to become somewhat conservative (i.e.
+lower than expected type I error rates ) in the case of high-performance predictive models (when the $R^2$ between the predicted and observed values is greater than 0.7).
 
-These simulations have confirmed that the proposed implementation provides a valid type I error control for both the full and the partial confound test and revealed that both tests have a high power for a wide variety of scenarios, with different strength of the association between the target and the confounder, varying predictive performance, various sample sizes and different true confound-to-signal ratios in the predictions.
-Both tests were found to become somewhat conservative (i.e.
-lower than expected type I error rates ) in the case of very powerful predictive models (when the $R^2$ between the predicted and observed values is greater than 0.7).
-However, the power of the test remains high with sufficiently large ($\ge 500$) sample sizes.
-
+Even tough the tests valid type I error control is justified by theory and reinforced by the simulations, a low statistical power could significantly limit their usefulness.
+The simulations presented in this paper revealed that both tests have a high, practically relevant power in a wide variety of scenarios; with different strength of the association between the target and the confounder, varying predictive performance, various sample sizes and different true confound-to-signal ratios in the predictions.
 Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confounder bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confounder explaining 1-2\% of the variance in the prediction).
@@
 Strong biases  (>25\%) can be very robustly detected with sample sizes as low as 50.
-
-As expected, the power of the full confounder test is very similar to the partial test, hence the only difference is that the conditional contribution of the target to the prediction is investigated, instead that of the confounder.
+As expected, the power of the full confounder test is very similar to the partial test, since the only difference is that the conditional contribution of the target to the prediction is investigated, instead that of the confounder.
 
@@
 
-Importantly, the simulations strongly suggest that both tests are very robust to the violation of the normality assumption for the conditional distribution between the target and the confounder.
+Next to their statistical power, the practical applicability of the proposed tests is also highly affected by their robustness of the normality assumptions on the conditional distribution of $\y$ and $\c$.
+Importantly, the simulations strongly suggest that both tests are strikingly robust to the violation of this assumption.
 Type I error is well controlled even in case of highly skewed and heavy tailed distributions (kurtosis > 5, skewness > 2) and gets significantly inflated only with extreme violations of normality (see the purple distribution on Fig.
 \ref{fig:sim-non-normal}).
-Due to the conservatives of the tests, if predictive model is very accurate, even this extremely skewed and heavy tailed distribution resulted in lower than the nominal type I error rates.
+Due to the conservatives of the tests, if the predictive model is very accurate, even this extremely skewed and heavy tailed distribution resulted in lower than the nominal type I error rates.
 
-Non-normality has a negative effect on the power of the both tests, however, the decrease in power is in many simulation cases negligible and gets only relevant if the sample size and/or the confounder-effect is low.
+Non-normality has a negative effect on the power of both tests, however, the decrease in power is in most cases small or even negligible and gets only more pronounced if the sample size and/or the confounder-effect is low.
 
-Given the power and the robustness of the proposed tests, they might become a useful tool in diagnosing predictive models in a wide spectrum of applications.
-
+Validity, power and robustness renders the proposed tests as unprecedented model diagnostic tools in a wide spectrum of predictive modelling scenarios.
 For instance, in biomedical research, confounder-bias has often been investigated (if investigated at all) by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.
@@
 However, the significance of this association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable.
-In this case, namely, the model still might not be directly driven by the confounder, i.e.
-the dependence of the predictions on the confounder can be explained solely by the confounder-target association.
-Two examples for this situation can be found in the presented analysis of real data (when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data).
+In this case, namely, the model still might not be directly driven by the confounder and the dependence of the predictions on the confounder might be only marginal, to be explained solely by the confounder-target association.
+Two examples for this situation can be found in the presented analysis of real data; when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data (see Table \ref{tab:unconditional-pvals}).
 
@@
 the model might still use the confounder's effect on the features (e.g artifacts) to construct the corresponding part of the variance in the prediction.
-Nevertheless, in most of the cases, it does not seem likely that the model at the same time (i) misses that specific "part" of the target variable's variance in the features that is shared with the confoudner (ii) captures exclusively those components of the confounder's variance in the features that is needed to mimic the unbiased model.
-Usually a much more plausible explanation is that the direct confoudner effect is simply not present in the features or not captured by the model.
+Nevertheless, in most of the cases, it does not seem likely that the model ,at the same time, (i) misses that specific "part" of the target variable's variance in the features that is shared with the confounder (ii) captures exclusively those components of the confounder's variance in the features that is needed to mimic an unbiased model.
+Usually a much more plausible explanation is that a direct confounder effect in the model output is simply negligible  model (e.g because it is not to be found in the features after confound mitigation).
 
 Among the many potential areas of applications, the proposed tests could be certainly useful in translational neuroscience.
-A characteristic example is the field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds a great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools \citep{woo2017building, wager2013fmri, spisak2020pain}.
+A characteristic example is provided by the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds a great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools \citep{woo2017building, wager2013fmri, spisak2020pain}.
 However, the presence of confounders typical for biomedical research (e.g.
 sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g.
-imaging artifacts) present a great challenge to these efforts and are, currently, often neglected.
+imaging artifacts) presents a great challenge to these efforts and is, currently, often largely neglected.
 
 To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two research settings where specific types of confounding effects have previously been described.
-First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \cite{lohmann2021predicting, dubois2018distributed} and also a somewhat ignored aspect, the acquisition date of the data within the course of the data acquisition process.
+First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
 Second, functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.
@@
 
-The hypothesised confound effects were clearly present in both datasets.
+The hypothesised confounder effects were clearly present in both datasets.
 
 In case of the HCP dataset, the weak but statistically significant association between age and IQ predictions could likely exaggerate to a serious bias when testing the model on data of participants outside of the (relatively narrow)
-age range of the training sample.
-In this case, the bias could significantly harm out-of-sample generalizability of the model.
-The bias of the model for the acquisition batch (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset.
+age range of the HCP sample.
+In this case, the bias would likely significantly harm the out-of-sample generalizability of this model.
+The bias of the same model to the acquisition batch (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset.
 There can be manifold reasons for the observed associations.
-Fluid intelligence of the participants can be affected by a changing selection bias during requirement (e.g.
-as the human connectome project gradually received more public interest during its course the sociodemographic and educational characteristics of the volunteer sample might have changed).
-The model can utilise such changing characteristics of the data (caused by e.g.
-gradual changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff) to predict this part of the effect.
+Fluid intelligence of the participants can be affected by a changing selection bias during participant recruitment (e.g.
+as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed).
+To explain the associated change in the target variable, the model might utilise changing characteristics of the data (caused by e.g.
+slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff).
 This is obviously undesirable for any brain-based predictive model of fluid intelligence.
@@
 Similarly, a model biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with higher motion or altered motor characteristics (as known e.g.
-for ADHD \cite{eloyan2012automated}).
+for ADHD \citep{eloyan2012automated}).
 
 Confounder mitigation techniques, with one exception, tended to decrease confounder bias.
-The partial confoudner test provided detailed insights about the differences of the techniques for each confounder.
+The partial confounder test provided detailed insights into the effectiveness of the techniques for each confounder.
 It revealed that, in the case of both the acquisition-bias and the age-bias was very effectively removed by both feature regression and COMBAT (p>0.05 for all).
-Given the high power of the test at these sample sizes (N=999 and 866 for the HCP and the ABIDE datasets, respectively), any remaining confounder-bias is most probably negligible and out of the range of practical relevance.
+Given the high power of the test at these sample sizes (N=999 and 866 for the HCP and the ABIDE datasets, respectively), any remaining confounder-bias is most probably very safely negligible and well out of the range of practical relevance.
 
-The performance of the investigated confound mitigation techniques is, however, less convincing for the ABIDE dataset.
-The center-bias of the classification in this massively multi-center dataset was, namely, found to be significant, even after feature regression and COMBAT, although the p-values were larger with one order of magnitude, suggesting that a large part of the center effects was still successfully removed by both approaches.
-While determining the relevance of the remaining bias is to be performed based on the experimental context and the requirements towards the predictive marker, significant p-values of the partial confoudner test generally suggest that this model would either require more effective data harmonization or a validation on external data, in order to construct realistic estimates of the amount of bias.
+On the other hand, the partial confoudner test also showed, that the performance of the investigated confound mitigation techniques is much less convincing for the ABIDE dataset.
+The center-bias of the classification in this massively multi-center dataset was, namely, found to be significant, even after feature regression and COMBAT.
+Nevertheless, the p-values returned by the partial confounder test were larger with one order of magnitude than without any confound mitigation, suggesting that a large part of the center effects was still successfully removed by both approaches.
+While determining the relevance of the remaining bias is to be performed based on the experimental context and the requirements towards the predictive marker, significant p-values of the partial confounder test generally suggest that these models would either require more effective data harmonization or a validation on external data, that confirms the low practical relevance of the amount of bias.
 
@@
 Interestingly, in this case, the association between the confounder and the predictions become even stronger than before but with an opposite sign.
-Meanwhile, the prediction performance significantly dropped.
- A possible explanation to these phenomena is that in scanner motion is not simply a confounder but, as previously described \citep{fournier2010motor, anzulewicz2016toward}, it might have a complex relationship to ASD.
+Meanwhile, the prediction performance significantly dropped (but still remained statistically significant).
+A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD.
 Therefore, regressing out motion estimates from the connectivity features might eliminate significant amount of signal-of-interest (especially in functional connections that are associated with motor deficits in ASD).
 On the other hand, motion artefacts were previously shown to display complex spatiotemporal patterns in fMRI-based functional connectivity estimates \cite{spisak2014voxel} and simply regressing them out of the features is far from being sufficient for removing their effect on the model.
-The previously described "residual artefacts" and their regional interactions might be responsible for the paradox behavior of the predictive model and raises caution for using feature regression to mitigate the effect of similarly complex confounders.
+The so-called "residual motion artefacts" and their regional interactions, as described in the referenced work, might be responsible for the paradox behavior of this model, which raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
 
- As COMBAT was originally developed for harmonizing effects of categorical variables in the dataset (e.g.
+ As COMBAT was originally developed for harmonizing effects of categorical variables (e.g.
 center or batch), using it to remove the effect of continuous variables is not trivial.
@@
  Nevertheless, inputting discretized versions of continuous variables into COMBAT might be sub-optimal and raises further questions e.g regarding the optimal number of bins used during the discretization.
- The application of the partial confounder test has, therefore, pointed out that 
- behavior of existing confound elimination techniques in such situations and the applicability of COMBAT with continuous numeric confiders is an important target of further research, where the proposed conditional permutation based confounder tests can be deployed.
 
-The full confounder test did not provide any evidence that the paradox motion-regression model in the ABIDE dataset captures any extra variance over that explained by the confoudner.
-In other words, this model might be exclusively driven by motion artifacts.
-For all other models, the full confound test confirmed that - even if there was a significant confounder-bias, the model was not fully driven by the confounder (in other words, the target and the prediction were not independent, given the confounders).
-This example highlight that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased and the questions is rather if there is any biomedical relevant signal driving the model.
+ In sum, the application of the partial confounder test suggests that confounder-bias should be much more carefully investigated and reported in the predictive modelling literature as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT.
+The partial confoudner test can be considered as a useful benchmark to guide the search for a suitable confounder mitigation approach.
+
+While the real data examples in this paper were not typical cases of the potential application of the full confounder test, it has still provided important insights into model biases.
+Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confoudner test did not provide any evidence that the model captures any extra variance over that explained by the confounder.
+In other words, the test indicated that this model might be, in fact, exclusively driven by motion artifacts.
+For all other models, the full confound test confirmed that - even if there was a significant confounder-bias - the model was not fully driven by the confounder (in other words, the target and the prediction were not independent, given the confounders).
+This examples highlights that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased and the questions is whether there is any biomedical relevant signal captured by the model, at all.
 
@@
 
-The lack of rigorous statistical tests of confounder-bias significantly hampers the development of predictive models in many fields of research.
-Even tough several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias in the resulting model.
+The lack of rigorous statistical tests for confounder-bias significantly hampers the development of predictive models in many fields of research.
+Even tough several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder signal and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias in the resulting model.
 
-To fill this critical gap in predictive model development, I propose two novel tests, the partial and the full confounder tests, which probe the null hypotheses of 'no confounder-bias' and 'full confounder-bias', respectively.
-Both approaches are based on a recently proposed conditional permutation framework, with a mathematical justification of valid type I error control.
-The tests were constructed so that they do not place any assumption about the distribution of the model predictions and require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed without needing to re-fit the model.
+To fill this critical gap in predictive model development, here I proposed two novel tests, the partial and the full confounder tests, which probe the null hypotheses of 'no confounder-bias' and 'full confounder-bias', respectively.
+Both approaches are based on a recently published framework for conditional permutation testing, with a mathematical justification of valid type I error control.
+The tests were constructed so that they do not place any assumption on the distribution of the model predictions and require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed very efficiently, without needing to re-fit the model.
 
-Given the easy application, the high statistical power, the robustness to the violation of the assumptions on the data and the computationally effective implementation (python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}) the tests are deployable in a wide variety of applications.
+Given their simplicity, high statistical power, robustness to the violation of the assumptions and thanks to the computationally effective implementation (python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the tests are deployable in a wide variety of applications.
 
-As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of confound mitigation and expected to improve the assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research, thereby paving the way for the development of clinically useful machine learning biomarkers.
+As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and significantly enhances the quantitative assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research, thereby paving the way for the development of clinically useful machine learning biomarkers.
 

### 2021-08-15 manuscript.tex
polished (inner review)

@@
 The partial confounder test revealed that the model was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig.
-\ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.084$ and $0.041$, respectively) associated to larger predicted values.
+\ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.
 
 After applying confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p > 0.05$ for all, second and third columns of Fig.
-\ref{fig:hcp}).
-Both feature regression and COMBAT decreased the predictive performance in the case of both confounders, with COMBAT providing better performances ($R^2=0.75$ and $0.76$ when applied to remove the effect of acquisition and age, respectively).
+\ref{fig:hcp}), neither for acquisition batch nor for age.
+Both feature regression and COMBAT increased the predictive performance in the case of both confounders, with COMBAT providing best performances ($R^2=0.122$ and $0.121$ when applied to remove the effect of acquisition and age, respectively).
 
@@
 Several sites (e.g.
-Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to a lower probability for the participant to be assigned to the ASD group by the model.
+Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately ($R^2=0.169$ and $0.111$, respectively) associated to a lower probability for the participant to be assigned to the ASD group by the model.
 
-Both feature regression and COMBAT seemed to significantly attenuate the effect of center, however the partial confound test still provided evidence for a significant center-bias ($p < 0.04$ and $0.01$ for feature regression and COMBAT, respectively, see the second and third columns of Fig \ref{fig:abide}).
+Both feature regression and COMBAT seemed to significantly attenuate the effect of center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.01$ for feature regression and COMBAT, respectively, see the second and third columns of the first row on Fig \ref{fig:abide}).
 
 When trying to mitigate the effect of in-scanner motion (bottom row on Fig \ref{fig:abide}), feature regression obviously failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates.
-The partial confounder successfully detected this strong bias ($p<0.0001$).
+The partial confounder successfully detected the resulting strong bias ($p<0.0001$).
 Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$, bottom right panel of Fig \ref{fig:abide}).
 
-Both feature regression and COMBAT slightly improved the predictive performance when mitigating center-effects ($R^2=0.111$ and $0.132$ and $AUC=0.76$ and $0.75$, respectively).
-With both feature regression and COMBAT, the effort of mitigating motion effects happened at the cost of a drop in predictive performance ($R^2=0.02$ and $0.111$ and $AUC=0.59$ and $0.70$, respectively)
+Both feature regression and COMBAT considerably improved the predictive performance when mitigating center-effects ($R^2=0.111$ and $0.132$ and $AUC=0.76$ and $0.75$, respectively).
+With both feature regression and COMBAT, however, the effort of mitigating motion effects happened at the cost of a drop in predictive performance ($R^2=0.02$ and $0.111$ and $AUC=0.59$ and $0.70$, respectively)
 
@@
 \hline
-HCP & acq.  & raw      & 0.032 & <0.001 & 0.084 & <0.001  & 0.095 & <0.001 & <0.001 & <0.001 \\
-    &              & f.reg.    & & &0.003 & 0.99 & 0.066  & <0.001  & 0.99 & <0.001 \\
-    &              & COMBAT    & & &0.008 & 0.75 & 0.075  & <0.001  & 0.88 & <0.001\\
+HCP & acq.  & raw      & 0.032 & <0.001 & 0.071 & <0.001  & 0.095 & <0.001 & <0.0001 & <0.0001 \\
+    &              & f.reg.    & & & 0.011 & 0.477 & 0.105  & <0.001  & 0.72 & <0.0001 \\
+    &              & COMBAT    & & &0.013 & 0.4 & 0.122  & <0.001  & 0.88 & <0.0001\\
 \hline
-    & age   & raw       & 0.011 & 0.001  & 0.041 & <0.001  & 0.095  & <0.001 & <0.001 & <0.001 \\
-    &       & f.reg.    & & &0.001 & 0.3 & 0.07  & <0.001  & 0.5 & <0.001 \\
-    &       & COMBAT    & & &0.004 & 0.055 & 0.076 & <0.001 & 0.12 & <0.001 \\
+    & age   & raw       & 0.011 & 0.001  & 0.034 & <0.001  & 0.095  & <0.001 & <0.0001 & <0.0001 \\
+    &       & f.reg.    & & &0.004 & 0.052 & 0.114  & <0.001  & 0.23 & <0.0001 \\
+    &       & COMBAT    & & &0.005 & 0.048 & 0.121 & <0.001 & 0.17 & <0.0001 \\
 \hline
-ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & <0.001 & <0.001 \\
-        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & 0.04 & <0.001 \\
-        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & 0.01 & <0.001 \\
+ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & <0.0001 & <0.0001 \\
+        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & 0.04 & <0.0001 \\
+        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & 0.01 & <0.0001 \\
 \hline
-        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & <0.001& <0.001 \\
-        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & <0.001 & 0.09 \\
-        &          & COMBAT    & & &  0.002 & 0.202 & 0.111     & <0.001 & 0.64 & <0.001 \\
+        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & <0.0001& <0.0001 \\
+        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & <0.0001 & 0.09 \\
+        &          & COMBAT    & & &  0.002 & 0.19 & 0.111     & <0.001 & 0.64 & <0.0001 \\
 
@@
 The tests are made available in the python package \emph{mlconfound} which builds upon a parallelized, computationally effective implementation of the Markov-chain Monte-Carlo based parallel pairwise sampler of \cite{berrett2020conditional}.
-It utilizes, coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal.
+It utilizes coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal.
 The partial and full confounder tests available in \emph{mlconfound} can be uniformly applied for any predictive model, given that realistic (e.g.
@@
 
-As expected from theoretical justification, both tests displayed a valid type I error control in the simulation with the necessary assumptions, that is, even if $\yhat$ is not normally distributed.
+As expected from theoretical justification, both tests displayed a valid type I error control in the simulations with the necessary assumptions, that is, even if $\yhat$ is not normally distributed.
 In fact, the tests were found to become somewhat conservative (i.e.
@@
 
-Next to their statistical power, the practical applicability of the proposed tests is also highly affected by their robustness of the normality assumptions on the conditional distribution of $\y$ and $\c$.
+Next to statistical power, the practical applicability of the proposed tests is also highly affected by their robustness to violations of the normality assumptions on the conditional distribution of $\y$ and $\c$.
 Importantly, the simulations strongly suggest that both tests are strikingly robust to the violation of this assumption.
@@
 
-Validity, power and robustness renders the proposed tests as unprecedented model diagnostic tools in a wide spectrum of predictive modelling scenarios.
-For instance, in biomedical research, confounder-bias has often been investigated (if investigated at all) by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.
+Validity, power and robustness renders the proposed tests as unprecedented tools for model diagnostics in a wide spectrum of predictive modelling scenarios.
+For instance, in biomedical research, confounder-bias has previously often been investigated (if investigated at all) by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.
 \citep{spisak2020pain}.
-However, the significance of this association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable.
+However, the significance of the unconditioned association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable.
 In this case, namely, the model still might not be directly driven by the confounder and the dependence of the predictions on the confounder might be only marginal, to be explained solely by the confounder-target association.
@@
 
-Arguably, in certain applications, the researcher might aim to fully eliminate confounder-related variance of the predictions and in this case testing the association (unconditional independence) between the confounder and the predictions might be sufficient.
+Arguably, in certain applications, the researcher might aim to fully eliminate confounder-related variance of the predictions and, in this case, testing unconditional independence between the confounder and the predictions might be sufficient.
 However, a much more realistic aim is to construct models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e.
@@
 the model might still use the confounder's effect on the features (e.g artifacts) to construct the corresponding part of the variance in the prediction.
-Nevertheless, in most of the cases, it does not seem likely that the model ,at the same time, (i) misses that specific "part" of the target variable's variance in the features that is shared with the confounder (ii) captures exclusively those components of the confounder's variance in the features that is needed to mimic an unbiased model.
-Usually a much more plausible explanation is that a direct confounder effect in the model output is simply negligible  model (e.g because it is not to be found in the features after confound mitigation).
+Nevertheless, in most of the cases, it does not seem likely that the model, at the same time, (i) misses that specific "part" of the target variable's variance in the features that is shared with the confounder (ii) captures exclusively that component of the confounder's variance in the features that is needed to mimic an unbiased model.
+Usually a much more plausible explanation is that a direct confounder effect in the model output is simply negligible (e.g because it is not to be found in the features after confound mitigation).
 
@@
 
-To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two research settings where specific types of confounding effects have previously been described.
+To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two research settings (a regression and a classification problem) where specific types of confounding effects have previously been described.
 First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
@@
 In this case, the bias would likely significantly harm the out-of-sample generalizability of this model.
-The bias of the same model to the acquisition batch (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset.
-There can be manifold reasons for the observed associations.
-Fluid intelligence of the participants can be affected by a changing selection bias during participant recruitment (e.g.
-as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed).
-To explain the associated change in the target variable, the model might utilise changing characteristics of the data (caused by e.g.
+The bias of the same model to 'acquisition batch' (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset.
+There can be manifold reasons for the observed association.
+Fluid intelligence of the participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g.
+as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed over time).
+To explain the associated change in the target variable, the model might utilise changing characteristics of the imaging data (caused by e.g.
 slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff).
@@
 
-In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of autism.
-For instance the model trained on the raw features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g.
+In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD.
+For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g.
 the CMU (Carnegie Mellon University) center as neurotypical control participants.
@@
 The partial confounder test provided detailed insights into the effectiveness of the techniques for each confounder.
-It revealed that, in the case of both the acquisition-bias and the age-bias was very effectively removed by both feature regression and COMBAT (p>0.05 for all).
-Given the high power of the test at these sample sizes (N=999 and 866 for the HCP and the ABIDE datasets, respectively), any remaining confounder-bias is most probably very safely negligible and well out of the range of practical relevance.
+In the HCP data, it revealed that both the acquisition-bias and the age-bias was very effectively removed by both feature regression and COMBAT (p>0.05 for all).
+Given the high power of the test at the sample sizes of the HCP dataset ($N=999$), any remaining confounder-bias is most probably very safely negligible and well out of the range of practical relevance.
 
@@
 The center-bias of the classification in this massively multi-center dataset was, namely, found to be significant, even after feature regression and COMBAT.
-Nevertheless, the p-values returned by the partial confounder test were larger with one order of magnitude than without any confound mitigation, suggesting that a large part of the center effects was still successfully removed by both approaches.
+Nevertheless, the p-values returned by the partial confounder test were larger with two orders of magnitude than without any confound mitigation, suggesting that a large part of the center effects was still successfully removed by both approaches.
 While determining the relevance of the remaining bias is to be performed based on the experimental context and the requirements towards the predictive marker, significant p-values of the partial confounder test generally suggest that these models would either require more effective data harmonization or a validation on external data, that confirms the low practical relevance of the amount of bias.
@@
 Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
-Interestingly, in this case, the association between the confounder and the predictions become even stronger than before but with an opposite sign.
-Meanwhile, the prediction performance significantly dropped (but still remained statistically significant).
-A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD.
-Therefore, regressing out motion estimates from the connectivity features might eliminate significant amount of signal-of-interest (especially in functional connections that are associated with motor deficits in ASD).
-On the other hand, motion artefacts were previously shown to display complex spatiotemporal patterns in fMRI-based functional connectivity estimates \cite{spisak2014voxel} and simply regressing them out of the features is far from being sufficient for removing their effect on the model.
+Interestingly, in the latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign.
+Meanwhile, the prediction performance substantially dropped (although still remained statistically significant).
+A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
+On the other hand, motion artefacts were previously shown to display complex spatiotemporal patterns in fMRI-based functional connectivity estimates \citep{spisak2014voxel} and simply regressing them out of the features is far from being sufficient for removing their effect on the model.
 The so-called "residual motion artefacts" and their regional interactions, as described in the referenced work, might be responsible for the paradox behavior of this model, which raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
@@
 
- In sum, the application of the partial confounder test suggests that confounder-bias should be much more carefully investigated and reported in the predictive modelling literature as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT.
-The partial confoudner test can be considered as a useful benchmark to guide the search for a suitable confounder mitigation approach.
+ In sum, the application of the partial confounder test suggests that confounder-bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT.
+The partial confounder test can be considered as a useful objective benchmark to guide the search for a suitable confounder mitigation approach.
 
-While the real data examples in this paper were not typical cases of the potential application of the full confounder test, it has still provided important insights into model biases.
-Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confoudner test did not provide any evidence that the model captures any extra variance over that explained by the confounder.
-In other words, the test indicated that this model might be, in fact, exclusively driven by motion artifacts.
-For all other models, the full confound test confirmed that - even if there was a significant confounder-bias - the model was not fully driven by the confounder (in other words, the target and the prediction were not independent, given the confounders).
+While the real data examples in this paper were not typical cases of the potential applications of the full confounder test, it has still provided important insights into model biases.
+Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder.
+In other words, the test accepted the null hypothesis of full model bias, suggesting that this model might be, in fact, almost exclusively driven by motion artifacts.
+For all other models, the full confounder test confirmed that - even if there was a significant confounder-bias - the model was not fully driven by the confounder (in other words, the target and the prediction were not independent, given the confounders).
 This examples highlights that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased and the questions is whether there is any biomedical relevant signal captured by the model, at all.
@@
 The lack of rigorous statistical tests for confounder-bias significantly hampers the development of predictive models in many fields of research.
-Even tough several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder signal and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias in the resulting model.
+Even tough several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder signal and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias.
 
@@
 Both approaches are based on a recently published framework for conditional permutation testing, with a mathematical justification of valid type I error control.
-The tests were constructed so that they do not place any assumption on the distribution of the model predictions and require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed very efficiently, without needing to re-fit the model.
+The tests were constructed so that they do not place any assumption on the distribution of the model predictions and require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed with a minimal computational overhead; without needing to re-fit the model.
 
-Given their simplicity, high statistical power, robustness to the violation of the assumptions and thanks to the computationally effective implementation (python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the tests are deployable in a wide variety of applications.
+Given their simplicity, high statistical power, robustness to the violation of the assumptions and thanks to the computationally effective implementation (available in python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the tests are deployable in a wide variety of applications.
 
-As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and significantly enhances the quantitative assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research, thereby paving the way for the development of clinically useful machine learning biomarkers.
+As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and allows quantitative, statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research, thereby paving the way for the development of clinically useful machine learning biomarkers.
 

### 2021-08-16 manuscript.tex
small changes

@@
 Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high computational complexity (especially when hyperparameters are optimized in a nested cross-validation).
-The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the the predictors are a direct cause of the target variable or, alternatively, there exists an unobserved random variable that is the cause of both.
+The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the predictors are a direct cause of the target variable or, alternatively, there exists an unobserved random variable that is the cause of both.
 It's computation assumes 'faithfulness' (if two variables are independent, there is no direct influence between the two in the underlying graph) a strong condition that might not be fulfilled in many real applications (e.g.
@@
 
-Analysis code is available as jupyter notebook\footnote{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb} in the github repository of the 'mlconfound' package.
+Analysis code is available as jupyter notebook\footnote{\href{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb}{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb}} in the github repository of the 'mlconfound' package.
 
@@
 
-Tangent correlation across the timeseries of the n=122 regions of the BASC \citep{bellec2010multi} atlas) was computed with nilearn\footnote{http://nilearn.github.io/} \citep{huntenburg2017loading, esteve2015big}.
+Tangent correlation across the timeseries of the n=122 regions of the BASC \citep{bellec2010multi} atlas) was computed with nilearn\footnote{\href{http://nilearn.github.io/}{http://nilearn.github.io/}} \citep{huntenburg2017loading, esteve2015big}.
 
@@
 
-Analysis code is available as jupyter notebook\footnote{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_abide.ipynb} in the github repository of the 'mlconfound' package.
+Analysis code is available as jupyter notebook\footnote{\href{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_abide.ipynb}{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_abide.ipynb}} in the github repository of the 'mlconfound' package.
 

### 2021-10-06 simulated/run_all.sh
simulations

@@ -1,0 +1,3 @@
+DIR=`dirname $0`
+
+python3 $DIR/simulate.py --out_prefix=$DIR/results/ccc_partial_d1e0_linear --mode=partial --delta=1 --epsilon=0 --nonlin-trf=identity

### 2021-10-06 simulated/simulate.py

@@ -1,0 +1,209 @@
+#!/usr/bin/env python
+import argparse
+import itertools
+import numpy as np
+import pandas as pd
+import pingouin as pg
+from tqdm import tqdm
+from joblib import Parallel, delayed
+import os
+import sys
+from mlconfound.stats import full_confound_test, partial_confound_test
+from mlconfound.simulate import simulate_y_c_yhat, identity, polynomial, sigmoid
+
+path = os.path.dirname(os.path.abspath(__file__))
+sys.path.insert(0, sys.path[0] + '/..')
+
+parser = argparse.ArgumentParser(description="Validate 'mlconfound' on simulated data.")
+
+parser.add_argument("--repetitions", help="Number of repetitions. Default: 100.",
+                    action="store", type=int, default=100)
+
+parser.add_argument("--mode", help="Confound testing mode: 'partial' or 'full'. Default: 'partial'. ",
+                    choices=['partial', 'full'],
+                    type=str, default='partial')
+
+parser.add_argument("--delta", dest="delta", action="store", default=1.0,
+                    help="Delta of the sinh_arcsinh transformation to set kurtosis. "
+                         "Default: 1: no transformation",
+                    type=float)
+parser.add_argument("--epsilon", dest="epsilon", action="store", default=0.0,
+                    help="Epsilon of the sinh_arcsinh transformation to set skewness. "
+                         "Default: 0: no transformation",
+                    type=float)
+
+parser.add_argument("--nonlin-trf", dest="nonlin_trf",
+                    help="Transformation to simulate non-linear dependnecies. "
+                         "Default: identity",
+                    choices=['identity', 'squared', 'sigmoid'],
+                    type=str, default='identity')
+
+parser.add_argument("--cat-y", help="y is categorical (binary). (default: continuous)",
+                    action="store_true")
+parser.add_argument("--cat-yhat", help="yhat is categorical (binary). (default: continuous)",
+                    action="store_true")
+parser.add_argument("--cat-c", help="c is categorical (binary). (default: continuous)",
+                    action="store_true")
+
+parser.add_argument("--random-seed", dest="random_seed", action="store", default=42,
+                    help="Random seed. Default: 42", type=int)
+parser.add_argument("--n-jobs", dest="n_jobs", action="store", default=-1,
+                    help="Number of cores to use. Default: -1 (all available cores)", type=int)
+
[…]

### 2021-10-06 simulated/simulate.py
simulation ready to run

@@ -101,9 +101,7 @@ if __name__ == '__main__':
 
     rng = np.random.default_rng(args.random_seed)
 
-    results = pd.DataFrame(columns=["p_cpt_lin", "p_cpt_gam", "p_pc_pearson", "p_pc_pearson", "r2_y_c", "r2_yhat_c", "r2_y_yhat",
-                                    "n", "w_yc", "w_yyhat", "w_cyhat",
-                                    "num_perms", "random_seed"])
+    results = pd.DataFrame()
 
     for w_yc, w_yyhat, w_cyhat, n in tqdm(list(all_param_configurations)):
 

### 2021-10-18 simulated/run_all.sh
simulation ready

@@ -2,19 +2,19 @@ DIR=`dirname $0`
 
 # H0
 
-python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_linear --mode=partial --delta=1 --epsilon=0 --nonlin-trf=identity --repetitions=1000
+python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_linear.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=identity --repetitions=1000
 
-python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d.5e0_linear --mode=partial --delta=0.5 --epsilon=0 --nonlin-trf=identity --repetitions=1000
+python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e0_linear.csv --mode=partial --delta=0.1 --epsilon=0 --nonlin-trf=identity --repetitions=1000
 
-python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_sigmoid --mode=partial --delta=1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=1000
+python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_sigmoid.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=1000
 
 
 # H0+H1
 
-python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1e0_linear --mode=partial --delta=1 --epsilon=0 --nonlin-trf=identity --repetitions=100
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1e0_linear.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=identity --repetitions=100
 
-python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d.5e0_linear --mode=partial --delta=0.5 --epsilon=0 --nonlin-trf=identity --repetitions=100
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d.5e0_linear.csv --mode=partial --delta=0.5 --epsilon=0 --nonlin-trf=identity --repetitions=100
 
[…]

### 2021-10-18 simulated/simulate_H1.py

@@ -95,11 +95,14 @@ if __name__ == '__main__':
     all_n = [50, 100, 500, 1000]
 
     #     w  0    0.33  0.66  1    2
-    #     r2 0%   10%   25%  50%  80%
-    #     r  0    .3    .5   .7  .9
-    all_yc = [0, 0.33, 0.66, 1, 2]
-    all_yyhat = [0, 0.33, 0.66, 1, 2]
-    all_cyhat = [0, 0.33, 0.66, 1]  # H0 or H1
+    #     r2 0%   10%   30%  50%  80%
+    #     r  0    .3    .55   .7  .9
+    all_yc = [0, 0.33, 0.66, 1]
+    all_yyhat = [0, 0.33, 0.66, 1]
+    #     w  0    0.2  0.4  0.6
+    #     r2 0%   4%   12%  25%
+    #     r  0    .2    .35   .75
+    all_cyhat = [0, 0.2, 0.4, 0.6]  # H0 or H1
     #########################################################################
 
     print('Number of simulated:', np.prod([len(i) for i in [

### 2021-10-18 simulated/run_all.sh
new simulation cases added

@@ -15,6 +15,12 @@ python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1e0_linear.csv --
 
 python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d.5e0_linear.csv --mode=partial --delta=0.5 --epsilon=0 --nonlin-trf=identity --repetitions=100
 
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d.5e.5_linear.csv --mode=partial --delta=0.5 --epsilon=0.5 --nonlin-trf=identity --repetitions=100
+
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d01e0_linear.csv --mode=partial --delta=0.1 --epsilon=0 --nonlin-trf=identity --repetitions=100
+
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d01e.5_linear.csv --mode=partial --delta=0.1 --epsilon=0.5 --nonlin-trf=identity --repetitions=100
+
 python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1e0_sigmoid.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=100
 
 

### 2021-10-19 simulated/run_all.sh
new simulation cases added

@@ -6,6 +6,8 @@ python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_linear.csv --
 
 python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e0_linear.csv --mode=partial --delta=0.1 --epsilon=0 --nonlin-trf=identity --repetitions=1000
 
+python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e5_linear.csv --mode=partial --delta=0.1 --epsilon=5 --nonlin-trf=identity --repetitions=1000
+
 python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_sigmoid.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=1000
 
 

### 2021-10-19 simulated/run_all.sh
new simulation cases added

@@ -6,7 +6,7 @@ python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_linear.csv --
 
 python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e0_linear.csv --mode=partial --delta=0.1 --epsilon=0 --nonlin-trf=identity --repetitions=1000
 
-python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e5_linear.csv --mode=partial --delta=0.1 --epsilon=5 --nonlin-trf=identity --repetitions=1000
+python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e3_linear.csv --mode=partial --delta=0.1 --epsilon=3 --nonlin-trf=identity --repetitions=1000
 
 python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_sigmoid.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=1000
 

### 2021-10-19 simulated/run_all.sh
- new simulation cases added - new figures added

@@ -6,7 +6,13 @@ python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_linear.csv --
 
 python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e0_linear.csv --mode=partial --delta=0.1 --epsilon=0 --nonlin-trf=identity --repetitions=1000
 
+python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e1_linear.csv --mode=partial --delta=0.1 --epsilon=1 --nonlin-trf=identity --repetitions=1000
+
+python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e2_linear.csv --mode=partial --delta=0.1 --epsilon=2 --nonlin-trf=identity --repetitions=1000
+
 python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e3_linear.csv --mode=partial --delta=0.1 --epsilon=3 --nonlin-trf=identity --repetitions=1000
+
+python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e4_linear.csv --mode=partial --delta=0.1 --epsilon=4 --nonlin-trf=identity --repetitions=1000
 
 python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_sigmoid.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=1000
 

### 2021-10-20 fig/raw/emp_abide_motion_comb_partial.tex
emp figures patched

\documentclass{article}
\usepackage[x11names, svgnames, rgb]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usetikzlibrary{snakes,arrows,shapes}
\usepackage{amsmath}
%
%

\usepackage[active,tightpage]{preview}
\PreviewEnvironment{tikzpicture}
\setlength\PreviewBorder{0pt}%

%

\begin{document}
\pagestyle{empty}
%
%
%

\enlargethispage{100cm}
% Start of code
\begin{tikzpicture}[>=latex',line join=bevel,]
%%
\node (c) at (27.0bp,18.0bp) [draw,ellipse] {c};
  \node (y) at (185.5bp,53.0bp) [draw,ellipse] {y};
  \node (yhat) at (344.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
  \draw [] (c) ..controls (82.016bp,30.08bp) and (130.28bp,40.874bp)  .. (y);
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (92.5bp,43.5bp) node {0.028};
  \draw [dashed] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  .. (131.0bp,11.0bp) .. controls (179.42bp,9.3567bp) and (191.58bp,9.3567bp)  .. (240.0bp,11.0bp) .. controls (266.33bp,11.894bp) and (296.39bp,14.069bp)  .. (yhat);
  \draw (185.5bp,18.5bp) node {0.002 (p=0.59)};
  \draw [] (y) ..controls (240.52bp,40.92bp) and (288.78bp,30.126bp)  .. (yhat);
  \draw (278.5bp,43.5bp) node {0.111};
%
\end{tikzpicture}
% End of code

%
\end{document}
%

### 2021-10-20 fig/raw/emp_abide_motion_raw_full.tex
emp figures patched

--- 2021-10-20 fig/raw/emp_abide_motion_comb_partial.tex
@@
 \node (c) at (27.0bp,18.0bp) [draw,ellipse] {c};
-  \node (y) at (185.5bp,53.0bp) [draw,ellipse] {y};
-  \node (yhat) at (344.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
-  \draw [] (c) ..controls (82.016bp,30.08bp) and (130.28bp,40.874bp)  .. (y);
+  \node (y) at (158.0bp,53.0bp) [draw,ellipse] {y};
+  \node (yhat) at (381.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
+  \draw [] (c) ..controls (75.327bp,30.841bp) and (109.82bp,40.199bp)  .. (y);
   \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
   \pgfsetstrokecolor{strokecol}
-  \draw (92.5bp,43.5bp) node {0.028};
-  \draw [dashed] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  ..
-(131.0bp,11.0bp) ..
-controls (179.42bp,9.3567bp) and (191.58bp,9.3567bp)  ..
-(240.0bp,11.0bp) ..
-controls (266.33bp,11.894bp) and (296.39bp,14.069bp)  ..
-(yhat);
-  \draw (185.5bp,18.5bp) node {0.002 (p=0.59)};
-  \draw [] (y) ..controls (240.52bp,40.92bp) and (288.78bp,30.126bp)  .. (yhat);
-  \draw (278.5bp,43.5bp) node {0.111};
+  \draw (92.5bp,47.5bp) node {0.028};
+  \draw [] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  .. (131.0bp,11.0bp) .. controls (212.26bp,8.242bp) and (308.45bp,13.31bp)  .. (yhat);
+  \draw (158.0bp,18.5bp) node {0.111};
+  \draw [dashed] (y) ..controls (226.87bp,42.252bp) and (312.06bp,28.76bp)  .. (yhat);
+  \draw (269.5bp,52.5bp) node {0.126 (p<0.0001*)};
 %

### 2021-10-20 fig/raw/emp_abide_motion_raw_partial.tex
emp figures patched

--- 2021-10-20 fig/raw/emp_abide_motion_comb_partial.tex
@@
 \node (c) at (27.0bp,18.0bp) [draw,ellipse] {c};
-  \node (y) at (185.5bp,53.0bp) [draw,ellipse] {y};
-  \node (yhat) at (344.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
-  \draw [] (c) ..controls (82.016bp,30.08bp) and (130.28bp,40.874bp)  .. (y);
+  \node (y) at (197.5bp,53.0bp) [draw,ellipse] {y};
+  \node (yhat) at (368.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
+  \draw [] (c) ..controls (85.084bp,29.857bp) and (139.57bp,41.174bp)  .. (y);
   \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
   \pgfsetstrokecolor{strokecol}
-  \draw (92.5bp,43.5bp) node {0.028};
+  \draw (92.5bp,42.5bp) node {0.028};
   \draw [dashed] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  ..
 (131.0bp,11.0bp) ..
-controls (179.42bp,9.3567bp) and (191.58bp,9.3567bp)  ..
-(240.0bp,11.0bp) ..
-controls (266.33bp,11.894bp) and (296.39bp,14.069bp)  ..
+controls (190.08bp,8.9948bp) and (204.92bp,8.9948bp)  ..
+(264.0bp,11.0bp) ..
+controls (290.33bp,11.894bp) and (320.39bp,14.069bp)  ..
 (yhat);
-  \draw (185.5bp,18.5bp) node {0.002 (p=0.59)};
-  \draw [] (y) ..controls (240.52bp,40.92bp) and (288.78bp,30.126bp)  .. (yhat);
-  \draw (278.5bp,43.5bp) node {0.111};
+  \draw (197.5bp,18.5bp) node {0.111 (p<0.0001*)};
+  \draw [] (y) ..controls (255.58bp,41.143bp) and (310.07bp,29.826bp)  .. (yhat);
+  \draw (302.5bp,42.5bp) node {0.126};
 %

### 2021-10-20 fig/raw/emp_abide_motion_reg_full.tex
emp figures patched

--- 2021-10-20 fig/raw/emp_abide_motion_raw_full.tex
@@
   \node (y) at (158.0bp,53.0bp) [draw,ellipse] {y};
-  \node (yhat) at (381.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
+  \node (yhat) at (339.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
   \draw [] (c) ..controls (75.327bp,30.841bp) and (109.82bp,40.199bp)  .. (y);
@@
   \draw (92.5bp,47.5bp) node {0.028};
-  \draw [] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  .. (131.0bp,11.0bp) .. controls (212.26bp,8.242bp) and (308.45bp,13.31bp)  .. (yhat);
-  \draw (158.0bp,18.5bp) node {0.111};
-  \draw [dashed] (y) ..controls (226.87bp,42.252bp) and (312.06bp,28.76bp)  .. (yhat);
-  \draw (269.5bp,52.5bp) node {0.126 (p<0.0001*)};
+  \draw [] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  .. (131.0bp,11.0bp) .. controls (196.05bp,8.7922bp) and (272.59bp,13.206bp)  .. (yhat);
+  \draw (158.0bp,18.5bp) node {0.359};
+  \draw [dashed] (y) ..controls (218.38bp,41.389bp) and (278.81bp,29.573bp)  .. (yhat);
+  \draw (248.5bp,50.5bp) node {0.02 (p=0.1)};
 %

### 2021-10-20 fig/raw/emp_abide_motion_reg_partial.tex
emp figures patched

--- 2021-10-20 fig/raw/emp_abide_motion_raw_partial.tex
@@
   \node (y) at (197.5bp,53.0bp) [draw,ellipse] {y};
-  \node (yhat) at (368.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
+  \node (yhat) at (359.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
   \draw [] (c) ..controls (85.084bp,29.857bp) and (139.57bp,41.174bp)  .. (y);
@@
 (131.0bp,11.0bp) ..
-controls (190.08bp,8.9948bp) and (204.92bp,8.9948bp)  ..
+controls (190.08bp,8.9948bp) and (204.93bp,8.8036bp)  ..
 (264.0bp,11.0bp) ..
-controls (290.33bp,11.894bp) and (320.39bp,14.069bp)  ..
+controls (287.06bp,11.857bp) and (313.23bp,13.894bp)  ..
 (yhat);
-  \draw (197.5bp,18.5bp) node {0.111 (p<0.0001*)};
-  \draw [] (y) ..controls (255.58bp,41.143bp) and (310.07bp,29.826bp)  .. (yhat);
-  \draw (302.5bp,42.5bp) node {0.126};
+  \draw (197.5bp,18.5bp) node {0.359 (p<0.0001*)};
+  \draw [] (y) ..controls (253.39bp,40.956bp) and (303.37bp,29.989bp)  .. (yhat);
+  \draw (298.0bp,41.5bp) node {0.02};
 %

### 2021-10-20 fig/raw/emp_abide_site_comb_full.tex
emp figures patched

--- 2021-10-20 fig/raw/emp_abide_motion_reg_full.tex
@@
   \node (y) at (158.0bp,53.0bp) [draw,ellipse] {y};
-  \node (yhat) at (339.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
+  \node (yhat) at (372.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
   \draw [] (c) ..controls (75.327bp,30.841bp) and (109.82bp,40.199bp)  .. (y);
@@
   \pgfsetstrokecolor{strokecol}
-  \draw (92.5bp,47.5bp) node {0.028};
-  \draw [] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  .. (131.0bp,11.0bp) .. controls (196.05bp,8.7922bp) and (272.59bp,13.206bp)  .. (yhat);
-  \draw (158.0bp,18.5bp) node {0.359};
-  \draw [dashed] (y) ..controls (218.38bp,41.389bp) and (278.81bp,29.573bp)  .. (yhat);
-  \draw (248.5bp,50.5bp) node {0.02 (p=0.1)};
+  \draw (92.5bp,47.5bp) node {0.019};
+  \draw [] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  .. (131.0bp,11.0bp) .. controls (208.9bp,8.3558bp) and (301.06bp,13.31bp)  .. (yhat);
+  \draw (158.0bp,18.5bp) node {0.05};
+  \draw [dashed] (y) ..controls (225.43bp,42.033bp) and (304.96bp,28.903bp)  .. (yhat);
+  \draw (265.0bp,52.5bp) node {0.17 (p<0.0001*)};
 %

### 2021-10-20 fig/raw/emp_abide_site_comb_partial.tex
emp figures patched

--- 2021-10-20 fig/raw/emp_abide_motion_comb_partial.tex
@@
 \node (c) at (27.0bp,18.0bp) [draw,ellipse] {c};
-  \node (y) at (185.5bp,53.0bp) [draw,ellipse] {y};
-  \node (yhat) at (344.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
-  \draw [] (c) ..controls (82.016bp,30.08bp) and (130.28bp,40.874bp)  .. (y);
+  \node (y) at (188.5bp,53.0bp) [draw,ellipse] {y};
+  \node (yhat) at (341.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
+  \draw [] (c) ..controls (82.887bp,30.044bp) and (132.87bp,41.011bp)  .. (y);
   \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
   \pgfsetstrokecolor{strokecol}
-  \draw (92.5bp,43.5bp) node {0.028};
+  \draw (92.5bp,43.5bp) node {0.019};
   \draw [dashed] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  ..
 (131.0bp,11.0bp) ..
-controls (179.42bp,9.3567bp) and (191.58bp,9.3567bp)  ..
-(240.0bp,11.0bp) ..
-controls (266.33bp,11.894bp) and (296.39bp,14.069bp)  ..
+controls (182.08bp,9.2662bp) and (194.92bp,9.1009bp)  ..
+(246.0bp,11.0bp) ..
+controls (269.06bp,11.857bp) and (295.23bp,13.894bp)  ..
 (yhat);
-  \draw (185.5bp,18.5bp) node {0.002 (p=0.59)};
-  \draw [] (y) ..controls (240.52bp,40.92bp) and (288.78bp,30.126bp)  .. (yhat);
-  \draw (278.5bp,43.5bp) node {0.111};
+  \draw (188.5bp,18.5bp) node {0.05 (p=0.009*)};
+  \draw [] (y) ..controls (242.32bp,40.716bp) and (287.48bp,30.214bp)  .. (yhat);
+  \draw (280.0bp,42.5bp) node {0.17};
 %

### 2021-10-20 fig/raw/emp_abide_site_reg_partial.tex
emp figures patched

--- 2021-10-20 fig/raw/emp_abide_site_comb_partial.tex
@@
   \node (y) at (188.5bp,53.0bp) [draw,ellipse] {y};
-  \node (yhat) at (341.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
+  \node (yhat) at (350.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
   \draw [] (c) ..controls (82.887bp,30.044bp) and (132.87bp,41.011bp)  .. (y);
@@
 (131.0bp,11.0bp) ..
-controls (182.08bp,9.2662bp) and (194.92bp,9.1009bp)  ..
+controls (182.08bp,9.2662bp) and (194.92bp,9.2662bp)  ..
 (246.0bp,11.0bp) ..
-controls (269.06bp,11.857bp) and (295.23bp,13.894bp)  ..
+controls (272.33bp,11.894bp) and (302.39bp,14.069bp)  ..
 (yhat);
-  \draw (188.5bp,18.5bp) node {0.05 (p=0.009*)};
-  \draw [] (y) ..controls (242.32bp,40.716bp) and (287.48bp,30.214bp)  .. (yhat);
-  \draw (280.0bp,42.5bp) node {0.17};
+  \draw (188.5bp,18.5bp) node {0.042 (p=0.04*)};
+  \draw [] (y) ..controls (244.39bp,40.956bp) and (294.37bp,29.989bp)  .. (yhat);
+  \draw (284.5bp,43.5bp) node {0.173};
 %

### 2021-10-20 fig/raw/emp_hcp_age_reg_partial.tex
emp figures patched

--- 2021-10-20 fig/raw/emp_abide_motion_comb_partial.tex
@@
 \node (c) at (27.0bp,18.0bp) [draw,ellipse] {c};
-  \node (y) at (185.5bp,53.0bp) [draw,ellipse] {y};
-  \node (yhat) at (344.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
-  \draw [] (c) ..controls (82.016bp,30.08bp) and (130.28bp,40.874bp)  .. (y);
+  \node (y) at (181.0bp,53.0bp) [draw,ellipse] {y};
+  \node (yhat) at (335.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
+  \draw [] (c) ..controls (81.025bp,30.21bp) and (126.99bp,40.794bp)  .. (y);
   \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
   \pgfsetstrokecolor{strokecol}
-  \draw (92.5bp,43.5bp) node {0.028};
+  \draw (92.5bp,43.5bp) node {0.011};
   \draw [dashed] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  ..
 (131.0bp,11.0bp) ..
-controls (179.42bp,9.3567bp) and (191.58bp,9.3567bp)  ..
-(240.0bp,11.0bp) ..
-controls (266.33bp,11.894bp) and (296.39bp,14.069bp)  ..
+controls (175.42bp,9.4924bp) and (186.58bp,9.4924bp)  ..
+(231.0bp,11.0bp) ..
+controls (257.33bp,11.894bp) and (287.39bp,14.069bp)  ..
 (yhat);
-  \draw (185.5bp,18.5bp) node {0.002 (p=0.59)};
-  \draw [] (y) ..controls (240.52bp,40.92bp) and (288.78bp,30.126bp)  .. (yhat);
-  \draw (278.5bp,43.5bp) node {0.111};
+  \draw (181.0bp,18.5bp) node {0.004 (p=0.2)};
+  \draw [] (y) ..controls (235.03bp,40.79bp) and (280.99bp,30.206bp)  .. (yhat);
+  \draw (269.5bp,43.5bp) node {0.114};
 %

### 2021-10-21 manuscript.tex
Updates from Overleaf

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!b]
+  \centering
+  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
+  \caption{\textbf{Type-I error control and power of the conditional permutation-based confound testing for predictive modelling in normally distributed simulated data}. \\
+  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette at the bottom) in various simulations settings (100 simulations per tile) with different covariance between the target and the confounder (given as $R^2_{y,c}$ on the vertical axes of the heatmaps), for different sample sizes (N, horizontal axes).
+  Rows and columns correspond to simulations with different signal-in-prediction (y+c) and confound-to-signal values, respectively.
+The first row and the first column contain simulations under H0 (no confounder bias) and separated form the simulations under the alternative hypothesis with dashed line.
+Positive rates for the simulations under the null and alternative hypotheses can be interpreted as type I error rate and statistical power, respectively.
+The higher confidence limit for the type I error rates is 0.11 for each tile, with 5 tiles expected to pass it form all H0 tiles.
+  The actual weights $w_{Y}$, $w_{C}$ and $w_{e}$ are depicted at the bars on top of each heatmap and color coded as shown in the legend at the bottom.
+Above each heatmap, the strengths of the associations of $\hat{y}$ to $Y$ and $C$, respectively, are given with $R^2$ values averaged over all simulation cases of the heatmap.
+  }
+  \label{fig:sim-normal}
+\end{figure}
+
 \subsection{Validation on simulated data}
@@
   \centering
+
   \includegraphics[width=0.40\paperwidth]{fig/sim_non-norm.eps}
@@
   \centering
-  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.pdf}
+  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
   \caption{\textbf{Acquisition- and age-bias of fluid IQ prediction in the HCP dataset.} \\
@@
   \centering
-  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.pdf}
+  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.png}
   \caption{\textbf{Center- and motion age-bias of ASD diagnosis prediction in the ABIDE dataset.} \\
@@
 \hline
-HCP & acq.  & raw      & 0.032 & <0.001 & 0.071 & <0.001  & 0.095 & <0.001 & <0.0001 & <0.0001 \\
-    &              & f.reg.    & & & 0.011 & 0.477 & 0.105  & <0.001  & 0.72 & <0.0001 \\
-    &              & COMBAT    & & &0.013 & 0.4 & 0.122  & <0.001  & 0.88 & <0.0001\\
+HCP & acq.  & raw      & 0.032 & <0.001 & 0.071 & <0.001  & 0.095 & <0.001 & \textbf{<0.0001} & <0.0001 \\
+    &              & f.reg.    & & & 0.011 & 0.477 & 0.105  & <0.001  & 0.73 & <0.0001 \\
+    &              & COMBAT    & & &0.013 & 0.4 & 0.122  & <0.001  & 0.65 & <0.0001\\
 \hline
-    & age   & raw       & 0.011 & 0.001  & 0.034 & <0.001  & 0.095  & <0.001 & <0.0001 & <0.0001 \\
-    &       & f.reg.    & & &0.004 & 0.052 & 0.114  & <0.001  & 0.23 & <0.0001 \\
-    &       & COMBAT    & & &0.005 & 0.048 & 0.121 & <0.001 & 0.17 & <0.0001 \\
+    & age   & raw       & 0.011 & 0.001  & 0.034 & <0.001  & 0.095  & <0.001 & \textbf{<0.0001} & <0.0001 \\
+    &       & f.reg.    & & &0.004 & 0.052 & 0.114  & <0.001  & 0.2 & <0.0001 \\
+    &       & COMBAT    & & &0.005 & 0.048 & 0.121 & <0.001 & 0.16 & <0.0001 \\
 \hline
-ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & <0.0001 & <0.0001 \\
-        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & 0.04 & <0.0001 \\
-        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & 0.01 & <0.0001 \\
+ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & \textbf{<0.0001} & <0.0001 \\
+        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & \textbf{0.04} & <0.0001 \\
+        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & \textbf{0.009} & <0.0001 \\
 \hline
-        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & <0.0001& <0.0001 \\
-        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & <0.0001 & 0.09 \\
-        &          & COMBAT    & & &  0.002 & 0.19 & 0.111     & <0.001 & 0.64 & <0.0001 \\
+        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & \textbf{<0.0001} & <0.0001 \\
+        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & \textbf{<0.0001} & 0.1 \\
+        &          & COMBAT    & & &  0.002 & 0.19 & 0.111     & <0.001 & 0.59 & <0.0001 \\
 
 \end{tabular}
-\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated models.}
+\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated models.
+Significant confoudner bias is denoted by bold font.}
 \end{table}

### 2021-10-21 simulated/run_all.sh
new simulation cases added

@@ -31,4 +31,14 @@ python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d01e.5_linear.csv 
 
 python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1e0_sigmoid.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=100
 
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1e1_linear.csv --mode=partial --delta=1 --epsilon=1 --nonlin-trf=identity --repetitions=100
+
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1.05e3_linear.csv --mode=partial --delta=1.05 --epsilon=3 --nonlin-trf=identity --repetitions=100
+
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1.05e3_linear.csv --mode=partial --delta=1.5 --epsilon=5 --nonlin-trf=identity --repetitions=100
+
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1.05e3_linear.csv --mode=partial --delta=5 --epsilon=10 --nonlin-trf=identity --repetitions=100
+
+
+
 

### 2021-10-21 simulated/normality_and_linearity_violation.tex
example notebook

\hypertarget{an-example-of-model-non-linearity-in-the-presence-of-a-linear-predictor.}{%
\section{An example of model non-linearity in the presence of a linear
predictor.}\label{an-example-of-model-non-linearity-in-the-presence-of-a-linear-predictor.}}

Notebook available on-line:
https://github.com/spisakt/mlconfound\_manuscript/tree/main/simulated/normality\_and\_linearity\_violation.ipynb

\begin{lstlisting}[language=Python]
import numpy as np
import seaborn as sns
import matplotlib.pyplot as plt
from sklearn.linear_model import Ridge
sns.set(rc={"figure.figsize":(6, 2)})
sns.set_style("white")
\end{lstlisting}

\hypertarget{five-features-one-linear-four-sigmoid}{%
\subsubsection{Five features: one linear, four
sigmoid}\label{five-features-one-linear-four-sigmoid}}

\begin{lstlisting}[language=Python]
n=300
p=5
rng = np.random.default_rng(42)
y = np.arange(n)
y = (y - y.mean())/y.std()

X_true = np.repeat(y, p).reshape(n,p)
for i in range(1,4):
    X_true[:,i] = np.tanh(X_true[:,i]*2)

X=X_true + rng.normal(0,0.1, (n,p))

for i in range(4):
    sns.lineplot(x=y, y=X_true[:,i])
    sns.scatterplot(x=y, y=X[:,i])
    plt.show()
\end{lstlisting}

\begin{figure}
\centering
\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_3_0.png}
\caption{png}
\end{figure}

\begin{figure}
\centering
\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_3_1.png}
\caption{png}
\end{figure}

\begin{figure}
\centering
\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_3_2.png}
\caption{png}
\end{figure}

\begin{figure}
\centering
\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_3_3.png}
\caption{png}
\end{figure}

\hypertarget{model-predictions-without-regularization}{%
\subsubsection{\texorpdfstring{Model predictions \textbf{without}
regularization}{Model predictions without regularization}}\label{model-predictions-without-regularization}}

\begin{lstlisting}[language=Python]
model = Ridge(alpha=0)
model.fit(y=y, X=X)
yhat = model.predict(X)
sns.scatterplot(x=y, y=yhat, alpha=0.3)
#sns.lineplot(x=y, y=y, linestyle="--")
model.coef_
\end{lstlisting}

\begin{lstlisting}
array([ 0.47902429,  0.00946499,  0.07378478, -0.02828672,  0.47070535])
\end{lstlisting}

\begin{figure}
\centering
\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_5_1.png}
\caption{png}
\end{figure}

\hypertarget{model-predictions-with-regularization}{%
\subsubsection{\texorpdfstring{Model predictions \textbf{with}
regularization}{Model predictions with regularization}}\label{model-predictions-with-regularization}}

\begin{lstlisting}[language=Python]
model = Ridge(alpha=2000)
model.fit(y=y, X=X)
yhat = model.predict(X)
sns.scatterplot(x=y, y=yhat, alpha=0.3)
#sns.lineplot(x=y, y=y, linestyle="--")
model.coef_
\end{lstlisting}

\begin{lstlisting}
array([0.09472368, 0.07472972, 0.07338482, 0.07440682, 0.09463765])
\end{lstlisting}

\begin{figure}
\centering
\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_7_1.png}
\caption{png}
\end{figure}

\begin{lstlisting}[language=Python]
\end{lstlisting}

### 2021-10-21 manuscript.tex

@@
 
-\title{A conditional permutation-based approach to test confounder effect and center-bias in machine learning models}
+\title{A conditional permutation-based approach to test confounder and center effects in machine learning models}
 
@@
 \begin{abstract} % 150 words
-The lack of rigorous non-parametric statistical tests of confounder-bias significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
-Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unbiased} and \emph{fully biased model}, respectively.
-The proposed tests set no assumptions about the distribution of the predictive model output that is often non-normal and display a remarkable robustness and a high statistical power in simulations.
-Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset provided statistical evidence for several models being significantly confounded.
-Some of these confounder-biases were previously unreported and found to be hard to fully mitigate with state-of-the-art confound mitigation approaches.
+The lack of rigorous non-parametric statistical tests of confounder effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
+Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unconfounded} and \emph{fully confounded model}, respectively.
+The proposed methods provide a strict control for Type-I errors and high statistical power, even in the presence of typical non-linear and non-normal dependencies between the target variable and the model output.
+Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset reveals confoudner effects that were previously unreported and found to be hard to correct for with state-of-the-art confound mitigation approaches.
 The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment and improvement of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
@@
 
-Predictive modelling and supervised learning methods have recently became increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized}.
+Predictive modelling and supervised learning methods have recently become increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized}.
 When evaluating the usefulness and applicability of such markers, predictive performance (i.e.
 prognostic/diagnostic value) is far from being the only important consideration.
-Generalizability across contexts and populations and biomedical validity (i.e.
-whether the model is driven by biomedically relevant signal) are also crucial requirements for candidate biomarkers \citep{woo2017building}.
+Biomedical validity (i.e.
+whether the model is driven by biomedically relevant signal) and generalizability across contexts and populations are also crucial requirements for candidate biomarkers \citep{woo2017building}.
 
-Variables that can induce spurious associations between the predictor variables (features) and the prediction target - often referred to as confounders - can be detrimental to the model's biomedical validity and generalizability.
-Many types of confounders can be distinguished.
+Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - induced by so-called confounder effects - can be detrimental to the model's biomedical validity and generalizability.
+Many types of confounder effects can be distinguished.
+
 First, measurement artifacts can obviously be considered as confounders.
-For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with movement kinematics or motor function, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive} attention deficit hyperactivity disorder (ADHD) \citep{couvy2016head, eloyan2012automated} or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}.
-Similarly, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.
+For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with motor function, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive} attention deficit hyperactivity disorder (ADHD) \citep{couvy2016head, eloyan2012automated} or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}.
+Likewise, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.
+
 Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling.
-For instance, resting-state functional connectivity is known to be strongly predictive to  demographic age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}.
-As fluid intelligence is known to decline with aging, \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by picking up on solely age-related variance \citep{lohmann2021predicting, dubois2018distributed}.
+For instance, resting-state functional connectivity has been shown to be strongly predictive to age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}.
+As fluid intelligence is known to decline with aging \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by picking up on solely age-related variance \citep{lohmann2021predicting, dubois2018distributed}.
+
 Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance} as the models can utilize such effects to explain (random or true) center/batch-differences in the predictive target.
@@
 
-Various data harmonization approaches have been proposed to mitigate confounder-bias, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}.
+Various data cleansing, correction and harmonization approaches have been proposed to mitigate detrimental effects of confounders, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}.
 However, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.
 
-Powerful and robust statistical tests to quantify the confounding effect of certain variables in predictive models could largely foster both the the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation and data harmonization approaches.
-Such methods, however, have to tackle the fact that the outputs of predictive models often follow non-normal distributions (due to e.g.
-regularization) \citep{garcia2009study, kristensen2017whole}.
+Powerful and robust statistical tests to quantify the confounding effect of certain variables in predictive models could substantially foster both the the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation and data harmonization approaches.
+Such methods, however, have to tackle the fact that the conditional distribution of the models' output, given the target, is often non-normal and non-linear, as a consequence of e.g.
+model regularization \citep{garcia2009study, kristensen2017whole}.
 

### 2021-10-21 simulated/normality_and_linearity_violation.tex
example notebook

@@
 
-\begin{lstlisting}[language=Python]
-import numpy as np
-import seaborn as sns
-import matplotlib.pyplot as plt
-from sklearn.linear_model import Ridge
-sns.set(rc={"figure.figsize":(6, 2)})
-sns.set_style("white")
-\end{lstlisting}
+\begin{Shaded}
+\begin{Highlighting}[]
+\ImportTok{import}\NormalTok{ numpy }\ImportTok{as}\NormalTok{ np}
+\ImportTok{import}\NormalTok{ seaborn }\ImportTok{as}\NormalTok{ sns}
+\ImportTok{import}\NormalTok{ matplotlib.pyplot }\ImportTok{as}\NormalTok{ plt}
+\ImportTok{from}\NormalTok{ sklearn.linear_model }\ImportTok{import}\NormalTok{ Ridge}
+\NormalTok{sns.}\BuiltInTok{set}\NormalTok{(rc}\OperatorTok{=}\NormalTok{\{}\StringTok{"figure.figsize"}\NormalTok{:(}\DecValTok{6}\NormalTok{, }\DecValTok{2}\NormalTok{)\})}
+\NormalTok{sns.set_style(}\StringTok{"white"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-\begin{lstlisting}[language=Python]
-n=300
-p=5
-rng = np.random.default_rng(42)
-y = np.arange(n)
-y = (y - y.mean())/y.std()
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{n}\OperatorTok{=}\DecValTok{300}
+\NormalTok{p}\OperatorTok{=}\DecValTok{5}
+\NormalTok{rng }\OperatorTok{=}\NormalTok{ np.random.default_rng(}\DecValTok{42}\NormalTok{)}
+\NormalTok{y }\OperatorTok{=}\NormalTok{ np.arange(n)}
+\NormalTok{y }\OperatorTok{=}\NormalTok{ (y }\OperatorTok{-}\NormalTok{ y.mean())}\OperatorTok{/}\NormalTok{y.std()}
 
-X_true = np.repeat(y, p).reshape(n,p)
-for i in range(1,4):
-    X_true[:,i] = np.tanh(X_true[:,i]*2)
+\NormalTok{X_true }\OperatorTok{=}\NormalTok{ np.repeat(y, p).reshape(n,p)}
+\ControlFlowTok{for}\NormalTok{ i }\KeywordTok{in} \BuiltInTok{range}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{4}\NormalTok{):}
+\NormalTok{    X_true[:,i] }\OperatorTok{=}\NormalTok{ np.tanh(X_true[:,i]}\OperatorTok{*}\DecValTok{2}\NormalTok{)}
 
-X=X_true + rng.normal(0,0.1, (n,p))
+\NormalTok{X}\OperatorTok{=}\NormalTok{X_true }\OperatorTok{+}\NormalTok{ rng.normal(}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{, (n,p))}
 
-for i in range(4):
-    sns.lineplot(x=y, y=X_true[:,i])
-    sns.scatterplot(x=y, y=X[:,i])
-    plt.show()
-\end{lstlisting}
+\ControlFlowTok{for}\NormalTok{ i }\KeywordTok{in} \BuiltInTok{range}\NormalTok{(}\DecValTok{4}\NormalTok{):}
+\NormalTok{    sns.lineplot(x}\OperatorTok{=}\NormalTok{y, y}\OperatorTok{=}\NormalTok{X_true[:,i])}
+\NormalTok{    sns.scatterplot(x}\OperatorTok{=}\NormalTok{y, y}\OperatorTok{=}\NormalTok{X[:,i])}
+\NormalTok{    plt.show()}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-\begin{lstlisting}[language=Python]
-model = Ridge(alpha=0)
-model.fit(y=y, X=X)
-yhat = model.predict(X)
-sns.scatterplot(x=y, y=yhat, alpha=0.3)
-#sns.lineplot(x=y, y=y, linestyle="--")
-model.coef_
-\end{lstlisting}
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{model }\OperatorTok{=}\NormalTok{ Ridge(alpha}\OperatorTok{=}\DecValTok{0}\NormalTok{)}
+\NormalTok{model.fit(y}\OperatorTok{=}\NormalTok{y, X}\OperatorTok{=}\NormalTok{X)}
+\NormalTok{yhat }\OperatorTok{=}\NormalTok{ model.predict(X)}
+\NormalTok{sns.scatterplot(x}\OperatorTok{=}\NormalTok{y, y}\OperatorTok{=}\NormalTok{yhat, alpha}\OperatorTok{=}\FloatTok{0.3}\NormalTok{)}
+\CommentTok{#sns.lineplot(x=y, y=y, linestyle="--")}
+\NormalTok{model.coef_}
+\end{Highlighting}
+\end{Shaded}
 
-\begin{lstlisting}
+\begin{verbatim}
 array([ 0.47902429,  0.00946499,  0.07378478, -0.02828672,  0.47070535])
-\end{lstlisting}
+\end{verbatim}
 
@@
 
-\begin{lstlisting}[language=Python]
-model = Ridge(alpha=2000)
-model.fit(y=y, X=X)
-yhat = model.predict(X)
-sns.scatterplot(x=y, y=yhat, alpha=0.3)
-#sns.lineplot(x=y, y=y, linestyle="--")
-model.coef_
-\end{lstlisting}
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{model }\OperatorTok{=}\NormalTok{ Ridge(alpha}\OperatorTok{=}\DecValTok{2000}\NormalTok{)}
+\NormalTok{model.fit(y}\OperatorTok{=}\NormalTok{y, X}\OperatorTok{=}\NormalTok{X)}
+\NormalTok{yhat }\OperatorTok{=}\NormalTok{ model.predict(X)}
+\NormalTok{sns.scatterplot(x}\OperatorTok{=}\NormalTok{y, y}\OperatorTok{=}\NormalTok{yhat, alpha}\OperatorTok{=}\FloatTok{0.3}\NormalTok{)}
+\CommentTok{#sns.lineplot(x=y, y=y, linestyle="--")}
+\NormalTok{model.coef_}
+\end{Highlighting}
+\end{Shaded}
 
-\begin{lstlisting}
+\begin{verbatim}
 array([0.09472368, 0.07472972, 0.07338482, 0.07440682, 0.09463765])
-\end{lstlisting}
+\end{verbatim}
 
@@
 
-\begin{lstlisting}[language=Python]
-\end{lstlisting}
+\begin{Shaded}
+\begin{Highlighting}[]
+
+\end{Highlighting}
+\end{Shaded}

### 2021-10-21 simulated/normality_and_linearity_violation.tex
example notebook

--- 2021-10-21 simulated/normality_and_linearity_violation.tex
@@
 \centering
-\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_3_0.png}
+\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_0.png}
 \caption{png}
@@
 \centering
-\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_3_1.png}
+\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_1.png}
 \caption{png}
@@
 \centering
-\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_3_2.png}
+\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_2.png}
 \caption{png}
@@
 \centering
-\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_3_3.png}
+\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_3.png}
 \caption{png}
@@
 sns.scatterplot(x=y, y=yhat, alpha=0.3)
-#sns.lineplot(x=y, y=y, linestyle="--")
 model.coef_
@@
 \centering
-\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_5_1.png}
+\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_5_1.png}
 \caption{png}
@@
 sns.scatterplot(x=y, y=yhat, alpha=0.3)
-#sns.lineplot(x=y, y=y, linestyle="--")
 model.coef_
+
 \end{lstlisting}
@@
 \centering
-\includegraphics{normality_and_linearity_violation_files/normality_and_linearity_violation_7_1.png}
+\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_7_1.png}
 \caption{png}
 \end{figure}
-
-\begin{lstlisting}[language=Python]
-\end{lstlisting}

### 2021-10-21 manuscript.tex
patched

@@
 
-Powerful and robust statistical tests to quantify the confounding effect of certain variables in predictive models could substantially foster both the the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation and data harmonization approaches.
+Powerful and robust statistical tests to quantify confounding effects in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches.
 Such methods, however, have to tackle the fact that the conditional distribution of the models' output, given the target, is often non-normal and non-linear, as a consequence of e.g.
-model regularization \citep{garcia2009study, kristensen2017whole}.
+model regularization \citep{garcia2009study, kristensen2017whole}; see Supplementary Material \ref{sup:nomlinviol} for a simplistic example.
 
-Recently, three dedicated techniques \citep{chaibub2019permutation, ferrari2020measuring, wachinger2021detect} have been proposed for quantifying confounder bias in predictive modelling.
+Recently, three dedicated techniques \citep{chaibub2019permutation, ferrari2020measuring, wachinger2021detect} have been proposed for quantifying confounder effects in predictive modelling.
 
-The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via "restricted permutations" that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature) However, as shown by \cite{southworth2009properties} and further discussed by \cite{hemerik2018exact}, balanced permutations do not provide a correct null behavior and can be invalid in certain circumstances.
+The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via "restricted permutations" that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature).
+However, as shown by \cite{southworth2009properties} and further discussed by \cite{hemerik2018exact}, balanced permutations do not provide a correct null behavior and can be invalid in certain circumstances.
 Indeed, \cite{ferrari2020measuring} demonstrated that this method does not guarantee a proper type I error control with biased classification problems.
-The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, by design, does not provide any quantification of statistical significance.
+The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, on the other hand, does not provide any quantification of statistical significance.
 Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high computational complexity (especially when hyperparameters are optimized in a nested cross-validation).
-The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the predictors are a direct cause of the target variable or, alternatively, there exists an unobserved random variable that is the cause of both.
-It's computation assumes 'faithfulness' (if two variables are independent, there is no direct influence between the two in the underlying graph) a strong condition that might not be fulfilled in many real applications (e.g.
-functional MRI is an \emph{indirect} measure of neural activation, rendering the alternative hypothesis of this test true in absence of any real confounder).
+The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the predictors are a direct cause of the target variable or, alternatively, there is an unobserved random variable that is the cause of both.
+However, this null hypothesis might not be of interest in many real applications.
+For instance, causation between several psychological or clinical states and functional MRI activity-connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation.
+In such cases, the alternative hypothesis of this test might be true in absence of any real confounder.
 
@@
 This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ("no free lunch" theorem).
+As an important consequence, partial correlation - a trivial candidate for investigating the dependence between the predictions and the confounder (given the target variable) - might fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved.
+These conditions are very common for case of predictive modelling \citep{garcia2009study, kristensen2017whole} (Supplementary Material \ref{sup:nomlinviol})
 
-Next, I utilize the permutation-based conditional independence framework of \cite{berrett2020conditional}, to construct two novel tests for evaluating the confounder-bias in predictive models.
+To overcome the limitations of partial correlation (demonstrated on Fig.
+\ref{fig:sim-h0-demo}), I build on the permutation-based conditional independence framework of \cite{berrett2020conditional}, to construct two novel tests for evaluating the confounder-bias in predictive models.
 The \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder.
@@
 The tests set no assumption about the distribution of the model predictions.
-The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality).
-The proposed tests are then applied on functional brain connectivity data to provide evidence of motion- and center-bias in a model trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} and to evaluate the age and acquisition-batch dependence of fluid intelligence predictions on data from the Human Connectome Project.
+The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality and non-linearity).
+The proposed tests are then applied on functional brain connectivity data to provide evidence of motion- and center-bias in a model trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} and to evaluate the age and acquisition-batch dependence of fluid intelligence predictions on data from the Human Connectome Project \citep{van2013wu}.
 
@@
 \ref{fig:schematic}B).
-Let us assume, furthermore, that $\y$, $\yhat$ and $\c$ consist of independent and identically distributed (IID) data points $(y_i, \hat{y}_i, c_i) \in Y \times \\hat{Y} \times C$ for $i=1, \dots , n$ so that $\y=(y_1, \dots ,y_n)$, $\yhat=(\hat{y}_1, \dots, \hat{y}_n)$ and $\c=(c_1, \dots, c_n)$.
+Let us assume, furthermore, that $\y$, $\yhat$ and $\c$ consist of independent and identically distributed (IID) data points $(y_i, \hat{y}_i, c_i) \in Y \times \hat{Y} \times C$ for $i=1, \dots , n$ so that $\y=(y_1, \dots ,y_n)$, $\yhat=(\hat{y}_1, \dots, \hat{y}_n)$ and $\c=(c_1, \dots, c_n)$.
 
@@
 
-Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; it's predictions are driven by the confounder rather than information about the target variable.
+Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; its predictions are driven by the confounder rather than information about the target variable.
 Testing unconditional independence ($\yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
 For instance, even if $\yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to the dependence of both on $\y$.
-In other words, if the target variable $\y$ is significantly associated to the confounder variable $\c$, a model that is completely blind to $\c$ (and, therefore, unbiased) might still provide outputs $\yhat$ that are significantly associated with $\c$.
+In other words, if the target variable $\y$ is significantly associated to the confounder variable $\c$, a model that is completely blind to $\c$ (i.e not confounded) might still provide outputs $\yhat$ that are significantly associated with $\c$.
 
@@
 
-Instead of focusing on the unconditional independence between the confounder and the predictions, we shall consider the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, by definition  \citep{dawid1979conditional}, means that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$.
+Instead of focusing on the unconditional independence between the confounder and the predictions, we shall consider the \emph{conditional independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, by definition  \citep{dawid1979conditional}, means that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$.
 The statistical test with this null hypothesis will be referred to as the \emph{partial confounder test}.
 
-Alternatively, one might be also interested in testing $\yhat \independent \y | \c$. We will refer to the corresponding test as the \emph{full confounder test}.
+Alternatively, one might also be interested in testing $\yhat \independent \y | \c$. We will refer to the corresponding test as the \emph{full confounder test}.
 
@@
  Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional independence testing is effectively impossible.
-In other words, neither the full nor the partial confounder tests can be constructed so that - for all distributions - they provides a valid type I error control and, at the same time, a non-trivial statistical power.
+In other words, neither the full nor the partial confounder tests can be constructed so that - for all distributions - they provide a valid type I error control and, at the same time, a non-trivial statistical power.
 
-This result is in strong contrast to unconditional independence testing - where permutation tests  \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs, depending on the applied machine learning model, is unknown and often heavily non-normal.
+This result stands in strong contrast to unconditional independence testing - where permutation tests  \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs, depending on the applied machine learning model, is unknown and often heavily non-normal.
 
-Even if constructing fully non-parametric confounder tests was proven to be impossible, as demonstrated by \citep{candes2016panning} and \cite{berrett2020conditional}, valid and powerful conditional independence tests can be constructed with inputting distributional information about only two (out of the three) variables.
-Specifically, the conditional permutation test (CPT) of Berret and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable.
+One of the trivial candidate for the task, partial correlation, for instance assumes that all involved variables are multivariate Gaussian and - as to be shown below - even its Spearman-based variant is unable to tolerate relatively small deviations from normality and linearity.
+
+Recently, \cite{candes2016panning} and \cite{berrett2020conditional} have demonstrated that valid and powerful conditional independence tests can be constructed with inputting distributional information about only two (out of the three) variables.
+Specifically, the conditional permutation test (CPT) of Berrett and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable.
 Thereby, it incorporates the information available on the conditional distribution of interest into the permutation-based inference in a statistically valid manner.
+Moreover, as shown in this paper, CPT can be combined with a general additive model to handle non-linear relationships, rendering it as an appealing candidate for testing confounder effects in predictive modelling.
+
+As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is a set of predictors (features), $Y$ is the target variable and $Z$ is a potential confounder.
+In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether the features $X$ still affect $Y$, when controlling for $Z$.
+For instance, in genome-wide association studies, CPT can be used to determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.
+
+In this paper, a different setting is considered, where the supervised learning model is already fitted and we are focusing on model diagnostics, by testing the triplet $(\y,\yhat, \c)$. 
+The conceptual difference between the original and the proposed application of CPT is depicted on Fig. \ref{fig:schematic}.
 
@@
 
-As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is a set of predictors (features), $Y$ is the target variable and $Z$ is a potential confounder.
-In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether the features $X$ still affect $Y$, when controlling for $Z$.
-For instance, in genome-wide association studies, CPT can be used to determine whether a particular genetic variant $X$ affects a response $Y$ such as disease status or some other phenotype, even after controlling for the rest of the genome, encoded in $Z$.
-
-In this paper, a different setting is considered, where the supervised learning model is already fitted and we are focusing on model diagnostics, by testing the triplet $(\y,\yhat, \c)$.
-The conceptual difference between the original and the proposed application of CPT is depicted on Fig.
-\ref{fig:schematic}.
-
 Within this setting, conditional independence testing and, specifically, the framework of conditional permutation testing allows investigating three different null hypotheses corresponding to the $(\y, \yhat, \c)$ triplet.
@@
 
-\subsubsection*{Gaussian conditional log-likelihood}
+\subsubsection*{Conditional log-likelihood}
 
@@
   }
-  \label{fig:sim-normal}
+  \label{fig:sim-h0-demo}
 \end{figure}

### 2021-10-22 simulated/normality_and_linearity_violation.tex

@@
 predictor.}\label{an-example-of-model-non-linearity-in-the-presence-of-a-linear-predictor.}}
+\label{sup:nomlinviol}
 
 Notebook available on-line:
+
 https://github.com/spisakt/mlconfound\_manuscript/tree/main/simulated/normality\_and\_linearity\_violation.ipynb
@@
 
-\begin{figure}
+\begin{figure}[H]
 \centering
-\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_0.png}
-\caption{png}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_1.png}
-\caption{png}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_2.png}
-\caption{png}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_3.png}
-\caption{png}
+\includegraphics[width=0.15\paperwidth]{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_0.png}
+\includegraphics[width=0.15\paperwidth]{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_1.png}
+\includegraphics[width=0.15\paperwidth]{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_2.png}
+\includegraphics[width=0.15\paperwidth]{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_3_3.png}
 \end{figure}
@@
 
-\begin{figure}
+\begin{figure}[H]
 \centering
-\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_5_1.png}
-\caption{png}
+\includegraphics[width=0.5\paperwidth]{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_5_1.png}
 \end{figure}
@@
 
-\begin{figure}
+\begin{figure}[H]
 \centering
-\includegraphics{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_7_1.png}
-\caption{png}
+\includegraphics[width=0.5\paperwidth]{fig/normality_and_linearity_violation_files/normality_and_linearity_violation_7_1.png}
 \end{figure}

### 2021-10-22 simulated/run_all.sh
new simulation cases: sigmoid

@@ -29,8 +29,6 @@ python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d01e0_linear.csv -
 
 python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d01e.5_linear.csv --mode=partial --delta=0.1 --epsilon=0.5 --nonlin-trf=identity --repetitions=100
 
-python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1e0_sigmoid.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=100
-
 python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1e1_linear.csv --mode=partial --delta=1 --epsilon=1 --nonlin-trf=identity --repetitions=100
 
 python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1.05e3_linear.csv --mode=partial --delta=1.05 --epsilon=3 --nonlin-trf=identity --repetitions=100
@@ -40,5 +38,15 @@ python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1.05e3_linear.csv
 python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1.05e3_linear.csv --mode=partial --delta=5 --epsilon=10 --nonlin-trf=identity --repetitions=100
 
 
+
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1e0_sigmoid.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=100
+
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1e1_sigmoid.csv --mode=partial --delta=1 --epsilon=1 --nonlin-trf=sigmoid --repetitions=100
+
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1.05e3_sigmoid.csv --mode=partial --delta=1.05 --epsilon=3 --nonlin-trf=sigmoid --repetitions=100
+
[…]

### 2021-10-23 simulated/run_all.sh
new simulation cases: sigmoid

@@ -49,4 +49,6 @@ python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1.5e5_sigmoid.csv
 
 python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d5e10_sigmoid.csv --mode=partial --delta=5 --epsilon=10 --nonlin-trf=sigmoid --repetitions=100
 
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d01e0_sigmoid.csv --mode=partial --delta=0.1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=100
+
 

### 2021-10-23 simulated/h1_simulate_normviol_lin.py
normviol sim

@@ -1,0 +1,123 @@
+import argparse
+import itertools
+import numpy as np
+import pandas as pd
+import pingouin as pg
+from tqdm import tqdm
+from joblib import Parallel, delayed
+import os
+import sys
+from mlconfound.stats import full_confound_test, partial_confound_test
+from mlconfound.simulate import simulate_y_c_yhat, identity, polynomial, sigmoid
+
+
+def run(e, d, w_yc, wyyhat, w_cyhat, Ns):
+    def workhorse(_random_state):
+        # simulate
+        ps = []
+        for n in Ns:
+            y, c, yhat = simulate_y_c_yhat(w_yc=w_yc,
+                                           w_yyhat=w_yyhat, w_cyhat=w_cyhat,
+                                           n=n,
+                                           random_state=_random_state,
+                                           delta=d,
+                                           epsilon=e,
+                                           nonlin_trf_fun=identity)
+
+            res_gam = partial_confound_test(y, yhat, c,
+                                            cat_y=False,
+                                            cat_yhat=False,
+                                            cat_c=False,
+                                            num_perms=1000,
+                                            cond_dist_method='gam',
+                                            random_state=_random_state,
+                                            n_jobs=1,
+                                            progress=False)
+
+            ps.append(res_gam.p)
+        return np.array(ps)
+
+    rng = np.random.default_rng(42)
+
+    random_sates = rng.integers(np.iinfo(np.int32).max, size=100)
+    p50, p100, p500, p1000 = zip(*
+                                 Parallel(n_jobs=-1)(
+                                     delayed(
+                                         workhorse)(
+                                         rs) for rs in
+                                     random_sates)
+                                 )
+    ret = [np.sum(np.array(p50) < 0.05)/100,
+           np.sum(np.array(p100) < 0.05)/100,
+           np.sum(np.array(p500) < 0.05)/100,
[…]

### 2021-10-23 simulated/h0_simulate_normviol_lin.py
normviol sim

@@ -1,12 +1,6 @@
-import argparse
-import itertools
 import numpy as np
 import pandas as pd
-import pingouin as pg
-from tqdm import tqdm
 from joblib import Parallel, delayed
-import os
-import sys
 from mlconfound.stats import full_confound_test, partial_confound_test
 from mlconfound.simulate import simulate_y_c_yhat, identity, polynomial, sigmoid
 
@@ -65,7 +59,7 @@ d = 0.1
 e = 2
 w_yc = 0.1
 w_yyhat = 0.1
-w_cyhat = 0.06
+w_cyhat = 0
 Ns = [50, 100, 500, 1000]
 
 res[0, :] = run(e, d, w_yc, w_yyhat, w_cyhat, Ns)
@@ -76,7 +70,7 @@ d = 1
 e = 0
 w_yc = 0.33
 w_yyhat = 0.33
-w_cyhat = 0.2
+w_cyhat = 0
 Ns = [50, 100, 500, 1000]
 
 res[1, :] = run(e, d, w_yc, w_yyhat, w_cyhat, Ns)
@@ -87,7 +81,7 @@ d = 1.05
 e = -3
 w_yc = 4
 w_yyhat = 4
-w_cyhat = 2.5
+w_cyhat = 0
 Ns = [50, 100, 500, 1000]
 
 res[2, :] = run(e, d, w_yc, w_yyhat, w_cyhat, Ns)
@@ -98,7 +92,7 @@ d = 1.5
 e = -5
 w_yc = 55
 w_yyhat = 55
-w_cyhat = 40
+w_cyhat = 0
 Ns = [50, 100, 500, 1000]
 
 res[3, :] = run(e, d, w_yc, w_yyhat, w_cyhat, Ns)
@@ -110,7 +104,7 @@ d = 5
 e = -10
 w_yc = 3000000
 w_yyhat = 3000000
-w_cyhat = 1000000
+w_cyhat = 0
 Ns = [50, 100, 500, 1000]
 
 res[4, :] = run(e, d, w_yc, w_yyhat, w_cyhat, Ns)
@@ -120,4 +114,4 @@ results = pd.DataFrame(res, columns=['n=50', 'n=100', 'n=500', 'n=1000'], index=
                                                                                  'e=1.05,d=-3',
                                                                                  'e=1.5,d=-5',
                                                                                  'e=5,d=-10'])
-results.to_csv('simulated/results/h1_simulate_normviol_lin.csv')
[…]

### 2021-10-23 simulated/h0_simulate_normviol_sig.py

@@ -1,14 +1,8 @@
-import argparse
-import itertools
 import numpy as np
 import pandas as pd
-import pingouin as pg
-from tqdm import tqdm
 from joblib import Parallel, delayed
-import os
-import sys
 from mlconfound.stats import full_confound_test, partial_confound_test
-from mlconfound.simulate import simulate_y_c_yhat, identity, polynomial, sigmoid
+from mlconfound.simulate import simulate_y_c_yhat, polynomial, sigmoid
 
 
 def run(e, d, w_yc, wyyhat, w_cyhat, Ns):
@@ -22,7 +16,7 @@ def run(e, d, w_yc, wyyhat, w_cyhat, Ns):
                                            random_state=_random_state,
                                            delta=d,
                                            epsilon=e,
-                                           nonlin_trf_fun=identity)
+                                           nonlin_trf_fun=sigmoid)
 
             res_gam = partial_confound_test(y, yhat, c,
                                             cat_y=False,
@@ -65,7 +59,7 @@ d = 0.1
 e = 2
 w_yc = 0.1
 w_yyhat = 0.1
-w_cyhat = 0.06
+w_cyhat = 0
 Ns = [50, 100, 500, 1000]
 
 res[0, :] = run(e, d, w_yc, w_yyhat, w_cyhat, Ns)
@@ -76,7 +70,7 @@ d = 1
 e = 0
 w_yc = 0.33
 w_yyhat = 0.33
-w_cyhat = 0.2
+w_cyhat = 0
 Ns = [50, 100, 500, 1000]
 
 res[1, :] = run(e, d, w_yc, w_yyhat, w_cyhat, Ns)
@@ -87,7 +81,7 @@ d = 1.05
 e = -3
 w_yc = 4
 w_yyhat = 4
-w_cyhat = 2.5
+w_cyhat = 0
 Ns = [50, 100, 500, 1000]
 
 res[2, :] = run(e, d, w_yc, w_yyhat, w_cyhat, Ns)
@@ -98,7 +92,7 @@ d = 1.5
 e = -5
 w_yc = 55
 w_yyhat = 55
-w_cyhat = 40
+w_cyhat = 0
 Ns = [50, 100, 500, 1000]
 
 res[3, :] = run(e, d, w_yc, w_yyhat, w_cyhat, Ns)
@@ -110,7 +104,7 @@ d = 5
 e = -10
[…]

### 2021-10-23 simulated/h1_simulate_normviol_sig.py

@@ -1,14 +1,9 @@
-import argparse
-import itertools
+
 import numpy as np
 import pandas as pd
-import pingouin as pg
-from tqdm import tqdm
 from joblib import Parallel, delayed
-import os
-import sys
 from mlconfound.stats import full_confound_test, partial_confound_test
-from mlconfound.simulate import simulate_y_c_yhat, identity, polynomial, sigmoid
+from mlconfound.simulate import simulate_y_c_yhat, polynomial, sigmoid
 
 
 def run(e, d, w_yc, wyyhat, w_cyhat, Ns):
@@ -22,7 +17,7 @@ def run(e, d, w_yc, wyyhat, w_cyhat, Ns):
                                            random_state=_random_state,
                                            delta=d,
                                            epsilon=e,
-                                           nonlin_trf_fun=identity)
+                                           nonlin_trf_fun=sigmoid)
 
             res_gam = partial_confound_test(y, yhat, c,
                                             cat_y=False,
@@ -120,4 +115,4 @@ results = pd.DataFrame(res, columns=['n=50', 'n=100', 'n=500', 'n=1000'], index=
                                                                                  'e=1.05,d=-3',
                                                                                  'e=1.5,d=-5',
                                                                                  'e=5,d=-10'])
-results.to_csv('simulated/results/h1_simulate_normviol_lin.csv')
+results.to_csv('simulated/results/h1_simulate_normviol_sig.csv')

### 2021-10-23 manuscript.tex

@@
 Thereby, it incorporates the information available on the conditional distribution of interest into the permutation-based inference in a statistically valid manner.
-Moreover, as shown in this paper, CPT can be combined with a general additive model to handle non-linear relationships, rendering it as an appealing candidate for testing confounder effects in predictive modelling.
+Moreover, as shown in this paper, CPT can be combined with a generalized additive model (GAM) to handle non-linear relationships, rendering it as an appealing candidate for testing confounder effects in predictive modelling.
 
@@
 \hline
-\textbf{1.} & $\y \independent \yhat | \c$ \quad  & model exclusively driven by confounder (full confounder test) & $Q(\cdot|\c)$ & $\yhat$ or $\y$ \\
+\textbf{1.} & $\y \independent \yhat | \c$ \quad  & full confounder test: model exclusively driven by confounder & $Q(\cdot|\c)$ & $\yhat$ or $\y$ \\
 %\hline
@@
 %\hline
-\textbf{3.} & $\yhat \independent \c | \y$  \quad &  model does not capture more confound than what is present in the target (partial test) & $Q(\cdot|\y)$ & $\yhat$ or $\c$ \\
+\textbf{3.} & $\yhat \independent \c | \y$  \quad &  partial confounder  test: model does not capture more confound than what is present in the target & $Q(\cdot|\y)$ & $\yhat$ or $\c$ \\
 \end{tabular}
@@
 
-Below, CPT is adapted for partial confounder testing (option 3). The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
+In the following section, CPT is adapted for partial confounder testing (option 3).
+The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
 \subsubsection*{The partial confounder test}
@@
 Obtaining a relatively accurate, independent estimate of $D(\cdot|\y)$ (of any shape) for CPT inference is important.
-Berrett and colleagues recommend to use a large independent sample to obtain the log-likelihood matrix that represents the conditional distribution $D(\cdot|Z)$.
-This approach might be often viable in semi-supervised learning settings where unlabelled data $(X, Z)$ are easier to obtain than labelled data $(X, Y, Z)$.
-Since in the proposed predictive modelling setting, normality might be often a reasonable assumption for $\y$ and $\c$ (but not for $\yhat$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|\y)$.
-
-In case of continuous target, a linear regression model  can be fitted:
+Berrett and colleagues recommend to use a large independent sample to obtain the log-likelihood matrix that represents the conditional distribution $D(\cdot|Z)$ or, alternatively, to re-use the data by fitting a least squares linear regression:
 
@@
     \label{eq:linreg}
-    \c = \alpha + \beta \y + \boldsymbol{e}
+    \c = \alpha + \beta f(\y) + \boldsymbol{e}
 \end{equation}
 
-with the assumption that the residual error follows a Gaussian distribution.
-If we write $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma$ denotes the standard deviation of the residual $\boldsymbol{e}$, then the conditional distribution of interest can also be assumed to be normal with the parameters:
+As the linear regression-based method, obviously, does not handle non-linear relationships I propose to fit a generalized additive model (GAM), instead:
+
+\begin{equation}
+    \label{eq:gam}
+    \c = \alpha + \beta f(\y) + \boldsymbol{e}
+\end{equation}
+
+where the feature functions $f$ is built using penalized B splines, which allow us to automatically model non-linear relationships without having to manually try out many different transformations on each variable.
+
+If we write $\boldsymbol{\mu} = \alpha + \beta f(\y)$ and $\sigma$ denotes the standard deviation of the residual $\boldsymbol{e}$, then the conditional distribution of interest can be assumed to be normal with the parameters:
 
@@
 
-In the case of categorical $c$, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$, with the extra assumption of non-complete separation if $y$ is also categorical (for an invertible Hessian, see e.g.
+In the case of categorical $c$, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$, with the extra assumption of \emph{complete separation} if $y$ is also categorical (for an invertible Hessian, see e.g.
 \citep{bohning1992multinomial}).
 
-Importantly, both the linear and the multinomial logistic regression approaches guarantee that the numerator of Eq.
+Importantly, both the GAM-based and the multinomial logistic regression approaches guarantee that the numerator of Eq.
 \ref{eq-pperm} is always greater than zero and the group property for the permutations holds.
 
-Note that from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, for option 2, the proposed approach can not provide a test that is assumption-free about $yhat$, as the variable on which the independence is conditional must be always the predictor variable in Eq.
-\ref{eq:linreg}.
+Note that from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, for option 2, the proposed approach can not provide a test that is assumption-free about $\yhat$, as the variable, on which the independence is conditional, must be always the predictor variable in Eq.
+\ref{eq:gam}.
 However, as discussed above, this option is of low practical relevance, anyway.
@@
 
-Using CPT to test confounder-bias in predictive modelling allows relaxing assumptions on $\yhat$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions on - the joint distribution of the other two variables ($\yhat$ and $\c$).
-\cite{berrett2020conditional} give a detailed analysis of the robustness of their CPT approach when estimating the conditional distribution with re-using the tested data and, also, against misspecifying the conditional distribution of interest.
+Using CPT to test confounder-bias in predictive modelling allows relaxing assumptions on $\yhat$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions on - the joint distribution of the other two variables ($\y$ and $\c$).
+\cite{berrett2020conditional} give a detailed analysis of the robustness of their CPT approach when estimating the conditional distribution with re-using the tested data via linear regression and, also, against misspecifying the conditional distribution of interest to introduce non-linearity.
 
-While their findings are largely applicable to the proposed tests, here I extend those results by performing simulations in a form that is accessible for power calculations in predictive modelling (considering various weights of the confounder and the target signals in $\yhat$ and in the case of varying levels of covariance between $\y$ and $\c$).
-Moreover, I perform a thorough investigation of the robustness of the test against the violation of normality (for both $(\y, \c)$ and $\yhat$).
+Here I extend these results by performing simulations that evaluate the GAM- and multinomial liner model-based approaches, in a form that is accessible for power calculations in predictive modelling (considering various weights of the target signal in $c$ and the confounder and the target signals in $\yhat$).
+Moreover, I investigate the robustness of the tests against the violation of normality and linearity of the conditional distributions $D(\c|\y)$ and $D(\yhat|\y)$
 
@@
 
-\subsubsection*{Gaussian Simulations}
-As  a first step, the target variable $\y$ and the confounder variable $\c$ are drawn randomly from a multivariate normal distribution:
+\subsubsection*{Simulation approach}
+As  a first step, the target variable $\y$ is drawn randomly from a normal distribution:
 
-$$ \y, \c \sim \mathcal{N}(\mu_0, \Sigma_\sigma) $$
+$$ \y \sim \mathcal{N}(0, 1) $$
 
-where $\mu_0=(0, 0)$ and the covariance matrix is given as:
-$\Sigma_\sigma = \big(\begin{smallmatrix}
-  1 & \sigma\\
-  \sigma & 1
-\end{smallmatrix}\big)$ with $\sigma$ (i.e. the covariance between $\y$ and $\c$) being a free parameter of the simulation.
+Next, the confounder signal is simulated as:
 
-Next, the simulated predicted values are constructed:
- with an additive model:
-$$ \yhat = w_y \y + w_c \c + w_e \boldsymbol{e}$$
+$$ \c \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yc} \quad g(\y) $$
 
-so that $w_y + w_c + w_e = 1$, where $\boldsymbol{e} \sim \mathcal{N}(0,1)$ is the noise component.
+where $f$ is a function to introduce non-normality, namely the \emph{sinh-arcsinh} transformation of \cite{jones2009sinh} defiend as:
 
-To test the implementation for categorical variables, simulated $\y$, $\yhat$ and $\c$ variables are binarized by thresholding at 0.
-
-For more straightforward interpretation and visualization of the simulation results, the 'confound-to-target' and the 'signal-in-prediction' parameters are defined as
-
-$$ w_{c/y} = w_c / w_y $$
-
-and
-
-$$ w_{c+y} = w_c + w_y $$
-
-respectively.
-
-\subsubsection*{Introducing non-normality}
-
-Non-normality is introduced by applying the \emph{sinh-arcsinh} transformation of \cite{jones2009sinh} on the simulated variables:
-
-$$\boldsymbol{x}' = sinh(\delta sinh^{-1}(\boldsymbol{x}) - \epsilon)$$
+$$f_{\delta, \epsilon}(\boldsymbol{x}) = sinh(\delta sinh^{-1}(\boldsymbol{x}) - \epsilon)$$
 
@@
 
-\subsubsection*{Simulation cases}
+Moreover, non-linearity can be introduced with the function $g$, which can be simply the identity function (no non-linearity is introduced) or e.g. a sigmoid-shaped function, in pour case:
 
-In the simulation procedure for the 'partial' confound test, 100 repetitions were performed of all combination of the following parameter values: 
+$$ g(\boldsymbol{x}) = tanh(\boldsymbol{x}) $$
+
+The simulated predicted values are constructed analogously:
+
+$$ \yhat \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yyhat} \quad g(\y)  + w_{c\hat{y}} \quad g(\y)$$
+
+To test the implementation for categorical variables, simulated $\y$, $\yhat$ and $\c$ variables are binarized by thresholding at 0.
+
+\subsubsection*{Simulations for comparison with partial Spearman correlation and linear CPT}
+
+To demonstrate the need for the proposed GAM-based CPT approach for partial confounder testing, its validity was contrasted to partial Spearman correlation and the linear variety of CPT (based on eq.
+\ref{eq:linreg}, as described by \citep{berrett2020conditional}) in the following parameters:
+\begin{itemize}
+    \item sample size: $n = 1000$
+    \item $w_{c\hat{y}} = 0$ (H0 simulations only)
+    \item all combinations of:
+    \begin{itemize}
+        \item $w_{yc} \in \{0.5, 1, 2, 3\}$ and
+        \item $w_{y\hat{y}} \in \{0.5, 1, 2, 3\}$
+    \end{itemize}
+    \item non-normality ($f_{\delta = 0.1, \epsilon = 2)}$ and non-linearity (sigmoid $g$)
+\end{itemize}
+
+For each parameter combination, 1000 repetitions were performed and false positive rate was calculated as the ratio of p-values smaller than $\alpha = 0.05$.
+
+These simulation cases are visualized (with $w_{yc} = w_{y\hat{y}} = 2$) on the left of Figure \ref{fig:sim-h0-demo}.
+
+\subsubsection*{Simulations for evaluating power.}
+
+Both for the partial and the full confound test, 100 repetitions were performed of all combination of the following parameter values: 
 \begin{itemize}

### 2021-10-23 simulated/run_all.sh
new simulation cases added

@@ -3,52 +3,29 @@ DIR=`dirname $0`
 # H0
 
 python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_linear.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=identity --repetitions=1000
-
-python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e0_linear.csv --mode=partial --delta=0.1 --epsilon=0 --nonlin-trf=identity --repetitions=1000
-
-python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e1_linear.csv --mode=partial --delta=0.1 --epsilon=1 --nonlin-trf=identity --repetitions=1000
-
 python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e2_linear.csv --mode=partial --delta=0.1 --epsilon=2 --nonlin-trf=identity --repetitions=1000
-
-python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e3_linear.csv --mode=partial --delta=0.1 --epsilon=3 --nonlin-trf=identity --repetitions=1000
-
-python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d01e4_linear.csv --mode=partial --delta=0.1 --epsilon=4 --nonlin-trf=identity --repetitions=1000
-
 python3 $DIR/simulate_H0.py --out=$DIR/results/h0_ccc_partial_d1e0_sigmoid.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=1000
 
 
 # H0+H1
 
 python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d1e0_linear.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=identity --repetitions=100
+python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d01e0_sigmoid.csv --mode=partial --delta=1 --epsilon=0 --nonlin-trf=sigmoid --repetitions=100
 
-python3 $DIR/simulate_H1.py --out=$DIR/results/h1_ccc_partial_d.5e0_linear.csv --mode=partial --delta=0.5 --epsilon=0 --nonlin-trf=identity --repetitions=100
[…]

### 2021-10-23 manuscript.tex
Updates from Overleaf

@@
     \label{eq:linreg}
-    \c = \alpha + \beta f(\y) + \boldsymbol{e}
+    \c = \alpha + \beta \y + \boldsymbol{e}
 \end{equation}
 
-As the linear regression-based method, obviously, does not handle non-linear relationships I propose to fit a generalized additive model (GAM), instead:
+As the linear regression-based method, obviously, does not handle non-linear relationships I propose to fit a generalized additive model (GAM) \citep{hastie1987generalized}, instead:
 
@@
 
-where the feature functions $f$ is built using penalized B splines, which allow us to automatically model non-linear relationships without having to manually try out many different transformations on each variable.
+where the feature functions $f$ is built using penalized B-splines, which allow us to automatically model non-linear relationships without having to manually try out many different transformations on each variable.
+
+Smoothness of the GAM model is optimized with a grid-search by picking the model with the lowest generalized cross-validation score form the models defined by the default parameters as implemented in PyGAM \citep{serven2018generalized}.
 
@@
   \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
-  \caption{\textbf{Type-I error control and power of the conditional permutation-based confound testing for predictive modelling in normally distributed simulated data}. \\
-  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette at the bottom) in various simulations settings (100 simulations per tile) with different covariance between the target and the confounder (given as $R^2_{y,c}$ on the vertical axes of the heatmaps), for different sample sizes (N, horizontal axes).
-  Rows and columns correspond to simulations with different signal-in-prediction (y+c) and confound-to-signal values, respectively.
-The first row and the first column contain simulations under H0 (no confounder bias) and separated form the simulations under the alternative hypothesis with dashed line.
-Positive rates for the simulations under the null and alternative hypotheses can be interpreted as type I error rate and statistical power, respectively.
-The higher confidence limit for the type I error rates is 0.11 for each tile, with 5 tiles expected to pass it form all H0 tiles.
-  The actual weights $w_{Y}$, $w_{C}$ and $w_{e}$ are depicted at the bars on top of each heatmap and color coded as shown in the legend at the bottom.
-Above each heatmap, the strengths of the associations of $\hat{y}$ to $Y$ and $C$, respectively, are given with $R^2$ values averaged over all simulation cases of the heatmap.
+  \caption{\textbf{Type-I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
+  Type-I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row).
+Non-normal conditional distribution on the second plot is illustrated with blue density diagrams and dependency is depicted with blue dotted lines an all plots.
+False positive rates for the three investigated tests is shown in heatmaps, with darker red meaning higher values.
+The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065.
+Values below this threshold indicate a valid type-I error control and colored white.
   }
@@
 \begin{itemize}
-    \item covariance of $\c$ and $\y$: $\sigma \in \{0, 0.2, 0.4, 0.6, 0.8\}$
-    \item confound-to-target: $w_{c/y} \in \{0, 0.1, 0.3, 0.5, 1\}$
-    \item signal-in-prediction: $w_{c+y} \in \{0, 0.3, 0.6, 0.9\}$
-    \item sample size: $n \in \{50, 100, 500, 1000\}$
+    \item $w_{yc} \in \{0.5, 1, 2, 3\}$
+    \item $w_{y\hat{y}} \in \{0.5, 1, 2, 3\}$
+    \item $w_{c\hat{y}} \in \{0, 0.2, 0.4, 0.6\}$
+    \item $n \in \{50, 100, 500, 1000\}$
+    \item linear and sigmoid dependence
+    \item normal conditional distribution
 \end{itemize}
 
-For testing 'full' confounder bias, identical parameters were used, except that instead of the 'confound-to-target' parameter $ w_{c/y}$ we used the 'target-to-confound' parameter $ w_{y/c} = w_y/ w_c $, but with the same parameter values $w_{y/c} \in \{0, 0.1, 0.3, 0.5, 1\}$.
+The 'partial' and 'full' confound tests, as implemented in version 0.13.0 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.
 
-Next to the above described normally distributed simulation scenario, four analogous, but \emph{sinh-arcsinh} transformed scenarios were also investigated, with $\delta \in \{1, 1.05, 1.5, 5\}$ $\epsilon \in \{1, 3, 5, 10\}$
-
-The 'partial' and 'full' confound tests, as implemented in version 0.9.0 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.
-
-Analysis code used for simulations is available\footnote{https://github.com/pni-lab/mlconfound/blob/master/validation/simulation.py} in the github repository of the 'mlconfound' package.
+All analysis code used for the simulations is available on github\footnote{https://github.com/spisakt/mlconfound_manuscript}.
 
@@
 
-  \includegraphics[width=0.40\paperwidth]{fig/sim_non-norm.eps}
-  \caption{\textbf{Robustness of conditional permutation based confound testing to the violation of normality assumptions for $\y$ and $\c$.} \\
-  Simulations included variables with five different degrees of non-normality (A), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation, from normal (left) to severely skewed and heavy tailed (right).
-Fisher's kurtosis and skewness is given for each at the bottom of panel A.
-Panel (B) depicts false positive rates in the simulations under H0, separately for different signal-in-confounder ratios (here identical to $w_{y}$).
-  Vertical axis shows type I error rate.
-The nominal type I error rate of $alpha=0.05$ is shown with a vertical dashed line.
-Panel (C) shows how violation of the normality assumption affects power.
-Boxplots show the mean power over simulation cases, grouped by confound-to-target ratio and sample size.}
+  \includegraphics[width=0.5\paperwidth]{fig/sim_non-norm.eps}
+  \caption{\textbf{Robustness of conditional permutation based confound testing to non-normality and non-linearity.} \\
+  Simulations included variables with five different degrees of non-normality (top panel), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation (yellow: normally distributed).
+Fisher's kurtosis and skewness is given for each distribution.
+False and true positive rates in the simulations under H0 and H1, respectively, for each investigated sample size (N), are depicted by barplots for both linear and sigmoid dependency structure.
+Upper 95\% binomial confidence interval corresponding to $alpha=0.05$ is shown with a vertical dashed line.}
   \label{fig:sim-non-normal}
@@
 
-Simulations based on normally distributed data justified that the proposed confounder tests - in line with theory - provide a valid type I error control (For detailed results, see \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-full}, \ref{fig:sim-ccb-partial}, \ref{fig:sim-ccb-full}, \ref{fig:sim-bbc-partial}, \ref{fig:sim-bbc-full}, \ref{fig:sim-bbb-partial} and \ref{fig:sim-bbc-full} for the partial test on categorical data and the full tests on both numeric and categorical data).
+Simulations based on normally distributed data justified that the proposed confounder tests - in line with theory - provide a valid type I error control (For detailed results, see \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-lin-full}, \ref{fig:sim-bbb-sig-full}, for the partial test on categorical data and the full tests on both numeric and categorical data).
 In case of a valid test, the expected type I error rate in the null hypothesis simulation cases (first row and first column of the figures) is less then or equal to the alpha level (0.05).

### 2021-10-23 supplement.tex
polished
Updates from Overleaf
patched

@@
 \beginsupplement
+
+\include{simulated/normality_and_linearity_violation}
 
@@
 
-\begin{equation}
-    \mathbb{P}(\pi^{(j)} = \pi | \y,\yhat,\c) = \frac{q^n(\y_\pi | \c)}{\sum_{\pi' \in S_n} q^n(\y_{\pi'} | \c)}
-    \label{eq-pperm}
-\end{equation}
+    $$\mathbb{P}(\pi^{(j)} = \pi | \y,\yhat,\c) = \frac{q^n(\y_\pi | \c)}{\sum_{\pi' \in S_n} q^n(\y_{\pi'} | \c)}$$
 
@@
   \fbox{
-   \includegraphics[width=0.75\paperwidth]{fig/supplement/sim_ccc_full_normal_all_heatmap.pdf}}
-  \caption{Heatmaps showing the positive rates of the 'full' confounder test, with normally distributed numerical simulated variables.}
-  \label{fig:sim-ccc-full}
+   \includegraphics[width=0.75\paperwidth]{fig/raw/sim_h1_ccc_partial_d10e0_sig_cpt_gam.pdf}}
+  \caption{Heatmaps showing the positive rates of the 'partial' confounder test, with normal conditional distribution and sigmoid dependence.}
+  \label{fig:sim-ccc-sig-partial}
 \end{figure}
-
-\begin{figure}[H]
-  \centering
-  \fbox{
-   \includegraphics[width=0.75\paperwidth]{fig/supplement/sim_ccb_partial_normal_all_heatmap.pdf}}
-  \caption{Heatmaps showing the positive rates of the 'partial' confounder test, with normally distributed numeric $Y$ and $\hat{Y}$ and categorical $C$.}
-  \label{fig:sim-ccb-partial}
-\end{figure}

### 2021-10-25 manuscript.tex

@@
 
-The 'partial' and 'full' confound tests, as implemented in version 0.13.0 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.
+The 'partial' and 'full' confound tests, as implemented in version 0.20.0 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.
 
-All analysis code used for the simulations is available on github\footnote{https://github.com/spisakt/mlconfound_manuscript}.
+All analysis code used for the simulations is available on github\footnote{\href{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}}.
 
@@
 
-'Partial' and 'full' confound testing was performed with default parameters (10000 permutations, 50 Markov-chain Monte Carlo steps) as implemented in version 0.9.0 of the package \emph{'mlconfound'}.
+'Partial' and 'full' confound testing was performed with 10000 permutations and 50 Markov-chain Monte Carlo steps, as implemented in version 0.20.0 of the package \emph{'mlconfound'}.
 Unconditional dependence across the involved variables was investigated with conventional permutation test on the $R^2$ values, with 1000 permutations.
+
+All empirical analyses are available as jupyter notebooks on github\footnote{\href{https://github.com/pni-lab/mlconfound-manuscript/tree/main/empirical}{https://github.com/pni-lab/mlconfound-manuscript/tree/main/empirical}}.
 
@@
 
-Analysis code is available as jupyter notebook\footnote{\href{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb}{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb}} in the github repository of the 'mlconfound' package.
-
 \subsubsection*{ABIDE: testing motion- and center-bias in predictive models of autism spectrum disorder diagnosis}
@@
 The resulting functional connectivity estimates were considered as features either (i) in their raw form or after applying (ii) feature regression \citep{rao2017predictive} or (iii) COMBAT \citep{johnson2007adjusting, fortin2018harmonization}.
-The investigated confounder variables were 'imaging center' and in-scanner motion, as measured by the mean framewise displacement of \cite{power2014methods}).
+The investigated confounder variables were 'imaging center' and 'in-scanner motion', as measured by the mean framewise displacement of \cite{power2014methods})
 Mean FD was non-linearly transformed to normal distribution with the quantile transformation \citep{beasley2009rank} as implemented in \emph{scikit-learn} \citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:abide-hist} for details).
 
-As COMBAT is not able to handle continuous variables (since it was primarily designed to remove "batch-effects"), motion was binned into 10 groups, based on equdistant data quantiles ranging from 0 to 1.
+As COMBAT is not able to handle continuous variables (since it was primarily designed to remove "batch-effects"), motion was binned into 10 groups, based on equidistant data quantiles ranging from 0 to 1.
 
@@
 
-Analysis code is available as jupyter notebook\footnote{\href{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_abide.ipynb}{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_abide.ipynb}} in the github repository of the 'mlconfound' package.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   \caption{\textbf{Type-I error control and power of the conditional permutation-based confound testing for predictive modelling in normally distributed simulated data}. \\
-  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette at the bottom) in various simulations settings (100 simulations per tile) with different covariance between the target and the confounder (given as $R^2_{y,c}$ on the vertical axes of the heatmaps), for different sample sizes (N, horizontal axes).
-  Rows and columns correspond to simulations with different signal-in-prediction (y+c) and confound-to-signal values, respectively.
-The first row and the first column contain simulations under H0 (no confounder bias) and separated form the simulations under the alternative hypothesis with dashed line.
-Positive rates for the simulations under the null and alternative hypotheses can be interpreted as type I error rate and statistical power, respectively.
-The higher confidence limit for the type I error rates is 0.11 for each tile, with 5 tiles expected to pass it form all H0 tiles.
-  The actual weights $w_{Y}$, $w_{C}$ and $w_{e}$ are depicted at the bars on top of each heatmap and color coded as shown in the legend at the bottom.
-Above each heatmap, the strengths of the associations of $\hat{y}$ to $Y$ and $C$, respectively, are given with $R^2$ values averaged over all simulation cases of the heatmap.
+  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette at the bottom) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\yhat}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\yhat}$ (rows) and for different sample sizes (N, columns).
+Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%.
+First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounder bias).
+  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively.
+The higher 95\% confidence limit for the type I error rates is 0.11 for each tile.
   }
@@
 
-The conditional permutation based 'partial' and 'full' confound tests for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.
+The conditional permutation based \emph{partial} and \emph{full confounder tests} for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.
 
@@
 False and true positive rates in the simulations under H0 and H1, respectively, for each investigated sample size (N), are depicted by barplots for both linear and sigmoid dependency structure.
-Upper 95\% binomial confidence interval corresponding to $alpha=0.05$ is shown with a vertical dashed line.}
+Upper 95\% binomial confidence limit corresponding to $alpha=0.05$ is shown with a vertical dashed line.}
   \label{fig:sim-non-normal}

### 2021-10-25 manuscript.tex

@@
 Simulations with non-normal data showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $\y$ and the confounder $\c$ and, as expected, completely non-parametric regarding the distribution of the predictions $\yhat$).
-The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal}, involved various \emph{sinh-arcsinh} distributions, form normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
+The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal}, involved various \emph{sinh-arcsinh} distributions, from normal (Fisher's kurtosis: 0, skewness: 0) to severely skewed and heavy tailed (kurtosis: 42, skewness: -6).
 The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to kurtosis and skewness values as high as 5 and 2 respectively.

### 2021-10-26 manuscript.tex

@@
 
-\title{A conditional permutation-based approach to test confounder and center effects in machine learning models}
+\title{Detecting confounder and center bias in machine learning models}
 
@@
 
+%TC:ignore
 \begin{abstract} % 150 words
-The lack of rigorous non-parametric statistical tests of confounder effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
-Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unconfounded} and \emph{fully confounded model}, respectively.
-The proposed methods provide a strict control for Type-I errors and high statistical power, even in the presence of typical non-linear and non-normal dependencies between the target variable and the model output.
-Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset reveals confoudner effects that were previously unreported and found to be hard to correct for with state-of-the-art confound mitigation approaches.
+The lack of rigorous non-parametric statistical tests for confounder effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
+
+Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and, for a given confounder variable, probe the null hypotheses of \emph{unconfounded} and \emph{fully confounded models}, respectively.
+
+The tests provide a strict control for Type-I errors and high statistical power, even for non-linear and non-normal dependencies typical in machine learning.
+Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset reveals confounders that were previously partly unreported and found to be hard to correct for with state-of-the-art confound mitigation approaches.
+
 The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment and improvement of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
 \end{abstract}
+%TC:endignore
 
@@
 
-Predictive modelling and supervised learning methods have recently become increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized}.
+Predictive modelling and supervised learning methods have recently become increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized, spisak2020pain, walsh2021dome}.
 When evaluating the usefulness and applicability of such markers, predictive performance (i.e.
@@
 Biomedical validity (i.e.
-whether the model is driven by biomedically relevant signal) and generalizability across contexts and populations are also crucial requirements for candidate biomarkers \citep{woo2017building}.
+whether the model is driven by biomedically relevant signal) as well as generalizability and fairness across contexts and populations are crucial requirements for candidate biomarkers \citep{woo2017building, obermeyer2019dissecting, mehrabi2021survey}.
 
-Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - induced by so-called confounder effects - can be detrimental to the model's biomedical validity and generalizability.
-Many types of confounder effects can be distinguished.
+Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - i.e the so-called confounder and collider biases \citep{prosperi2020causal} - can be detrimental to the model's biomedical validity and generalizability.
+Such biases can originate form many sources.
 
-First, measurement artifacts can obviously be considered as confounders.
-For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with motor function, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive} attention deficit hyperactivity disorder (ADHD) \citep{couvy2016head, eloyan2012automated} or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}.
+First, measurement artifacts can obviously be considered as a potential source of confounder bias.
+For instance, magnetic resonance imaging (MRI)-based predictive models can be driven by in-scanner head motion artifacts, next to (or instead of) signal of neural origin, especially, if the prediction target is also associated with motor function and motion characteristics, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive}, attention deficit hyperactivity disorder (ADHD) \citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD) \citep{gotts2013perils, spisak2014voxel, spisak2019optimal}.
 Likewise, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.
@@
 Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling.
-For instance, resting-state functional connectivity has been shown to be strongly predictive to age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}.
-As fluid intelligence is known to decline with aging \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by picking up on solely age-related variance \citep{lohmann2021predicting, dubois2018distributed}.
+For instance, resting-state functional connectivity has been shown to be strongly predictive to age \citep{dukart2011age, wang2012decoding} and moderately to fluid intelligence \citep{ cole2012global, he2020deep}.
+As fluid intelligence is known to decline with aging \citep{kievit2018neural}, models trained to predict intelligence might provide a statistically significant predictive performance by picking up solely on age-related variance \citep{dubois2018distributed, lohmann2021predicting}.
+Importantly, the target variable must not necessarily display a true association to the confoudner to produce a biased model.
+Model training can be sensitive to sampling bias and stochastic group differences in the training set; a phenomenon that might have calamitous consequences, as increasingly recognised, for instance, in case of racially biased machine learning models \citep{ obermeyer2019dissecting, lwowski2021risk}.
 
 Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance} as the models can utilize such effects to explain (random or true) center/batch-differences in the predictive target.
-Such models will display worse-than-expected performance when applied on data from new centers/batches.
+Such models are hard to calibrate, might output systematically biased predictions and might display a worse-than-expected performance when applied on data from new centers/batches.
 
-Various data cleansing, correction and harmonization approaches have been proposed to mitigate detrimental effects of confounders, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}.
+Various data cleansing, correction and harmonization approaches have been proposed to mitigate detrimental effects of confounders, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{dukart2011age, spisak2014voxel, abdulkadir2014reduction, rao2017predictive} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}.
 However, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.
@@
 
-The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via "restricted permutations" that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature).
+The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via 'restricted permutations' that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature).
 However, as shown by \cite{southworth2009properties} and further discussed by \cite{hemerik2018exact}, balanced permutations do not provide a correct null behavior and can be invalid in certain circumstances.
-Indeed, \cite{ferrari2020measuring} demonstrated that this method does not guarantee a proper type I error control with biased classification problems.
+Indeed, \cite{ferrari2020measuring} demonstrated that this method does not guarantee a proper type I error control in certain, realistic scenarios.
 The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, on the other hand, does not provide any quantification of statistical significance.
-Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high computational complexity (especially when hyperparameters are optimized in a nested cross-validation).
+Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high a computational cost (especially when hyperparameters are optimized in a nested cross-validation).
 The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the predictors are a direct cause of the target variable or, alternatively, there is an unobserved random variable that is the cause of both.
 However, this null hypothesis might not be of interest in many real applications.
-For instance, causation between several psychological or clinical states and functional MRI activity-connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation.
+For instance, causation between several psychological or clinical states and functional MRI activity/connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation.
 In such cases, the alternative hypothesis of this test might be true in absence of any real confounder.
@@
 In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}.
-This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ("no free lunch" theorem).
-As an important consequence, partial correlation - a trivial candidate for investigating the dependence between the predictions and the confounder (given the target variable) - might fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved.
+This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem).
+As an important consequence, partial correlation - a trivial candidate for investigating the dependence between the predictions and the confounder (given the target variable) - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (demonstrated on Fig.
+\ref{fig:sim-h0-demo}).
 These conditions are very common for case of predictive modelling \citep{garcia2009study, kristensen2017whole} (Supplementary Material \ref{sup:nomlinviol})
 
-To overcome the limitations of partial correlation (demonstrated on Fig.
-\ref{fig:sim-h0-demo}), I build on the permutation-based conditional independence framework of \cite{berrett2020conditional}, to construct two novel tests for evaluating the confounder-bias in predictive models.
-The \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder.
-The \emph{partial confounder test} investigates whether the model utilizes any confounder-variance, when controlled for the target variable.
+To overcome these limitations, I extend the permutation-based conditional independence framework of \cite{berrett2020conditional} with a generalized additive model based conditional distribution estimation, which allows avoiding to place any assumptions about the most problematic variable, the prediction output.
+The framework allows constructing two tests: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
 Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
-The tests set no assumption about the distribution of the model predictions.
 The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality and non-linearity).
-The proposed tests are then applied on functional brain connectivity data to provide evidence of motion- and center-bias in a model trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} and to evaluate the age and acquisition-batch dependence of fluid intelligence predictions on data from the Human Connectome Project \citep{van2013wu}.
+
+The usefulness of the proposed tests in detecting various types of confounder bias is demonstrated in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.
+First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} is used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{ dubois2018distributed, lohmann2021predicting} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
+Second, functional connectivity data from the ABIDE \citep{di2014autism} database is used to investigate the potential motion- and center-bias \cite{gotts2013perils, spisak2014voxel, spisak2019optimal} when training models that aim to predict ASD diagnosis.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%TC:ignore
 \section{Methods}
@@
 For instance, even if $\yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to the dependence of both on $\y$.
-In other words, if the target variable $\y$ is significantly associated to the confounder variable $\c$, a model that is completely blind to $\c$ (i.e not confounded) might still provide outputs $\yhat$ that are significantly associated with $\c$.
+In other words, if the target variable $\y$ displays a true association to the confounder variable $\c$, a model that is completely blind to $\c$ (i.e not confounded at all) might still provide outputs $\yhat$ that are significantly associated with $\c$.
 
@@
 
- Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications \citep{spirtes2000causation, peters2016causal, fiedler2011mediation, candes2016panning}.
+ Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications \citep{spirtes2000causation, fiedler2011mediation, peters2016causal, candes2016panning}.
 Recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
@@
 
-This result stands in strong contrast to unconditional independence testing - where permutation tests  \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs, depending on the applied machine learning model, is unknown and often heavily non-normal.
+This result stands in strong contrast to \emph{unconditional} independence testing - where permutation tests  \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs (conditioned on the target variable) - depending on the applied machine learning model - is unknown and often non-normal and non-linear.
 
-One of the trivial candidate for the task, partial correlation, for instance assumes that all involved variables are multivariate Gaussian and - as to be shown below - even its Spearman-based variant is unable to tolerate relatively small deviations from normality and linearity.
+One of the trivial candidates for the task, partial correlation, for instance assumes that all involved variables are multivariate Gaussian and - as to be shown below - even its Spearman-based variant is unable to tolerate relatively small deviations from normality and linearity.
 
@@
 Confounder testing was performed on the class probabilities (instead of the predicted labels).
+%TC:endignore
 
@@
 This bias is robustly detected by the proposed 'partial' confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
-  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$.
+  Relation between the observed ($Gf$) and predicted ($\hat{y}$) IQ scores and the confounder variables is shown by the graphs as $R^2 values$.
 Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
@@
 
-Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}).
+Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $p=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}).
 The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
@@
 
-Both feature regression and COMBAT seemed to significantly attenuate the effect of center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.01$ for feature regression and COMBAT, respectively, see the second and third columns of the first row on Fig \ref{fig:abide}).
+Both feature regression and COMBAT seemed to significantly attenuate the effect of imaging center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.01$ for feature regression and COMBAT, respectively, see the second and third columns of the first row on Fig \ref{fig:abide}).
 
@@
 \end{tabular}
-\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated models.
+\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated confounder-mitigation methods.
 Significant confoudner bias is denoted by bold font.}
@@
 Testing confounder-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confounder variable.
-These two scenarios result in two different test, testing whether the model is partially or fully driven by the given confounder.
+These two scenarios result in two different tests, probing whether the model is partially or fully driven by the given confounder.
 Importantly, in predictive modelling, the model predictions often follow a non-normal distribution, thus any test for confounder bias must be, at least partly, non-parametric.
@@
 Here I propose the use of the conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models.
-Adapting this approach for the confounder testing problem makes it possible to place assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output.
-The tests are made available in the python package \emph{mlconfound} which builds upon a parallelized, computationally effective implementation of the Markov-chain Monte-Carlo based parallel pairwise sampler of \cite{berrett2020conditional}.
+Adapting this approach for the confounder testing problem allows placing assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output.
+The tests are made available in the python package \emph{mlconfound} which builds upon a parallelized and computationally effective implementation of the Markov-chain Monte-Carlo based parallel pairwise sampler of \cite{berrett2020conditional}.
 It utilizes coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal.
@@
 As expected from theoretical justification, both tests displayed a valid type I error control in the simulations with the necessary assumptions, that is, even if $\yhat$ is not normally distributed.
-In fact, the tests were found to become somewhat conservative (i.e.
+In fact, the tests were found to become more conservative (i.e.
 lower than expected type I error rates ) in the case of high-performance predictive models (when the $R^2$ between the predicted and observed values is greater than 0.7).
 
-Even tough the tests valid type I error control is justified by theory and reinforced by the simulations, a low statistical power could significantly limit their usefulness.
-The simulations presented in this paper revealed that both tests have a high, practically relevant power in a wide variety of scenarios; with different strength of the association between the target and the confounder, varying predictive performance, various sample sizes and different true confound-to-signal ratios in the predictions.
+Even though the tests' valid type I error control is justified by theory and reinforced by the simulations, a low statistical power could significantly limit their usefulness.
+The simulations presented in this paper revealed that both tests have a high, practically relevant power in a wide variety of scenarios.
 Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confounder bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confounder explaining 1-2\% of the variance in the prediction).
@@
 
-While different biomedical applications might consider different amounts of bias to be relevant, an a-priory power calculation can help identifying the necessary sample size in order to exclude the possibility of a relevant confounder bias with a high probability.
+While different biomedical applications might consider different amounts of bias to be relevant, an a-priori power calculation can help identifying the necessary sample size in order to exclude the possibility of a relevant confounder bias with a high probability.
 
@@
 
-Among the many potential areas of applications, the proposed tests could be certainly useful in translational neuroscience.
-A characteristic example is provided by the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds a great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools \citep{woo2017building, wager2013fmri, spisak2020pain}.
+Among the many potential areas of applications, the proposed tests could be particularly useful in translational neuroscience.
+A characteristic example is provided by the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds a great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making \citep{woo2017building, wager2013fmri, spisak2020pain}.
 However, the presence of confounders typical for biomedical research (e.g.
@@
 
-To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two research settings (a regression and a classification problem) where specific types of confounding effects have previously been described.
+To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.
 First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
@@
 There can be manifold reasons for the observed association.
-Fluid intelligence of the participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g.
+Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g.
 as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed over time).
@@
 A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
-On the other hand, motion artefacts were previously shown to display complex spatiotemporal patterns in fMRI-based functional connectivity estimates \citep{spisak2014voxel} and simply regressing them out of the features is far from being sufficient for removing their effect on the model.
-The so-called "residual motion artefacts" and their regional interactions, as described in the referenced work, might be responsible for the paradox behavior of this model, which raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
+On the other hand, motion artifacts were previously shown to display complex spatiotemporal patterns in fMRI-based functional connectivity estimates \citep{spisak2014voxel} and simply regressing them out of the features is far from being sufficient for removing their effect on the model.
+The so-called "residual motion artifacts" and their regional interactions, as described in the referenced work, might be responsible for the paradox behavior of this model, which raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
 
  As COMBAT was originally developed for harmonizing effects of categorical variables (e.g.
-center or batch), using it to remove the effect of continuous variables is not trivial.
+center or batch), its application for continuous confounder variables is not trivial.
 The presented method of inputting the binned "motion groups" into COMBAT resulted in an efficient removal of motion-bias and, at the same time, preserved a significant amount of signal-of-interest.
@@
 
- In sum, the application of the partial confounder test suggests that confounder-bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT.
+ In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounder-bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT.
 The partial confounder test can be considered as a useful objective benchmark to guide the search for a suitable confounder mitigation approach.
-
-While the real data examples in this paper were not typical cases of the potential applications of the full confounder test, it has still provided important insights into model biases.
-Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder.
+ Regarding the \emph{full} confounder test; while the real data examples in this paper were not typical cases of its potential applications, it has still provided important insights into model biases.
+Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder ($p=0.1$).
 In other words, the test accepted the null hypothesis of full model bias, suggesting that this model might be, in fact, almost exclusively driven by motion artifacts.
@@
 The lack of rigorous statistical tests for confounder-bias significantly hampers the development of predictive models in many fields of research.
-Even tough several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder signal and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias.
+Although several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder signal and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias.
 
@@
 Both approaches are based on a recently published framework for conditional permutation testing, with a mathematical justification of valid type I error control.
-The tests were constructed so that they do not place any assumption on the distribution of the model predictions and require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed with a minimal computational overhead; without needing to re-fit the model.
+The tests were constructed so that they do not place any assumption on the conditional distribution of the model predictions and can be performed with a minimal computational overhead; without needing to re-fit the model.
 
-Given their simplicity, high statistical power, robustness to the violation of the assumptions and thanks to the computationally effective implementation (available in python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the tests are deployable in a wide variety of applications.
+Given their simplicity, high statistical power, robustness to the violation of the assumptions and thanks to the computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the tests are deployable in a wide variety of applications.
 
-As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and allows quantitative, statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research, thereby paving the way for the development of clinically useful machine learning biomarkers.
+As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and allow quantitative, statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research, thereby paving the way for the development of clinically useful machine learning biomarkers.
 
+\section{Acknowledgements}
+I am thankful to Ulrike Bingel (University Hospital Essen, Germany) and Robert Englert (University Hospital Essen, Germany) for their valuable insights and comments on the manuscript.
+This research was supported by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – Projektnummer 316803389 – SFB 1280  and TRR 289 Treatment Expectation - Projektnummer 422744262.
+
+%TC:ignore
 \bibliographystyle{apalike}  
@@
 \input{supplement}
+%TC:endignore
 

### 2021-10-26 manuscript.tex
Updates from Overleaf

@@
 \hline
-\textbf{1.} & $\y \independent \yhat | \c$ \quad  & full confounder test: model exclusively driven by confounder & $Q(\cdot|\c)$ & $\yhat$ or $\y$ \\
+\textbf{1.} & $\yhat \independent \y | \c$ \quad  & full confounder test: model exclusively driven by the confounder & $Q(\y|\c)$ & $(\yhat,\y), (\yhat, \c)$ \\
 %\hline
-\textbf{2.} & $\y \independent \c | \yhat$ \quad & model captures all variance in confounder (not of interest) & $Q(\cdot|\yhat)$ & $\y$ or $\c$ \\
+\textbf{2.} & $\y \independent \c | \yhat$ \quad & model captures all variance in the confounder (not of interest) & $Q(\c|\yhat)$ & $(\y, \c), (\y, \yhat)$ \\
 %\hline
-\textbf{3.} & $\yhat \independent \c | \y$  \quad &  partial confounder  test: model does not capture more confound than what is present in the target & $Q(\cdot|\y)$ & $\yhat$ or $\c$ \\
+\textbf{3.} & $\yhat \independent \c | \y$  \quad &  partial confounder  test: model not directly driven by the confoudner & $Q(\c|\y)$ & $(\yhat, \c), (\yhat, \y)$ \\
 \end{tabular}
@@
 \\The table lists the three possible null hypotheses (H0), and the variables where assumption about the joint/conditional distributions is required/not required.
-($y$: prediction target, $\yhat$: predictions, $\c$: confounder variable) }
+($\y$: prediction target, $\yhat$: predictions, $\c$: confounder variable) }
 \end{table}
@@
 
-The partial confounder test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based "copies" of $\c$,
+The partial confounder test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based 'copy' of $\c$,
 
@@
 
-where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y$ and $j=1,\dots, m$ indexes the "copy" of $\c$ so that
+where, $Q(.|y)$ denotes the conditional distribution of $\c$ given $\y=y_i$ and $j=1,\dots, m$ indexes the 'copy' of $\c$ so that
 
@@
 
-is a permutation of the original vector $\c = (c_1, \dots, c_n)$, with its elements reordered according to the permutation $\quad \boldsymbol{\pi} \in S_n$ where $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$.
+is a permutation of the original vector $\c = (c_1, \dots, c_n)$, with its elements reordered according to the permutation $ \boldsymbol{\pi} \in S_n$ where $S_n$ denote the set of all permutations on the indices $\{1,\dots,n\}$.
 
@@
 
-While the required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$ at random, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times.
+The required permutations could be theoretically sampled with a simple Metropolis-Hastings algorithm that draws uniformly from $S_n$ at random.
+However, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times.
 The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of \cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!b]
-  \centering
-  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
-  \caption{\textbf{Type-I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
-  Type-I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row).
-Non-normal conditional distribution on the second plot is illustrated with blue density diagrams and dependency is depicted with blue dotted lines an all plots.
-False positive rates for the three investigated tests is shown in heatmaps, with darker red meaning higher values.
-The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065.
-Values below this threshold indicate a valid type-I error control and colored white.
-  }
-  \label{fig:sim-h0-demo}
-\end{figure}
-
 \subsection{Validation on simulated data}
@@
 Here I extend these results by performing simulations that evaluate the GAM- and multinomial liner model-based approaches, in a form that is accessible for power calculations in predictive modelling (considering various weights of the target signal in $c$ and the confounder and the target signals in $\yhat$).
-Moreover, I investigate the robustness of the tests against the violation of normality and linearity of the conditional distributions $D(\c|\y)$ and $D(\yhat|\y)$
+Moreover, I investigate the robustness of the tests against the violation of normality and linearity of the conditional distributions $D(\c|\y)$ and $D(\yhat|\y)$.
 
@@
 
-$$ \c \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yc} \quad g(\y) $$
+$$ \c \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yc} \, g(\y) $$
 
-where $f$ is a function to introduce non-normality, namely the \emph{sinh-arcsinh} transformation of \cite{jones2009sinh} defiend as:
+where $f$ is a function to introduce non-normality, namely the \emph{sinh-arcsinh} transformation of \cite{jones2009sinh} definend as:
 
@@
 
-Moreover, non-linearity can be introduced with the function $g$, which can be simply the identity function (no non-linearity is introduced) or e.g. a sigmoid-shaped function, in pour case:
+Moreover, non-linearity can be introduced with the function $g$, which can be simply the identity function (no non-linearity is introduced) or e.g. a sigmoid-shaped function, in our case:
 
@@
 
-$$ \yhat \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yyhat} \quad g(\y)  + w_{c\hat{y}} \quad g(\y)$$
+$$ \yhat \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yyhat} \: g(\y) \, + \, w_{c\hat{y}} \: g(\y)$$
 
@@
 To demonstrate the need for the proposed GAM-based CPT approach for partial confounder testing, its validity was contrasted to partial Spearman correlation and the linear variety of CPT (based on eq.
-\ref{eq:linreg}, as described by \citep{berrett2020conditional}) in the following parameters:
+\ref{eq:linreg}, as described by \cite{berrett2020conditional}) with the following simulation parameters:
 \begin{itemize}
@@
 \section{Results}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!b]
+  \centering
+  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
+  \caption{\textbf{Type-I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
+  Type-I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row).
+Non-normal conditional distribution on the second plot is illustrated with blue density diagrams and dependency is depicted with blue dotted lines an all plots.
+False positive rates for the three investigated tests is shown in heatmaps, with darker red meaning higher values.
+The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065.
+Values below this threshold indicate a valid type-I error control and colored white.
+  }
+  \label{fig:sim-h0-demo}
+\end{figure}
 

### 2021-10-28 simulated/simulate_H1.py
lin CPT omitter for more efficiency

@@ -154,15 +154,16 @@ if __name__ == '__main__':
                                     n_jobs=1,
                                     progress=False)
 
-            res_lin = confound_test(y, yhat, c,
-                                    cat_y=args.cat_y,
-                                    cat_yhat=args.cat_yhat,
-                                    cat_c=args.cat_c,
-                                    num_perms=num_perms,
-                                    cond_dist_method='linear',
-                                    random_state=_random_state,
-                                    n_jobs=1,
-                                    progress=False)
+            # for efficiency
+            res_lin = np.nan #confound_test(y, yhat, c,
+                             #       cat_y=args.cat_y,
+                             #       cat_yhat=args.cat_yhat,
+                             #       cat_c=args.cat_c,
+                             #       num_perms=num_perms,
+                             #       cond_dist_method='linear',
+                             #       random_state=_random_state,
+                             #       n_jobs=1,
+                             #       progress=False)
 
             # do partial correlation, too
             if args.mode.startswith('partial'):

### 2021-10-28 manuscript.tex

@@
 
-The tests provide a strict control for Type-I errors and high statistical power, even for non-linear and non-normal dependencies typical in machine learning.
+The tests provide a strict control for Type I errors and high statistical power, even for non-linear and non-normal dependencies typical in machine learning.
 Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset reveals confounders that were previously partly unreported and found to be hard to correct for with state-of-the-art confound mitigation approaches.
@@
 This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem).
-As an important consequence, partial correlation - a trivial candidate for investigating the dependence between the predictions and the confounder (given the target variable) - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (demonstrated on Fig.
-\ref{fig:sim-h0-demo}).
-These conditions are very common for case of predictive modelling \citep{garcia2009study, kristensen2017whole} (Supplementary Material \ref{sup:nomlinviol})
+As an important consequence, methods like partial correlation - and even its non-parametric variety - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (see e.g.
+\citep{korn1984ranges} or Fig.
+\ref{fig:sim-h0-demo} in this paper), as commonly seen in machine learning.
 
@@
 In the following section, CPT is adapted for partial confounder testing (option 3).
+For an overview, see Fig.
+\ref{fig:overview}.
 The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
@@
 In the case of categorical $c$, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$, with the extra assumption of \emph{complete separation} if $y$ is also categorical (for an invertible Hessian, see e.g.
-\citep{bohning1992multinomial}).
+\citep{bennett1966multiple, jones1975proability}).
 
@@
 
-Moreover, non-linearity can be introduced with the function $g$, which can be simply the identity function (no non-linearity is introduced) or e.g. a sigmoid-shaped function, in our case:
+Moreover, non-linearity can be introduced with the function $g$, which can be simply the identity function (no non-linearity is introduced in this case) or e.g. a sigmoid-shaped function, in our case:
 
@@
 
-The simulated predicted values are constructed analogously:
+The simulated predicted values are constructed similarly:
 
-$$ \yhat \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yyhat} \: g(\y) \, + \, w_{c\hat{y}} \: g(\y)$$
+$$ \yhat \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{y\hat{y}} \: g(\y) \, + \, w_{c\hat{y}} \: g(\y)$$
 
@@
 
-To demonstrate the need for the proposed GAM-based CPT approach for partial confounder testing, its validity was contrasted to partial Spearman correlation and the linear variety of CPT (based on eq.
+To demonstrate the need for the proposed GAM-based CPT approach for partial confounder testing (Fig.
+\ref{fig:sim-h0-demo}), its validity was contrasted to partial Spearman correlation and the linear variety of CPT (based on eq.
 \ref{eq:linreg}, as described by \cite{berrett2020conditional}) with the following simulation parameters:
@@
     \item sample size: $n = 1000$
-    \item $w_{c\hat{y}} = 0$ (H0 simulations only)
+    \item $w_{c\hat{y}} = 0$ (i.e. H0 simulations only)
     \item all combinations of:
@@
     \end{itemize}
-    \item non-normality ($f_{\delta = 0.1, \epsilon = 2)}$ and non-linearity (sigmoid $g$)
+    \item non-normality ($f_{\delta = 0.1, \epsilon = 2}$) and non-linearity (sigmoid $g$)
 \end{itemize}
 
-For each parameter combination, 1000 repetitions were performed and false positive rate was calculated as the ratio of p-values smaller than $\alpha = 0.05$.
+For each parameter combination, 1000 repetitions were performed and false positive rates were calculated as the ratio of p-values smaller than $\alpha = 0.05$.
 
-These simulation cases are visualized (with $w_{yc} = w_{y\hat{y}} = 2$) on the left of Figure \ref{fig:sim-h0-demo}.
+The simulation cases are exemplified (with $w_{yc} = w_{y\hat{y}} = 2$) on the left of Figure \ref{fig:sim-h0-demo}.
 
@@
 
-Both for the partial and the full confound test, 100 repetitions were performed of all combination of the following parameter values: 
+Both for the partial and the full confounder test, 100 repetitions were performed of all combination of the following parameter values: 
 \begin{itemize}
@@
 
-The 'partial' and 'full' confound tests, as implemented in version 0.20.0 of the package \emph{'mlconfound'} were run with default parameters, that is 1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations and by implying categorical variables, where needed.
+The partial and full confound tests, as implemented in version 0.20.1 of the package \emph{'mlconfound'} were run with default parameters (1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations) and by implying categorical variables, where needed.
 
-All analysis code used for the simulations is available on github\footnote{\href{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}}.
+All code used for the simulations is available on github\footnote{\href{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}{https://github.com/pni-lab/mlconfound-manuscript/tree/main/simulated}}.
 
@@
 
-\subsubsection*{HCP: testing age- and acquisition batch-bias in fluid intelligence prediction}
+\subsubsection*{HCP: testing age and acquisition batch bias in fluid intelligence prediction}
 
-Human Connectome Project contains imaging and behavioral data of approximately 1200 healthy subjects \citep{van2013wu}.
+The Human Connectome Project dataset contains imaging and behavioral data of approximately 1200 healthy subjects \citep{van2013wu}.
 Preprocessed resting state fMRI connectivity data (partial correlation matrices) \citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores (IQ), measured with Penn Progressive Matrices \citep{duncan2000neural}.
@@
 
-The total of 5 types of features (raw, regressing out acquisition batch, regressing out age group, COMBAT with acquisition batch, COMBAT with age group) were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the individual fluid intelligence scores with a ridge-regression \citep{hoerl1970ridge}.
+Each of the 5 types of features (raw, regressing out acquisition batch, regressing out age group, COMBAT with acquisition batch, COMBAT with age group) was independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the individual fluid intelligence scores with a ridge regression \citep{hoerl1970ridge}.
 The $\alpha$ parameter of the ridge model was considered as a hyperparameter ($\alpha \in \{0.00001, 0.0001, 0.001, 0.01, 0.1, 1, 10, 100, 1000, 10000, 100000\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric.
@@
 The total of five types (raw, feature regression of site, feature regression of motion, COMBAT with site, COMBAT with motion) of features were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the diagnosis (DX: ASD vs.
-HC) with a L2-regularized logistic regression (as e.g.
-in \citep{dadi2019benchmarking}).
+HC) with a L2-regularized logistic regression (see rationale in \citep{dadi2019benchmarking}).
 The $C$ parameter of the model was considered as a hyperparameter ($C \in \{0.1, 1, 10\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with area under the receiver operator curve (AUC under ROC) as optimization metric.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% fig: overview
 \begin{figure}[!b]
   \centering
-  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
-  \caption{\textbf{Type-I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
-  Type-I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row).
-Non-normal conditional distribution on the second plot is illustrated with blue density diagrams and dependency is depicted with blue dotted lines an all plots.
-False positive rates for the three investigated tests is shown in heatmaps, with darker red meaning higher values.
-The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065.
-Values below this threshold indicate a valid type-I error control and colored white.
+  \includegraphics[width=0.75\paperwidth]{fig/overview.eps}
+  \caption{\textbf{Overview of the proposed partial confoudner test}. \\
+    The partial confoudner test models the conditional distribution of the confounder, given the target variable, with a generalized additive model (GAM).
+The parallel-pairwise Markov-chain Monte-Carlo sampler draws permutations of the confounder variable that comply with the GAM-based conditional distribution (permutation 1, 2, ..., m).
+The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted confoudner variables.
+The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original.
+Figure source code available as jupyter notebook: \href{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}
   }
-  \label{fig:sim-h0-demo}
+  \label{fig:overview}
 \end{figure}
 
-\begin{figure}[!b]
-  \centering
-  \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
-  \caption{\textbf{Type-I error control and power of the conditional permutation-based confound testing for predictive modelling in normally distributed simulated data}. \\
-  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette at the bottom) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\yhat}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\yhat}$ (rows) and for different sample sizes (N, columns).
-Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%.
-First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounder bias).
-  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively.
-The higher 95\% confidence limit for the type I error rates is 0.11 for each tile.
-  }
-  \label{fig:sim-normal}
-\end{figure}
+\subsection{The partial and full confoudner tests}
 
-The conditional permutation based \emph{partial} and \emph{full confounder tests} for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.
+The proposed \emph{partial} and \emph{full confoudner tests} for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.
+The inner workings of the tests is summarized on Fig.
+\ref{fig:overview}.
+In short, the tests model the conditional distribution between the confounder and the target variable with a generalized additive model (GAM) and then use a parallel-pairwise Markov-chain Monte-Carlo sampler that draws permutations of the confounder or target variable, for the partial and full confounder test, respectively.
+As a result, the permuted variables comply with the GAM-based conditional distribution.
+The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted variables.
+The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original.
+The strength of this approach lies in avoiding to place any assumptions on the conditional distributions of the model output.
+This property ensures valid model diagnostics even in cases where Pearson and Spearman partial correlation and many other methods fail (as shown on Fig.
+\ref{fig:sim-h0-demo}).
 
@@
 
+As suggested by theory and shown by the simulations, both of the proposed tests provide a valid Type I error control (Fig.
+\ref{fig:sim-normal}), even in case of non-linearity and non-normality (Figs.
+\ref{fig:sim-h0-demo} and \ref{fig:sim-non-normal}), except when non-normality is very extreme (purple distribution on Fig.
+\ref{fig:sim-non-normal}, kurtosis: 42, skewness: -6).
+With this the proposed tests overcome the limitations of partial Spearman correlation and linear conditional permutations, demonstrated on Fig.
+\ref{fig:sim-h0-demo}.
+
+For detailed results, see also Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-lin-full} and \ref{fig:sim-bbb-sig-full}.
+
+%fig:sim-h0-demo
+\begin{figure}[!b]
+  \centering
+  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
+  \caption{\textbf{Type I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
+  Type I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), slightly non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row).
+Non-normal conditional distribution on the second plot is illustrated with blue density diagrams (kurtosis: -0.8, skewness: -0.1) and dependency of the mean is depicted with blue dotted lines an all plots.
+False positive rates for the three investigated tests is shown in heatmaps, with darker red meaning higher values.
+The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065.
+Values below this threshold indicate a valid type I error control and colored white.
+  }
+  \label{fig:sim-h0-demo}
+\end{figure}
+
+%fig:sim-normal
+\begin{figure}[!b]
+  \centering
+  \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
+  \caption{\textbf{Type I error control and power of the partial confoudner test based on simulations with normal conditional distribution and linear dependencies.} \\
+  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette on the right) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\hat{y}}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\hat{y}}$ (rows) and for different sample sizes (N, columns).
+Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1.0 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%.
+First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounder bias).
+  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively.
+The higher 95\% confidence limit for a positive rate of $alpha=0.05$ is 0.11 for each tile.
+  }
+  \label{fig:sim-normal}
+\end{figure}
+
+%fig:fig:sim-non-normal
 \begin{figure}[!b]
@@
 
-Simulations based on normally distributed data justified that the proposed confounder tests - in line with theory - provide a valid type I error control (For detailed results, see \ref{fig:sim-normal} for the 'partial' test on numerical data and Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-lin-full}, \ref{fig:sim-bbb-sig-full}, for the partial test on categorical data and the full tests on both numeric and categorical data).
-In case of a valid test, the expected type I error rate in the null hypothesis simulation cases (first row and first column of the figures) is less then or equal to the alpha level (0.05).
-The upper 95\% confidence limit for the type I error rates for one simulation case (i.e.
-a single tile on the figures) is 0.11 (binomial confidence intervals with 100 repetitions).
-The results suggest, that the proposed tests are somewhat overly conservative if the model has a high predictive performance (type I error rates closer to 0 than expected).
-
-Simulations with non-normal data showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $\y$ and the confounder $\c$ and, as expected, completely non-parametric regarding the distribution of the predictions $\yhat$).
-The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal}, involved various \emph{sinh-arcsinh} distributions, from normal (Fisher's kurtosis: 0, skewness: 0) to severely skewed and heavy tailed (kurtosis: 42, skewness: -6).
-The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to kurtosis and skewness values as high as 5 and 2 respectively.
-Furthermore, in terms of type I error rate, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).
-
 \subsubsection*{Power}
 
-Power estimates for the 'partial' test (with normality) are shown on Figure \ref{fig:sim-normal}.
-Notably, with sample sizes as large as 1000, a confounder contributing to only 0.4-0.6\% to the variance of the predictions can already be successfully detected ($w_c = 0.05 - 0.07$) with a power of 60-90\%.
-With a sample size of 500, a confounder contributing to only 1.4-3\% to the variance of the predictions can be detected ($w_c = 0.1 - 0.14$) with a power greater than 80\% in most of the simulation cases.
-A sample size of 100 yields an approximate power of 60-100\% for a confounder-variance of 6-7\% ($w_c = 0.2 - 0.21$) in most of the cases.
+Power estimates for the 'partial' test (with normal and linear simulations) are shown on Figure \ref{fig:sim-normal}.
+Notably, with sample sizes as large as 1000, a confounder contributing to only ~4\% to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%.
+With a sample size of 500, the same confounder bias is still detected with a power greater than 84\% in all of the simulation cases.
+A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) for a robust detection, with a power of 75-98\%.
 Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%.
-If the confounder explains more than 15\% of variance, it is almost certainly detected with $n \geq 50$.
+If the confounder explains more than 25\% of variance, it is almost certainly detected even with the low sample size of $n \geq 50$.
 
-Violation of the normality assumptions for $\y$ and $\c$ has a negative effect on the power of the tests, however the loss of power is low or even negligible, if normality is not extremely violated (first 4 distributions on Fig.
-\ref{fig:sim-non-normal}A) and the confounder-to-signal ratio is greater or equal then 0.3 (Fig.
-\ref{fig:sim-non-normal}C).
+Simulations with non-normal and non-linear dependencies has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}).
 
@@
   \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
-  \caption{\textbf{Acquisition- and age-bias of fluid IQ prediction in the HCP dataset.} \\
+  \caption{\textbf{Acquisition- and age-bias of fluid intelligence prediction in the HCP dataset.} \\
   Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies.
 Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data.
-This bias is robustly detected by the proposed 'partial' confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
-  Relation between the observed ($Gf$) and predicted ($\hat{y}$) IQ scores and the confounder variables is shown by the graphs as $R^2 values$.
+This bias is robustly detected by the partial confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
+  Relation between the observed ($Gf$) and predicted ($\hat{y}$) IQ scores and the confounder variables is given on the graphs via $R^2 values$.
 Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
   Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
-P-values are determined with the 'partial confound test'.
+P-values are determined with the partial confound test.
 P-values of the 'full' confound test were all less then 0.0001, i.e.
-confounder did not fully drive prediction for any of the models.
+the confounders did not fully drive prediction for any of the models.
   }
@@
 
-After applying confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p > 0.05$ for all, second and third columns of Fig.
+After applying confound mitigation approaches (feature regression or COMBAT) the partial confounder test did not provide evidence of confounder bias anymore ($p > 0.05$ for all; shown in the second and third columns of Fig.
 \ref{fig:hcp}), neither for acquisition batch nor for age.
@@
 Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
-P-values are determined with the 'partial confound test'.
-P-values of the 'full' confound test were all less then 0.0001, i.e.
+P-values are determined with the partial confound test.
+P-values of the full confound test were all less then 0.0001, i.e.
 confounder did not fully drive prediction for any of the models.
@@
 Several sites (e.g.
-Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately ($R^2=0.169$ and $0.111$, respectively) associated to a lower probability for the participant to be assigned to the ASD group by the model.
+Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately associated ($R^2=0.169$ and $0.111$, respectively) to a lower probability for the participant to be assigned to the ASD group by the model.
 
-Both feature regression and COMBAT seemed to significantly attenuate the effect of imaging center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.01$ for feature regression and COMBAT, respectively, see the second and third columns of the first row on Fig \ref{fig:abide}).
+Both feature regression and COMBAT seemed to significantly attenuate the effect of imaging center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.009$ for feature regression and COMBAT, respectively (second and third columns of the first row on Fig \ref{fig:abide}).
 
-When trying to mitigate the effect of in-scanner motion (bottom row on Fig \ref{fig:abide}), feature regression obviously failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates.
-The partial confounder successfully detected the resulting strong bias ($p<0.0001$).
+When trying to mitigate the effect of in-scanner motion (bottom row on Fig \ref{fig:abide}), feature regression failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates.
+The partial confounder test successfully detected the resulting strong bias ($p<0.0001$).
 Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$, bottom right panel of Fig \ref{fig:abide}).
@@
 
-The full confounder test was highly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$) Indicating that this model might have been (almost) exclusively driven by motion artifacts.
+The full confounder test was strongly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$), indicating that this model might have been (almost) exclusively driven by motion artifacts.
 
@@
 Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models.
-Testing confounder-bias of predictive models can be achieved by considering the "confounder-problem" as assessing \emph{conditional independence} between the predicted values and either the target or the confounder variable.
-These two scenarios result in two different tests, probing whether the model is partially or fully driven by the given confounder.
-Importantly, in predictive modelling, the model predictions often follow a non-normal distribution, thus any test for confounder bias must be, at least partly, non-parametric.
+The problem of testing confounder-bias in predictive models can considered as assessing \emph{conditional independence} between the predicted values, the the target and the confounder variable.
+Importantly, in predictive modelling, the model predictions are often characterized by a non-normal and/or non-linear conditional dependence on the target variable, thus any test for confounder bias must be, at least partly, non-parametric.
 However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
-Here I propose the use of the conditional permutation test (CPT) of \cite{berrett2020conditional} to test confounder bias in predictive models.
-Adapting this approach for the confounder testing problem allows placing assumptions about the joint distribution of only two out of the three variables involved; that is, it allows constructing both the partial and full confounder tests so that no assumption is made about the distribution of the model output.
-The tests are made available in the python package \emph{mlconfound} which builds upon a parallelized and computationally effective implementation of the Markov-chain Monte-Carlo based parallel pairwise sampler of \cite{berrett2020conditional}.
-It utilizes coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic and assumes that the conditional distribution between the target variable and the confounder is normal.
-The partial and full confounder tests available in \emph{mlconfound} can be uniformly applied for any predictive model, given that realistic (e.g.
-cross-validated) model predictions can be obtained.
-Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost; re-fitting the model is not required.
+In practice, this means that all candidate tests, that assume normality and/or linearity of the joint distribution, will necessarily fail to properly control for Type I errors in many realistic predictive modelling scenarios.
+This is clearly demonstrated on Fig.
+\ref{fig:sim-h0-demo} in case of partial Spearman correlation.
+Although warnings about these issues were given from early on \citep{korn1984ranges} and received a fair amount of attention recently \citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated within the predictive modelling community.
 
-As expected from theoretical justification, both tests displayed a valid type I error control in the simulations with the necessary assumptions, that is, even if $\yhat$ is not normally distributed.
-In fact, the tests were found to become more conservative (i.e.
-lower than expected type I error rates ) in the case of high-performance predictive models (when the $R^2$ between the predicted and observed values is greater than 0.7).
+From the three involved variables, the most problematic one is clearly the prediction output, known to often display non-normal and non-linear conditional dependencies on the other two.
+Therefore, a method being distribution-free for this variable, but inputting custom distributional information about the other two variables might provide a sufficient flexibility required for predictive model diagnostics.
 
+To achieve this, here I propose to combine the conditional permutation test (CPT) of \cite{berrett2020conditional} with generalized additive model (GAM)-based conditional distribution modelling \citep{hastie1987generalized}.
+The proposed mathematical framework offers two novel tests for probing confounder bias: the partial and full confounder tests.
+These tests can handle both numerical and categorical data and can be applied for any predictive model, without having to re-fit the model, that is, with a negligible extra computational cost.
+
+As expected from theory, both tests displayed a valid type I error control in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable.
 Even though the tests' valid type I error control is justified by theory and reinforced by the simulations, a low statistical power could significantly limit their usefulness.
 The simulations presented in this paper revealed that both tests have a high, practically relevant power in a wide variety of scenarios.
-Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confounder bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confounder explaining 1-2\% of the variance in the prediction).
-A lower sample size of 100 will still very likely identify medium sized confounder-biases (around 9\%).
+Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confounder bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confounder explaining about 4\% of the variance in the prediction).
+A lower sample size of 100 will still very likely identify medium sized confounder-biases (around 12\%).
 Strong biases  (>25\%) can be very robustly detected with sample sizes as low as 50.
-As expected, the power of the full confounder test is very similar to the partial test, since the only difference is that the conditional contribution of the target to the prediction is investigated, instead that of the confounder.
+As expected, the power of the full confounder test is very similar to that of the partial confounder test, since the only difference is that the conditional contribution of the target to the prediction is investigated, instead that of the confounder.
+While different biomedical applications might consider different amounts of bias to be relevant, an a-priori power calculation can help identifying the necessary sample size for proper model diagnostics.
 
-While different biomedical applications might consider different amounts of bias to be relevant, an a-priori power calculation can help identifying the necessary sample size in order to exclude the possibility of a relevant confounder bias with a high probability.
+Simulations also confirmed that non-normality and non-linearity of the conditional distribution of the predictions and the confounder (given the target) has a negligible effect on the power of both tests, except in case of extreme violations of normality.
+This result confirms that the proposed tests, unlike e.g.
+partial correlation, remain valid, independent of the distributional characteristics of the applied model's output.
+Validity, power and robustness renders the proposed tests as unprecedented tools for model diagnostics in a wide spectrum of predictive modelling scenarios.
 
-Next to statistical power, the practical applicability of the proposed tests is also highly affected by their robustness to violations of the normality assumptions on the conditional distribution of $\y$ and $\c$.
-Importantly, the simulations strongly suggest that both tests are strikingly robust to the violation of this assumption.
-Type I error is well controlled even in case of highly skewed and heavy tailed distributions (kurtosis > 5, skewness > 2) and gets significantly inflated only with extreme violations of normality (see the purple distribution on Fig.
-\ref{fig:sim-non-normal}).
-Due to the conservatives of the tests, if the predictive model is very accurate, even this extremely skewed and heavy tailed distribution resulted in lower than the nominal type I error rates.
-
-Non-normality has a negative effect on the power of both tests, however, the decrease in power is in most cases small or even negligible and gets only more pronounced if the sample size and/or the confounder-effect is low.
-
-Validity, power and robustness renders the proposed tests as unprecedented tools for model diagnostics in a wide spectrum of predictive modelling scenarios.
-For instance, in biomedical research, confounder-bias has previously often been investigated (if investigated at all) by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.
+In biomedical research, confounder-bias has previously often been investigated (if investigated at all) simply by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.
 \citep{spisak2020pain}.
@@
 Two examples for this situation can be found in the presented analysis of real data; when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data (see Table \ref{tab:unconditional-pvals}).
-
-Arguably, in certain applications, the researcher might aim to fully eliminate confounder-related variance of the predictions and, in this case, testing unconditional independence between the confounder and the predictions might be sufficient.
-However, a much more realistic aim is to construct models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e.
-the prediction-confounder association is proportional to, and can likely be fully explained by, the target-confounder association.
+These examples clearly indicate that guiding model diagnostics based on unconditioned associations between the confounder and the model output might result in overly strict decisions.
+More realistic is to aim at models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e.
+the prediction-confounder association is proportional to, and can be fully explained by, the target-confounder association.
 Put simply, this is what is tested by the proposed partial confounder test.
 
-Obviously, a true confounder bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder association, i.e.
+%Obviously, a true confounder bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder association, i.e.
 the model might still use the confounder's effect on the features (e.g artifacts) to construct the corresponding part of the variance in the prediction.
 Nevertheless, in most of the cases, it does not seem likely that the model, at the same time, (i) misses that specific "part" of the target variable's variance in the features that is shared with the confounder (ii) captures exclusively that component of the confounder's variance in the features that is needed to mimic an unbiased model.
-Usually a much more plausible explanation is that a direct confounder effect in the model output is simply negligible (e.g because it is not to be found in the features after confound mitigation).
+%Usually a much more plausible explanation is that a direct confounder effect in the model output is simply negligible (e.g because it is not to be found in the features after confound mitigation).
 
-Among the many potential areas of applications, the proposed tests could be particularly useful in translational neuroscience.
-A characteristic example is provided by the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds a great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making \citep{woo2017building, wager2013fmri, spisak2020pain}.
+Among the many potential areas of applications, the partial and full confounder tests could be particularly useful in translational neuroscience.
+A characteristic example is provided by the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making \citep{woo2017building, wager2013fmri, spisak2020pain}.
 However, the presence of confounders typical for biomedical research (e.g.
@@
 imaging artifacts) presents a great challenge to these efforts and is, currently, often largely neglected.
-
 To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.
@@
 \cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
-
 The hypothesised confounder effects were clearly present in both datasets.
@@
 the CMU (Carnegie Mellon University) center as neurotypical control participants.
-It's easy to see that this bias can possibly lead to a very low predictive performance on out-of-center data.
-Similarly, a model biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with higher motion or altered motor characteristics (as known e.g.
+It is easy to see that this bias can lead to calibration issues and, possibly, a very low predictive performance on out-of-sample data.
+Similarly, a model biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with a tendency for higher in-scanner motion, e.g.
+due to atypical motor function (as known e.g.
 for ADHD \citep{eloyan2012automated}).
@@
 Confounder mitigation techniques, with one exception, tended to decrease confounder bias.
-The partial confounder test provided detailed insights into the effectiveness of the techniques for each confounder.
+The partial confounder test provided quantitative, statistically rigorous metrics for assessing the effectiveness of the techniques for each confounder.
 In the HCP data, it revealed that both the acquisition-bias and the age-bias was very effectively removed by both feature regression and COMBAT (p>0.05 for all).
@@
 In other words, the test accepted the null hypothesis of full model bias, suggesting that this model might be, in fact, almost exclusively driven by motion artifacts.
-For all other models, the full confounder test confirmed that - even if there was a significant confounder-bias - the model was not fully driven by the confounder (in other words, the target and the prediction were not independent, given the confounders).
-This examples highlights that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased and the questions is whether there is any biomedical relevant signal captured by the model, at all.
+This example highlight that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased by various confounders and the questions is whether there is any biomedically relevant signal captured by the model, at all.
 

### 2021-10-28 simulated/simulate_H1.py
lin CPT omitter for more efficiency

@@ -149,7 +149,7 @@ if __name__ == '__main__':
                                     progress=False)
 
             # for efficiency
-            res_lin = np.nan #confound_test(y, yhat, c,
+            res_lin = res_gam.copy() #confound_test(y, yhat, c,
                              #       cat_y=args.cat_y,
                              #       cat_yhat=args.cat_yhat,
                              #       cat_c=args.cat_c,
@@ -158,6 +158,7 @@ if __name__ == '__main__':
                              #       random_state=_random_state,
                              #       n_jobs=1,
                              #       progress=False)
+            res_lin.p = np.nan
 
             # do partial correlation, too
             if args.mode.startswith('partial'):

### 2021-10-28 simulated/simulate_H1.py
lin CPT omitter for more efficiency

@@ -149,7 +149,7 @@ if __name__ == '__main__':
                                     progress=False)
 
             # for efficiency
-            res_lin = res_gam.copy() #confound_test(y, yhat, c,
+            res_lin = res_gam #confound_test(y, yhat, c,
                              #       cat_y=args.cat_y,
                              #       cat_yhat=args.cat_yhat,
                              #       cat_c=args.cat_c,
@@ -158,7 +158,6 @@ if __name__ == '__main__':
                              #       random_state=_random_state,
                              #       n_jobs=1,
                              #       progress=False)
-            res_lin.p = np.nan
 
             # do partial correlation, too
             if args.mode.startswith('partial'):

### 2021-10-28 manuscript.tex
Updates from Overleaf

@@
 
-Predictive modelling and supervised learning methods have recently become increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized, spisak2020pain, walsh2021dome}.
+Predictive modelling has recently become increasingly important in biomedical research and holds promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized, spisak2020pain, walsh2021dome}.
 When evaluating the usefulness and applicability of such markers, predictive performance (i.e.
@@
 Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - i.e the so-called confounder and collider biases \citep{prosperi2020causal} - can be detrimental to the model's biomedical validity and generalizability.
-Such biases can originate form many sources.
+Examples include models picking up on measurement artifacts (in-scanner head motion artifacts in magnetic resonance imaging-based predictive models of Alzheimer's \citep{rao2017predictive}, attention deficit hyperactivity disorder \citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD) \citep{gotts2013perils, spisak2014voxel, spisak2019optimal}), demographic and psychometric variables (e.g models trained to predict intelligence \citep{ cole2012global, he2020deep} might provide a statistically significant predictive performance by picking up solely on age-related variance \citep{dubois2018distributed, lohmann2021predicting}), sampling bias and stochastic group differences (racially biased machine learning models \citep{ obermeyer2019dissecting, lwowski2021risk}), batch effects or, in multi-center studies, center-effects.
 
-First, measurement artifacts can obviously be considered as a potential source of confounder bias.
-For instance, magnetic resonance imaging (MRI)-based predictive models can be driven by in-scanner head motion artifacts, next to (or instead of) signal of neural origin, especially, if the prediction target is also associated with motor function and motion characteristics, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive}, attention deficit hyperactivity disorder (ADHD) \citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD) \citep{gotts2013perils, spisak2014voxel, spisak2019optimal}.
-Likewise, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.
-
-Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling.
-For instance, resting-state functional connectivity has been shown to be strongly predictive to age \citep{dukart2011age, wang2012decoding} and moderately to fluid intelligence \citep{ cole2012global, he2020deep}.
-As fluid intelligence is known to decline with aging \citep{kievit2018neural}, models trained to predict intelligence might provide a statistically significant predictive performance by picking up solely on age-related variance \citep{dubois2018distributed, lohmann2021predicting}.
-Importantly, the target variable must not necessarily display a true association to the confoudner to produce a biased model.
-Model training can be sensitive to sampling bias and stochastic group differences in the training set; a phenomenon that might have calamitous consequences, as increasingly recognised, for instance, in case of racially biased machine learning models \citep{ obermeyer2019dissecting, lwowski2021risk}.
-
-Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance} as the models can utilize such effects to explain (random or true) center/batch-differences in the predictive target.
-Such models are hard to calibrate, might output systematically biased predictions and might display a worse-than-expected performance when applied on data from new centers/batches.
-
-Various data cleansing, correction and harmonization approaches have been proposed to mitigate detrimental effects of confounders, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{dukart2011age, spisak2014voxel, abdulkadir2014reduction, rao2017predictive} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}.
-However, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.
+While various data cleaning methods might help in mitigating confounder bias \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, rao2017predictive, johnson2007adjusting}, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.
 
 Powerful and robust statistical tests to quantify confounding effects in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches.
-Such methods, however, have to tackle the fact that the conditional distribution of the models' output, given the target, is often non-normal and non-linear, as a consequence of e.g.
-model regularization \citep{garcia2009study, kristensen2017whole}; see Supplementary Material \ref{sup:nomlinviol} for a simplistic example.
+Although confounder bias can be expressed in terms of conditional independence of the model predictions on the target and the confoudner variables, the proper evaluation of the conditional independence among these variables is challenging.
+Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation) are not valid measures of conditional independence.
+Although warnings about this issue were given from early on \citep{korn1984ranges}, and received a fair amount of attention recently \citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g.
+feature-set characteristics and model regularization \citep{garcia2009study, kristensen2017whole}; see Supplementary Material \ref{sup:nomlinviol} for a simplistic example.
 
-Recently, three dedicated techniques \citep{chaibub2019permutation, ferrari2020measuring, wachinger2021detect} have been proposed for quantifying confounder effects in predictive modelling.
+Alternative techniques for quantifying confoudner bias either fail to control type error (as  the known in the case of balanced permutations \cite{southworth2009properties, hemerik2018exact}, used in \cite{chaibub2019permutation}), or do not aim at all to do so \cite{ferrari2020measuring, wachinger2021detect}.
+Moreover, without some modifications, they are applicable for either only numeric \cite{wachinger2021detect} or only categorical variables \citep{chaibub2019permutation, ferrari2020measuring} and involve re-fitting the model \citep{chaibub2019permutation, ferrari2020measuring}, which might not be feasible for models with high a computational cost (e.g.
+with nested cross-validation).
+Some techniques \cite{wachinger2021detect}, furthermore, probe a null hypothesis (direct causation between the features and the prediction target) that is not applicable for for many kinds of data and models \footnote{For instance, causation between several psychological or clinical states and functional MRI activity/connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation.
+In such cases, the alternative hypothesis of this test might be true in absence of any real confounder.
+}.
 
-The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via 'restricted permutations' that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature).
-However, as shown by \cite{southworth2009properties} and further discussed by \cite{hemerik2018exact}, balanced permutations do not provide a correct null behavior and can be invalid in certain circumstances.
-Indeed, \cite{ferrari2020measuring} demonstrated that this method does not guarantee a proper type I error control in certain, realistic scenarios.
-The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, on the other hand, does not provide any quantification of statistical significance.
-Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high a computational cost (especially when hyperparameters are optimized in a nested cross-validation).
-The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the predictors are a direct cause of the target variable or, alternatively, there is an unobserved random variable that is the cause of both.
-However, this null hypothesis might not be of interest in many real applications.
-For instance, causation between several psychological or clinical states and functional MRI activity/connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation.
-In such cases, the alternative hypothesis of this test might be true in absence of any real confounder.
+This work aims to construct a test of confounder bias which (i) can be applied post-hoc, without re-fitting the model (i.e.
+requires only the target variable, the confoudner and the model predictions), (ii) guarantees valid type-I error control even if the involved conditional distributions are non-normal and/or non-linear, (iii) is applicable for classification and prediction problems, both with numerical and categorical confounders.
 
@@
 
-In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}.
+%In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}.
 This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem).
-As an important consequence, methods like partial correlation - and even its non-parametric variety - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (see e.g.
+%As an important consequence, methods like partial correlation - and even its non-parametric variety - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (see e.g.
 \citep{korn1984ranges} or Fig.
@@
 
-To overcome these limitations, I extend the permutation-based conditional independence framework of \cite{berrett2020conditional} with a generalized additive model based conditional distribution estimation, which allows avoiding to place any assumptions about the most problematic variable, the prediction output.
-The framework allows constructing two tests: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
-Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
-The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality and non-linearity).
+%To overcome these limitations, I extend the permutation-based conditional independence framework of \cite{berrett2020conditional} with a generalized additive model based conditional distribution estimation, which allows avoiding to place any assumptions about the most problematic variable, the prediction output.
+%The framework allows constructing two tests: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
+%Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
+%The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality and non-linearity).
 
-The usefulness of the proposed tests in detecting various types of confounder bias is demonstrated in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.
-First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} is used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{ dubois2018distributed, lohmann2021predicting} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
-Second, functional connectivity data from the ABIDE \citep{di2014autism} database is used to investigate the potential motion- and center-bias \cite{gotts2013perils, spisak2014voxel, spisak2019optimal} when training models that aim to predict ASD diagnosis.
+%The usefulness of the proposed tests in detecting various types of confounder bias is demonstrated in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.
+%First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} is used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{ dubois2018distributed, lohmann2021predicting} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
+%Second, functional connectivity data from the ABIDE \citep{di2014autism} database is used to investigate the potential motion- and center-bias \cite{gotts2013perils, spisak2014voxel, spisak2019optimal} when training models that aim to predict ASD diagnosis.
+
+\section{Results}
+
+% fig: overview
+\begin{figure}[!b]
+  \centering
+  \includegraphics[width=0.75\paperwidth]{fig/overview.eps}
+  \caption{\textbf{Overview of the proposed partial confoudner test}. \\
+    The partial confoudner test models the conditional distribution of the confounder, given the target variable, with a generalized additive model (GAM).
+The parallel-pairwise Markov-chain Monte-Carlo sampler draws permutations of the confounder variable that comply with the GAM-based conditional distribution (permutation 1, 2, ..., m).
+The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted confoudner variables.
+The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original.
+Figure source code available as jupyter notebook: \href{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}
+  }
+  \label{fig:overview}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{The partial and full confoudner tests}
+
+%fig:sim-h0-demo
+\begin{figure}[!b]
+  \centering
+  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
+  \caption{\textbf{Type I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
+  Type I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), slightly non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row).
+Non-normal conditional distribution on the second plot is illustrated with blue density diagrams (kurtosis: -0.8, skewness: -0.1) and dependency of the mean is depicted with blue dotted lines an all plots.
+False positive rates for the three investigated tests is shown in heatmaps, with darker red meaning higher values.
+The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065.
+Values below this threshold indicate a valid type I error control and colored white.
+  }
+  \label{fig:sim-h0-demo}
+\end{figure}
+
+The idea of conditional independence provides a straightforward framework for assessing the confounder bias in predictive models.
+However, handling the non-normal and/or non-linear conditional dependence often seen in predictive models \citep{garcia2009study, kristensen2017whole} poses a great challenge.
+In fact, as recently shown by Shah and colleagues \cite{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
+In practice, this means that all candidate tests, that assume normality and/or linearity of the joint distribution, will necessarily fail to properly control for Type I errors in many realistic predictive modelling scenarios.
+For instance, a trivial candidate, partial Spearman correlation, exhibits inflated type I errors even with slight violations of normality and linearity, as clearly demonstrated with simulated data on Fig.
+\ref{fig:sim-h0-demo}.
+
+As the most problematic from the three variables (target, prediction, confounder) is clearly the model prediction \citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics.
+Exactly this can be achieved with the combination of the novel framework of conditional permutation testing (CPT) \cite{berrett2020conditional} and the use of generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}-based conditional distribution estimation.
+The proposed approach offers two novel tests for probing confounder bias: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
+These tests place no assumptions on the conditional distributions of the model predictions, ensuring valid model diagnostics even in cases of non-normally and non-linearly dependent predictions.
+
+The inner workings of the \emph{partial confoudner test} is summarized on Fig.
+\ref{fig:overview}.
+In short, the test models the conditional distribution between the confounder and the target variable with a generalized additive model (GAM) - or with a multinomial logistic regression, in case of categorical confounder - and then uses a so-called parallel-pairwise Markov-chain Monte-Carlo sampler that draws permutations of the confounder, so that the permuted variables still comply with the GAM-based conditional distribution.
+The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted variables.
+The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original.
+The \emph{full confoudner test} works in an analogous way, with the difference that it creates permuted copies of the target variable, instead of the confoudner.
+
+These tests can handle both numerical and categorical data and can be applied for any predictive model, without having to re-fit the model, that is, with a negligible extra computational cost.
+The proposed \emph{partial} and \emph{full confoudner tests} have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.
+
+\subsection{Simulations}
+
+%fig:sim-normal
+\begin{figure}[!b]
+  \centering
+  \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
+  \caption{\textbf{Type I error control and power of the partial confoudner test based on simulations with normal conditional distribution and linear dependencies.} \\
+  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette on the right) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\hat{y}}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\hat{y}}$ (rows) and for different sample sizes (N, columns).
+Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1.0 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%.
+First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounder bias).
+  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively.
+The higher 95\% confidence limit for a positive rate of $alpha=0.05$ is 0.11 for each tile.
+  }
+  \label{fig:sim-normal}
+\end{figure}
+
+%fig:fig:sim-non-normal
+\begin{figure}[!b]
+  \centering
+
+  \includegraphics[width=0.5\paperwidth]{fig/sim_non-norm.eps}
+  \caption{\textbf{Robustness of conditional permutation based confound testing to non-normality and non-linearity.} \\
+  Simulations included variables with five different degrees of non-normality (top panel), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation (yellow: normally distributed).
+Fisher's kurtosis and skewness is given for each distribution.
+False and true positive rates in the simulations under H0 and H1, respectively, for each investigated sample size (N), are depicted by barplots for both linear and sigmoid dependency structure.
+Upper 95\% binomial confidence limit corresponding to $alpha=0.05$ is shown with a vertical dashed line.}
+  \label{fig:sim-non-normal}
+\end{figure}
+
+\subsubsection*{Type I error}
+
+As suggested by theory (see \hyperref[sec:methods]{Methods} for details) and shown by the simulations with a wide range of settings, both of the proposed tests provide a valid Type I error control (Fig.
+\ref{fig:sim-normal}), even in case of non-linearity and non-normality (Figs.
+\ref{fig:sim-h0-demo} and \ref{fig:sim-non-normal}), except when non-normality is very extreme (purple distribution on Fig.
+\ref{fig:sim-non-normal}, kurtosis: 42, skewness: -6).
+
+\subsubsection*{Power}
+
+Estimates of statistical power for the partial confounder test (with normal and linear simulations, for  a wide range of parameters) are shown on Figure \ref{fig:sim-normal}.
+Notably, with sample sizes as large as 1000, a confounder contributing to only ~4\% to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%.
+With a sample size of 500, the same confounder bias is still detected with a power greater than 84\% in all of the simulation cases.
+A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) for a robust detection, with a power of 75-98\%.
+Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%.
+If the confounder explains more than 25\% of variance, it is almost certainly detected even with the low sample size of $n \geq 50$.
+
+Simulations with non-normal and non-linear dependencies has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}).
+Type I error and power was found to be highly similar in case of categorical variables and for the \emph{full} confounder test, as well (Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-lin-full} and \ref{fig:sim-bbb-sig-full}).
+
+\subsection{Neuroimaging data}
+
+To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, they have been deployed in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.
+
+\subsubsection*{HCP dataset}
+
+\begin{figure}[!b]
+  \centering
+  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
+  \caption{\textbf{Acquisition- and age-bias of fluid intelligence prediction in the HCP dataset.} \\
+  Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies.
+Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data.
+This bias is robustly detected by the partial confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
+  Relation between the observed ($Gf$) and predicted ($\hat{y}$) IQ scores and the confounder variables is given on the graphs via $R^2 values$.
+Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
+  Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
+P-values are determined with the partial confound test.
+P-values of the 'full' confound test were all less then 0.0001, i.e.
+the confounders did not fully drive prediction for any of the models.
+  }
+  \label{fig:hcp}
+\end{figure}
+
+Functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
+
+Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $p=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}).
+The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
+The partial confounder test revealed that the model was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig.
+\ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.
+
+After applying confound mitigation approaches (feature regression or COMBAT) the partial confounder test did not provide evidence of confounder bias anymore ($p > 0.05$ for all; shown in the second and third columns of Fig.
+\ref{fig:hcp}), neither for acquisition batch nor for age.
+Both feature regression and COMBAT increased the predictive performance in the case of both confounders, with COMBAT providing best performances ($R^2=0.122$ and $0.121$ when applied to remove the effect of acquisition and age, respectively).
+
+The full confounder test was highly significant ($p<0.0001$) for all models, indicating that neither of the models were exclusively driven by the confounds.
+
+\subsubsection*{ABIDE dataset}
+
+\begin{figure}[!b]
+  \centering
+  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.png}
+  \caption{\textbf{Center- and motion age-bias of ASD diagnosis prediction in the ABIDE dataset.} \\
+  Boxplots and points show the predicted class probabilities (0: HC, 1: ASD), separately for the HC and ASD groups.
+In the top panel, predictions are plotted for each center separately and color indicates the true diagnosis (DX).
+AT the bottom plot, color indicates the normalized index of in-scanner motion (based on mean framewise displacement, mean FD).
+The proposed confounder test reveals significant center- and motion-bias in the model trained on the raw data (p<0.0001).
+While center bias seems to be effectively mitigated by feature regression, the same technique fails to remove motion-bias (actually introduces a paradoxical negative bias).
+COMBAT seems to effectively remove the effect of both confounders on the predictions and improves the predictive performance of the models in both cases.
+  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$.
+Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
+P-values are determined with the partial confound test.
+P-values of the full confound test were all less then 0.0001, i.e.
+confounder did not fully drive prediction for any of the models.
+  }
+  \label{fig:abide}
+\end{figure}
+
+Functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.
+\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
+
+Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.019$ and $0.028$, respectively, $p<0.001$ for both, see also Table \ref{tab:unconditional-pvals}).
+The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).
+The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$, see first column on Fig.
+\ref{fig:abide}).
+Several sites (e.g.
+Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately associated ($R^2=0.169$ and $0.111$, respectively) to a lower probability for the participant to be assigned to the ASD group by the model.
+
+Both feature regression and COMBAT seemed to significantly attenuate the effect of imaging center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.009$ for feature regression and COMBAT, respectively (second and third columns of the first row on Fig \ref{fig:abide}).
+
+When trying to mitigate the effect of in-scanner motion (bottom row on Fig \ref{fig:abide}), feature regression failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates.
+The partial confounder test successfully detected the resulting strong bias ($p<0.0001$).
+Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$, bottom right panel of Fig \ref{fig:abide}).
+
+Both feature regression and COMBAT considerably improved the predictive performance when mitigating center-effects ($R^2=0.111$ and $0.132$ and $AUC=0.76$ and $0.75$, respectively).
+With both feature regression and COMBAT, however, the effort of mitigating motion effects happened at the cost of a drop in predictive performance ($R^2=0.02$ and $0.111$ and $AUC=0.59$ and $0.70$, respectively)
+
+The full confounder test was strongly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$), indicating that this model might have been (almost) exclusively driven by motion artifacts.
+
+% tab:unconditional-pvals
+\renewcommand{\arraystretch}{1.2}
+\begin{table}
+\centering
+\begin{tabular}{lll|ll|ll|ll|ll} 
+dataset & conf. & method & $R^2_{y, c}$ & $p_{y, c}$ & $R^2_{\hat{y}, c}$ & $p_{\hat{y}, c}$ & $R^2_{\hat{y}, y}$ & $p_{\hat{y}, y}$ & partial test & full test  \\
+\hline
+HCP & acq.  & raw      & 0.032 & <0.001 & 0.071 & <0.001  & 0.095 & <0.001 & \textbf{<0.0001} & <0.0001 \\
+    &              & f.reg.    & & & 0.011 & 0.477 & 0.105  & <0.001  & 0.73 & <0.0001 \\
+    &              & COMBAT    & & &0.013 & 0.4 & 0.122  & <0.001  & 0.65 & <0.0001\\
+\hline
+    & age   & raw       & 0.011 & 0.001  & 0.034 & <0.001  & 0.095  & <0.001 & \textbf{<0.0001} & <0.0001 \\
+    &       & f.reg.    & & &0.004 & 0.052 & 0.114  & <0.001  & 0.2 & <0.0001 \\
+    &       & COMBAT    & & &0.005 & 0.048 & 0.121 & <0.001 & 0.16 & <0.0001 \\
+\hline
+ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & \textbf{<0.0001} & <0.0001 \\
+        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & \textbf{0.04} & <0.0001 \\
+        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & \textbf{0.009} & <0.0001 \\
+\hline
+        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & \textbf{<0.0001} & <0.0001 \\
+        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & \textbf{<0.0001} & 0.1 \\
+        &          & COMBAT    & & &  0.002 & 0.19 & 0.111     & <0.001 & 0.59 & <0.0001 \\
+
+\end{tabular}
+\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated confounder-mitigation methods.
+Significant confoudner bias is denoted by bold font.}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Discussion}
+
+The assessment of confoudner bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings.
+However, statistical testing of conditional independence is challenging: analogs of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity \citep{korn1984ranges} (see an example on Fig.
+\ref{fig:sim-h0-demo}).
+While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the "no free lunch theorem" (i.e.
+the nonexistence of a uniformly valid conditional independence test) \cite{shah2020hardness} clearly implies that conditional independence-based confoudner bias testing must be designed so that its suitability for the particular problem may be judged easily.
+
+In this paper, I have proposed a solution for confounder testing, which was designed to be suitable for the specific characteristics of the 'confoudner problem' in predictive modelling.
+Namely, the proposed approach places assumptions only about the conditional distribution of the confoudner and the target variables, as required by the applied generalized additive \cite{hastie1987generalized} or multinomial logistic model \cite{bennett1966multiple, jones1975proability}.
+In biomedical research practice, these are variables that are often subject to descriptive (and often parametric) statistical analysis, as well.
+Accordingly, usually there is sufficient expert knowledge about the applicability of the required assumptions and data acquisition is probably already designed so that the assumptions are likely fulfilled.
+Of key importance, the proposed solution is \emph{absolutely assumption-free} regarding the distribution and dependency of the prediction output, which is, in most of the cases, unknown and can often be expected to be non-normal and/or non-linear (due to e.g model regularization) \citep{garcia2009study, kristensen2017whole}.
+This property distinguishes the approach from other alternatives as it guarantees a valid type-I error control even in cases of non-normally and non-linearly dependent predictions, i.e.
+in cases where Pearson and Spearman partial correlations, and many other methods fail.
+
+The proposed approach gives rise to two different statistical tests for testing confoudner bias: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
+The tests can be applied for arbitrary classification or regression models, without having to re-fit the model, that is, with a negligible extra computational cost.
+
+As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except extreme non-normality), confirming the tests' can be deployed in most predictive modelling scenarios.
+While different biomedical applications might consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculation in order to identify the necessary sample size for proper model diagnostics.
+Validity, power and robustness renders the proposed tests as unprecedented tools for model diagnostics in a wide spectrum of predictive modelling scenarios.
+
+%In biomedical research, confounder-bias has previously often been investigated (if investigated at all) simply by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.
+\citep{spisak2020pain}.
+However, the significance of the unconditioned association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable.
+In this case, namely, the model still might not be directly driven by the confounder and the dependence of the predictions on the confounder might be only marginal, to be explained solely by the confounder-target association.
+%Two examples for this situation can be found in the presented analysis of real data; when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data (see Table \ref{tab:unconditional-pvals}).
+%These examples clearly indicate that guiding model diagnostics based on unconditioned associations between the confounder and the model output might result in overly strict decisions.
+More realistic is to aim at models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e.
+the prediction-confounder association is proportional to, and can be fully explained by, the target-confounder association.
+%Put simply, this is what is tested by the proposed partial confounder test.
+
+%Obviously, a true confounder bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder association, i.e.
+the model might still use the confounder's effect on the features (e.g artifacts) to construct the corresponding part of the variance in the prediction.
+Nevertheless, in most of the cases, it does not seem likely that the model, at the same time, (i) misses that specific "part" of the target variable's variance in the features that is shared with the confounder (ii) captures exclusively that component of the confounder's variance in the features that is needed to mimic an unbiased model.
+%Usually a much more plausible explanation is that a direct confounder effect in the model output is simply negligible (e.g because it is not to be found in the features after confound mitigation).
+
+A characteristic example for the potential areas of applications is the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making \citep{woo2017building, wager2013fmri, spisak2020pain}.
+However, the presence of confounders typical for biomedical research (e.g.
+sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g.
+imaging artifacts) presents a great challenge to these efforts.
+The usefulness of the proposed tests is demonstrated in two such examples, using the HCP \citep{van2013wu}  and the ABIDE \citep{di2014autism} datasets.
+%First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
+$Second, functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.
+\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
+The hypothesised confounder effects were clearly present in both datasets.
+
+In case of the HCP dataset, the weak but statistically significant confounder effect of age on fluid intelligence is in line with previous findings \citep{dubois2018distributed, lohmann2021predicting}) could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample.
+In this case, the bias would likely significantly harm the out-of-sample generalizability of this model.
+The bias of the same model to 'acquisition batch' (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset.
+There can be manifold reasons for the observed bias.
+Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g.
+as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed over time).
+To explain the associated change in the target variable, the model might utilise changing characteristics of the imaging data (caused by e.g.
+slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff).
+This is obviously undesirable for any brain-based predictive model of fluid intelligence.
+
+In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD.
+For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g.
+the CMU (Carnegie Mellon University) center as neurotypical control participants.
+Similarly, the models biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with a tendency for higher in-scanner motion, e.g.
+due to atypical motor function (as known e.g.
+for ADHD \citep{eloyan2012automated}).
+
+ The partial confounder test provided quantitative, statistically rigorous metrics for assessing the effectiveness of the investigated confounder mitigation techniques.
+In the HCP data, it revealed that both the acquisition bias and the age bias was very effectively removed by both feature regression and COMBAT (p>0.05 for all).
+Given the high power of the test at the sample sizes of the HCP dataset ($N=999$), any remaining confounder-bias is most probably very safely negligible and well out of the range of practical relevance.
+
+On the other hand, the partial confoudner test also showed, that the performance of the investigated confound mitigation techniques is much less convincing for the ABIDE dataset.
+The center bias of the classification in this massively multi-center dataset was, namely, found to be significant, even after feature regression and COMBAT.
+Nevertheless, the p-values returned by the partial confounder test increased, in both cases, with two orders of magnitude as compared to the case with raw features, suggesting that a large part of center effects was still successfully removed by both approaches.
+Although determining the relevance of the remaining bias is out of the scope of this paper, significant p-values of the partial confounder test generally suggest that these models require more effective data harmonization.
+
+Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
+Interestingly, in the latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign.
+Meanwhile, the prediction performance substantially dropped (although still remained statistically significant).
+A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
+This effect might be boosted by the previously reported "residual motion artifacts" and their regional interactions \citep{spisak2014voxel} and raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
+ As COMBAT was originally developed for harmonizing effects of categorical variables (e.g.
+center or batch), its application for continuous confounder variables is not trivial.
+The presented method of inputting the binned "motion groups" into COMBAT resulted in an efficient removal of motion-bias and, at the same time, preserved a significant amount of signal-of-interest.
+ Nevertheless, inputting discretized versions of continuous variables into COMBAT might be sub-optimal and raises further questions e.g regarding the optimal number of bins used during the discretization.
+
+ In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounder-bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT and (iii) in some cases confoudner mitigation can - paradoxically - introduce more bias.
+The partial confounder test can be considered as a useful, objective benchmark to guide the search for a suitable confounder mitigation approach for every dataset.
+ Regarding the \emph{full} confounder test; while the real data examples in this paper were not typical cases of its potential applications, it has still provided important insights into model biases.
+Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder ($p=0.1$).
+In other words, the test accepted the null hypothesis of full model bias, suggesting that this model might be, in fact, almost exclusively driven by motion artifacts.
+This example highlights that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased by various confounders and the questions is whether there is any biomedically relevant signal captured by the model, at all.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+
+The lack of rigorous statistical tests for confounder-bias significantly hampers the development of predictive models in many fields of research.
+Although several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder signal and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias.
+
+To fill this critical gap in predictive model development, here I proposed two novel tests, the \emph{partial} and the \emph{full confounder tests}, which probe the null hypotheses of 'no confounder-bias' and 'full confounder-bias', respectively.
+Both approaches are based on a recently published framework for conditional permutation testing, with a mathematical justification of valid type I error control and with assumptions that are easy to meet in most predictive modelling scenarios.
+The tests have a minimal computational overhead; as re-fitting the model is not required.
+
+As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and allow quantitative, statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research.
+Given their simplicity, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confoudner tests can become widespread tools that pave the way for the development of clinically useful machine learning biomarkers.
+
+\section{Acknowledgements}
+I am thankful to Ulrike Bingel (University Hospital Essen, Germany) and Robert Englert (University Hospital Essen, Germany) for their valuable insights and comments on the manuscript.
+This research was supported by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – Projektnummer 316803389 – SFB 1280  and TRR 289 Treatment Expectation - Projektnummer 422744262.
 
@@
 \section{Methods}
+\label{sec:methods}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% fig: overview
-\begin{figure}[!b]
-  \centering
-  \includegraphics[width=0.75\paperwidth]{fig/overview.eps}
-  \caption{\textbf{Overview of the proposed partial confoudner test}. \\
-    The partial confoudner test models the conditional distribution of the confounder, given the target variable, with a generalized additive model (GAM).
-The parallel-pairwise Markov-chain Monte-Carlo sampler draws permutations of the confounder variable that comply with the GAM-based conditional distribution (permutation 1, 2, ..., m).
-The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted confoudner variables.
-The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original.
-Figure source code available as jupyter notebook: \href{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}{https://github.com/pni-lab/mlconfound-manuscript/blob/main/simulated/overview-fig.ipynb}
-  }
-  \label{fig:overview}
-\end{figure}
-
-\subsection{The partial and full confoudner tests}
-
-The proposed \emph{partial} and \emph{full confoudner tests} for predictive modelling have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.
-The inner workings of the tests is summarized on Fig.
-\ref{fig:overview}.
-In short, the tests model the conditional distribution between the confounder and the target variable with a generalized additive model (GAM) and then use a parallel-pairwise Markov-chain Monte-Carlo sampler that draws permutations of the confounder or target variable, for the partial and full confounder test, respectively.
-As a result, the permuted variables comply with the GAM-based conditional distribution.
-The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted variables.
-The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original.
-The strength of this approach lies in avoiding to place any assumptions on the conditional distributions of the model output.
-This property ensures valid model diagnostics even in cases where Pearson and Spearman partial correlation and many other methods fail (as shown on Fig.
-\ref{fig:sim-h0-demo}).
-
-\subsection{Simulations}
-
-\subsubsection*{Type I error}
-
-As suggested by theory and shown by the simulations, both of the proposed tests provide a valid Type I error control (Fig.
-\ref{fig:sim-normal}), even in case of non-linearity and non-normality (Figs.
-\ref{fig:sim-h0-demo} and \ref{fig:sim-non-normal}), except when non-normality is very extreme (purple distribution on Fig.
-\ref{fig:sim-non-normal}, kurtosis: 42, skewness: -6).
-With this the proposed tests overcome the limitations of partial Spearman correlation and linear conditional permutations, demonstrated on Fig.
-\ref{fig:sim-h0-demo}.
-
-For detailed results, see also Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-lin-full} and \ref{fig:sim-bbb-sig-full}.
-
-%fig:sim-h0-demo
-\begin{figure}[!b]
-  \centering
-  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
-  \caption{\textbf{Type I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
-  Type I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), slightly non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row).
-Non-normal conditional distribution on the second plot is illustrated with blue density diagrams (kurtosis: -0.8, skewness: -0.1) and dependency of the mean is depicted with blue dotted lines an all plots.
-False positive rates for the three investigated tests is shown in heatmaps, with darker red meaning higher values.
-The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065.
-Values below this threshold indicate a valid type I error control and colored white.
-  }
-  \label{fig:sim-h0-demo}
-\end{figure}
-
-%fig:sim-normal
-\begin{figure}[!b]
-  \centering
-  \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
-  \caption{\textbf{Type I error control and power of the partial confoudner test based on simulations with normal conditional distribution and linear dependencies.} \\
-  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette on the right) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\hat{y}}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\hat{y}}$ (rows) and for different sample sizes (N, columns).
-Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1.0 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%.
-First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounder bias).
-  Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively.
-The higher 95\% confidence limit for a positive rate of $alpha=0.05$ is 0.11 for each tile.
-  }
-  \label{fig:sim-normal}
-\end{figure}
-
-%fig:fig:sim-non-normal
-\begin{figure}[!b]
-  \centering
-
-  \includegraphics[width=0.5\paperwidth]{fig/sim_non-norm.eps}
-  \caption{\textbf{Robustness of conditional permutation based confound testing to non-normality and non-linearity.} \\
-  Simulations included variables with five different degrees of non-normality (top panel), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation (yellow: normally distributed).
-Fisher's kurtosis and skewness is given for each distribution.
-False and true positive rates in the simulations under H0 and H1, respectively, for each investigated sample size (N), are depicted by barplots for both linear and sigmoid dependency structure.
-Upper 95\% binomial confidence limit corresponding to $alpha=0.05$ is shown with a vertical dashed line.}
-  \label{fig:sim-non-normal}
-\end{figure}
-
-\subsubsection*{Power}
-
-Power estimates for the 'partial' test (with normal and linear simulations) are shown on Figure \ref{fig:sim-normal}.
-Notably, with sample sizes as large as 1000, a confounder contributing to only ~4\% to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%.
-With a sample size of 500, the same confounder bias is still detected with a power greater than 84\% in all of the simulation cases.
-A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) for a robust detection, with a power of 75-98\%.
-Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%.
-If the confounder explains more than 25\% of variance, it is almost certainly detected even with the low sample size of $n \geq 50$.
-
-Simulations with non-normal and non-linear dependencies has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}).
-
-\subsection{Neuroimaging data}
-\subsubsection*{HCP dataset}
-
-\begin{figure}[!b]
-  \centering
-  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
-  \caption{\textbf{Acquisition- and age-bias of fluid intelligence prediction in the HCP dataset.} \\
-  Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies.
-Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data.
-This bias is robustly detected by the partial confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
-  Relation between the observed ($Gf$) and predicted ($\hat{y}$) IQ scores and the confounder variables is given on the graphs via $R^2 values$.
-Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
-  Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
-P-values are determined with the partial confound test.
-P-values of the 'full' confound test were all less then 0.0001, i.e.
-the confounders did not fully drive prediction for any of the models.
-  }
-  \label{fig:hcp}
-\end{figure}
-
-Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $p=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}).
-The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
-The partial confounder test revealed that the model was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig.
-\ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.
-
-After applying confound mitigation approaches (feature regression or COMBAT) the partial confounder test did not provide evidence of confounder bias anymore ($p > 0.05$ for all; shown in the second and third columns of Fig.
-\ref{fig:hcp}), neither for acquisition batch nor for age.
-Both feature regression and COMBAT increased the predictive performance in the case of both confounders, with COMBAT providing best performances ($R^2=0.122$ and $0.121$ when applied to remove the effect of acquisition and age, respectively).
-
-The full confounder test was highly significant ($p<0.0001$) for all models, indicating that neither of the models were exclusively driven by the confounds.
-
-\subsubsection*{ABIDE dataset}
-
-\begin{figure}[!b]
-  \centering
-  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.png}
-  \caption{\textbf{Center- and motion age-bias of ASD diagnosis prediction in the ABIDE dataset.} \\
-  Boxplots and points show the predicted class probabilities (0: HC, 1: ASD), separately for the HC and ASD groups.
-In the top panel, predictions are plotted for each center separately and color indicates the true diagnosis (DX).
-AT the bottom plot, color indicates the normalized index of in-scanner motion (based on mean framewise displacement, mean FD).
-The proposed confounder test reveals significant center- and motion-bias in the model trained on the raw data (p<0.0001).
-While center bias seems to be effectively mitigated by feature regression, the same technique fails to remove motion-bias (actually introduces a paradoxical negative bias).
-COMBAT seems to effectively remove the effect of both confounders on the predictions and improves the predictive performance of the models in both cases.
-  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$.
-Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
-P-values are determined with the partial confound test.
-P-values of the full confound test were all less then 0.0001, i.e.
-confounder did not fully drive prediction for any of the models.
-  }
-  \label{fig:abide}
-\end{figure}
-
-Imaging center and in-scanner motion (normalized mean framewise displacement) were statistically significantly associated with ASD diagnosis ($R^2=0.019$ and $0.028$, respectively, $p<0.001$ for both, see also Table \ref{tab:unconditional-pvals}).
-The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).
-The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$, see first column on Fig.
-\ref{fig:abide}).
-Several sites (e.g.
-Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately associated ($R^2=0.169$ and $0.111$, respectively) to a lower probability for the participant to be assigned to the ASD group by the model.
-
-Both feature regression and COMBAT seemed to significantly attenuate the effect of imaging center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.009$ for feature regression and COMBAT, respectively (second and third columns of the first row on Fig \ref{fig:abide}).
-
-When trying to mitigate the effect of in-scanner motion (bottom row on Fig \ref{fig:abide}), feature regression failed to remove the motion-bias from the model and, in fact, it introduced a paradoxical inverse dependence of the predictions on motion estimates.
-The partial confounder test successfully detected the resulting strong bias ($p<0.0001$).
-Applying COMBAT to remove the effect of motion (by using binned motion estimates), in turn, seemed to effectively mitigate motion-bias, as suggested by the partial confound test ($p=0.64$, bottom right panel of Fig \ref{fig:abide}).
-
-Both feature regression and COMBAT considerably improved the predictive performance when mitigating center-effects ($R^2=0.111$ and $0.132$ and $AUC=0.76$ and $0.75$, respectively).
-With both feature regression and COMBAT, however, the effort of mitigating motion effects happened at the cost of a drop in predictive performance ($R^2=0.02$ and $0.111$ and $AUC=0.59$ and $0.70$, respectively)
-
-The full confounder test was strongly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$), indicating that this model might have been (almost) exclusively driven by motion artifacts.
-
-% tab:unconditional-pvals
-\renewcommand{\arraystretch}{1.2}
-\begin{table}
-\centering
-\begin{tabular}{lll|ll|ll|ll|ll} 
-dataset & conf. & method & $R^2_{y, c}$ & $p_{y, c}$ & $R^2_{\hat{y}, c}$ & $p_{\hat{y}, c}$ & $R^2_{\hat{y}, y}$ & $p_{\hat{y}, y}$ & partial test & full test  \\
-\hline
-HCP & acq.  & raw      & 0.032 & <0.001 & 0.071 & <0.001  & 0.095 & <0.001 & \textbf{<0.0001} & <0.0001 \\
-    &              & f.reg.    & & & 0.011 & 0.477 & 0.105  & <0.001  & 0.73 & <0.0001 \\
-    &              & COMBAT    & & &0.013 & 0.4 & 0.122  & <0.001  & 0.65 & <0.0001\\
-\hline
-    & age   & raw       & 0.011 & 0.001  & 0.034 & <0.001  & 0.095  & <0.001 & \textbf{<0.0001} & <0.0001 \\
-    &       & f.reg.    & & &0.004 & 0.052 & 0.114  & <0.001  & 0.2 & <0.0001 \\
-    &       & COMBAT    & & &0.005 & 0.048 & 0.121 & <0.001 & 0.16 & <0.0001 \\
-\hline
-ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & \textbf{<0.0001} & <0.0001 \\
-        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & \textbf{0.04} & <0.0001 \\
-        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & \textbf{0.009} & <0.0001 \\
-\hline
-        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & \textbf{<0.0001} & <0.0001 \\
-        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & \textbf{<0.0001} & 0.1 \\
-        &          & COMBAT    & & &  0.002 & 0.19 & 0.111     & <0.001 & 0.59 & <0.0001 \\
-
-\end{tabular}
-\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated confounder-mitigation methods.
-Significant confoudner bias is denoted by bold font.}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
-
-Confounder-bias can significantly harm the generalizability and biomedical validity of predictive models.
-The problem of testing confounder-bias in predictive models can considered as assessing \emph{conditional independence} between the predicted values, the the target and the confounder variable.
-Importantly, in predictive modelling, the model predictions are often characterized by a non-normal and/or non-linear conditional dependence on the target variable, thus any test for confounder bias must be, at least partly, non-parametric.
-However, as shown by \citet{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
-In practice, this means that all candidate tests, that assume normality and/or linearity of the joint distribution, will necessarily fail to properly control for Type I errors in many realistic predictive modelling scenarios.
-This is clearly demonstrated on Fig.
-\ref{fig:sim-h0-demo} in case of partial Spearman correlation.
-Although warnings about these issues were given from early on \citep{korn1984ranges} and received a fair amount of attention recently \citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated within the predictive modelling community.
-
-From the three involved variables, the most problematic one is clearly the prediction output, known to often display non-normal and non-linear conditional dependencies on the other two.
-Therefore, a method being distribution-free for this variable, but inputting custom distributional information about the other two variables might provide a sufficient flexibility required for predictive model diagnostics.
-
-To achieve this, here I propose to combine the conditional permutation test (CPT) of \cite{berrett2020conditional} with generalized additive model (GAM)-based conditional distribution modelling \citep{hastie1987generalized}.
-The proposed mathematical framework offers two novel tests for probing confounder bias: the partial and full confounder tests.
-These tests can handle both numerical and categorical data and can be applied for any predictive model, without having to re-fit the model, that is, with a negligible extra computational cost.
-
-As expected from theory, both tests displayed a valid type I error control in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable.
-Even though the tests' valid type I error control is justified by theory and reinforced by the simulations, a low statistical power could significantly limit their usefulness.
-The simulations presented in this paper revealed that both tests have a high, practically relevant power in a wide variety of scenarios.
-Considering the partial test, if we apply \citeauthor{cohen2013statistical}'s terminology for the effect size of the confounder bias, sample sizes around 500-1000 can robustly identify a very weak confounder bias (e.g a confounder explaining about 4\% of the variance in the prediction).
-A lower sample size of 100 will still very likely identify medium sized confounder-biases (around 12\%).
-Strong biases  (>25\%) can be very robustly detected with sample sizes as low as 50.
-As expected, the power of the full confounder test is very similar to that of the partial confounder test, since the only difference is that the conditional contribution of the target to the prediction is investigated, instead that of the confounder.
-While different biomedical applications might consider different amounts of bias to be relevant, an a-priori power calculation can help identifying the necessary sample size for proper model diagnostics.
-
-Simulations also confirmed that non-normality and non-linearity of the conditional distribution of the predictions and the confounder (given the target) has a negligible effect on the power of both tests, except in case of extreme violations of normality.
-This result confirms that the proposed tests, unlike e.g.
-partial correlation, remain valid, independent of the distributional characteristics of the applied model's output.
-Validity, power and robustness renders the proposed tests as unprecedented tools for model diagnostics in a wide spectrum of predictive modelling scenarios.
-
-In biomedical research, confounder-bias has previously often been investigated (if investigated at all) simply by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.
-\citep{spisak2020pain}.
-However, the significance of the unconditioned association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable.
-In this case, namely, the model still might not be directly driven by the confounder and the dependence of the predictions on the confounder might be only marginal, to be explained solely by the confounder-target association.
-Two examples for this situation can be found in the presented analysis of real data; when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data (see Table \ref{tab:unconditional-pvals}).
-These examples clearly indicate that guiding model diagnostics based on unconditioned associations between the confounder and the model output might result in overly strict decisions.
-More realistic is to aim at models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e.
-the prediction-confounder association is proportional to, and can be fully explained by, the target-confounder association.
-Put simply, this is what is tested by the proposed partial confounder test.
-
-%Obviously, a true confounder bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder association, i.e.
-the model might still use the confounder's effect on the features (e.g artifacts) to construct the corresponding part of the variance in the prediction.
-Nevertheless, in most of the cases, it does not seem likely that the model, at the same time, (i) misses that specific "part" of the target variable's variance in the features that is shared with the confounder (ii) captures exclusively that component of the confounder's variance in the features that is needed to mimic an unbiased model.
-%Usually a much more plausible explanation is that a direct confounder effect in the model output is simply negligible (e.g because it is not to be found in the features after confound mitigation).
-
-Among the many potential areas of applications, the partial and full confounder tests could be particularly useful in translational neuroscience.
-A characteristic example is provided by the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making \citep{woo2017building, wager2013fmri, spisak2020pain}.
-However, the presence of confounders typical for biomedical research (e.g.
-sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g.
-imaging artifacts) presents a great challenge to these efforts and is, currently, often largely neglected.
-To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, here they have been deployed in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.
-First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
-Second, functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.
-\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
-The hypothesised confounder effects were clearly present in both datasets.
-
-In case of the HCP dataset, the weak but statistically significant association between age and IQ predictions could likely exaggerate to a serious bias when testing the model on data of participants outside of the (relatively narrow)
-age range of the HCP sample.
-In this case, the bias would likely significantly harm the out-of-sample generalizability of this model.
-The bias of the same model to 'acquisition batch' (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset.
-There can be manifold reasons for the observed association.
-Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g.
-as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed over time).
-To explain the associated change in the target variable, the model might utilise changing characteristics of the imaging data (caused by e.g.
-slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff).
-This is obviously undesirable for any brain-based predictive model of fluid intelligence.
-
-In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD.
-For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g.
-the CMU (Carnegie Mellon University) center as neurotypical control participants.
-It is easy to see that this bias can lead to calibration issues and, possibly, a very low predictive performance on out-of-sample data.
-Similarly, a model biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with a tendency for higher in-scanner motion, e.g.
-due to atypical motor function (as known e.g.
-for ADHD \citep{eloyan2012automated}).
-
-Confounder mitigation techniques, with one exception, tended to decrease confounder bias.
-The partial confounder test provided quantitative, statistically rigorous metrics for assessing the effectiveness of the techniques for each confounder.
-In the HCP data, it revealed that both the acquisition-bias and the age-bias was very effectively removed by both feature regression and COMBAT (p>0.05 for all).
-Given the high power of the test at the sample sizes of the HCP dataset ($N=999$), any remaining confounder-bias is most probably very safely negligible and well out of the range of practical relevance.
-
-On the other hand, the partial confoudner test also showed, that the performance of the investigated confound mitigation techniques is much less convincing for the ABIDE dataset.
-The center-bias of the classification in this massively multi-center dataset was, namely, found to be significant, even after feature regression and COMBAT.
-Nevertheless, the p-values returned by the partial confounder test were larger with two orders of magnitude than without any confound mitigation, suggesting that a large part of the center effects was still successfully removed by both approaches.
-While determining the relevance of the remaining bias is to be performed based on the experimental context and the requirements towards the predictive marker, significant p-values of the partial confounder test generally suggest that these models would either require more effective data harmonization or a validation on external data, that confirms the low practical relevance of the amount of bias.
-
-Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
-Interestingly, in the latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign.
-Meanwhile, the prediction performance substantially dropped (although still remained statistically significant).
-A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
-On the other hand, motion artifacts were previously shown to display complex spatiotemporal patterns in fMRI-based functional connectivity estimates \citep{spisak2014voxel} and simply regressing them out of the features is far from being sufficient for removing their effect on the model.
-The so-called "residual motion artifacts" and their regional interactions, as described in the referenced work, might be responsible for the paradox behavior of this model, which raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
-
- As COMBAT was originally developed for harmonizing effects of categorical variables (e.g.
-center or batch), its application for continuous confounder variables is not trivial.
-The presented method of inputting the binned "motion groups" into COMBAT resulted in an efficient removal of motion-bias and, at the same time, preserved a significant amount of signal-of-interest.
- Nevertheless, inputting discretized versions of continuous variables into COMBAT might be sub-optimal and raises further questions e.g regarding the optimal number of bins used during the discretization.
-
- In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounder-bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT.
-The partial confounder test can be considered as a useful objective benchmark to guide the search for a suitable confounder mitigation approach.
- Regarding the \emph{full} confounder test; while the real data examples in this paper were not typical cases of its potential applications, it has still provided important insights into model biases.
-Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder ($p=0.1$).
-In other words, the test accepted the null hypothesis of full model bias, suggesting that this model might be, in fact, almost exclusively driven by motion artifacts.
-This example highlight that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased by various confounders and the questions is whether there is any biomedically relevant signal captured by the model, at all.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-
-The lack of rigorous statistical tests for confounder-bias significantly hampers the development of predictive models in many fields of research.
-Although several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder signal and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias.
-
-To fill this critical gap in predictive model development, here I proposed two novel tests, the partial and the full confounder tests, which probe the null hypotheses of 'no confounder-bias' and 'full confounder-bias', respectively.
-Both approaches are based on a recently published framework for conditional permutation testing, with a mathematical justification of valid type I error control.
-The tests were constructed so that they do not place any assumption on the conditional distribution of the model predictions and can be performed with a minimal computational overhead; without needing to re-fit the model.
-
-Given their simplicity, high statistical power, robustness to the violation of the assumptions and thanks to the computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the tests are deployable in a wide variety of applications.
-
-As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and allow quantitative, statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research, thereby paving the way for the development of clinically useful machine learning biomarkers.
-
-\section{Acknowledgements}
-I am thankful to Ulrike Bingel (University Hospital Essen, Germany) and Robert Englert (University Hospital Essen, Germany) for their valuable insights and comments on the manuscript.
-This research was supported by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – Projektnummer 316803389 – SFB 1280  and TRR 289 Treatment Expectation - Projektnummer 422744262.
 
 %TC:ignore 
-\bibliographystyle{apalike}  
+\bibliographystyle{naturemag} %apalike}  
 \bibliography{references}

### 2021-10-28 manuscript.tex
fix
Updates from Overleaf

@@
 
-Predictive modelling has recently become increasingly important in biomedical research and holds promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized, spisak2020pain, walsh2021dome}.
+Predictive modelling has recently become increasingly important in biomedical research and holds promise for delivering biomarkers that substantially impact clinical practice and public health\citep{kent2018personalized, spisak2020pain, walsh2021dome}.
 When evaluating the usefulness and applicability of such markers, predictive performance (i.e.
@@
 Biomedical validity (i.e.
-whether the model is driven by biomedically relevant signal) as well as generalizability and fairness across contexts and populations are crucial requirements for candidate biomarkers \citep{woo2017building, obermeyer2019dissecting, mehrabi2021survey}.
+whether the model is driven by biomedically relevant signal) as well as generalizability and fairness across contexts and populations are crucial requirements for candidate biomarkers\citep{woo2017building, obermeyer2019dissecting, mehrabi2021survey}.
 
-Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - i.e the so-called confounder and collider biases \citep{prosperi2020causal} - can be detrimental to the model's biomedical validity and generalizability.
-Examples include models picking up on measurement artifacts (in-scanner head motion artifacts in magnetic resonance imaging-based predictive models of Alzheimer's \citep{rao2017predictive}, attention deficit hyperactivity disorder \citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD) \citep{gotts2013perils, spisak2014voxel, spisak2019optimal}), demographic and psychometric variables (e.g models trained to predict intelligence \citep{ cole2012global, he2020deep} might provide a statistically significant predictive performance by picking up solely on age-related variance \citep{dubois2018distributed, lohmann2021predicting}), sampling bias and stochastic group differences (racially biased machine learning models \citep{ obermeyer2019dissecting, lwowski2021risk}), batch effects or, in multi-center studies, center-effects.
+Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - i.e the so-called confounder and collider biases\citep{prosperi2020causal} - can be detrimental to the model's biomedical validity and generalizability.
+Examples include models picking up on measurement artifacts (in-scanner head motion artifacts in magnetic resonance imaging-based predictive models of Alzheimer's\citep{rao2017predictive}, attention deficit hyperactivity disorder\citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD)\citep{gotts2013perils, spisak2014voxel, spisak2019optimal}), demographic and psychometric variables (e.g models trained to predict intelligence\citep{ cole2012global, he2020deep} might provide a statistically significant predictive performance by picking up solely on age-related variance\citep{dubois2018distributed, lohmann2021predicting}), sampling bias and stochastic group differences (racially biased machine learning models\citep{ obermeyer2019dissecting, lwowski2021risk}), batch effects or, in multi-center studies, center-effects.
 
-While various data cleaning methods might help in mitigating confounder bias \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, rao2017predictive, johnson2007adjusting}, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.
+While various data cleaning methods might help in mitigating confounder bias\citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, rao2017predictive, johnson2007adjusting}, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest\citep{wachinger2021detect}.
 
@@
 Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation) are not valid measures of conditional independence.
-Although warnings about this issue were given from early on \citep{korn1984ranges}, and received a fair amount of attention recently \citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g.
-feature-set characteristics and model regularization \citep{garcia2009study, kristensen2017whole}; see Supplementary Material \ref{sup:nomlinviol} for a simplistic example.
+Although warnings about this issue were given from early on\citep{korn1984ranges}, and received a fair amount of attention recently\citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g.
+feature-set characteristics and model regularization\citep{garcia2009study, kristensen2017whole}; see Supplementary Material \ref{sup:nomlinviol} for a simplistic example.
 
-Alternative techniques for quantifying confoudner bias either fail to control type error (as  the known in the case of balanced permutations \cite{southworth2009properties, hemerik2018exact}, used in \cite{chaibub2019permutation}), or do not aim at all to do so \cite{ferrari2020measuring, wachinger2021detect}.
-Moreover, without some modifications, they are applicable for either only numeric \cite{wachinger2021detect} or only categorical variables \citep{chaibub2019permutation, ferrari2020measuring} and involve re-fitting the model \citep{chaibub2019permutation, ferrari2020measuring}, which might not be feasible for models with high a computational cost (e.g.
+Alternative techniques for quantifying confoudner bias either fail to control type error (as  the known in the case of balanced permutations\cite{southworth2009properties, hemerik2018exact}, used in\cite{chaibub2019permutation}), or do not aim at all to do so\cite{ferrari2020measuring, wachinger2021detect}.
+Moreover, without some modifications, they are applicable for either only numeric\cite{wachinger2021detect} or only categorical variables\citep{chaibub2019permutation, ferrari2020measuring} and involve re-fitting the model\citep{chaibub2019permutation, ferrari2020measuring}, which might not be feasible for models with high a computational cost (e.g.
 with nested cross-validation).
-Some techniques \cite{wachinger2021detect}, furthermore, probe a null hypothesis (direct causation between the features and the prediction target) that is not applicable for for many kinds of data and models \footnote{For instance, causation between several psychological or clinical states and functional MRI activity/connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation.
+Some techniques\cite{wachinger2021detect}, furthermore, probe a null hypothesis (direct causation between the features and the prediction target) that is not applicable for for many kinds of data and models \footnote{For instance, causation between several psychological or clinical states and functional MRI activity/connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation.
 In such cases, the alternative hypothesis of this test might be true in absence of any real confounder.
@@
 
-%In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing \citep{dawid1979conditional}.
-This allows relating the problem to the seminal work of \cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem).
-%As an important consequence, methods like partial correlation - and even its non-parametric variety - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (see e.g.
-\citep{korn1984ranges} or Fig.
+%In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing\citep{dawid1979conditional}.
+This allows relating the problem to the seminal work of\cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem).
+%As an important consequence, methods like partial correlation - and even its non-parametric variety - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (see e.g.\citep{korn1984ranges} or Fig.
 \ref{fig:sim-h0-demo} in this paper), as commonly seen in machine learning.
 
-%To overcome these limitations, I extend the permutation-based conditional independence framework of \cite{berrett2020conditional} with a generalized additive model based conditional distribution estimation, which allows avoiding to place any assumptions about the most problematic variable, the prediction output.
+%To overcome these limitations, I extend the permutation-based conditional independence framework of\cite{berrett2020conditional} with a generalized additive model based conditional distribution estimation, which allows avoiding to place any assumptions about the most problematic variable, the prediction output.
 %The framework allows constructing two tests: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
@@
 %The usefulness of the proposed tests in detecting various types of confounder bias is demonstrated in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.
-%First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} is used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{ dubois2018distributed, lohmann2021predicting} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
-%Second, functional connectivity data from the ABIDE \citep{di2014autism} database is used to investigate the potential motion- and center-bias \cite{gotts2013perils, spisak2014voxel, spisak2019optimal} when training models that aim to predict ASD diagnosis.
+%First, functional connectivity data from the Human Connectome Project (HCP)\citep{van2013wu} is used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age\citep{ dubois2018distributed, lohmann2021predicting} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
+%Second, functional connectivity data from the ABIDE\citep{di2014autism} database is used to investigate the potential motion- and center-bias\cite{gotts2013perils, spisak2014voxel, spisak2019optimal} when training models that aim to predict ASD diagnosis.
 
@@
 The idea of conditional independence provides a straightforward framework for assessing the confounder bias in predictive models.
-However, handling the non-normal and/or non-linear conditional dependence often seen in predictive models \citep{garcia2009study, kristensen2017whole} poses a great challenge.
-In fact, as recently shown by Shah and colleagues \cite{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
+However, handling the non-normal and/or non-linear conditional dependence often seen in predictive models\citep{garcia2009study, kristensen2017whole} poses a great challenge.
+In fact, as recently shown by Shah and colleagues\cite{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
 In practice, this means that all candidate tests, that assume normality and/or linearity of the joint distribution, will necessarily fail to properly control for Type I errors in many realistic predictive modelling scenarios.
@@
 
-As the most problematic from the three variables (target, prediction, confounder) is clearly the model prediction \citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics.
-Exactly this can be achieved with the combination of the novel framework of conditional permutation testing (CPT) \cite{berrett2020conditional} and the use of generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}-based conditional distribution estimation.
+As the most problematic from the three variables (target, prediction, confounder) is clearly the model prediction\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics.
+Exactly this can be achieved with the combination of the novel framework of conditional permutation testing (CPT)\cite{berrett2020conditional} and the use of generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}-based conditional distribution estimation.
 The proposed approach offers two novel tests for probing confounder bias: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
@@
 
-Functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
+Functional connectivity data from the Human Connectome Project (HCP)\citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age\citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
 
@@
 
-Functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.
-\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
+Functional connectivity data from the ABIDE\citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
 
@@
 The assessment of confoudner bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings.
-However, statistical testing of conditional independence is challenging: analogs of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity \citep{korn1984ranges} (see an example on Fig.
+However, statistical testing of conditional independence is challenging: analogs of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity\citep{korn1984ranges} (see an example on Fig.
 \ref{fig:sim-h0-demo}).
 While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the "no free lunch theorem" (i.e.
-the nonexistence of a uniformly valid conditional independence test) \cite{shah2020hardness} clearly implies that conditional independence-based confoudner bias testing must be designed so that its suitability for the particular problem may be judged easily.
+the nonexistence of a uniformly valid conditional independence test)\cite{shah2020hardness} clearly implies that conditional independence-based confoudner bias testing must be designed so that its suitability for the particular problem may be judged easily.
 
 In this paper, I have proposed a solution for confounder testing, which was designed to be suitable for the specific characteristics of the 'confoudner problem' in predictive modelling.
-Namely, the proposed approach places assumptions only about the conditional distribution of the confoudner and the target variables, as required by the applied generalized additive \cite{hastie1987generalized} or multinomial logistic model \cite{bennett1966multiple, jones1975proability}.
+Namely, the proposed approach places assumptions only about the conditional distribution of the confoudner and the target variables, as required by the applied generalized additive\cite{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}.
 In biomedical research practice, these are variables that are often subject to descriptive (and often parametric) statistical analysis, as well.
 Accordingly, usually there is sufficient expert knowledge about the applicability of the required assumptions and data acquisition is probably already designed so that the assumptions are likely fulfilled.
-Of key importance, the proposed solution is \emph{absolutely assumption-free} regarding the distribution and dependency of the prediction output, which is, in most of the cases, unknown and can often be expected to be non-normal and/or non-linear (due to e.g model regularization) \citep{garcia2009study, kristensen2017whole}.
+Of key importance, the proposed solution is \emph{absolutely assumption-free} regarding the distribution and dependency of the prediction output, which is, in most of the cases, unknown and can often be expected to be non-normal and/or non-linear (due to e.g model regularization)\citep{garcia2009study, kristensen2017whole}.
 This property distinguishes the approach from other alternatives as it guarantees a valid type-I error control even in cases of non-normally and non-linearly dependent predictions, i.e.
@@
 
-%In biomedical research, confounder-bias has previously often been investigated (if investigated at all) simply by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.
-\citep{spisak2020pain}.
+%In biomedical research, confounder-bias has previously often been investigated (if investigated at all) simply by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.\citep{spisak2020pain}.
 However, the significance of the unconditioned association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable.
@@
 
-A characteristic example for the potential areas of applications is the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making \citep{woo2017building, wager2013fmri, spisak2020pain}.
+A characteristic example for the potential areas of applications is the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making\citep{woo2017building, wager2013fmri, spisak2020pain}.
 However, the presence of confounders typical for biomedical research (e.g.
@@
 imaging artifacts) presents a great challenge to these efforts.
-The usefulness of the proposed tests is demonstrated in two such examples, using the HCP \citep{van2013wu}  and the ABIDE \citep{di2014autism} datasets.
-%First, functional connectivity data from the Human Connectome Project (HCP) \citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age \citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
-$Second, functional connectivity data from the ABIDE \citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.
-\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
-The hypothesised confounder effects were clearly present in both datasets.
+The usefulness of the proposed tests is demonstrated in two such examples, using the HCP\citep{van2013wu}  and the ABIDE\citep{di2014autism} datasets.
+%First, functional connectivity data from the Human Connectome Project (HCP)\citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age\citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
+%Second, functional connectivity data from the ABIDE\citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
+%The hypothesised confounder effects were clearly present in both datasets.
 
-In case of the HCP dataset, the weak but statistically significant confounder effect of age on fluid intelligence is in line with previous findings \citep{dubois2018distributed, lohmann2021predicting}) could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample.
+In case of the HCP dataset, the weak but statistically significant confounder effect of age on fluid intelligence is in line with previous findings\citep{dubois2018distributed, lohmann2021predicting}) could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample.
 In this case, the bias would likely significantly harm the out-of-sample generalizability of this model.
@@
 due to atypical motor function (as known e.g.
-for ADHD \citep{eloyan2012automated}).
+for ADHD\citep{eloyan2012automated}).
 
@@
 Meanwhile, the prediction performance substantially dropped (although still remained statistically significant).
-A possible explanation to these phenomena is that in-scanner motion, as previously described \citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
-This effect might be boosted by the previously reported "residual motion artifacts" and their regional interactions \citep{spisak2014voxel} and raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
+A possible explanation to these phenomena is that in-scanner motion, as previously described\citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
+This effect might be boosted by the previously reported "residual motion artifacts" and their regional interactions\citep{spisak2014voxel} and raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
  As COMBAT was originally developed for harmonizing effects of categorical variables (e.g.
@@
 
-\section{Acknowledgements}
-I am thankful to Ulrike Bingel (University Hospital Essen, Germany) and Robert Englert (University Hospital Essen, Germany) for their valuable insights and comments on the manuscript.
-This research was supported by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – Projektnummer 316803389 – SFB 1280  and TRR 289 Treatment Expectation - Projektnummer 422744262.
-
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
- Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications \citep{spirtes2000causation, fiedler2011mediation, peters2016causal, candes2016panning}.
-Recently, \cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
+ Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications\citep{spirtes2000causation, fiedler2011mediation, peters2016causal, candes2016panning}.
+Recently,\cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
  Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional independence testing is effectively impossible.
@@
 
-Recently, \cite{candes2016panning} and \cite{berrett2020conditional} have demonstrated that valid and powerful conditional independence tests can be constructed with inputting distributional information about only two (out of the three) variables.
+Recently,\cite{candes2016panning} and\cite{berrett2020conditional} have demonstrated that valid and powerful conditional independence tests can be constructed with inputting distributional information about only two (out of the three) variables.
 Specifically, the conditional permutation test (CPT) of Berrett and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable.
@@
 
-As many related papers, the work of \cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is a set of predictors (features), $Y$ is the target variable and $Z$ is a potential confounder.
+As many related papers, the work of\cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is a set of predictors (features), $Y$ is the target variable and $Z$ is a potential confounder.
 In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether the features $X$ still affect $Y$, when controlling for $Z$.
@@
 
-As shown in \citep{berrett2020conditional}, to ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:
+As shown in\citep{berrett2020conditional}, to ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:
 
@@
 As long as the numerator of Eq.
-\ref{eq-pperm} is non-zero for all $c_\pi \in C$ and $y \in Y$, the conditional permutations constitute an algebraic group, thus, as shown by \citep{hemerik2018exact}, an unbiased estimate of the p-value under the null can be obtained as:
+\ref{eq-pperm} is non-zero for all $c_\pi \in C$ and $y \in Y$, the conditional permutations constitute an algebraic group, thus, as shown by\citep{hemerik2018exact}, an unbiased estimate of the p-value under the null can be obtained as:
 $$ p= \frac{\sum_{j=1}^m \mathbb{1} \{T(\y, \yhat, \c^{(j)}) \geq T(\y, \yhat, \c) \}  }{m}$$
 
-While the group property of the conditioned permutations provides a straightforward proof for the validity of the approach, for an alternative verification see the proof of Theorem 1 in \citep{berrett2020conditional}.
+While the group property of the conditioned permutations provides a straightforward proof for the validity of the approach, for an alternative verification see the proof of Theorem 1 in\citep{berrett2020conditional}.
 
@@
 However, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times.
-The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of \cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data.
+The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of\cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data.
 The acceptance odds ratio of swapping indices $i$ and $j$ is:
@@
 
-In their Theorem 2, \cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.
+In their Theorem 2,\cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.
 
@@
 
-As the linear regression-based method, obviously, does not handle non-linear relationships I propose to fit a generalized additive model (GAM) \citep{hastie1987generalized}, instead:
+As the linear regression-based method, obviously, does not handle non-linear relationships I propose to fit a generalized additive model (GAM)\citep{hastie1987generalized}, instead:
 
@@
 
-Smoothness of the GAM model is optimized with a grid-search by picking the model with the lowest generalized cross-validation score form the models defined by the default parameters as implemented in PyGAM \citep{serven2018generalized}.
+Smoothness of the GAM model is optimized with a grid-search by picking the model with the lowest generalized cross-validation score form the models defined by the default parameters as implemented in PyGAM\citep{serven2018generalized}.
 
@@
 
-In the case of categorical $c$, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$, with the extra assumption of \emph{complete separation} if $y$ is also categorical (for an invertible Hessian, see e.g.
-\citep{bennett1966multiple, jones1975proability}).
+In the case of categorical $c$, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$, with the extra assumption of \emph{complete separation} if $y$ is also categorical (for an invertible Hessian, see e.g.\citep{bennett1966multiple, jones1975proability}).
 
@@
 In theory, any predefined test statistic $T$ can be used with the proposed approach.
-The python package \emph{mlconfound}, implementing the proposed full and partial confoudner tests, utilizes the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification \citep{starkweather2011multinomial}) as a test statistic:
+The python package \emph{mlconfound}, implementing the proposed full and partial confoudner tests, utilizes the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification\citep{starkweather2011multinomial}) as a test statistic:
 
@@
 
-where $f$ is a function to introduce non-normality, namely the \emph{sinh-arcsinh} transformation of \cite{jones2009sinh} definend as:
+where $f$ is a function to introduce non-normality, namely the \emph{sinh-arcsinh} transformation of\cite{jones2009sinh} definend as:
 
@@
 \ref{fig:sim-h0-demo}), its validity was contrasted to partial Spearman correlation and the linear variety of CPT (based on eq.
-\ref{eq:linreg}, as described by \cite{berrett2020conditional}) with the following simulation parameters:
+\ref{eq:linreg}, as described by\cite{berrett2020conditional}) with the following simulation parameters:
 \begin{itemize}
@@
 
-The Human Connectome Project dataset contains imaging and behavioral data of approximately 1200 healthy subjects \citep{van2013wu}.
-Preprocessed resting state fMRI connectivity data (partial correlation matrices) \citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores (IQ), measured with Penn Progressive Matrices \citep{duncan2000neural}.
+The Human Connectome Project dataset contains imaging and behavioral data of approximately 1200 healthy subjects\citep{van2013wu}.
+Preprocessed resting state fMRI connectivity data (partial correlation matrices)\citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores (IQ), measured with Penn Progressive Matrices\citep{duncan2000neural}.
 
-To ensure normality, fluid intelligence scores were non-linearly transformed to normal distribution with the quantile transformation \citep{beasley2009rank} as implemented in \emph{scikit-learn} \citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:hcp-hist} for details).
+To ensure normality, fluid intelligence scores were non-linearly transformed to normal distribution with the quantile transformation\citep{beasley2009rank} as implemented in \emph{scikit-learn}\citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:hcp-hist} for details).
 
-Features (functional connectivities across 100 group-independent component analysis based regions) were either (i) considered in their raw form or were subject to confound mitigation approaches by (ii) feature regression \citep{rao2017predictive} or (iii) COMBAT \citep{johnson2007adjusting, fortin2018harmonization}.
+Features (functional connectivities across 100 group-independent component analysis based regions) were either (i) considered in their raw form or were subject to confound mitigation approaches by (ii) feature regression\citep{rao2017predictive} or (iii) COMBAT\citep{johnson2007adjusting, fortin2018harmonization}.
 The feature mitigation strategies were separately applied for acquisition batch and age group as confounder variable.
 
-Each of the 5 types of features (raw, regressing out acquisition batch, regressing out age group, COMBAT with acquisition batch, COMBAT with age group) was independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the individual fluid intelligence scores with a ridge regression \citep{hoerl1970ridge}.
+Each of the 5 types of features (raw, regressing out acquisition batch, regressing out age group, COMBAT with acquisition batch, COMBAT with age group) was independently incorporated into a scikit-learn-based\citep{pedregosa2011scikit} machine learning procedure aiming to predict the individual fluid intelligence scores with a ridge regression\citep{hoerl1970ridge}.
 The $\alpha$ parameter of the ridge model was considered as a hyperparameter ($\alpha \in \{0.00001, 0.0001, 0.001, 0.01, 0.1, 1, 10, 100, 1000, 10000, 100000\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric.
@@
 
-The proposed tests were applied to provide evidence of center- and  motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464.
-Preprocessed regional timeseries data was obtained as shared with the paper \citep{dadi2019benchmarking} which was based on preprocessed image data provided by the Preprocessed Connectome Project \citep{craddock2013neuro}.
+The proposed tests were applied to provide evidence of center- and  motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset\citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464.
+Preprocessed regional timeseries data was obtained as shared with the paper\citep{dadi2019benchmarking} which was based on preprocessed image data provided by the Preprocessed Connectome Project\citep{craddock2013neuro}.
 
-Tangent correlation across the timeseries of the n=122 regions of the BASC \citep{bellec2010multi} atlas) was computed with nilearn\footnote{\href{http://nilearn.github.io/}{http://nilearn.github.io/}} \citep{huntenburg2017loading, esteve2015big}.
+Tangent correlation across the timeseries of the n=122 regions of the BASC\citep{bellec2010multi} atlas) was computed with nilearn\footnote{\href{http://nilearn.github.io/}{http://nilearn.github.io/}}\citep{huntenburg2017loading, esteve2015big}.
 
-The resulting functional connectivity estimates were considered as features either (i) in their raw form or after applying (ii) feature regression \citep{rao2017predictive} or (iii) COMBAT \citep{johnson2007adjusting, fortin2018harmonization}.
-The investigated confounder variables were 'imaging center' and 'in-scanner motion', as measured by the mean framewise displacement of \cite{power2014methods})
-Mean FD was non-linearly transformed to normal distribution with the quantile transformation \citep{beasley2009rank} as implemented in \emph{scikit-learn} \citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:abide-hist} for details).
+The resulting functional connectivity estimates were considered as features either (i) in their raw form or after applying (ii) feature regression\citep{rao2017predictive} or (iii) COMBAT\citep{johnson2007adjusting, fortin2018harmonization}.
+The investigated confounder variables were 'imaging center' and 'in-scanner motion', as measured by the mean framewise displacement of\cite{power2014methods})
+Mean FD was non-linearly transformed to normal distribution with the quantile transformation\citep{beasley2009rank} as implemented in \emph{scikit-learn}\citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:abide-hist} for details).
 
@@
 
-The total of five types (raw, feature regression of site, feature regression of motion, COMBAT with site, COMBAT with motion) of features were independently incorporated into a scikit-learn-based \citep{pedregosa2011scikit} machine learning procedure aiming to predict the diagnosis (DX: ASD vs.
-HC) with a L2-regularized logistic regression (see rationale in \citep{dadi2019benchmarking}).
+The total of five types (raw, feature regression of site, feature regression of motion, COMBAT with site, COMBAT with motion) of features were independently incorporated into a scikit-learn-based\citep{pedregosa2011scikit} machine learning procedure aiming to predict the diagnosis (DX: ASD vs.
+HC) with a L2-regularized logistic regression, as previously recommended\citep{dadi2019benchmarking}.
 The $C$ parameter of the model was considered as a hyperparameter ($C \in \{0.1, 1, 10\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with area under the receiver operator curve (AUC under ROC) as optimization metric.
@@
 Confounder testing was performed on the class probabilities (instead of the predicted labels).
-%TC:endignore
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Acknowledgements}
+I am thankful to Ulrike Bingel (University Hospital Essen, Germany) and Robert Englert (University Hospital Essen, Germany) for their valuable insights and comments on the manuscript.
+This research was supported by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – Projektnummer 316803389 – SFB 1280  and TRR 289 Treatment Expectation - Projektnummer 422744262.
 
-%TC:ignore 
-\bibliographystyle{naturemag} %apalike}  
+\bibliographystyle{naturemag} %apalike
 \bibliography{references}

### 2021-10-29 manuscript.tex

@@
 
-\title{Detecting confounder and center bias in machine learning models}
+\title{Statistical quantification of confounding bias in predictive modelling}
 
@@
 \begin{abstract} % 150 words
-The lack of rigorous non-parametric statistical tests for confounder effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
+The lack of non-parametric statistical tests for confounding bias significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
 
-Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and, for a given confounder variable, probe the null hypotheses of \emph{unconfounded} and \emph{fully confounded models}, respectively.
+Here I propose the \emph{partial} and \emph{full confounder tests}, which, for a given confounder variable, probe the null hypotheses of \emph{unconfounded} and \emph{fully confounded models}, respectively.
 
-The tests provide a strict control for Type I errors and high statistical power, even for non-linear and non-normal dependencies typical in machine learning.
-Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset reveals confounders that were previously partly unreported and found to be hard to correct for with state-of-the-art confound mitigation approaches.
+The tests provide a strict control for Type I errors and high statistical power, even for non-normally and non-linearly dependent predictions, often seen in machine learning.
+Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset reveals confounders that were previously unreported or found to be hard to correct for with state-of-the-art confound mitigation approaches.
 
-The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment and improvement of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
+The tests (implemented in the package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}) can aid the assessment and improvement of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
 \end{abstract}
@@
 Predictive modelling has recently become increasingly important in biomedical research and holds promise for delivering biomarkers that substantially impact clinical practice and public health\citep{kent2018personalized, spisak2020pain, walsh2021dome}.
-When evaluating the usefulness and applicability of such markers, predictive performance (i.e.
-prognostic/diagnostic value) is far from being the only important consideration.
+When evaluating the usefulness and applicability of such markers, predictive performance is far from being the only important consideration.
 Biomedical validity (i.e.
-whether the model is driven by biomedically relevant signal) as well as generalizability and fairness across contexts and populations are crucial requirements for candidate biomarkers\citep{woo2017building, obermeyer2019dissecting, mehrabi2021survey}.
+whether the model is driven by biomedically relevant signal) as well as generalizability and fairness across contexts and populations are also fundamental requirements for candidate biomarkers\citep{woo2017building, obermeyer2019dissecting, mehrabi2021survey}.
 
-Spurious, out-of-interest associations between the predictor variables (features) and the prediction target - i.e the so-called confounder and collider biases\citep{prosperi2020causal} - can be detrimental to the model's biomedical validity and generalizability.
-Examples include models picking up on measurement artifacts (in-scanner head motion artifacts in magnetic resonance imaging-based predictive models of Alzheimer's\citep{rao2017predictive}, attention deficit hyperactivity disorder\citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD)\citep{gotts2013perils, spisak2014voxel, spisak2019optimal}), demographic and psychometric variables (e.g models trained to predict intelligence\citep{ cole2012global, he2020deep} might provide a statistically significant predictive performance by picking up solely on age-related variance\citep{dubois2018distributed, lohmann2021predicting}), sampling bias and stochastic group differences (racially biased machine learning models\citep{ obermeyer2019dissecting, lwowski2021risk}), batch effects or, in multi-center studies, center-effects.
+Spurious, out-of-interest associations between the predictor variables (features) and the prediction target can be detrimental to the model's biomedical validity and generalizability.
+This phenomenon is often called confounding bias\citep{prosperi2020causal}.
+Confounding bias can be driven by - among others - measurement artifacts (e.g motion artifacts in magnetic resonance imaging-based predictive models of Alzheimer's\citep{rao2017predictive}, attention deficit hyperactivity disorder\citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD)\citep{gotts2013perils, spisak2014voxel, spisak2019optimal}), demographic and psychometric variables (e.g models trained to predict intelligence\citep{ cole2012global, he2020deep} might provide a statistically significant predictive performance by picking up solely on age-related variance\citep{dubois2018distributed, lohmann2021predicting}), sampling bias and stochastic group differences (e.g.
+racially biased machine learning models\citep{ obermeyer2019dissecting, lwowski2021risk}), batch effects or, in multi-center studies, center-effects.
 
-While various data cleaning methods might help in mitigating confounder bias\citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, rao2017predictive, johnson2007adjusting}, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest\citep{wachinger2021detect}.
+While various data cleaning methods might help in mitigating confounding bias\citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, johnson2007adjusting}, it is often unclear which variables should be considered as confounders and such approaches hold risks of eliminating signal-of-interest\citep{wachinger2021detect}.
 
-Powerful and robust statistical tests to quantify confounding effects in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches.
-Although confounder bias can be expressed in terms of conditional independence of the model predictions on the target and the confoudner variables, the proper evaluation of the conditional independence among these variables is challenging.
+Powerful and robust statistical tests for quantifying confounding bias in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches.
+Although it is tempting to think about confounding bias as the \emph{conditional dependence} of the model predictions on the confounder, given the target variable, the proper evaluation of the conditional independence among these variables is challenging.
 Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation) are not valid measures of conditional independence.
 Although warnings about this issue were given from early on\citep{korn1984ranges}, and received a fair amount of attention recently\citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g.
-feature-set characteristics and model regularization\citep{garcia2009study, kristensen2017whole}; see Supplementary Material \ref{sup:nomlinviol} for a simplistic example.
+feature-set characteristics and model regularization\citep{garcia2009study, kristensen2017whole} (see Supplementary Material \ref{sup:nomlinviol} for a simplistic example).
 
-Alternative techniques for quantifying confoudner bias either fail to control type error (as  the known in the case of balanced permutations\cite{southworth2009properties, hemerik2018exact}, used in\cite{chaibub2019permutation}), or do not aim at all to do so\cite{ferrari2020measuring, wachinger2021detect}.
-Moreover, without some modifications, they are applicable for either only numeric\cite{wachinger2021detect} or only categorical variables\citep{chaibub2019permutation, ferrari2020measuring} and involve re-fitting the model\citep{chaibub2019permutation, ferrari2020measuring}, which might not be feasible for models with high a computational cost (e.g.
-with nested cross-validation).
-Some techniques\cite{wachinger2021detect}, furthermore, probe a null hypothesis (direct causation between the features and the prediction target) that is not applicable for for many kinds of data and models \footnote{For instance, causation between several psychological or clinical states and functional MRI activity/connectivity is obviously not \emph{direct}, as functional MRI itself is an \emph{indirect} measure of neural activation.
-In such cases, the alternative hypothesis of this test might be true in absence of any real confounder.
-}.
+Recently, two different approaches were proposed for quantifying confounding bias \citep{chaibub2019permutation, ferrari2020measuring}.
+However, these methods either fail to control type I error (as known in the case of balanced permutations\cite{southworth2009properties, hemerik2018exact}, used in ref.\cite{chaibub2019permutation}), or do not provide p-values at all\cite{ferrari2020measuring}.
+Moreover, without some modifications, they are only applicable for categorical variables and involve re-fitting the model, which might not be feasible for models with high computational cost (e.g.
+when trained with nested cross-validation).
 
-This work aims to construct a test of confounder bias which (i) can be applied post-hoc, without re-fitting the model (i.e.
-requires only the target variable, the confoudner and the model predictions), (ii) guarantees valid type-I error control even if the involved conditional distributions are non-normal and/or non-linear, (iii) is applicable for classification and prediction problems, both with numerical and categorical confounders.
-
-% todo: RO and CI: classification, KC: regression
-
-%In this paper, I formulate the 'confounder problem' as a special case of conditional independence testing\citep{dawid1979conditional}.
-This allows relating the problem to the seminal work of\cite{shah2020hardness}, who showed that, without placing some assumptions on the joint distribution of the involved variables, establishing a conditional independence test with a valid type I error control and non-trivial power is effectively impossible ('no free lunch' theorem).
-%As an important consequence, methods like partial correlation - and even its non-parametric variety - will fail if the corresponding conditional distributions are non-normal or if non-linear relationships are involved (see e.g.\citep{korn1984ranges} or Fig.
-\ref{fig:sim-h0-demo} in this paper), as commonly seen in machine learning.
-
-%To overcome these limitations, I extend the permutation-based conditional independence framework of\cite{berrett2020conditional} with a generalized additive model based conditional distribution estimation, which allows avoiding to place any assumptions about the most problematic variable, the prediction output.
-%The framework allows constructing two tests: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
-%Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
-%The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality and non-linearity).
-
-%The usefulness of the proposed tests in detecting various types of confounder bias is demonstrated in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.
-%First, functional connectivity data from the Human Connectome Project (HCP)\citep{van2013wu} is used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age\citep{ dubois2018distributed, lohmann2021predicting} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
-%Second, functional connectivity data from the ABIDE\citep{di2014autism} database is used to investigate the potential motion- and center-bias\cite{gotts2013perils, spisak2014voxel, spisak2019optimal} when training models that aim to predict ASD diagnosis.
+This work aims to construct a statistical test for confounding bias that (i) can be applied post-hoc, without re-fitting the model (i.e.
+requires only the target variable, the confounder and the model predictions), (ii) guarantees valid type-I error control even if the involved conditional distributions are non-normal and/or non-linear, (iii) is applicable for classification as well as for prediction problems, both with numerical and categorical confounders.
 
@@
   \includegraphics[width=0.75\paperwidth]{fig/overview.eps}
-  \caption{\textbf{Overview of the proposed partial confoudner test}. \\
-    The partial confoudner test models the conditional distribution of the confounder, given the target variable, with a generalized additive model (GAM).
+  \caption{\textbf{Overview of the proposed partial confounder test}. \\
+    The partial confounder test models the conditional distribution of the confounder, given the target variable, with a generalized additive model (GAM).
 The parallel-pairwise Markov-chain Monte-Carlo sampler draws permutations of the confounder variable that comply with the GAM-based conditional distribution (permutation 1, 2, ..., m).
-The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted confoudner variables.
+The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted confounder variables.
 The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original.
@@
 
-\subsection{The partial and full confoudner tests}
+\subsection{The partial and full confounder tests}
 
@@
 
-The idea of conditional independence provides a straightforward framework for assessing the confounder bias in predictive models.
+The idea of conditional independence provides a straightforward framework for assessing confounding bias in predictive models.
 However, handling the non-normal and/or non-linear conditional dependence often seen in predictive models\citep{garcia2009study, kristensen2017whole} poses a great challenge.
-In fact, as recently shown by Shah and colleagues\cite{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and non-trivial power.
-In practice, this means that all candidate tests, that assume normality and/or linearity of the joint distribution, will necessarily fail to properly control for Type I errors in many realistic predictive modelling scenarios.
-For instance, a trivial candidate, partial Spearman correlation, exhibits inflated type I errors even with slight violations of normality and linearity, as clearly demonstrated with simulated data on Fig.
-\ref{fig:sim-h0-demo}.
+In fact, as recently shown by Shah and colleagues\cite{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and a non-trivial power.
+In practice, this means that no fully assumption-free test for assessing confounding bias can be constructed.
+Indeed - perhaps somewhat surprisingly, but not totally unexpectedly\cite{korn1984ranges} - partial correlation-like analogs of the widely used bivariate non-parametric test, like partial Spearman correlation, exhibits inflated type I errors even with slight violations of normality and/or linearity (as clearly demonstrated with simulated data on Fig.
+\ref{fig:sim-h0-demo}).
+Such tests are therefore poor choices for testing confounding bias in machine learning.
 
-As the most problematic from the three variables (target, prediction, confounder) is clearly the model prediction\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics.
-Exactly this can be achieved with the combination of the novel framework of conditional permutation testing (CPT)\cite{berrett2020conditional} and the use of generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}-based conditional distribution estimation.
-The proposed approach offers two novel tests for probing confounder bias: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
+As, in terms of its conditional distribution on the others, the most intractable from the three involved variables (target, prediction, confounder) is clearly the model output\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics.
+Exactly this can be achieved with the proposed approach, which extends the novel framework of conditional permutation testing (CPT)\cite{berrett2020conditional} with conditional distribution estimation via generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic models\cite{bennett1966multiple, jones1975proability}.
+The proposed approach offers two novel tests for probing confounding bias: the \emph{full confounder test} probes whether the model's predictive performance can be exclusively attributed to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
 These tests place no assumptions on the conditional distributions of the model predictions, ensuring valid model diagnostics even in cases of non-normally and non-linearly dependent predictions.
 
-The inner workings of the \emph{partial confoudner test} is summarized on Fig.
+The inner workings of the \emph{partial confounder test} is summarized on Fig.
 \ref{fig:overview}.
@@
 The original and the permuted test statistics construct the p-value as the ratio of permuted test statistics more extreme than the original.
-The \emph{full confoudner test} works in an analogous way, with the difference that it creates permuted copies of the target variable, instead of the confoudner.
+The \emph{full confounder test} works in an analogous way, with the difference that it creates permuted copies of the target variable, instead of the confounder.
 
-These tests can handle both numerical and categorical data and can be applied for any predictive model, without having to re-fit the model, that is, with a negligible extra computational cost.
-The proposed \emph{partial} and \emph{full confoudner tests} have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.
+These tests can handle both numerical and categorical data and can be applied for any predictive model, without needing to re-fit the model, that is, with a negligible extra computational cost.
+The proposed \emph{partial} and \emph{full confounder tests} have been implemented in the python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}.
 
@@
   \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
-  \caption{\textbf{Type I error control and power of the partial confoudner test based on simulations with normal conditional distribution and linear dependencies.} \\
+  \caption{\textbf{Type I error control and power of the partial confounder test based on simulations with normal conditional distribution and linear dependencies.} \\
   Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette on the right) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\hat{y}}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\hat{y}}$ (rows) and for different sample sizes (N, columns).
 Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1.0 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%.
-First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounder bias).
+First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounding bias).
   Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively.
@@
 Notably, with sample sizes as large as 1000, a confounder contributing to only ~4\% to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%.
-With a sample size of 500, the same confounder bias is still detected with a power greater than 84\% in all of the simulation cases.
+With a sample size of 500, the same confounding bias is still detected with a power greater than 84\% in all of the simulation cases.
 A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) for a robust detection, with a power of 75-98\%.
@@
 Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
-  Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
+  Solid red line between the confounder and the prediction means significant confounding bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
 P-values are determined with the partial confound test.
@@
 
-After applying confound mitigation approaches (feature regression or COMBAT) the partial confounder test did not provide evidence of confounder bias anymore ($p > 0.05$ for all; shown in the second and third columns of Fig.
+After applying confound mitigation approaches (feature regression or COMBAT) the partial confounder test did not provide evidence of confounding bias anymore ($p > 0.05$ for all; shown in the second and third columns of Fig.
 \ref{fig:hcp}), neither for acquisition batch nor for age.
@@
   Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$.
-Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
+Solid red line between the confounder and the prediction means significant confounding bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
 P-values are determined with the partial confound test.
@@
 \caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated confounder-mitigation methods.
-Significant confoudner bias is denoted by bold font.}
+Significant confounding bias is denoted by bold font.}
 \end{table}
@@
 
-The assessment of confoudner bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings.
+The assessment of confounding bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings.
 However, statistical testing of conditional independence is challenging: analogs of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity\citep{korn1984ranges} (see an example on Fig.
@@
 While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the "no free lunch theorem" (i.e.
-the nonexistence of a uniformly valid conditional independence test)\cite{shah2020hardness} clearly implies that conditional independence-based confoudner bias testing must be designed so that its suitability for the particular problem may be judged easily.
+the nonexistence of a uniformly valid conditional independence test)\cite{shah2020hardness} clearly implies that conditional independence-based confounding bias testing must be designed so that its suitability for the particular problem may be judged easily.
 
-In this paper, I have proposed a solution for confounder testing, which was designed to be suitable for the specific characteristics of the 'confoudner problem' in predictive modelling.
-Namely, the proposed approach places assumptions only about the conditional distribution of the confoudner and the target variables, as required by the applied generalized additive\cite{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}.
+In this paper, I have proposed a solution for confounder testing, which was designed to be suitable for the specific characteristics of the 'confounder problem' in predictive modelling.
+Namely, the proposed approach places assumptions only about the conditional distribution of the confounder and the target variables, as required by the applied generalized additive\cite{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}.
 In biomedical research practice, these are variables that are often subject to descriptive (and often parametric) statistical analysis, as well.
@@
 
-The proposed approach gives rise to two different statistical tests for testing confoudner bias: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
+The proposed approach gives rise to two different statistical tests for testing confounding bias: the \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
 The tests can be applied for arbitrary classification or regression models, without having to re-fit the model, that is, with a negligible extra computational cost.
@@
 
-%Obviously, a true confounder bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder association, i.e.
+%Obviously, a true confounding bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder association, i.e.
 the model might still use the confounder's effect on the features (e.g artifacts) to construct the corresponding part of the variance in the prediction.
@@
 
-On the other hand, the partial confoudner test also showed, that the performance of the investigated confound mitigation techniques is much less convincing for the ABIDE dataset.
+On the other hand, the partial confounder test also showed, that the performance of the investigated confound mitigation techniques is much less convincing for the ABIDE dataset.
 The center bias of the classification in this massively multi-center dataset was, namely, found to be significant, even after feature regression and COMBAT.
@@
 
- In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounder-bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT and (iii) in some cases confoudner mitigation can - paradoxically - introduce more bias.
+ In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounder-bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT and (iii) in some cases confounder mitigation can - paradoxically - introduce more bias.
 The partial confounder test can be considered as a useful, objective benchmark to guide the search for a suitable confounder mitigation approach for every dataset.
@@
 As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and allow quantitative, statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research.
-Given their simplicity, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confoudner tests can become widespread tools that pave the way for the development of clinically useful machine learning biomarkers.
+Given their simplicity, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confounder tests can become widespread tools that pave the way for the development of clinically useful machine learning biomarkers.
 
@@
 
-\subsubsection*{Conditional independence for testing confounder bias}
+\subsubsection*{Conditional independence for testing confounding bias}
 
@@
 Testing the null hypothesis $\yhat \independent \c | \y$ (option 3, partial confounder testing) examines, whether the dependence of the model output on the confounder can likely be explained by the confounder's dependence on the target variable, i.e.
-whether there is any confounder bias in the model.
+whether there is any confounding bias in the model.
 
@@
 %\hline
-\textbf{3.} & $\yhat \independent \c | \y$  \quad &  partial confounder  test: model not directly driven by the confoudner & $Q(\c|\y)$ & $(\yhat, \c), (\yhat, \y)$ \\
+\textbf{3.} & $\yhat \independent \c | \y$  \quad &  partial confounder  test: model not directly driven by the confounder & $Q(\c|\y)$ & $(\yhat, \c), (\yhat, \y)$ \\
 \end{tabular}
@@
 Option 3, i.e.
-partial confounder testing is typically of interest when testing confounder bias of predictive models.
+partial confounder testing is typically of interest when testing confounding bias of predictive models.
 Option 1, i.e.
@@
 In theory, any predefined test statistic $T$ can be used with the proposed approach.
-The python package \emph{mlconfound}, implementing the proposed full and partial confoudner tests, utilizes the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification\citep{starkweather2011multinomial}) as a test statistic:
+The python package \emph{mlconfound}, implementing the proposed full and partial confounder tests, utilizes the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification\citep{starkweather2011multinomial}) as a test statistic:
 
@@
 
-which allows interpretable, two tailed inference about confoudner-bias in predictive modelling.
+which allows interpretable, two tailed inference about confounder-bias in predictive modelling.
 

### 2021-10-29 manuscript.tex
Updates from Overleaf

@@
 % keywords can be removed
-\keywords{machine learning, predictive modelling, confounder test, conditional independence, conditional permutation}
+\keywords{machine learning, predictive modelling, confoudner bias, confounder test, conditional independence, conditional permutation}
 
@@
 Although it is tempting to think about confounding bias as the \emph{conditional dependence} of the model predictions on the confounder, given the target variable, the proper evaluation of the conditional independence among these variables is challenging.
-Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation) are not valid measures of conditional independence.
+Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation, see Fig.
+\ref{fig:sim-h0-demo}) are not valid measures of conditional independence.
 Although warnings about this issue were given from early on\citep{korn1984ranges}, and received a fair amount of attention recently\citep{bergsma2010nonparametric, candes2016panning, peters2016causal,  shah2020hardness, berrett2020conditional}, the magnitude of the problem may not be fully appreciated in case of predictive model diagnostics, where non-normality and non-linearity of the model output can be frequently seen, as a consequence of e.g.
@@
 
+%fig:sim-h0-demo
+\begin{figure}[!b]
+  \centering
+  \includegraphics[width=0.5\paperwidth]{fig/sim_h0_demo.eps}
+  \caption{\textbf{Type I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
+  Type I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), slightly non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row).
+Non-normal conditional distribution on the second plot is illustrated with blue density diagrams (kurtosis: -0.8, skewness: -0.1).
+False positive rates for various simulation settings is shown in heatmaps.
+The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065.
+Values below this threshold indicate a valid type I error control and colored white.
+  }
+  \label{fig:sim-h0-demo}
+\end{figure}
+
 \section{Results}
@@
   \centering
-  \includegraphics[width=0.75\paperwidth]{fig/overview.eps}
+  \includegraphics[width=0.6\paperwidth]{fig/overview.eps}
   \caption{\textbf{Overview of the proposed partial confounder test}. \\
@@
 \subsection{The partial and full confounder tests}
-
-%fig:sim-h0-demo
-\begin{figure}[!b]
-  \centering
-  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
-  \caption{\textbf{Type I error control of partial Spearman correlation, linear and GAM-based conditional permutation test}. \\
-  Type I error control was investigated in three example cases: normal conditional distribution with linear dependency (first row), slightly non-normal conditional distribution with linear dependency (second row) and normal conditional distribution with non-normal (sigmoid) dependency (third row).
-Non-normal conditional distribution on the second plot is illustrated with blue density diagrams (kurtosis: -0.8, skewness: -0.1) and dependency of the mean is depicted with blue dotted lines an all plots.
-False positive rates for the three investigated tests is shown in heatmaps, with darker red meaning higher values.
-The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065.
-Values below this threshold indicate a valid type I error control and colored white.
-  }
-  \label{fig:sim-h0-demo}
-\end{figure}
 
@@
 
-\subsection{Simulations}
-
 %fig:sim-normal
@@
   \centering
-  \includegraphics[width=0.75\paperwidth]{fig/sim_normal.eps}
+  \includegraphics[width=0.6\paperwidth]{fig/sim_normal.eps}
   \caption{\textbf{Type I error control and power of the partial confounder test based on simulations with normal conditional distribution and linear dependencies.} \\
@@
 \end{figure}
+
+\subsection{Simulations}
 
@@
 
-As suggested by theory (see \hyperref[sec:methods]{Methods} for details) and shown by the simulations with a wide range of settings, both of the proposed tests provide a valid Type I error control (Fig.
-\ref{fig:sim-normal}), even in case of non-linearity and non-normality (Figs.
-\ref{fig:sim-h0-demo} and \ref{fig:sim-non-normal}), except when non-normality is very extreme (purple distribution on Fig.
+As suggested by theory (see \hyperref[sec:methods]{Methods} \ref{sec:methods-pct} for details) and shown by the simulations with a wide range of settings, both of the proposed tests provide a valid Type I error control (Fig.
+\ref{fig:sim-normal} and Supplementary Figures \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-ccc-lin-full} and \ref{fig:sim-bbb-lin-full}), even in case of non-linearity and non-normality (Figs.
+\ref{fig:sim-h0-demo}, \ref{fig:sim-non-normal} and Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-sig-full}), except when non-normality is very extreme (purple distribution on Fig.
 \ref{fig:sim-non-normal}, kurtosis: 42, skewness: -6).
@@
 Estimates of statistical power for the partial confounder test (with normal and linear simulations, for  a wide range of parameters) are shown on Figure \ref{fig:sim-normal}.
-Notably, with sample sizes as large as 1000, a confounder contributing to only ~4\% to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%.
-With a sample size of 500, the same confounding bias is still detected with a power greater than 84\% in all of the simulation cases.
-A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) for a robust detection, with a power of 75-98\%.
+Notably, with sample sizes as large as 1000, a confounder contributing only $\sim 4\%$ to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%.
+With a sample size of 500, the same confounding bias is still detected with a power greater than 84-100\% in all of the simulation cases.
+A sample size of 100 requires a somewhat stronger bias with approximately 12\% of explained variance ($w_{c\hat{y}}=0.4$) to achieve a reasonable level of power (75-98\%).
 Finally, even with a relatively low sample size of 50, the same amount of confounder variance is detected with a power of at least 50\%.
-If the confounder explains more than 25\% of variance, it is almost certainly detected even with the low sample size of $n \geq 50$.
+If the confounder explains more than 25\% of variance, it is almost certainly detected even with a low sample size of $n \geq 50$.
 
-Simulations with non-normal and non-linear dependencies has minimal effect on the power of the tests, except in case of extreme non-normality. (Fig. \ref{fig:sim-non-normal}).
-Type I error and power was found to be highly similar in case of categorical variables and for the \emph{full} confounder test, as well (Supplementary Figures \ref{fig:sim-ccc-sig-partial}, \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-sig-partial}, \ref{fig:sim-ccc-sig-full}, \ref{fig:sim-bbb-lin-full} and \ref{fig:sim-bbb-sig-full}).
+Simulations show that non-normality has minimal effect on the power of the tests, except in case of extreme non-normality.
+(Fig.
+\ref{fig:sim-non-normal}).
+Simulations with sigmoid dependence resulted in an apparent loss of statistical power, however this is simply a consequence of the simulation methodology: with the same parameters, the sigmoid transformed confounder explains only approximately half the variance as compared to linear simulations.
+Type I error control was valid in case of categorical variables and for the \emph{full} confounder test, as well.
+Power in these cases was also highly similar to the partial confoudner test results with numerical variables.
+(Supplementary Figures \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-lin-full}, \ref{fig:sim-bbb-sig-partial} and \ref{fig:sim-bbb-sig-full}).
 
@@
 
-To demonstrate the usefulness of the proposed tests in detecting various types of confounder-bias, they have been deployed in two typical research scenarios (a regression and a classification problem) where confounder effects are known to harm building biomedically valid predictive models.
+To demonstrate the usefulness of the proposed tests in detecting various types of confounding bias, they have been deployed in two typical research scenarios - a regression and a classification problem - where confounder effects are known to hamper the development of biomedically valid predictive models.
 
@@
   \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
-  \caption{\textbf{Acquisition- and age-bias of fluid intelligence prediction in the HCP dataset.} \\
+  \caption{\textbf{Acquisition batch and age bias of fluid intelligence prediction in the HCP dataset.} \\
   Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies.
-Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data.
+Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder bias both for acquisition and age in the models trained on the raw data.
 This bias is robustly detected by the partial confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
-  Relation between the observed ($Gf$) and predicted ($\hat{y}$) IQ scores and the confounder variables is given on the graphs via $R^2 values$.
-Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
+  Relation between the observed ($Gf$) and predicted ($\hat{y}$) intelligence scores and the confounder variables is given on the graphs via $R^2 values$.
+Both confound mitigation techniques, but especially COMBAT, improve the predictive performance.
   Solid red line between the confounder and the prediction means significant confounding bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
-P-values are determined with the partial confound test.
-P-values of the 'full' confound test were all less then 0.0001, i.e.
+P-values are determined with the partial confounder test.
+P-values of the 'full' confounder test (not shown) were all less then 0.0001, i.e.
 the confounders did not fully drive prediction for any of the models.
@@
 
-Functional connectivity data from the Human Connectome Project (HCP)\citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age\citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
+Functional connectivity data from the Human Connectome Project\citep{van2013wu} (HCP) was used to build predictive models of fluid intelligence ($G_f$) and to test for the previously discussed confounding effect of age\citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
 
@@
 The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
-The partial confounder test revealed that the model was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig.
-\ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.071$ and $0.034$, respectively) associated to larger predicted values.
 
-After applying confound mitigation approaches (feature regression or COMBAT) the partial confounder test did not provide evidence of confounding bias anymore ($p > 0.05$ for all; shown in the second and third columns of Fig.
+The partial confounder test revealed that the 'raw' model (without confounder mitigation) was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig.
+\ref{fig:hcp}) with later phases of the acquisition and lower age being associated to larger predicted values.
+
+After applying confound mitigation approaches (feature regression or COMBAT) the partial confounder test did not provide evidence for confounding bias anymore ($p > 0.05$ for all; shown in the second and third columns of Fig.
 \ref{fig:hcp}), neither for acquisition batch nor for age.
-Both feature regression and COMBAT increased the predictive performance in the case of both confounders, with COMBAT providing best performances ($R^2=0.122$ and $0.121$ when applied to remove the effect of acquisition and age, respectively).
+Both feature regression and COMBAT increased the predictive performance, with COMBAT providing the overall best performances ($R^2=0.122$ and $0.121$ when applied to remove the effect of acquisition and age, respectively).
 
@@
   Boxplots and points show the predicted class probabilities (0: HC, 1: ASD), separately for the HC and ASD groups.
-In the top panel, predictions are plotted for each center separately and color indicates the true diagnosis (DX).
-AT the bottom plot, color indicates the normalized index of in-scanner motion (based on mean framewise displacement, mean FD).
-The proposed confounder test reveals significant center- and motion-bias in the model trained on the raw data (p<0.0001).
-While center bias seems to be effectively mitigated by feature regression, the same technique fails to remove motion-bias (actually introduces a paradoxical negative bias).
+In the top panel, predictions are plotted for each center separately.
+Color indicates the true diagnosis (DX).
+AT the bottom plot, color indicates the normalized index of in-scanner motion (normalized FD).
+The proposed confounder test reveals significant center and motion bias in the model trained on the raw data (p<0.0001).
+While feature regression seems to effectively mitigate center bias, it fails to remove motion bias (actually introduces a paradoxical negative bias).
 COMBAT seems to effectively remove the effect of both confounders on the predictions and improves the predictive performance of the models in both cases.
-  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$.
+  Relation between the true ($\hat{y}$) and predicted diagnosis scores and the confounder variables is shown by the graphs as $R^2$ values.
 Solid red line between the confounder and the prediction means significant confounding bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias.
 P-values are determined with the partial confound test.
-P-values of the full confound test were all less then 0.0001, i.e.
-confounder did not fully drive prediction for any of the models.
+P-values of the full confound test (not shown) were all less then 0.0001, i.e.
+the confounders did not fully drive the prediction for any of the models.
   }
@@
 
-Functional connectivity data from the ABIDE\citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
+Functional connectivity data from the ABIDE\citep{di2014autism} database was used to investigate the potential motion and center bias (as previously reported e.g.
+in refs\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
 
@@
 The model trained on the raw (unadjusted) connectivity features predicted diagnosis with a medium effect size ($R^2=0.126$, $ROC AUC = 0.71$, $p<0.001$).
+
 The partial confounder test revealed that the raw model was significantly biased both for age group and acquisition batch (both $p<0.0001$, see first column on Fig.
 \ref{fig:abide}).
-Several sites (e.g.
-Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) and lower motion being weakly-moderately associated ($R^2=0.169$ and $0.111$, respectively) to a lower probability for the participant to be assigned to the ASD group by the model.
+Predictions for several sites (e.g.
+Carnegine Mellon University, University of Leuven, Social Brain Lab UMC Groningen) were severely miscalibrated and higher motion was associated to a higher probability for ASD diagnosis.
 
-Both feature regression and COMBAT seemed to significantly attenuate the effect of imaging center, however the partial confound test still provided evidence for a significant center-bias ($p=0.04$ and $0.009$ for feature regression and COMBAT, respectively (second and third columns of the first row on Fig \ref{fig:abide}).
+Both feature regression and COMBAT seemed to significantly attenuate center bias, however the partial confound test still provided evidence for a significant bias ($p=0.04$ and $0.009$ for feature regression and COMBAT, respectively (second and third columns of the first row on Fig \ref{fig:abide}).
 
@@
         & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & \textbf{<0.0001} & <0.0001 \\
-        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & \textbf{<0.0001} & 0.1 \\
+        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & \textbf{<0.0001} & \textbf{0.1} \\
         &          & COMBAT    & & &  0.002 & 0.19 & 0.111     & <0.001 & 0.59 & <0.0001 \\
@@
 \end{tabular}
-\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated confounder-mitigation methods.
-Significant confounding bias is denoted by bold font.}
+\caption{\label{tab:unconditional-pvals} Coefficients-of-determination ($R^2$), the corresponding p-values and the p-values of the partial and full confounder tests, for all investigated confounders (conf.) confounder-mitigation methods (method).
+Bold numbers denote significant confounding bias identified by the partial confounder test or the lack of significant contribution over that of the confoudner, as identified by the full confounder test.}
 \end{table}
@@
 The assessment of confounding bias via conditional independence testing is a very appealing solution due to its simplicity and applicability in a wide variety of settings.
-However, statistical testing of conditional independence is challenging: analogs of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity\citep{korn1984ranges} (see an example on Fig.
+However, statistical testing of conditional independence is challenging: 'conditional analogs' of bivariate (unconditioned) non-parametric tests can give invalid results even by slight violations of normality and linearity\citep{korn1984ranges} (see an example on Fig.
 \ref{fig:sim-h0-demo}).
@@
 Namely, the proposed approach places assumptions only about the conditional distribution of the confounder and the target variables, as required by the applied generalized additive\cite{hastie1987generalized} or multinomial logistic model\cite{bennett1966multiple, jones1975proability}.
-In biomedical research practice, these are variables that are often subject to descriptive (and often parametric) statistical analysis, as well.
-Accordingly, usually there is sufficient expert knowledge about the applicability of the required assumptions and data acquisition is probably already designed so that the assumptions are likely fulfilled.
+In biomedical research practice, these are variables that are often subject to descriptive (and often parametric) statistical analyses, as well.
+Accordingly, usually there is sufficient expert knowledge about the applicability of the required assumptions and the data acquisition is probably already designed so that the assumptions are likely fulfilled.
 Of key importance, the proposed solution is \emph{absolutely assumption-free} regarding the distribution and dependency of the prediction output, which is, in most of the cases, unknown and can often be expected to be non-normal and/or non-linear (due to e.g model regularization)\citep{garcia2009study, kristensen2017whole}.
@@
 
-As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except extreme non-normality), confirming the tests' can be deployed in most predictive modelling scenarios.
+As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except extreme non-normality), confirming that the tests can be deployed in a wide variety of predictive modelling scenarios.
 While different biomedical applications might consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculation in order to identify the necessary sample size for proper model diagnostics.
-Validity, power and robustness renders the proposed tests as unprecedented tools for model diagnostics in a wide spectrum of predictive modelling scenarios.
-
-%In biomedical research, confounder-bias has previously often been investigated (if investigated at all) simply by testing the (unconditioned) association between the confounder variable and the predictive values, see e.g.\citep{spisak2020pain}.
-However, the significance of the unconditioned association does not necessarily imply a significant confound-bias of the model, especially if the confounder is also associated to the true target variable.
-In this case, namely, the model still might not be directly driven by the confounder and the dependence of the predictions on the confounder might be only marginal, to be explained solely by the confounder-target association.
-%Two examples for this situation can be found in the presented analysis of real data; when applying COMBAT to mitigate age-bias in the HCP data and when regressing out the center-effect from the ABIDE data (see Table \ref{tab:unconditional-pvals}).
-%These examples clearly indicate that guiding model diagnostics based on unconditioned associations between the confounder and the model output might result in overly strict decisions.
-More realistic is to aim at models that allow a reasonable degree of marginal confounder-association of the predictions, but are likely not directly driven by confounder-information, i.e.
-the prediction-confounder association is proportional to, and can be fully explained by, the target-confounder association.
-%Put simply, this is what is tested by the proposed partial confounder test.
-
-%Obviously, a true confounding bias might still be present in the model, even if the the target-confounder association likely explains the prediction-confounder association, i.e.
-the model might still use the confounder's effect on the features (e.g artifacts) to construct the corresponding part of the variance in the prediction.
-Nevertheless, in most of the cases, it does not seem likely that the model, at the same time, (i) misses that specific "part" of the target variable's variance in the features that is shared with the confounder (ii) captures exclusively that component of the confounder's variance in the features that is needed to mimic an unbiased model.
-%Usually a much more plausible explanation is that a direct confounder effect in the model output is simply negligible (e.g because it is not to be found in the features after confound mitigation).
 
@@
 The usefulness of the proposed tests is demonstrated in two such examples, using the HCP\citep{van2013wu}  and the ABIDE\citep{di2014autism} datasets.
-%First, functional connectivity data from the Human Connectome Project (HCP)\citep{van2013wu} was used to build predictive models of fluid intelligence and to test for the previously discussed confounding effect of age\citep{lohmann2021predicting, dubois2018distributed} and,  additionally, the - somewhat underdiscussed - batch-like effect of acquisition date of the data within the course of the data acquisition process.
-%Second, functional connectivity data from the ABIDE\citep{di2014autism} database was used to investigate the potential motion- and center-bias (as previously reported by e.g.\cite{spisak2014voxel, spisak2019optimal, gotts2013perils}) when training models that aim to predict ASD diagnosis.
-%The hypothesised confounder effects were clearly present in both datasets.
 
-In case of the HCP dataset, the weak but statistically significant confounder effect of age on fluid intelligence is in line with previous findings\citep{dubois2018distributed, lohmann2021predicting}) could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample.
+In case of the HCP dataset, the weak but statistically significant confounder effect of age on fluid intelligence is in line with previous findings\citep{dubois2018distributed, lohmann2021predicting} and could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample.
 In this case, the bias would likely significantly harm the out-of-sample generalizability of this model.
-The bias of the same model to 'acquisition batch' (phase of the study the acquisition belongs to) can also be problematic and it has not yet been thoroughly discussed in case of the HCP dataset.
+The bias of the same model for acquisition batch can also be problematic, especially as it has not yet been thoroughly discussed in case of the HCP dataset.
 There can be manifold reasons for the observed bias.
 Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g.
-as a consequence of the human connectome project receiving an increasing degree of public interest during its course, the sociodemographic and educational characteristics of the volunteer sample might have changed over time).
-To explain the associated change in the target variable, the model might utilise changing characteristics of the imaging data (caused by e.g.
-slight changes in the calibration settings of the measuring equipment or the changing level of experience of the research staff).
-This is obviously undesirable for any brain-based predictive model of fluid intelligence.
+as a consequence of the human connectome project receiving an increasing degree of public interest during its course).
 
-In the ABIDE dataset, neither the medium-strong center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD.
+In the ABIDE dataset, neither the center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD.
 For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g.
 the CMU (Carnegie Mellon University) center as neurotypical control participants.
-Similarly, the models biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with a tendency for higher in-scanner motion, e.g.
-due to atypical motor function (as known e.g.
-for ADHD\citep{eloyan2012automated}).
+Similarly, the models biased by motion - next to having questionable neuroscientific validity - might systematically fail in populations with a tendency for higher in-scanner motion (as known for many conditions, among others ADHD\citep{eloyan2012automated} or Alzheimer's disease \citep{rao2017predictive}).
 
@@
 
-On the other hand, the partial confounder test also showed, that the performance of the investigated confound mitigation techniques is much less convincing for the ABIDE dataset.
-The center bias of the classification in this massively multi-center dataset was, namely, found to be significant, even after feature regression and COMBAT.
-Nevertheless, the p-values returned by the partial confounder test increased, in both cases, with two orders of magnitude as compared to the case with raw features, suggesting that a large part of center effects was still successfully removed by both approaches.
-Although determining the relevance of the remaining bias is out of the scope of this paper, significant p-values of the partial confounder test generally suggest that these models require more effective data harmonization.
+On the other hand, the partial confounder test also showed, that the performance of the investigated confound mitigation techniques is much less convincing in the ABIDE dataset.
+The center bias of the classification in this massively multi-center dataset was found to be highly significant and, although mitigated, but not successfully removed by feature regression and COMBAT.
+Determining the relevance of the remaining bias is out of the scope of this paper, however significant p-values of the partial confounder test generally suggest that these models require more effective data harmonization.
 
 Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
-Interestingly, in the latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign.
-Meanwhile, the prediction performance substantially dropped (although still remained statistically significant).
-A possible explanation to these phenomena is that in-scanner motion, as previously described\citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest (especially in functional connections that are themselves directly associated with motor deficits in ASD).
-This effect might be boosted by the previously reported "residual motion artifacts" and their regional interactions\citep{spisak2014voxel} and raises caution for using feature regression to mitigate the effect of motion artifacts and similarly complex confounders.
+Interestingly, in this latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign.
+At the same time, the prediction performance substantially dropped (although still remained statistically significant).
+A possible explanation for this phenomenon is that in-scanner motion, as previously described\citep{fournier2010motor, anzulewicz2016toward}, has manifold links to ASD and, therefore, regressing out motion estimates from the connectivity features might eliminate a significant amount of signal-of-interest.
+This effect might be boosted by the previously reported 'residual motion artifacts' and their regional interactions\citep{spisak2014voxel} and raises caution for using feature regression to mitigate the effect of similar, complexly distributed confounders.
+
+ The success of COMBAT in eliminating motion bias is also not to be taken without any objections.
 As COMBAT was originally developed for harmonizing effects of categorical variables (e.g.
 center or batch), its application for continuous confounder variables is not trivial.
-The presented method of inputting the binned "motion groups" into COMBAT resulted in an efficient removal of motion-bias and, at the same time, preserved a significant amount of signal-of-interest.
- Nevertheless, inputting discretized versions of continuous variables into COMBAT might be sub-optimal and raises further questions e.g regarding the optimal number of bins used during the discretization.
+Inputting discretized versions of continuous variables into COMBAT might be sub-optimal and raises further questions e.g regarding the optimal number of bins used during the discretization.
 
- In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounder-bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT and (iii) in some cases confounder mitigation can - paradoxically - introduce more bias.
+ In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounding bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT and (iii) in some cases confounder mitigation can - paradoxically - introduce more bias.
 The partial confounder test can be considered as a useful, objective benchmark to guide the search for a suitable confounder mitigation approach for every dataset.
@@
 The lack of rigorous statistical tests for confounder-bias significantly hampers the development of predictive models in many fields of research.
-Although several approaches for mitigating the effects of counfounder variables in machine learning have been recently proposed, these can potentially remove signal-of-interest together with the confounder signal and an informed decision about confound mitigation strategy is impossible without a proper test of confounder-bias.
 
 To fill this critical gap in predictive model development, here I proposed two novel tests, the \emph{partial} and the \emph{full confounder tests}, which probe the null hypotheses of 'no confounder-bias' and 'full confounder-bias', respectively.
-Both approaches are based on a recently published framework for conditional permutation testing, with a mathematical justification of valid type I error control and with assumptions that are easy to meet in most predictive modelling scenarios.
-The tests have a minimal computational overhead; as re-fitting the model is not required.
+The tests are distinguished from alternative approaches by their robustness to non-normally and non-linearly dependent predictions, rendering them applicable with a wide variety of machine learning models.
+The tests have, moreover, a minimal computational overhead, as re-fitting the model is not required.
 
-As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism, the tests can guide the optimization of the confound mitigation strategy and allow quantitative, statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research.
-Given their simplicity, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confounder tests can become widespread tools that pave the way for the development of clinically useful machine learning biomarkers.
+As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism spectrum disorder, the tests can guide the optimization of the confound mitigation strategy and allow quantitative statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research.
+Given their simplicity, robustness, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confounder tests are novel tools in the methodological arsenal of predictive model development and might largely accelerate the development of clinically useful machine learning biomarkers.
 
@@
 In a predictive modelling setting, let $\y$ denote the target variable, $\yhat$ denote model output, i.e.
-the predictions for $\y$ and let $\c$ denote a variable which is considered as a confounder (see Fig.
+the predictions for $\y$ and let $\c$ denote a variable which is considered as a confounder (schematic representation on Fig.
 \ref{fig:schematic}B).
@@
  Conditional independence - in its general form - lies at the heart of several fundamental concepts in statistics and  plays an increasingly important role in various fields of applied statistics, particularly in biomedical applications\citep{spirtes2000causation, fiedler2011mediation, peters2016causal, candes2016panning}.
-Recently,\cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
+Recently, Shah and colleagues\cite{shah2020hardness} have raised important concerns regarding conditional independence testing.
  Their "no free lunch" theorem implies that, without placing some assumptions on the joint distribution of $(\y, \yhat, \c)$, conditional independence testing is effectively impossible.
@@
 
-Recently,\cite{candes2016panning} and\cite{berrett2020conditional} have demonstrated that valid and powerful conditional independence tests can be constructed with inputting distributional information about only two (out of the three) variables.
+Recently, Cand\`es et al.\cite{candes2016panning}, and, based on their work, Berret et al.\cite{berrett2020conditional}, have demonstrated that valid and powerful conditional independence tests can be constructed with inputting distributional information about only two (out of the three) variables.
 Specifically, the conditional permutation test (CPT) of Berrett and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable.
-Thereby, it incorporates the information available on the conditional distribution of interest into the permutation-based inference in a statistically valid manner.
-Moreover, as shown in this paper, CPT can be combined with a generalized additive model (GAM) to handle non-linear relationships, rendering it as an appealing candidate for testing confounder effects in predictive modelling.
+Thereby, it incorporates the information available about the conditional distribution of interest into the permutation-based inference in a statistically valid manner.
 
-As many related papers, the work of\cite{berrett2020conditional} was formalized as a (semi-)supervised learning approach, where $X$ is a set of predictors (features), $Y$ is the target variable and $Z$ is a potential confounder.
+As many related papers, the work of Berrett et al.
+was formalized as a (semi-)supervised learning approach, where $X$ is a set of predictors (features), $Y$ is the target variable and $Z$ is a potential confounder.
 In this setting, testing the null hypothesis $X \independent Y | Z$ aims to determine, whether the features $X$ still affect $Y$, when controlling for $Z$.
@@
 
-In the following section, CPT is adapted for partial confounder testing (option 3).
-For an overview, see Fig.
-\ref{fig:overview}.
+In the following section, CPT is adapted for partial confounder testing (option 3) and extended with general additive model\cite{hastie1987generalized} (GAM) and multinomial logistic regression\citep{bennett1966multiple, jones1975proability} based conditional distribution estimations, in order to make it handle categorical data and non-linear dependencies between the confoudner and the target variable.
+(For an overview of the method, see Fig.
+\ref{fig:overview}).
 The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
+
 \subsubsection*{The partial confounder test}
+\label{sec:methods-pct}
 
-The partial confounder test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based 'copy' of $\c$,
+In accordance with the CPT approach, the partial confounder test generates a null-distribution for an arbitrary predefined test statistic $T(\y,\yhat,\c)$ by sampling permutation based 'copies' of $\c$,
 
@@
 
-As shown in\citep{berrett2020conditional}, to ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:
+As shown by Berrett and colleagues\cite{berrett2020conditional}, to ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:
 
@@
 As long as the numerator of Eq.
-\ref{eq-pperm} is non-zero for all $c_\pi \in C$ and $y \in Y$, the conditional permutations constitute an algebraic group, thus, as shown by\citep{hemerik2018exact}, an unbiased estimate of the p-value under the null can be obtained as:
+\ref{eq-pperm} is non-zero for all $c_\pi \in C$ and $y \in Y$, the conditional permutations constitute an algebraic group, thus, as shown by Hemerik and Goeman\citep{hemerik2018exact}, an unbiased estimate of the p-value under the null can be obtained as:
 $$ p= \frac{\sum_{j=1}^m \mathbb{1} \{T(\y, \yhat, \c^{(j)}) \geq T(\y, \yhat, \c) \}  }{m}$$
 
-While the group property of the conditioned permutations provides a straightforward proof for the validity of the approach, for an alternative verification see the proof of Theorem 1 in\citep{berrett2020conditional}.
+While the group property of the conditioned permutations provides a straightforward proof for the validity of the approach, for an alternative verification see the proof of Theorem 1 in Ref.\citep{berrett2020conditional}.
 
@@
 However, this way the acceptance ratio would be extremely low, even for moderate $n$ (except there is very low dependence of $\c$ on $\y$), resulting in slow mixing times.
-The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of\cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data.
+The partial confounder test can be, however, efficiently implemented with the parallelized pairwise Markov-Chain Monte Carlo sampler of Berrett and colleagues\cite{berrett2020conditional} (Algorithm 1), that draws disjoint pairs in parallel and decides whether or not to swap them randomly, according to the odds ratio calculated from the conditional densities belonging to the original and swapped data.
 The acceptance odds ratio of swapping indices $i$ and $j$ is:
@@
 
-In their Theorem 2,\cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.
+In their Theorem 2, Berrett et al.\cite{berrett2020conditional} verify that the resulting Markov Chain yields the desired stationary distribution, even if the number of steps is small.
 
@@
 Using CPT to test confounder-bias in predictive modelling allows relaxing assumptions on $\yhat$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions on - the joint distribution of the other two variables ($\y$ and $\c$).
+Berrett et al.
 \cite{berrett2020conditional} give a detailed analysis of the robustness of their CPT approach when estimating the conditional distribution with re-using the tested data via linear regression and, also, against misspecifying the conditional distribution of interest to introduce non-linearity.
 
-Here I extend these results by performing simulations that evaluate the GAM- and multinomial liner model-based approaches, in a form that is accessible for power calculations in predictive modelling (considering various weights of the target signal in $c$ and the confounder and the target signals in $\yhat$).
+Here I extend these results by performing simulations that evaluate the GAM- and multinomial liner model-based approaches, in a form that is accessible for power calculations in predictive modelling (considering various weights of the target signal in $\c$ and the confounder and the target signals in $\yhat$).
 Moreover, I investigate the robustness of the tests against the violation of normality and linearity of the conditional distributions $D(\c|\y)$ and $D(\yhat|\y)$.
 
-Simulations are performed separately for the 'partial' and 'full' conditional confound tests.
+Simulations are performed separately for the two proposed tests.
 
@@
 
-$$ \c \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yc} \, g(\y) $$
+$$ \c | y_i \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{yc} \, g(y_i) $$
 
-where $f$ is a function to introduce non-normality, namely the \emph{sinh-arcsinh} transformation of\cite{jones2009sinh} definend as:
+where $f$ is a function to introduce non-normality, namely the \emph{sinh-arcsinh} transformation of Jones et al. \cite{jones2009sinh}, defined as:
 
@@
 
-Moreover, non-linearity can be introduced with the function $g$, which can be simply the identity function (no non-linearity is introduced in this case) or e.g. a sigmoid-shaped function, in our case:
+Moreover, non-linearity can be introduced with the function $g$, which can be simply the identity function (no non-linearity is introduced in this case) or, for instance, a sigmoid-shaped function, in our case:
 
@@
 
-The simulated predicted values are constructed similarly:
+The simulated predicted values are constructed in a similar fashion, but can depend on $\c$ as well:
 
-$$ \yhat \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{y\hat{y}} \: g(\y) \, + \, w_{c\hat{y}} \: g(\y)$$
+$$ \yhat | y_i, c_i \sim f_{\delta, \epsilon}(\mathcal{N}(0, 1)) + w_{y\hat{y}} \: g(y_i) \, + \, w_{c\hat{y}} \: c_i$$
+
+Note that simulations with $w_{c\hat{y}}=0$ produce data under the null hypothesis of no confoudner bias.
 
@@
 
-The partial and full confound tests, as implemented in version 0.20.1 of the package \emph{'mlconfound'} were run with default parameters (1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations) and by implying categorical variables, where needed.
+The partial and full confound tests, as implemented in version 0.20.0 of the package \emph{'mlconfound'} were run with default parameters (1000 permutations and 50 Markov-chain Monte-Carlo steps to generate the conditioned permutations) and by implying categorical variables, where needed.
 
@@
 
-'Partial' and 'full' confound testing was performed with 10000 permutations and 50 Markov-chain Monte Carlo steps, as implemented in version 0.20.0 of the package \emph{'mlconfound'}.
-Unconditional dependence across the involved variables was investigated with conventional permutation test on the $R^2$ values, with 1000 permutations.
+Partial and full confounder testing was performed with 10000 permutations and 50 Markov-chain Monte Carlo steps, as implemented in version 0.20.0 of the package \emph{'mlconfound'}.
+Unconditional dependence across the involved variables was investigated with conventional permutation tests on the $R^2$ values, with 1000 permutations.
 
@@
 The Human Connectome Project dataset contains imaging and behavioral data of approximately 1200 healthy subjects\citep{van2013wu}.
-Preprocessed resting state fMRI connectivity data (partial correlation matrices)\citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores (IQ), measured with Penn Progressive Matrices\citep{duncan2000neural}.
+Preprocessed resting state fMRI connectivity data (partial correlation matrices)\citep{glasser2013minimal} as published with the HCP1200 release (N=999 participants with functional connectivity data) were used to build models that predict individual fluid intelligence scores ($G_f$), measured with Penn Progressive Matrices\citep{duncan2000neural}.
 
-To ensure normality, fluid intelligence scores were non-linearly transformed to normal distribution with the quantile transformation\citep{beasley2009rank} as implemented in \emph{scikit-learn}\citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:hcp-hist} for details).
+To ensure normality, $G_f$ was non-linearly transformed to normal distribution with the quantile transformation\citep{beasley2009rank} as implemented in \emph{scikit-learn}\citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:hcp-hist} for details).
 
@@
 Each of the 5 types of features (raw, regressing out acquisition batch, regressing out age group, COMBAT with acquisition batch, COMBAT with age group) was independently incorporated into a scikit-learn-based\citep{pedregosa2011scikit} machine learning procedure aiming to predict the individual fluid intelligence scores with a ridge regression\citep{hoerl1970ridge}.
-The $\alpha$ parameter of the ridge model was considered as a hyperparameter ($\alpha \in \{0.00001, 0.0001, 0.001, 0.01, 0.1, 1, 10, 100, 1000, 10000, 100000\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with mean squared error as optimization metric.
+The $\alpha$ parameter of the ridge model was considered as a hyperparameter ($\alpha \in \{0.00001, 0.0001, 0.001, 0.01, 0.1, 1, 10, 100, 1000, 10000, 100000\}$) and optimized in a nested cross-validation with 10 folds both in the inner and the outer loop and with mean squared error as optimization metric.
 Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.
@@
 
-The proposed tests were applied to provide evidence of center- and  motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset\citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464.
-Preprocessed regional timeseries data was obtained as shared with the paper\citep{dadi2019benchmarking} which was based on preprocessed image data provided by the Preprocessed Connectome Project\citep{craddock2013neuro}.
+The proposed tests were applied to provide evidence of center- and  motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset\citep{di2014autism} involving 866 participants (ASD: 402, neurotypical control: 464).
+Preprocessed regional timeseries data was obtained as shared with the by Dadi et al.\citep{dadi2019benchmarking} which was based on preprocessed image data provided by the Preprocessed Connectome Project\citep{craddock2013neuro}.
 
-Tangent correlation across the timeseries of the n=122 regions of the BASC\citep{bellec2010multi} atlas) was computed with nilearn\footnote{\href{http://nilearn.github.io/}{http://nilearn.github.io/}}\citep{huntenburg2017loading, esteve2015big}.
+Tangent correlation across the timeseries of the n=122 regions of the BASC\citep{bellec2010multi} brain atlas was computed with nilearn\footnote{\href{http://nilearn.github.io/}{http://nilearn.github.io/}}\citep{huntenburg2017loading, esteve2015big}.
 
 The resulting functional connectivity estimates were considered as features either (i) in their raw form or after applying (ii) feature regression\citep{rao2017predictive} or (iii) COMBAT\citep{johnson2007adjusting, fortin2018harmonization}.
-The investigated confounder variables were 'imaging center' and 'in-scanner motion', as measured by the mean framewise displacement of\cite{power2014methods})
+The investigated confounder variables were 'imaging center' and 'in-scanner motion', as measured by the mean framewise displacement (FD), as defined by Power and colleagues\cite{power2014methods}.
 Mean FD was non-linearly transformed to normal distribution with the quantile transformation\citep{beasley2009rank} as implemented in \emph{scikit-learn}\citep{pedregosa2011scikit} (see Supplementary Figure \ref{fig:abide-hist} for details).
 
-As COMBAT is not able to handle continuous variables (since it was primarily designed to remove "batch-effects"), motion was binned into 10 groups, based on equidistant data quantiles ranging from 0 to 1.
+As COMBAT is not able to handle continuous variables (since it was primarily designed to remove categorical "batch-effects"), motion was binned into 10 groups, based on equidistant data quantiles ranging from 0 to 1.
 
 The total of five types (raw, feature regression of site, feature regression of motion, COMBAT with site, COMBAT with motion) of features were independently incorporated into a scikit-learn-based\citep{pedregosa2011scikit} machine learning procedure aiming to predict the diagnosis (DX: ASD vs.
-HC) with a L2-regularized logistic regression, as previously recommended\citep{dadi2019benchmarking}.
-The $C$ parameter of the model was considered as a hyperparameter ($C \in \{0.1, 1, 10\}$) and optimized in a nested cross-validation (cv) with 10 folds both in the inner and the outer cv-s and with area under the receiver operator curve (AUC under ROC) as optimization metric.
+neurotypical controls) with a L2-regularized logistic regression, as previously recommended\citep{dadi2019benchmarking}.
+The $C$ parameter of the model was considered as a hyperparameter ($C \in \{0.1, 1, 10\}$) and optimized in a nested cross-validation with 10 folds both in the inner and the outer cv-s and with area under the receiver operator curve (AUC under ROC) as optimization metric.
 Confound mitigation was performed inside of the outer cross-validation loop, to avoid leakage.
-
-Confounder testing was performed on the class probabilities (instead of the predicted labels).
+Confounder testing was performed on the predicted class probabilities.
 

### 2021-10-29 supplement.tex

@@
   \fbox{
+   \includegraphics[width=0.4\paperwidth]{fig/raw/sim_h1_bbb_partial_d10e0_lin_cpt_gam.pdf}}
+  \caption{Heatmaps showing the positive rates of the 'partial' confounder test, with categorical variables, normal conditional distribution and linear dependence.}
+  \label{fig:sim-bbb-lin-partial}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \fbox{
+   \includegraphics[width=0.4\paperwidth]{fig/raw/sim_h1_bbb_full_d1e0_linear_cpt_gam.pdf}}
+  \caption{Heatmaps showing the positive rates of the 'full' confounder test, with numerical variables, normal conditional distribution and linear dependence.}
+  \label{fig:sim-ccc-lin-full}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \fbox{
+   \includegraphics[width=0.4\paperwidth]{fig/raw/sim_h1_bbb_full_d1e0_linear_cpt_gam.pdf}}
+  \caption{Heatmaps showing the positive rates of the 'full' confounder test, with categorical variables, normal conditional distribution and linear dependence.}
+  \label{fig:sim-bbb-lin-full}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \fbox{
+   \includegraphics[width=0.4\paperwidth]{fig/raw/sim_h1_ccc_partial_d10e0_sig_cpt_gam.pdf}}
+  \caption{Heatmaps showing the positive rates of the 'partial' confounder test, with numerical variables, normal conditional distribution and sigmoid dependence.}
+  \label{fig:sim-ccc-sig-partial}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \fbox{
+   \includegraphics[width=0.4\paperwidth]{fig/raw/sim_h1_bbb_partial_d10e0_sig_cpt_gam.pdf}}
+  \caption{Heatmaps showing the positive rates of the 'partial' confounder test, with categorical variables, normal conditional distribution and sigmoid dependence.}
+  \label{fig:sim-bbb-sig-partial}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \fbox{
+   \includegraphics[width=0.4\paperwidth]{fig/raw/sim_h1_ccc_full_d1e0_sigmoid_cpt_gam.pdf}}
+  \caption{Heatmaps showing the positive rates of the 'full' confounder test, with numerical variables, normal conditional distribution and sigmoid dependence.}
+  \label{fig:sim-ccc-sig-full}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \fbox{
+   \includegraphics[width=0.4\paperwidth]{fig/raw/sim_h1_bbb_full_d1e0_sigmoid_cpt_gam.pdf}}
+  \caption{Heatmaps showing the positive rates of the 'full' confounder test, with categorical variables, normal conditional distribution and sigmoid dependence.}
+  \label{fig:sim-bbb-sig-full}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \fbox{
    \includegraphics[width=0.36\paperwidth]{fig/supplement/hcp_iq_nonnorm_hist.pdf}
@@
 \end{figure}
-
-\begin{figure}[H]
-  \centering
-  \fbox{
-   \includegraphics[width=0.75\paperwidth]{fig/raw/sim_h1_ccc_partial_d10e0_sig_cpt_gam.pdf}}
-  \caption{Heatmaps showing the positive rates of the 'partial' confounder test, with normal conditional distribution and sigmoid dependence.}
-  \label{fig:sim-ccc-sig-partial}
-\end{figure}

### 2021-10-30 manuscript.tex

@@
 % keywords can be removed
-\keywords{machine learning, predictive modelling, confoudner bias, confounder test, conditional independence, conditional permutation}
+\keywords{machine learning, predictive modelling, confounding bias, confounder test, conditional independence, conditional permutation}
 
@@
 Confounding bias can be driven by - among others - measurement artifacts (e.g motion artifacts in magnetic resonance imaging-based predictive models of Alzheimer's\citep{rao2017predictive}, attention deficit hyperactivity disorder\citep{eloyan2012automated, couvy2016head} or Autism Spectrum Disorder (ASD)\citep{gotts2013perils, spisak2014voxel, spisak2019optimal}), demographic and psychometric variables (e.g models trained to predict intelligence\citep{ cole2012global, he2020deep} might provide a statistically significant predictive performance by picking up solely on age-related variance\citep{dubois2018distributed, lohmann2021predicting}), sampling bias and stochastic group differences (e.g.
-racially biased machine learning models\citep{ obermeyer2019dissecting, lwowski2021risk}), batch effects or, in multi-center studies, center-effects.
+racially biased machine learning models\citep{ obermeyer2019dissecting, lwowski2021risk}) as well as batch effects or, in multi-center studies, center-effects.
 
@@
 Powerful and robust statistical tests for quantifying confounding bias in predictive models could substantially foster both the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation approaches.
-Although it is tempting to think about confounding bias as the \emph{conditional dependence} of the model predictions on the confounder, given the target variable, the proper evaluation of the conditional independence among these variables is challenging.
-Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the partial correlation analogs of bivariate non-parametric correlations (like the partial Spearman correlation, see Fig.
+It is tempting to think about confounding bias as the \emph{conditional dependence} of the model predictions on the confounder, given the target variable.
+However, the proper evaluation of conditional independence among these variables is challenging.
+Namely, even in the presence of a slight non-normality and/or non-linearity of the involved conditional distributions, the 'conditional' analogs of the most popular bivariate non-parametric tests (like the partial Spearman correlation, see Fig.
 \ref{fig:sim-h0-demo}) are not valid measures of conditional independence.
@@
 
-This work aims to construct a statistical test for confounding bias that (i) can be applied post-hoc, without re-fitting the model (i.e.
-requires only the target variable, the confounder and the model predictions), (ii) guarantees valid type-I error control even if the involved conditional distributions are non-normal and/or non-linear, (iii) is applicable for classification as well as for prediction problems, both with numerical and categorical confounders.
+This work aims to construct a statistical test for confounding bias that (i) guarantees valid type-I error control for arbitrary models, even if non-normal and/or non-linear dependencies are involved (ii) does not require re-fitting the model, (iii) is applicable for classification as well as for prediction problems and both with numerical and categorical confounders.
 
@@
 Non-normal conditional distribution on the second plot is illustrated with blue density diagrams (kurtosis: -0.8, skewness: -0.1).
-False positive rates for various simulation settings is shown in heatmaps.
+False positive rates for confounder contributions ($w_{yc}$) and predictive performances ($w_{y\hat{y}}$) is shown in heatmaps.
 The upper limit for the binomial confidence interval corresponding to $alpha=0.05$ is 0.065.
-Values below this threshold indicate a valid type I error control and colored white.
+Values below this threshold (colored white) indicate a valid type I error control.
   }
@@
   \includegraphics[width=0.6\paperwidth]{fig/overview.eps}
-  \caption{\textbf{Overview of the proposed partial confounder test}. \\
+  \caption{\textbf{Graphical representation of the proposed partial confounder test}. \\
     The partial confounder test models the conditional distribution of the confounder, given the target variable, with a generalized additive model (GAM).
-The parallel-pairwise Markov-chain Monte-Carlo sampler draws permutations of the confounder variable that comply with the GAM-based conditional distribution (permutation 1, 2, ..., m).
+The parallel-pairwise Markov-chain Monte-Carlo (MCMC) sampler draws permutations of the confounder variable that comply with the GAM-based conditional distribution (permutation 1, 2, ..., m).
 The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted confounder variables.
@@
 
-The idea of conditional independence provides a straightforward framework for assessing confounding bias in predictive models.
-However, handling the non-normal and/or non-linear conditional dependence often seen in predictive models\citep{garcia2009study, kristensen2017whole} poses a great challenge.
-In fact, as recently shown by Shah and colleagues\cite{shah2020hardness} in their 'no free lunch' theorem, it is effectively impossible to establish a fully non-parametric conditional independence test with a valid type I error control and a non-trivial power.
-In practice, this means that no fully assumption-free test for assessing confounding bias can be constructed.
-Indeed - perhaps somewhat surprisingly, but not totally unexpectedly\cite{korn1984ranges} - partial correlation-like analogs of the widely used bivariate non-parametric test, like partial Spearman correlation, exhibits inflated type I errors even with slight violations of normality and/or linearity (as clearly demonstrated with simulated data on Fig.
+The concept of conditional independence provides a straightforward framework for assessing confounding bias in predictive models.
+However, handling the non-normal and/or non-linear conditional dependencies often seen in predictive models\citep{garcia2009study, kristensen2017whole} poses a great challenge.
+In fact, as recently shown by Shah and colleagues in their 'no free lunch' theorem\cite{shah2020hardness}, it is effectively impossible to establish a \emph{fully non-parametric} conditional independence test with a valid type I error control and a non-trivial power.
+Indeed, perhaps somewhat surprisingly, but not totally unexpectedly\cite{korn1984ranges} - partial correlation-like analogs of widely used bivariate non-parametric test, like partial Spearman correlation, exhibit inflated type I errors even with slight violations of normality and/or linearity (as clearly demonstrated with simulated data on Fig.
 \ref{fig:sim-h0-demo}).
@@
 
-As, in terms of its conditional distribution on the others, the most intractable from the three involved variables (target, prediction, confounder) is clearly the model output\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics.
-Exactly this can be achieved with the proposed approach, which extends the novel framework of conditional permutation testing (CPT)\cite{berrett2020conditional} with conditional distribution estimation via generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic models\cite{bennett1966multiple, jones1975proability}.
+As, in terms of its conditional distribution on the others, the model output is clearly the most intractable from the three involved variables (target, prediction, confounder)\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics.
+Exactly this can be achieved with the proposed approach, which extends the novel framework of conditional permutation testing (CPT)\cite{berrett2020conditional} with conditional distribution estimation via generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic models (mnlogit)\cite{bennett1966multiple, jones1975proability}.
 The proposed approach offers two novel tests for probing confounding bias: the \emph{full confounder test} probes whether the model's predictive performance can be exclusively attributed to the confounder and the \emph{partial confounder test} investigates whether the model utilizes any confounder-variance in the predictions, when controlled for the target variable.
@@
 
-The inner workings of the \emph{partial confounder test} is summarized on Fig.
+The inner workings of the \emph{partial confounder test} are summarized on Fig.
 \ref{fig:overview}.
-In short, the test models the conditional distribution between the confounder and the target variable with a generalized additive model (GAM) - or with a multinomial logistic regression, in case of categorical confounder - and then uses a so-called parallel-pairwise Markov-chain Monte-Carlo sampler that draws permutations of the confounder, so that the permuted variables still comply with the GAM-based conditional distribution.
+In short, the test models the conditional distribution between the confounder and the target variable with a GAM - or with an \emph{mnlogit} regression, in case of categorical confounder - and then uses a so-called parallel-pairwise Markov-chain Monte-Carlo sampler\cite{berrett2020conditional} that draws permutations of the confounder, so that the permuted variables still comply with the estimated conditional distribution.
 The test statistic (coefficient of determination, $R^2$) is then computed between the model predictions and the original, as well as the permuted variables.
@@
   \caption{\textbf{Type I error control and power of the partial confounder test based on simulations with normal conditional distribution and linear dependencies.} \\
-  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette on the right) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\hat{y}}$ (horizontal axis on each heatmap), $w_{yc}$ (vertical axis on each heatmap), $w_{c\hat{y}}$ (rows) and for different sample sizes (N, columns).
-Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1.0 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%.
-First row contains simulations under the null hypothesis (H0, no confounder effect), rows 2-4 represent simulations from the alternative hypothesis (H1, confounding bias).
+  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette on the right) in various simulations settings (100 simulations per tile) with different simulation weights $w_{y\hat{y}}$ (predictive performance; horizontal axis on each heatmap), $w_{yc}$ (confounder-target association; vertical axis on each heatmap), $w_{c\hat{y}}$ (degree of confoudner bias; rows) and for different sample sizes (N, columns).
+Weights 0.2, 0.33, 0.4, 0.6, 0.66, 1.0 can be assigned to the following approximate explained variance values: 4\%, 10\%, 12\%, 25\%, 30\%, 50\%, respectively.
+First row contains simulations under the null hypothesis (H0, no confounder bias), rows 2-4 represent simulations from the alternative hypothesis (H1, confounding bias).
   Positive rates for the simulations under the null and the alternative hypotheses can be interpreted as type I error rate and statistical power, respectively.
@@
 Type I error control was valid in case of categorical variables and for the \emph{full} confounder test, as well.
-Power in these cases was also highly similar to the partial confoudner test results with numerical variables.
+Power in these cases was also highly similar to the partial confounder test results with numerical variables.
 (Supplementary Figures \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-lin-full}, \ref{fig:sim-bbb-sig-partial} and \ref{fig:sim-bbb-sig-full}).
@@
 \caption{\label{tab:unconditional-pvals} Coefficients-of-determination ($R^2$), the corresponding p-values and the p-values of the partial and full confounder tests, for all investigated confounders (conf.) confounder-mitigation methods (method).
-Bold numbers denote significant confounding bias identified by the partial confounder test or the lack of significant contribution over that of the confoudner, as identified by the full confounder test.}
+Bold numbers denote significant confounding bias identified by the partial confounder test or the lack of significant contribution over that of the confounder, as identified by the full confounder test.}
 \end{table}
@@
 
-In the following section, CPT is adapted for partial confounder testing (option 3) and extended with general additive model\cite{hastie1987generalized} (GAM) and multinomial logistic regression\citep{bennett1966multiple, jones1975proability} based conditional distribution estimations, in order to make it handle categorical data and non-linear dependencies between the confoudner and the target variable.
+In the following section, CPT is adapted for partial confounder testing (option 3) and extended with general additive model\cite{hastie1987generalized} (GAM) and multinomial logistic regression\citep{bennett1966multiple, jones1975proability} based conditional distribution estimations, in order to make it handle categorical data and non-linear dependencies between the confounder and the target variable.
 (For an overview of the method, see Fig.
@@
 
-Note that simulations with $w_{c\hat{y}}=0$ produce data under the null hypothesis of no confoudner bias.
+Note that simulations with $w_{c\hat{y}}=0$ produce data under the null hypothesis of no confounder bias.
 

### 2021-10-30 manuscript.tex
finalized for preprint

@@
 Estimates of statistical power for the partial confounder test (with normal and linear simulations, for  a wide range of parameters) are shown on Figure \ref{fig:sim-normal}.
-Notably, with sample sizes as large as 1000, a confounder contributing only $\sim 4\%$ to the variance of the predictions can already be robustly detected ($w_{c\hat{y}} = 0.2$) with a power of 94-100\%.
+Notably, with sample sizes as large as 1000, a confounder contributing only $\sim 4\%$ to the variance of the predictions  ($w_{c\hat{y}} = 0.2$) can already be robustly detected with a power of 94-100\%.
 With a sample size of 500, the same confounding bias is still detected with a power greater than 84-100\% in all of the simulation cases.
@@
 Type I error control was valid in case of categorical variables and for the \emph{full} confounder test, as well.
-Power in these cases was also highly similar to the partial confounder test results with numerical variables.
+Power in these cases was also highly similar to the that of the partial confounder test with numerical variables.
 (Supplementary Figures \ref{fig:sim-bbb-lin-partial}, \ref{fig:sim-bbb-lin-full}, \ref{fig:sim-bbb-sig-partial} and \ref{fig:sim-bbb-sig-full}).
@@
 
-Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $p=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}).
+Both acquisition batch and age group were statistically significantly associated with $G_f$ ($R^2=0.032$ and $0.011$ and $p<0.001$ and $p=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}).
 The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
@@
 
-The full confounder test was highly significant ($p<0.0001$) for all models, indicating that neither of the models were exclusively driven by the confounds.
+The full confounder test was highly significant ($p<0.0001$) for all models, indicating that neither of the models were exclusively driven by the confounders.
 
@@
 
-The full confounder test was strongly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$), indicating that this model might have been (almost) exclusively driven by motion artifacts.
+The full confounder test was strongly significant ($p<0.0001$) for all models, except when regressing-out motion from the features ($p=0.09$), indicating that this model may be (almost) exclusively driven by motion artifacts.
 
@@
 \end{tabular}
-\caption{\label{tab:unconditional-pvals} Coefficients-of-determination ($R^2$), the corresponding p-values and the p-values of the partial and full confounder tests, for all investigated confounders (conf.) confounder-mitigation methods (method).
+\caption{\label{tab:unconditional-pvals} Coefficients-of-determination ($R^2$), the corresponding p-values and the p-values of the partial and full confounder tests, for all investigated datasets, confounders (conf.) and confounder-mitigation methods (method).
 Bold numbers denote significant confounding bias identified by the partial confounder test or the lack of significant contribution over that of the confounder, as identified by the full confounder test.}
@@
 \ref{fig:sim-h0-demo}).
-While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the "no free lunch theorem" (i.e.
+While the magnitude of this problem may not be fully appreciated in case of predictive model diagnostics, the 'no free lunch theorem' (i.e.
 the nonexistence of a uniformly valid conditional independence test)\cite{shah2020hardness} clearly implies that conditional independence-based confounding bias testing must be designed so that its suitability for the particular problem may be judged easily.
@@
 
-As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except extreme non-normality), confirming that the tests can be deployed in a wide variety of predictive modelling scenarios.
-While different biomedical applications might consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculation in order to identify the necessary sample size for proper model diagnostics.
+As expected from theory, both tests displayed a valid type I error control and a high, practically relevant statistical power in the simulations, even if both the predictions and the confounder are non-normally and/or non-linearly dependent on the target variable (except by extreme non-normality).
+This result confirms that the tests can be deployed in a wide variety of predictive modelling scenarios.
+While different biomedical applications might consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculations, in order to identify the necessary sample size for proper model diagnostics.
 
 A characteristic example for the potential areas of applications is the novel field of "predictive neuroscience", where applying predictive modelling and machine learning on functional neuroimaging data holds great potential for both revolutionizing our understanding of the physical basis of mind and delivering clinically useful tools for diagnostics or therapeutic decision making\citep{woo2017building, wager2013fmri, spisak2020pain}.
-However, the presence of confounders typical for biomedical research (e.g.
-sample demographics, center-effects) and specific to the data acquisition and processing approach (e.g.
+However, the presence of confounders that are typical for biomedical research (e.g.
+sample demographics, center-effects) or specific to the data acquisition and processing approach (e.g.
 imaging artifacts) presents a great challenge to these efforts.
@@
 
-In case of the HCP dataset, the weak but statistically significant confounder effect of age on fluid intelligence is in line with previous findings\citep{dubois2018distributed, lohmann2021predicting} and could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample.
+In case of the HCP dataset, the statistically significant age bias of the 'raw' model for predicting fluid intelligence is in line with previous findings\citep{dubois2018distributed, lohmann2021predicting} and could likely exaggerate to a serious bias when testing the model on data of participants outside of the - relatively narrow - age range of the HCP sample.
 In this case, the bias would likely significantly harm the out-of-sample generalizability of this model.
 The bias of the same model for acquisition batch can also be problematic, especially as it has not yet been thoroughly discussed in case of the HCP dataset.
-There can be manifold reasons for the observed bias.
+There can be manifold reasons for the observed acquisition bias.
 Fluid intelligence of the included participants might be, for instance, affected by a changing selection bias during participant recruitment (e.g.
@@
 
-In the ABIDE dataset, neither the center-bias nor the age-bias is surprising but both are obviously problematic for a diagnostic model of ASD.
+In the ABIDE dataset, neither the center bias nor the age bias is surprising in the case of the 'raw' model but both would be obviously severely problematic for a diagnostic biomarker candidate of ASD.
 For instance the model trained on the raw (unadjusted) features - depending on the calibration of the predicted class probabilities - might classify all participants form e.g.
@@
 On the other hand, the partial confounder test also showed, that the performance of the investigated confound mitigation techniques is much less convincing in the ABIDE dataset.
-The center bias of the classification in this massively multi-center dataset was found to be highly significant and, although mitigated, but not successfully removed by feature regression and COMBAT.
+The center bias of the classification in this massively multi-center dataset was, although mitigated, but not successfully removed by feature regression and COMBAT.
 Determining the relevance of the remaining bias is out of the scope of this paper, however significant p-values of the partial confounder test generally suggest that these models require more effective data harmonization.
 
-Motion-bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
+Motion bias in the ABIDE dataset was not detectable anymore after COMBAT, but remained highly significant after feature regression.
 Interestingly, in this latter case, the association between the confounder and the predictions become even stronger than before but with an opposite sign.
@@
 
- In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounding bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant model-bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression and COMBAT and (iii) in some cases confounder mitigation can - paradoxically - introduce more bias.
+ In sum, the application of the \emph{partial} confounder test on the real data examples suggests that confounding bias should be much more carefully investigated and reported in studies utilizing predictive modelling and machine learning as (i) variables as trivial as the date of the acquisition can cause significant confounding bias and (ii) in certain situations, a sufficient mitigation of confounder-bias requires more effective solutions than feature regression or COMBAT and (iii) in some cases confounder mitigation can - paradoxically - introduce more bias.
 The partial confounder test can be considered as a useful, objective benchmark to guide the search for a suitable confounder mitigation approach for every dataset.
+
  Regarding the \emph{full} confounder test; while the real data examples in this paper were not typical cases of its potential applications, it has still provided important insights into model biases.
@@
 
-To fill this critical gap in predictive model development, here I proposed two novel tests, the \emph{partial} and the \emph{full confounder tests}, which probe the null hypotheses of 'no confounder-bias' and 'full confounder-bias', respectively.
+To fill this critical gap in predictive model development, here I proposed two novel tests, the \emph{partial} and the \emph{full confounder tests}, which probe the null hypotheses of 'no confounding bias' and 'full confounding bias', respectively.
 The tests are distinguished from alternative approaches by their robustness to non-normally and non-linearly dependent predictions, rendering them applicable with a wide variety of machine learning models.
@@
 As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism spectrum disorder, the tests can guide the optimization of the confound mitigation strategy and allow quantitative statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research.
-Given their simplicity, robustness, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confounder tests are novel tools in the methodological arsenal of predictive model development and might largely accelerate the development of clinically useful machine learning biomarkers.
+Given their simplicity, robustness, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confounder tests emerge as novel tools in the methodological arsenal of predictive modelling and might largely accelerate the development of clinically useful machine learning biomarkers.
 
@@
 Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; its predictions are driven by the confounder rather than information about the target variable.
-Testing unconditional independence ($\yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
+Assessing the simple bivariate (unconditioned) dependence ($H0: \yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
 For instance, even if $\yhat \independent \c$ is false, $\yhat$ might be only marginally dependent on $\c$, due to the dependence of both on $\y$.
@@
 
-Instead of focusing on the unconditional independence between the confounder and the predictions, we shall consider the \emph{conditional independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, by definition \citep{dawid1979conditional}, means that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$.
+Instead of focusing on the 'unconditioned' independence between the confounder and the predictions, we shall consider the \emph{conditional independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, by definition \citep{dawid1979conditional}, means that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$.
 The statistical test with this null hypothesis will be referred to as the \emph{partial confounder test}.
@@
 This result stands in strong contrast to \emph{unconditional} independence testing - where permutation tests \citep{pitman1937significance, fisher1942189}, provide a general, distribution-free solution - and it has important implications for confounder testing in predictive modelling where the distribution of the model outputs (conditioned on the target variable) - depending on the applied machine learning model - is unknown and often non-normal and non-linear.
-
-One of the trivial candidates for the task, partial correlation, for instance assumes that all involved variables are multivariate Gaussian and - as to be shown below - even its Spearman-based variant is unable to tolerate relatively small deviations from normality and linearity.
+One of the trivial candidates for the task, partial correlation, for instance assumes that all involved variables are multivariate Gaussian and - as to be shown below in a simulated example - even its Spearman-based variant is unable to tolerate relatively small deviations from normality and linearity.
 
@@
 
-In the following section, CPT is adapted for partial confounder testing (option 3) and extended with general additive model\cite{hastie1987generalized} (GAM) and multinomial logistic regression\citep{bennett1966multiple, jones1975proability} based conditional distribution estimations, in order to make it handle categorical data and non-linear dependencies between the confounder and the target variable.
+In the following section, CPT is adapted for \emph{partial} confounder testing (option 3) and extended with general additive model\cite{hastie1987generalized} (GAM) and multinomial logistic regression\citep{bennett1966multiple, jones1975proability} based conditional distribution estimations, in order to make it handle categorical data and non-linear dependencies between the confounder and the target variable.
 (For an overview of the method, see Fig.
 \ref{fig:overview}).
-The formulation of the full confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
+The formulation of the \emph{full} confounder test (option 1) is analogous and given in Supplementary Material \ref{sup:full-test}.
 
@@
 
-Smoothness of the GAM model is optimized with a grid-search by picking the model with the lowest generalized cross-validation score form the models defined by the default parameters as implemented in PyGAM\citep{serven2018generalized}.
+Smoothness of the GAM model is optimized with a grid-search by picking the model with the lowest generalized cross-validation score form the models defined by the default parameters as implemented in PyGAM\citep{serven2018generalized} (v0.8.0).
 
@@
 
-In the case of categorical $c$, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$, with the extra assumption of \emph{complete separation} if $y$ is also categorical (for an invertible Hessian, see e.g.\citep{bennett1966multiple, jones1975proability}).
+In the case of categorical $\c$, a multinomial logistic regression (\emph{mnlogit}) model can be used to obtain $D(\cdot|\y)$, with the extra assumption of \emph{complete separation} if $\y$ is also categorical (in order to ensure an invertible Hessian, see e.g.
+Ref.\citep{bennett1966multiple, jones1975proability}).
 
-Importantly, both the GAM-based and the multinomial logistic regression approaches guarantee that the numerator of Eq.
-\ref{eq-pperm} is always greater than zero and the group property for the permutations holds.
+Importantly, both the GAM- and the \emph{mnlogit}-based approaches guarantee that the numerator of Eq. \ref{eq-pperm} is always greater than zero and the group property for the permutations holds.
 
-Note that from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, for option 2, the proposed approach can not provide a test that is assumption-free about $\yhat$, as the variable, on which the independence is conditional, must be always the predictor variable in Eq.
+Note that from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, the proposed approach can not provide a test for option 2 that is assumption-free about $\yhat$, as the variable, on which the independence is conditional, must be always the predictor variable in Eq.
 \ref{eq:gam}.
@@
 
-Using CPT to test confounder-bias in predictive modelling allows relaxing assumptions on $\yhat$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions on - the joint distribution of the other two variables ($\y$ and $\c$).
+Using CPT to test confounding bias in predictive modelling allows relaxing assumptions on $\yhat$ (for both the partial and the full test) but - in line with the "no free lunch" theorem, requires knowing - or putting assumptions on - the joint distribution of the other two variables ($\y$ and $\c$).
 Berrett et al.
@@
 
-Here I extend these results by performing simulations that evaluate the GAM- and multinomial liner model-based approaches, in a form that is accessible for power calculations in predictive modelling (considering various weights of the target signal in $\c$ and the confounder and the target signals in $\yhat$).
+Here I extend these results by performing simulations that evaluate the GAM- and \emph{mnlogit}-based approaches, in a form that is accessible for power calculations in predictive modelling (considering various weights of the target signal in $\c$ and the confounder and the target signals in $\yhat$).
 Moreover, I investigate the robustness of the tests against the violation of normality and linearity of the conditional distributions $D(\c|\y)$ and $D(\yhat|\y)$.
@@
 
-The simulated predicted values are constructed in a similar fashion, but can depend on $\c$ as well:
+The simulated predicted values are constructed in a similar fashion, but may depend on $\c$ as well:
 
@@
 
-Note that simulations with $w_{c\hat{y}}=0$ produce data under the null hypothesis of no confounder bias.
+Note that simulations with $w_{c\hat{y}}=0$ produce data under the null hypothesis of no confounding bias.
 

### 2021-10-30 manuscript.tex
line numbers

@@
 
-Predictive modelling has recently become increasingly important in biomedical research and holds promise for delivering biomarkers that substantially impact clinical practice and public health\citep{kent2018personalized, spisak2020pain, walsh2021dome}.
+Predictive modelling has recently become increasingly important in biomedical research and holds promise for delivering biomarkers that substantially impact clinical practice and public health\citep{vogt2018machine, kent2018personalized, spisak2020pain, walsh2021dome}.
 When evaluating the usefulness and applicability of such markers, predictive performance is far from being the only important consideration.

### 2021-10-31 reproduce.sh
Create reproduce.sh

@@ -1,0 +1,31 @@
+#!/bin/bash
+# start this script from repository root as working directory
+
+##############
+# Simulations
+##############
+
+# for full reproduction, uncomment this line to delete results
+#rm simulated/results/*
+
+simulated/run_all.sh 
+
+simulated/h0_simulate_normviol_lin.py
+simulated/h0_simulate_normviol_sig.py
+simulated/h1_simulate_normviol_lin.py
+simulated/h1_simulate_normviol_sig.py
+
+jupyter nbconvert --to notebook --execute simulated/normality_and_linearity_violation.ipynb --output simulated/normality_and_linearity_violation.ipynb
+jupyter nbconvert --to notebook --execute simulated/overview-fig.ipynb --output simulated/overview-fig.ipynb
+jupyter nbconvert --to notebook --execute simulated/plot_simulations.ipynb --output simulated/plot_simulations.ipynb
+
+#####################
+# Empirical analysis
+#####################
+
+# data is not part of the repository for licensing reasons.
+# ABIDE data is downloaded automatically by the notebook analysis_abide.ipynb
+jupyter nbconvert --to notebook --execute empirical/notebook analysis_abide.ipynb --output empirical/notebook analysis_abide.ipynb
+
+# HCP data must be downloaded from the connectomeDB, as described in empirical/data/hcp/readme.md
+jupyter nbconvert --to notebook --execute empirical/notebook analysis_hcp.ipynb --output empirical/notebook analysis_hcp.ipynb

### 2021-10-31 manuscript.tex

@@
 
-While various data cleaning methods might help in mitigating confounding bias\citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, johnson2007adjusting}, it is often unclear which variables should be considered as confounders and such approaches hold risks of eliminating signal-of-interest\citep{wachinger2021detect}.
+While various data cleaning methods may help in mitigating confounding bias\citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction, johnson2007adjusting}, it is often unclear which variables should be considered as confounders and such approaches hold risks of eliminating signal-of-interest\citep{wachinger2021detect}.
 
@@
 However, these methods either fail to control type I error (as known in the case of balanced permutations\cite{southworth2009properties, hemerik2018exact}, used in ref.\cite{chaibub2019permutation}), or do not provide p-values at all\cite{ferrari2020measuring}.
-Moreover, without some modifications, they are only applicable for categorical variables and involve re-fitting the model, which might not be feasible for models with high computational cost (e.g.
+Moreover, without some modifications, they are only applicable for categorical variables and involve re-fitting the model, which may not be feasible for models with high computational cost (e.g.
 when trained with nested cross-validation).
@@
 
-As, in terms of its conditional distribution on the others, the model output is clearly the most intractable from the three involved variables (target, prediction, confounder)\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable might already provide a sufficient robustness for predictive model diagnostics.
+As, in terms of its conditional distribution on the others, the model output is clearly the most intractable from the three involved variables (target, prediction, confounder)\citep{garcia2009study, kristensen2017whole}, a method being distribution-free only for this variable may already provide a sufficient robustness for predictive model diagnostics.
 Exactly this can be achieved with the proposed approach, which extends the novel framework of conditional permutation testing (CPT)\cite{berrett2020conditional} with conditional distribution estimation via generalized additive (GAM)\citep{hastie1987generalized} or multinomial logistic models (mnlogit)\cite{bennett1966multiple, jones1975proability}.
@@
 This result confirms that the tests can be deployed in a wide variety of predictive modelling scenarios.
-While different biomedical applications might consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculations, in order to identify the necessary sample size for proper model diagnostics.
+While different biomedical applications may consider different amounts of bias to be relevant, the presented results can serve as a basis for power calculations, in order to identify the necessary sample size for proper model diagnostics.
 
@@
 Namely, in the case of the aforementioned paradox motion-regression model in the ABIDE dataset, the full confounder test did not provide any evidence that the model captures any extra variance over that explained by the confounder ($p=0.1$).
-In other words, the test accepted the null hypothesis of full model bias, suggesting that this model might be, in fact, almost exclusively driven by motion artifacts.
-This example highlights that the full confounder test might become useful in exploratory phases of model development, where models might be still severely biased by various confounders and the questions is whether there is any biomedically relevant signal captured by the model, at all.
+In other words, the test accepted the null hypothesis of full model bias, suggesting that this model may be, in fact, almost exclusively driven by motion artifacts.
+This example highlights that the full confounder test may become useful in exploratory phases of model development, where models might be still severely biased by various confounders and the questions is whether there is any biomedically relevant signal captured by the model, at all.
 
@@
 As demonstrated on functional brain connectivity-based predictive models of fluid intelligence and autism spectrum disorder, the tests can guide the optimization of the confound mitigation strategy and allow quantitative statistical assessment of the robustness, generalizability and neurobiological validity of predictive models in biomedical research.
-Given their simplicity, robustness, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confounder tests emerge as novel tools in the methodological arsenal of predictive modelling and might largely accelerate the development of clinically useful machine learning biomarkers.
+Given their simplicity, robustness, wide applicability, high statistical power and computationally effective implementation (available in the  python package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}}), the partial and full confounder tests emerge as novel tools in the methodological arsenal of predictive modelling and may largely accelerate the development of clinically useful machine learning biomarkers.
 
@@
 
-Obviously, a strong association between $\yhat$ and $\c$ might indicate that the model is biased; its predictions are driven by the confounder rather than information about the target variable.
+Obviously, a strong association between $\yhat$ and $\c$ may indicate that the model is biased; its predictions are driven by the confounder rather than information about the target variable.
 Assessing the simple bivariate (unconditioned) dependence ($H0: \yhat \independent \c$) between $\yhat$ and $\c$ (or any of the $\y$, $\yhat$, $\c$ variables) is, however, insufficient for the proper characterization of confounder-bias in predictive modelling.
@@
 Option 1, i.e.
-full confounder testing, might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction.
+full confounder testing, may be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction.
 Option 2 seems less appealing for model diagnostics and importantly, in this case the proposed variety of the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.

### 2022-04-22 fig/raw/emp_abide_motion_reg_full.tex
manuscript revised

@@
   \draw (92.5bp,47.5bp) node {0.028};
-  \draw [] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  .. (131.0bp,11.0bp) .. controls (196.05bp,8.7922bp) and (272.59bp,13.206bp)  .. (yhat);
-  \draw (158.0bp,18.5bp) node {0.359};
-  \draw [dashed] (y) ..controls (218.38bp,41.389bp) and (278.81bp,29.573bp)  .. (yhat);
-  \draw (248.5bp,50.5bp) node {0.02 (p=0.1)};
+  \draw [] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  .. (131.0bp,11.0bp) .. controls (208.9bp,8.3558bp) and (301.06bp,13.31bp)  .. (yhat);
+  \draw (158.0bp,18.5bp) node {0.062};
+  \draw [dashed] (y) ..controls (225.43bp,42.033bp) and (304.96bp,28.903bp)  .. (yhat);
+  \draw (265.0bp,52.5bp) node {0.031 (p=0.001*)};
 %

### 2022-04-22 fig/raw/emp_abide_site_reg_partial.tex
manuscript revised

--- 2021-10-20 fig/raw/emp_abide_motion_reg_partial.tex
@@
 \node (c) at (27.0bp,18.0bp) [draw,ellipse] {c};
-  \node (y) at (197.5bp,53.0bp) [draw,ellipse] {y};
+  \node (y) at (193.0bp,53.0bp) [draw,ellipse] {y};
   \node (yhat) at (359.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
-  \draw [] (c) ..controls (85.084bp,29.857bp) and (139.57bp,41.174bp)  .. (y);
+  \draw [] (c) ..controls (83.822bp,29.913bp) and (136.23bp,41.098bp)  .. (y);
   \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
   \pgfsetstrokecolor{strokecol}
-  \draw (92.5bp,42.5bp) node {0.028};
+  \draw (92.5bp,42.5bp) node {0.019};
   \draw [dashed] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  ..
 (131.0bp,11.0bp) ..
-controls (190.08bp,8.9948bp) and (204.93bp,8.8036bp)  ..
-(264.0bp,11.0bp) ..
-controls (287.06bp,11.857bp) and (313.23bp,13.894bp)  ..
+controls (186.08bp,9.1305bp) and (199.92bp,9.1305bp)  ..
+(255.0bp,11.0bp) ..
+controls (281.33bp,11.894bp) and (311.39bp,14.069bp)  ..
 (yhat);
-  \draw (197.5bp,18.5bp) node {0.359 (p<0.0001*)};
-  \draw [] (y) ..controls (253.39bp,40.956bp) and (303.37bp,29.989bp)  .. (yhat);
-  \draw (298.0bp,41.5bp) node {0.02};
+  \draw (193.0bp,18.5bp) node {0.052 (p=0.003*)};
+  \draw [] (y) ..controls (249.82bp,41.087bp) and (302.23bp,29.902bp)  .. (yhat);
+  \draw (293.5bp,42.5bp) node {0.161};
 %

### 2022-07-07 fig/raw/emp_abide_site_reg_partial.tex
manuscript revision

--- 2021-10-20 fig/raw/emp_abide_site_comb_partial.tex
@@
 \node (c) at (27.0bp,18.0bp) [draw,ellipse] {c};
-  \node (y) at (188.5bp,53.0bp) [draw,ellipse] {y};
-  \node (yhat) at (341.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
-  \draw [] (c) ..controls (82.887bp,30.044bp) and (132.87bp,41.011bp)  .. (y);
+  \node (y) at (181.0bp,53.0bp) [draw,ellipse] {y};
+  \node (yhat) at (335.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
+  \draw [] (c) ..controls (81.025bp,30.21bp) and (126.99bp,40.794bp)  .. (y);
   \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
@@
 (131.0bp,11.0bp) ..
-controls (182.08bp,9.2662bp) and (194.92bp,9.1009bp)  ..
-(246.0bp,11.0bp) ..
-controls (269.06bp,11.857bp) and (295.23bp,13.894bp)  ..
+controls (175.42bp,9.4924bp) and (186.58bp,9.4924bp)  ..
+(231.0bp,11.0bp) ..
+controls (257.33bp,11.894bp) and (287.39bp,14.069bp)  ..
 (yhat);
-  \draw (188.5bp,18.5bp) node {0.05 (p=0.009*)};
-  \draw [] (y) ..controls (242.32bp,40.716bp) and (287.48bp,30.214bp)  .. (yhat);
-  \draw (280.0bp,42.5bp) node {0.17};
+  \draw (181.0bp,18.5bp) node {0.004 (p=1.0)};
+  \draw [] (y) ..controls (235.03bp,40.79bp) and (280.99bp,30.206bp)  .. (yhat);
+  \draw (269.5bp,43.5bp) node {0.179};
 %

### 2022-07-07 fig/raw/emp_hcp_acq_reg_partial.tex
emp figures patched
manuscript revision

--- 2021-10-20 fig/raw/emp_abide_motion_comb_partial.tex
@@
 \node (c) at (27.0bp,18.0bp) [draw,ellipse] {c};
-  \node (y) at (185.5bp,53.0bp) [draw,ellipse] {y};
-  \node (yhat) at (344.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
-  \draw [] (c) ..controls (82.016bp,30.08bp) and (130.28bp,40.874bp)  .. (y);
+  \node (y) at (172.0bp,53.0bp) [draw,ellipse] {y};
+  \node (yhat) at (317.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
+  \draw [] (c) ..controls (78.775bp,30.427bp) and (120.01bp,40.519bp)  .. (y);
   \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
   \pgfsetstrokecolor{strokecol}
-  \draw (92.5bp,43.5bp) node {0.028};
+  \draw (92.5bp,45.5bp) node {0.032};
   \draw [dashed] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  ..
 (131.0bp,11.0bp) ..
-controls (179.42bp,9.3567bp) and (191.58bp,9.3567bp)  ..
-(240.0bp,11.0bp) ..
-controls (266.33bp,11.894bp) and (296.39bp,14.069bp)  ..
+controls (167.42bp,9.7637bp) and (176.58bp,9.7637bp)  ..
+(213.0bp,11.0bp) ..
+controls (239.33bp,11.894bp) and (269.39bp,14.069bp)  ..
 (yhat);
-  \draw (185.5bp,18.5bp) node {0.002 (p=0.59)};
-  \draw [] (y) ..controls (240.52bp,40.92bp) and (288.78bp,30.126bp)  .. (yhat);
-  \draw (278.5bp,43.5bp) node {0.111};
+  \draw (172.0bp,18.5bp) node {0.0 (p=1.0)};
+  \draw [] (y) ..controls (223.78bp,40.573bp) and (265.01bp,30.481bp)  .. (yhat);
+  \draw (251.5bp,45.5bp) node {0.114};
 %

### 2022-07-07 fig/raw/emp_hcp_age_reg_partial.tex
manuscript revision

@@
 \node (c) at (27.0bp,18.0bp) [draw,ellipse] {c};
-  \node (y) at (181.0bp,53.0bp) [draw,ellipse] {y};
-  \node (yhat) at (335.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
-  \draw [] (c) ..controls (81.025bp,30.21bp) and (126.99bp,40.794bp)  .. (y);
+  \node (y) at (176.5bp,53.0bp) [draw,ellipse] {y};
+  \node (yhat) at (326.0bp,18.0bp) [draw,ellipse] {<<y\&\#770;>>};
+  \draw [] (c) ..controls (79.916bp,30.319bp) and (123.47bp,40.654bp)  .. (y);
   \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
   \pgfsetstrokecolor{strokecol}
-  \draw (92.5bp,43.5bp) node {0.011};
+  \draw (92.5bp,44.5bp) node {0.011};
   \draw [dashed] (c) ..controls (74.615bp,14.069bp) and (104.67bp,11.894bp)  ..
 (131.0bp,11.0bp) ..
-controls (175.42bp,9.4924bp) and (186.58bp,9.4924bp)  ..
-(231.0bp,11.0bp) ..
-controls (257.33bp,11.894bp) and (287.39bp,14.069bp)  ..
+controls (171.42bp,9.628bp) and (181.58bp,9.628bp)  ..
+(222.0bp,11.0bp) ..
+controls (248.33bp,11.894bp) and (278.39bp,14.069bp)  ..
 (yhat);
-  \draw (181.0bp,18.5bp) node {0.004 (p=0.2)};
-  \draw [] (y) ..controls (235.03bp,40.79bp) and (280.99bp,30.206bp)  .. (yhat);
-  \draw (269.5bp,43.5bp) node {0.114};
+  \draw (176.5bp,18.5bp) node {0.0 (p=0.95)};
+  \draw [] (y) ..controls (229.42bp,40.681bp) and (272.97bp,30.346bp)  .. (yhat);
+  \draw (260.5bp,44.5bp) node {0.118};
 %
